---
title: "6000-yr APAL-Climate Regional Analysis"
author: "Alexis E Medina Valmaseda"
abstract: |
  This document is the executable record (Quarto/R) of all analyses reported in the
  main text. It uses two input files deposited in the repository: `ages_kernel_reg_updt2.xlsx`
  (dated *Acropora palmata* fossils, audited against the original age publications) and `df_GAM130326_regimes_v3.xlsx` (centennial climate
  predictors). From these it reproduces:

  1. binning of fossil ages into 0.1-ka (centennial) intervals, their alignment with
     climate predictors and the normalization of the response variable;
  2. kernel-density reconstructions of *A. palmata* age distributions for each reef
     site and for the regional composite, weighted so that they have the shape of the
     modelled series;
  3. exploratory analyses of predictor correlation and of abundance across
     climate-regime combinations;
  4. beta-regression generalized additive models (GAMs) fitted separately to four reef
     sites (Punta Maroma, Florida, Belize, Saint Croix) and to a regional composite,
     first with one climate predictor at a time (marginal models) and then as a nested
     sequence of multi-predictor models, ranked by AIC and Akaike weights;
  5. diagnostics of the selected models (residual plots, concurvity, residual temporal
     autocorrelation) and sensitivity analyses;
  6. validation and robustness tests of the selected models (cross-validation against an
     s(age)-only reference model, circular-shift test, drop-one influence).

  Code, outputs and a factual description of each result are shown together, so that
  every value reported in the main text can be traced to the code that produced it.
  Ecological interpretation is given in the main text.
format: html
execute:
  echo: true
---

# REGIONAL ACROPORA PALMATA ANALYSIS

We import the fossil chronology (`ages_kernel_reg_updt2.xlsx`, sheet `Registros`) and retain *A. palmata* records (`group == "palmata"`) dated 0–6 ka. The site colours defined here (`custom_colors`) are used in all figures.

**Age reference, calibration and data corrections.** The chronology was audited record by record against the original publications of each age (workbook `coral_ages_audit_M3.xlsx`; sheet `Trazabilidad` of `ages_kernel_reg_updt2.xlsx` lists, for every record, its source, method, published age, analytical uncertainty, calibration and the change with respect to the previous version, `ages_kernel_reg_updt.xlsx`). All ages are expressed in ka BP (before 1950 CE), the reference of the climate series, and all radiocarbon ages are calibrated with Marine20 using a local reservoir correction (ΔR), propagated in quadrature.

- **Punta Maroma.** U-Th ages (Medina-Valmaseda et al. 2025) are published relative to the year of chemical separation (2021) and are converted to BP (0.071 ka younger). Ages from Blanchon et al. (2017) are already in years BP. One sample (SURF7(a), ~1956 CE) post-dates 1950; it keeps its age (−0.006 ka) and is assigned to the youngest bin. Two samples dated on separate subsamples (M1(a), M5(b)) retain both ages.
- **Saint Croix.** Radiocarbon ages (Hubbard et al. 2005, 2013), originally calibrated with Marine98 and no local correction, are recalibrated with Marine20 and ΔR = 64 ± 32 ¹⁴C yr (DiNapoli et al. 2020). For samples without a published conventional age (Hubbard et al. 2013), the conventional age and its uncertainty are recovered as the values that reproduce the published calibrated age and 2σ range under the original calibration. One sample dated at two laboratories (BI2-05) enters once, with the error-weighted mean of its conventional ages; one published calibrated age inconsistent with its conventional age (BI1-13) is replaced by the recalibrated value; samples listed in both source publications are counted once. Four U-Th ages are not modified.
- **Belize.** U-Th ages (Gischler et al. 2023) are in years BP and are not modified. Radiocarbon ages are recalibrated with Marine20 and ΔR = −248 ± 35 ¹⁴C yr (Greer et al. 2020), from the conventional ages in Gischler & Hudson (2004) or, for the atoll cores, from conventional ages recovered by inverting the original Marine98 calibration. Because this ΔR is a single, strongly negative estimate, the analysis was repeated with a regional ΔR of −114 ± 70 ¹⁴C yr (inverse-variance weighted mean of the ten nearest entries of the Marine Reservoir Database, with their standard deviation as uncertainty). Seven *A. cervicornis* samples previously coded as *A. palmata* are reclassified, and one typographical error in the compiled table is corrected from the original publication.
- **Florida.** Radiocarbon ages recalibrated with Marine20 by L. T. Toth (pers. comm.) are used; three samples absent from that compilation are recalibrated with the same local correction. U-Th ages are in years BP. Seventeen rows that repeated a sample already present under another source database are removed. Records are standardized by collection (main reef tract, nearshore ridge, Dry Tortugas) as before.
- **Effect.** *A. palmata* records in 0–6 ka: 257 → 236 (Punta Maroma 64, Florida 111, Belize 39, Saint Croix 22). Radiocarbon ages carry 95 % calibrated ranges of 0.3–0.5 kyr (3–5 centennial bins), so for the radiocarbon-dated sites the 0.1-kyr bin is an aggregation unit, not the resolving power of the records.

<!-- Verificar antes de citar: Heaton et al. 2020 (Marine20), DiNapoli et al. 2020, Greer et al. 2020. -->

```{r datos-reg}
rm(list=ls(all=TRUE))
gc()
#install.packages("here")
library(here)
here()
library(readxl)
library(dplyr)
library(ggplot2)
library(stargazer)

set.seed(20260930)
datos_reg <- read_excel(here("ages_kernel_reg_updt2.xlsx"), sheet = "Registros")

datos_reg_pal <- datos_reg %>%
  # post-1950 ages (t_ka < 0) enter the youngest bin
  filter(group == "palmata", t_ka > -0.1 & t_ka <= 6)

custom_colors <- c(
  "Punta Maroma" = "#F2941A",
  "Regional (All spots)" = "#4D4D4D",
  "Regional"             = "#4D4D4D",
  "Saint Croix" = "#CC79A7",
  "Belize" = "#7D9E5F",
  "Florida" = "#0072B2"
)
```

Each record is assigned to a 0.1-ka bin (right-closed intervals; the first interval includes 0), and the upper bound of the bin (`interval_max`) is stored as its centennial age. A second copy of all records is labelled "Regional", and the two sets are stacked into `centennial_counts_all`. This object is still record-level (one row per dated fossil); counts per bin are computed in later chunks.

```{r cent-count}
library(dplyr)

breaks <- seq(0, 6, by = 0.1)

datos_reg_pal <- datos_reg %>%
  # post-1950 ages (t_ka < 0) enter the youngest bin
  filter(group == "palmata", t_ka > -0.1 & t_ka <= 6)

centennial_counts_sites <- datos_reg_pal %>%
  mutate(interval_bin = cut(
    pmax(t_ka, 0),
    breaks = breaks,
    include.lowest = TRUE,
    right = TRUE
  )) %>%
  mutate(interval_max = breaks[match(interval_bin, levels(interval_bin)) + 1])

centennial_counts_regional <- datos_reg %>%
  # post-1950 ages (t_ka < 0) enter the youngest bin
  filter(group == "palmata", t_ka > -0.1 & t_ka <= 6) %>%
  mutate(interval_bin = cut(
    pmax(t_ka, 0),
    breaks = breaks,
    include.lowest = TRUE,
    right = TRUE
  )) %>%
  mutate(
    spot = "Regional",
    interval_max = breaks[match(interval_bin, levels(interval_bin)) + 1]
  )

centennial_counts_all <- bind_rows(centennial_counts_sites, centennial_counts_regional)
```

## *1. Data Overview*

### 1.1 Summary of fossil records by site

We summarize the raw record-level ages by site (**Supplementary Table S1**): number of records, mean, SD, minimum and maximum age. `min_interval` and `max_interval` give the upper bound of the bins containing the youngest and oldest record.

```{r count-summ}
library(dplyr)
library(purrr)
library(gt)
library(stringr)

tabla_resumen_counts <- centennial_counts_all %>%
  group_by(spot) %>%
  summarise(
    n_records   = n(),                               
    mean_t_ka   = mean(t_ka, na.rm = TRUE),         
    sd_t_ka     = sd(t_ka, na.rm = TRUE),           
    min_t_ka    = min(t_ka, na.rm = TRUE),          
    max_t_ka    = max(t_ka, na.rm = TRUE),          
    min_interval = interval_max[which.min(t_ka)],   
    max_interval = interval_max[which.max(t_ka)],   
    .groups = "drop"  ) %>%
  arrange(spot)


tab_S1 <- tabla_resumen_counts %>%
  gt() %>%
  tab_header(
    title = "Summary of fossil records by site (raw data)",
    subtitle = "Temporal distribution and dispersion per 0.1 ka bin"
  ) %>%
  fmt_number(
    columns = c(mean_t_ka, sd_t_ka, min_t_ka, max_t_ka,
                min_interval, max_interval),
    decimals = 3
  )

tab_S1
gt::gtsave(tab_S1, "Supplementary_Table_S1.docx")   
```

The dataset comprises 257 dated records: **Florida** 128, **Punta Maroma** 64, **Belize** 41 and **Saint Croix** 24. Mean ages range from 2.26 ka (**Punta Maroma**) to 3.64 ka (**Florida**). **Punta Maroma** records span 0.065–5.46 ka and **Saint Croix** records 0.265–5.93 ka; **Belize** and **Florida** records cover the full 0–6 ka window.

### 1.2 Climatic predictor: import & alignment

We import the centennial climate table (`df_GAM130326_regimes_v3.xlsx`) and assign each row the upper bound of its 0.1-ka bin. Ages are in ka BP (before 1950 CE), the convention of the two continuous climate sources (Kaufman et al. 2020; Schmitt et al. 2025).

The three categorical regimes were coded from the published chronologies compiled in Supplementary Material 2, converted to ka BP, assigning each bin to the period that contains its midpoint (a midpoint on a boundary goes to the younger period):

- **Thermal** (published chronology, not the GMST series): warmer 0–0.1 ka (post-1850 CE); LIA 0.1–0.7 ka (\~1300–1850 CE); CWP&MWA 0.7–1.3 ka (700–1250 CE); cooler 1.3–4.8 ka; warmer 4.8–6.0 ka (mid-Holocene warmth).
- **Hurricane** (Schmitt et al. 2025 only, the source of `Hur100_av` and `Hur100_fqcy`): high 0–0.2 ka (rise after 1800 CE); calm 0.2–0.6 ka (1400–1800 CE); enhanced 0.6–1.1 and 1.2–1.4 ka (900–1400 and 600–800 CE); moderate 1.1–1.2 and 1.4–4.0 ka (mean 9–14 events per century); low 4.0–6.0 ka (\~7 events per century; the record begins at 5.7 ka).
- **Hydro**, by region. Yucatán (`Hydro`): wet 0–0.8 ka, dry 0.8–1.7 ka and wet 1.7–3.5 ka (Curtis et al. 1996, boundaries calibrated with IntCal20), wet 5.2–5.7 ka (Serrato Marks et al. 2021), no data elsewhere. Florida (`Hydro_FL`): dry 4.6–6.0 ka (Pollock et al. 2017; Donders 2014), wet 2.8–4.6 ka and dry 0–2.8 ka (Glaser et al. 2013), with multi-century droughts near 1.0 and 0.4 ka (Willard & Bernhardt 2011).
- **Hurricane, Florida** (`Hur_FL`, `FL_pact`, `FL_nrec`). Schmitt et al. (2025) record storms reaching the Yucatán coast, and Florida-proximal records do not follow it (agreement in 34 % of the bins younger than 3.5 ka). The Florida panel therefore uses a local series built from six records on the Atlantic side of Florida: the western Great Bahama Bank / Florida Straits (Toomey et al. 2013; 0–6 ka), the Everglades (Yao et al. 2020; 0–3.5 ka), southwest Florida (Ercolani et al., via Oliva et al. 2017), Andros and Abaco (van Hengstum et al., via Oliva et al. 2017) and Abaco (Winkler et al. 2020). All ages are cal BP (1950). For each bin, `FL_pact` is the proportion of the records covering the bin that report an active interval at its midpoint, `FL_nrec` the number of those records, and `Hur_FL` is "active" when `FL_pact` ≥ 0.5 and "quiet" otherwise. Older than 3.5 ka the series rests on a single record (Toomey et al. 2013). Records from the Gulf coast of Florida were not used, because they sample Gulf landfalls. For Hydro and the hurricane regime, the Florida panel uses the Florida series; Punta Maroma, Belize, Saint Croix and the regional composite use the Yucatán series (Hurricane = Schmitt et al. 2025). The continuous hurricane metrics (`Hur100_av`, `Hur100_fqcy`) are those of Schmitt et al. (2025) in every panel. In the Yucatán Hydro series, undocumented bins are coded as a separate level, "no data", so that no bins are lost. The mapping is positional: `interval_max` is taken from the order in which interval labels first appear, which is correct only if the file has exactly one row per bin, ordered from youngest to oldest.

Climate values are joined to every fossil record by `interval_max`, and the column is renamed `age` (ka). The resulting `merged_df` is record-level and therefore holds climate values only for bins that contain at least one fossil record.

```{r data-integ}
datos_cli <- read_excel(here("df_GAM130326_regimes_v3.xlsx"))
datos_cli<-as.data.frame(datos_cli)

library(dplyr)

breaks <- seq(0, 6, by = 0.1)

datos_cli <- datos_cli %>%
  mutate(interval = factor(interval, levels = unique(interval))) %>%
  mutate(interval_max = breaks[match(interval, levels(interval)) + 1])

datos_cli <- datos_cli %>%
  mutate(Hydro_YUC = ifelse(is.na(Hydro), "no data", Hydro))

# Hydro and hurricane regime by region: Florida records take the Florida series,
# all other records the Yucatán series (Hurricane = Schmitt et al. 2025)
merged_df <- centennial_counts_all %>%
  left_join(datos_cli, by = "interval_max") %>%
  mutate(Hydro     = ifelse(spot == "Florida", Hydro_FL, Hydro_YUC),
         Hurricane = ifelse(spot == "Florida", Hur_FL, Hurricane)) %>%
  select(-Hydro_FL, -Hydro_YUC, -Hur_FL, -FL_nrec) %>%
  rename( age = interval_max)
```

### 1.3 Response variable: centennial counts and normalization

The response variable of all models is built here, before any figure, so that Figs. 2 and 3 show the same series that enter the GAMs. This chunk makes three lasting changes to `merged_df`:

1.  **Reference levels.** Thermal, Hydro and Hurricane are converted to factors with reference levels "warmer", "wet" and "low" ("quiet" for the local Florida hurricane series). These levels carry through the aggregation below and define every categorical contrast reported in Sections 3.2 and 3.3.

2.  **Sites only.** Rows are restricted to the four reef sites, which drops the duplicated "Regional" records created in the binning step. The regional composite is rebuilt from the site series below.

3.  **Record counter.** A column `counts = 1` is added (one per dated record); it is summed per bin below.

```         
The per-site objects (`datos_pm`, `datos_fl`, `datos_bz`, `datos_sc`, `datos_reg`) and `merged_df_g` created in this chunk are intermediate: all of them are redefined below, before any model is fitted.
```

```{r data-prepa}
library(mgcv)
library(dplyr)
library(patchwork)

merged_df <- merged_df %>%
  mutate(
    Thermal   = relevel(as.factor(Thermal), ref = "warmer"),
    Hydro     = relevel(as.factor(Hydro), ref = "wet"),
    # "low" is the reference of the Schmitt regimes; in Florida (local series) the
    # first level present, "quiet", is the reference
    Hurricane = factor(Hurricane, levels = c("low", "moderate", "calm", "enhanced", "high",
                                             "quiet", "active"))
  )%>%
  filter(spot %in% c("Punta Maroma", "Florida", "Belize", "Saint Croix")) %>%
  mutate(panel = spot)

datos_pm <- merged_df %>% filter(spot == "Punta Maroma")
datos_fl <- merged_df %>% filter(spot == "Florida")
datos_bz <- merged_df %>% filter(spot == "Belize")
datos_sc <- merged_df %>% filter(spot == "Saint Croix")
datos_reg <- merged_df %>%
  filter(spot %in% c("Punta Maroma", "Florida", "Belize", "Saint Croix")) %>%
  mutate(spot = "Regional")

merged_df <- merged_df %>%
   mutate(counts = 1)

merged_df_g <- bind_rows(datos_pm, datos_fl, datos_bz, datos_sc, datos_reg)

```

#### Normalization

For each site, records are summed per bin (`counts`). Climate predictors are averaged within the bin, which leaves them unchanged because they are constant within a bin; regime labels are taken from the first record in the bin.

`normalize_counts()` divides counts by the site maximum (`counts_norm`, rounded to four decimals). Values exactly 0 or 1 are then transformed as y′ = (y(*n* − 1) + 0.5)/*n*, where *n* is the number of bins in the series (`counts_norm_sv`); all other values are unchanged. Only bins containing at least one record exist in the series, so no zeros occur and the correction affects only the maximum bin (e.g. 0.984 in **Punta Maroma**, *n* = 31).

The response is therefore the relative abundance within bins in which a site is represented. Bins without records are not part of any model: 29 of 60 in **Punta Maroma**, 33 in **Belize**, 42 in **Saint Croix**, 15 in **Florida** and 2 in the **regional pool**.

#### Florida source-standardization

Florida combines records from several source datasets (column `source`). Therefore, we first sum counts per source and bin, and further divided by each source's maximum and boundary-corrected. Next we sum the standardized values across sources per bin and normalize as above (`datos_fl_std`). Consequently, each source has equal weight at its maximum, independent of its number of records.

The regional composite used in every model is the per-bin sum of the four site-level `counts_norm_sv` series (**Florida** in its source-standardized form), normalized with the same function (`datos_reg`, *n* = 58 bins). Each site therefore enters the composite with its maximum bin equal to one, so its contribution depends on how its records are distributed among bins rather than on its number of records (Section 1.4). The five series are stacked into `merged_df_g`, which is used by every model and by Figs. 2 and 3a.

```{r norm-stand}

normalize_counts <- function(df, panel_name){
  
  df %>%
    mutate(
      counts_norm = round(
        counts / max(counts, na.rm = TRUE),
        4
      ),
      
      counts_norm_sv = ifelse(
        counts_norm %in% c(0,1),
        ((counts_norm * (n() - 1)) + 0.5) / n(),
        counts_norm
      ),
      
      panel = panel_name
    )
  
}

#---------------------------------------------------------
# PUNTA MAROMA
#---------------------------------------------------------

datos_pm <- merged_df %>%
  filter(spot == "Punta Maroma") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Punta Maroma")

#---------------------------------------------------------
# BELIZE
#---------------------------------------------------------

datos_bz <- merged_df %>%
  filter(spot == "Belize") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Belize")

#---------------------------------------------------------
# SAINT CROIX
#---------------------------------------------------------

datos_sc <- merged_df %>%
  filter(spot == "Saint Croix") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Saint Croix")

#---------------------------------------------------------
# FLORIDA
# Independently standardized by source dataset 
#---------------------------------------------------------

datos_fl_std <- merged_df %>%
  filter(spot == "Florida") %>%
  
  group_by(source, age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  
  group_by(source) %>%
  mutate(
    
    counts_std = counts / max(counts, na.rm = TRUE),
    
    counts_std = ifelse(
      counts_std %in% c(0,1),
      ((counts_std * (n() - 1)) + 0.5) / n(),
      counts_std
    )
    
  ) %>%
  ungroup() %>%
  
  group_by(age) %>%
  summarise(
    
    counts = sum(counts_std, na.rm = TRUE),
    
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    
    .groups = "drop"
    
  ) %>%
  normalize_counts("Florida")

#---------------------------------------------------------
# REGIONAL
# build upon normalized datasets
#---------------------------------------------------------

datos_reg <- bind_rows(
  datos_pm,
  datos_fl_std,
  datos_bz,
  datos_sc
) %>%
  group_by(age) %>%
  summarise(
    
    counts = sum(counts_norm_sv, na.rm = TRUE),
    
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    
    .groups = "drop"
    
  ) %>%
  select(-Hydro) %>%
  left_join(datos_cli %>% transmute(age = interval_max, Hydro = Hydro_YUC, Hur_YUC = Hurricane),
            by = "age") %>%
  mutate(Hydro = factor(Hydro, levels = levels(merged_df$Hydro)),
         Hurricane = factor(Hur_YUC, levels = levels(merged_df$Hurricane))) %>%
  select(-Hur_YUC) %>%
  normalize_counts("Regional")

#---------------------------------------------------------
# Ready-to use dataset
#---------------------------------------------------------

merged_df_g <- bind_rows(
  datos_pm,
  datos_fl_std,
  datos_bz,
  datos_sc,
  datos_reg
)
```

### 1.4 *Acropora palmata* age densities (weighted KDE)

Kernel density estimates (KDE) are computed from the individual ages, weighted so that the density has the shape of the modelled series of Section 1.3. Each record receives the weight it carries in `counts_norm_sv`: in Punta Maroma, Belize and Saint Croix, `counts_norm_sv / counts` of its bin; in Florida, the product of its source-standardization factor and the Florida normalization factor; in the regional composite, the site weight multiplied by the regional normalization factor. The weights of the records in each bin therefore add up to `counts_norm_sv` of that bin (checked in the code). For a site with a single source the weights are constant except in the maximum bin, so the weighted KDE has the same shape as the KDE of the raw ages; the weighting changes the shape only for Florida (source standardization) and for the regional composite.

Each density is a Gaussian KDE over 0–6 ka (bandwidth 0.09 ka, 512 points). Its uncertainty is estimated with 1,000 bootstrap resamples of the records with replacement, keeping their weights and stratified by source in Florida and by site in the regional composite. For each resample we keep the density curve and its mean in each 0.1-ka bin. `signal_all` holds the estimate and the 2.5–97.5 % and 25–75 % bootstrap envelopes of each curve; `centennial_density_all` holds the bin means of the estimate and their bootstrap SD and 95 % interval.

Because a KDE integrates to one, the regional density distributes a fixed total among the four sites in proportion to the sum of their weights, which depends on how their records are spread among bins, not on how many records they have. Saint Croix, whose 24 records occupy 18 bins with at most two records each, carries 27 % of the regional weight with 9 % of the records, and Punta Maroma 15 % with 25 % of the records (`site_share`). This is a property of the per-site normalization used in the models, not of the KDE, and it applies equally to the regional GAMs.

```{r dens}
library(dplyr)
library(gt)

## ---- record-level weights reproducing the modelled response -------------------
## The weights of the records in a bin add up to counts_norm_sv of that bin, so the
## weighted KDE has the shape of the series that enter the GAMs.
bin_upper <- function(t) breaks[as.integer(cut(pmax(t, 0), breaks, include.lowest = TRUE, right = TRUE)) + 1]

recs <- datos_reg_pal %>%
  mutate(t_ka = as.numeric(t_ka), age = round(bin_upper(t_ka), 1))

# Punta Maroma, Belize, Saint Croix: weight = counts_norm_sv / counts of the bin
w_single <- bind_rows(datos_pm, datos_bz, datos_sc) %>%
  transmute(spot = panel, age = round(age, 1), w_site = counts_norm_sv / counts)

# Florida: per-source standardization (as in datos_fl_std), then the Florida normalization
w_fl_src <- recs %>%
  filter(spot == "Florida") %>%
  count(source, age, name = "n_src") %>%
  group_by(source) %>%
  mutate(std = n_src / max(n_src),
         std = ifelse(std %in% c(0, 1), (std * (n() - 1) + 0.5) / n(), std),
         w_src = std / n_src) %>%
  ungroup() %>%
  select(source, age, w_src)
w_fl_bin <- datos_fl_std %>% transmute(age = round(age, 1), w_bin = counts_norm_sv / counts)

# Regional composite: site weights rescaled by the regional normalization
w_reg_bin <- datos_reg %>% transmute(age = round(age, 1), w_reg = counts_norm_sv / counts)

recs <- recs %>%
  left_join(w_single, by = c("spot", "age")) %>%
  left_join(w_fl_src, by = c("source", "age")) %>%
  left_join(w_fl_bin, by = "age") %>%
  left_join(w_reg_bin, by = "age") %>%
  mutate(w_site = ifelse(spot == "Florida", w_src * w_bin, w_site),
         w_regional = w_site * w_reg) %>%
  select(-w_src, -w_bin, -w_reg)

# check: the weights of each bin reproduce the modelled series
chk_site <- recs %>% group_by(spot, age) %>% summarise(w = sum(w_site), .groups = "drop") %>%
  inner_join(bind_rows(datos_pm, datos_fl_std, datos_bz, datos_sc) %>%
               transmute(spot = panel, age = round(age, 1), y = counts_norm_sv), by = c("spot", "age"))
chk_reg <- recs %>% group_by(age) %>% summarise(w = sum(w_regional), .groups = "drop") %>%
  inner_join(datos_reg %>% transmute(age = round(age, 1), y = counts_norm_sv), by = "age")
stopifnot(nrow(chk_site) == sum(merged_df_g$panel != "Regional"),
          nrow(chk_reg) == sum(merged_df_g$panel == "Regional"),
          max(abs(chk_site$w - chk_site$y)) < 1e-3, max(abs(chk_reg$w - chk_reg$y)) < 1e-3)

# share of each site in the regional density (weights sum to one)
site_share <- recs %>% group_by(spot) %>%
  summarise(records = n(), pct_records = 100 * n() / nrow(recs),
            pct_weight = 100 * sum(w_regional) / sum(recs$w_regional), .groups = "drop")
site_share

## ---- weighted KDE with stratified bootstrap -------------------------------------
## Records are resampled with replacement within strata (sources in Florida, sites in the
## regional composite), keeping their weights. bw = 0.09 ka, 512 points over 0-6 ka.
calc_density_summary <- function(t, w, spot_name, strata = rep(1, length(t)),
                                 B = 1000, bw = 0.09) {
  kde <- function(i) density(t[i], weights = w[i] / sum(w[i]), bw = bw,
                             from = 0, to = 6, n = 512)
  full <- kde(seq_along(t))
  idx  <- split(seq_along(t), strata)
  boot <- t(vapply(seq_len(B), function(b) {
    i <- unlist(lapply(idx, function(ix) ix[sample.int(length(ix), length(ix), replace = TRUE)]))
    kde(i)$y
  }, numeric(512)))

  signal_df <- data.frame(
    x = full$x, estimate = full$y,
    q025 = apply(boot, 2, quantile, 0.025), q975 = apply(boot, 2, quantile, 0.975),
    q25  = apply(boot, 2, quantile, 0.25),  q75  = apply(boot, 2, quantile, 0.75),
    spot = spot_name)

  # centennial summary: estimate and bootstrap SD of the bin-mean density
  bin <- cut(full$x, breaks = breaks, include.lowest = TRUE, right = TRUE)
  bin_mean <- function(y) tapply(y, bin, mean)
  boot_bins <- apply(boot, 1, bin_mean)                 # 60 bins x B
  centennial_summary <- data.frame(
    interval_mid = breaks[-length(breaks)] + 0.05,
    density      = as.numeric(bin_mean(full$y)),
    boot_sd      = apply(boot_bins, 1, sd),
    boot_q025    = apply(boot_bins, 1, quantile, 0.025),
    boot_q975    = apply(boot_bins, 1, quantile, 0.975),
    spot         = spot_name)

  list(signal_df = signal_df, centennial_summary = centennial_summary)
}

site_levels <- c("Punta Maroma", "Belize", "Saint Croix", "Florida")
results_list <- lapply(site_levels, function(sp) {
  r <- recs %>% filter(spot == sp)
  calc_density_summary(r$t_ka, r$w_site, sp,
                       strata = if (sp == "Florida") r$source else rep(1, nrow(r)))
})
results_list <- c(results_list, list(
  calc_density_summary(recs$t_ka, recs$w_regional, "Regional composite", strata = recs$spot)))

signal_all <- bind_rows(lapply(results_list, `[[`, "signal_df"))
centennial_density_all <- bind_rows(lapply(results_list, `[[`, "centennial_summary"))

## ---- Supplementary Table S2 ---------------------------------------------------
tabla_resumen <- centennial_density_all %>%
  mutate(spot = factor(spot, levels = c(site_levels, "Regional composite"))) %>%
  group_by(Site = spot) %>%
  summarise(
    mean_boot_sd  = mean(boot_sd),
    max_density   = max(density),
    max_interval  = interval_mid[which.max(density)],
    min_density   = min(density),
    min_interval  = interval_mid[which.min(density)],
    .groups = "drop")

tab_S2 <- tabla_resumen %>%
  gt() %>%
  tab_header(title = "Centennial weighted kernel-density summaries by site",
             subtitle = "Density of the modelled series and its bootstrap uncertainty") %>%
  fmt_number(columns = c(mean_boot_sd, max_density, min_density), decimals = 3) %>%
  fmt_number(columns = c(max_interval, min_interval), decimals = 2)
tab_S2
gt::gtsave(tab_S2, "Supplementary_Table_S2.docx")
```

**Supplementary Table S2. Centennial weighted kernel-density summaries.** For each site and the regional composite: mean bootstrap SD of the bin-mean density over the 60 bins (`mean_boot_sd`), and the maximum and minimum bin-mean density with the bin midpoint (ka) at which each occurs.

Maximum density occurs at 0.15 ka in **Punta Maroma**, **Belize** and the **regional composite**, at 1.05 ka in **Saint Croix** and at 4.25 ka in **Florida**. Minima fall at 5.95 ka (**Punta Maroma**), 2.35 ka (**Belize**), 2.65 ka (**Saint Croix**), 1.45 ka (**Florida**) and 0.55 ka (**regional composite**). Weighting moves the maximum of the regional curve from 4.25 ka, where the unweighted pool of all records peaked because of the Florida records, to 0.15 ka.

The intervals highlighted in Fig. 2 (`rects`) are: **Punta Maroma** 0.71–1.44 ka; **Belize** 1.99–2.72, 3.68–4.20 and 5.54–6.00 ka; **regional** 2.20–3.00 and 5.10–5.90 ka. They correspond to local *A. palmata* presence gaps reported for **Punta Maroma** (Medina-Valmaseda et al., 2025) and **Belize** (Hubbard et al., 2005, Gischler et al., 2023).

**Figure 2** (raw version; `Figure_2.pdf`). Weighted kernel-density curves of *A. palmata* ages. (a) Each site, with one colour and one line type per site. (b) Regional composite, with the 2.5–97.5 % and interquartile bootstrap envelopes. Horizontal bars at the top of each panel mark the intervals listed above.

```{r fig2m, fig.width=8, fig.height=8}
rects <- data.frame(
  xmin = c(0.71, 1.99, 2.2, 3.68, 5.1, 5.54),
  xmax = c(1.44, 2.72, 3.0, 4.20, 5.9, 6.00),
  spot = c("Punta Maroma", "Belize", "Regional (All spots)", "Belize",
           "Regional (All spots)", "Belize"))

library(patchwork)

site_lty <- c("Punta Maroma" = "solid", "Belize" = "longdash",
              "Saint Croix" = "dotdash", "Florida" = "dotted")

sig_sites <- signal_all %>%
  filter(spot %in% site_levels) %>%
  mutate(spot = factor(spot, levels = site_levels))
sig_reg <- signal_all %>% filter(spot == "Regional composite")

# gap bars drawn just above the highest curve of each panel
top_a <- max(sig_sites$estimate, na.rm = TRUE)
top_b <- max(sig_reg$q975, na.rm = TRUE)
rects_a <- rects %>% filter(spot %in% site_levels) %>%
  mutate(spot = factor(spot, levels = site_levels),
         ymin = top_a * 1.04, ymax = top_a * 1.12)
rects_b <- rects %>% filter(spot == "Regional (All spots)") %>%
  mutate(ymin = top_b * 1.04, ymax = top_b * 1.12)

x_scale <- list(
  scale_x_continuous(breaks = seq(0, 6, 0.5), expand = c(0.01, 0)),
  coord_cartesian(xlim = c(0, 6)),
  theme_classic(base_size = 11),
  theme(legend.position = "bottom", plot.tag = element_text(face = "bold"))
)

# (a) weighted KDE of each site: one colour and one line type per site
p2a <- ggplot(sig_sites, aes(x = x)) +
  geom_rect(data = rects_a,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = spot),
            inherit.aes = FALSE, alpha = 0.6, show.legend = FALSE) +
  geom_line(aes(y = estimate, colour = spot, linetype = spot), linewidth = 1) +
  scale_colour_manual(values = custom_colors[site_levels], name = "Site") +
  scale_linetype_manual(values = site_lty, name = "Site") +
  scale_fill_manual(values = custom_colors[site_levels], guide = "none") +
  labs(x = NULL, y = "Density") +
  x_scale + theme(axis.text.x = element_blank())

# (b) regional composite: weighted KDE with bootstrap 95 % and interquartile envelopes
p2b <- ggplot(sig_reg, aes(x = x)) +
  geom_rect(data = rects_b,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = FALSE, fill = custom_colors["Regional"], alpha = 0.6) +
  geom_ribbon(aes(ymin = q025, ymax = q975, fill = "95 % bootstrap interval"), alpha = 0.25) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = "Interquartile range"), alpha = 0.45) +
  geom_line(aes(y = estimate, colour = "Estimate"), linewidth = 1.2) +
  scale_fill_manual(values = c("95 % bootstrap interval" = "grey60", "Interquartile range" = "grey40"),
                    name = "Regional composite") +
  scale_colour_manual(values = c("Estimate" = unname(custom_colors["Regional"])), name = NULL) +
  labs(x = "Time (ka BP)", y = "Density") +
  x_scale

p <- p2a / p2b + plot_annotation(tag_levels = "a")
p
ggsave("Figure_2.pdf", plot = p, width = 8, height = 8, dpi = 300, device = cairo_pdf)
```

### 1.5 Abundance and Holocene trends in climatic variables

Panel (a) shows the modelled series of Section 1.3: the relative abundance `counts_norm_sv` of each site per 0.1-ka bin (Florida source-standardized), stacked by site. The height of each stack is the per-bin sum from which the regional composite is normalized. Bins without records have no bar. The legend gives the number of records per site.

**Figure 3a** (`Figure_3a.pdf`). Relative abundance of *A. palmata* per 100-year interval, stacked by reef site.

```{r count-by-site}
library(ggplot2)
library(dplyr)

n_records <- table(datos_reg_pal$spot)
site_labels <- setNames(paste0(site_levels, " (n = ", n_records[site_levels], ")"), site_levels)

norm_stack <- bind_rows(datos_pm, datos_fl_std, datos_bz, datos_sc) %>%
  transmute(age, spot = factor(panel, levels = rev(site_levels)), counts_norm_sv)

pr <- ggplot(norm_stack, aes(x = age, y = counts_norm_sv, fill = spot)) +
  geom_col(position = "stack", width = 0.09, alpha = 0.85) +
  scale_fill_manual(values = custom_colors[site_levels], labels = site_labels,
                    breaks = site_levels) +
  labs(x = "Time (ka BP)",
       y = "Relative abundance\n(stacked by site)",
       fill = "Site") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom") +
  scale_x_continuous(breaks = seq(0, 6, 0.5), minor_breaks = seq(0, 6, 0.5)) +
  coord_cartesian(xlim = c(0, 6))
pr
ggsave("Figure_3a.pdf", plot = pr, width = 8, height = 5, dpi = 300, device = cairo_pdf)
```

We assemble Fig. 3. Panel (a) is Fig. 3a. Below it, four panels show, for each of the 60 centennial bins:

- **(b)** the `GMHT_CPS0_30N` temperature anomaly (^o^C);
- **(c)** accumulated rate of change in thermal anomalies (ẟT,^o^C) calculated from the CPS model of GMST (`dT_raw`);
- **(d)** 100 year average reconstruction of hurricane impacts (`Hur100_av` ) as bars (dark grey where \> 10) and the 100 year TC frequency ( `Hur100_fqcy` ) as a step line;
- **(e)** local hurricane activity in Florida (`FL_pact`): proportion of the Florida-proximal paleotempestological records that report an active interval in each bin, as bars (dark blue where ≥ 0.5, the threshold that defines the "active" regime); bins older than 3.5 ka rest on a single record (Toomey et al. 2013);
- **(f)** the nominal thermal, hydroclimatic (Yucatán and Florida series) and hurricane (Yucatán and Florida series) regimes, drawn as contiguous time segments over the full span of each bin.

**Figure 3** (`Figure_3.pdf`). Reconstruction of regional coral abundance and paleoclimate trends. The climate panels (b–f) are drawn from the 60-bin climate table, so they cover the full 0–6 ka interval, including bins without fossil records.

```{r fig3m, fig.width=12, fig.height=11.5}

library(dplyr)
library(tidyr)
library(patchwork)

clim_bins <- datos_cli %>% rename(age = interval_max)

p1 <- clim_bins %>%
  group_by(age) %>%
  summarise(GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE)) %>%
  ggplot(aes(x = age, y = GMHT_CPS0_30N, fill = GMHT_CPS0_30N > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "#B8860B", "FALSE" = "#0072B2")) +
  coord_cartesian(ylim = c(-1.2, 1.2), xlim = c(0, 6)) +
  scale_x_continuous(position = "top", breaks = seq(0, 6)) +
  geom_hline(yintercept = 1.2, color = "black", linewidth = 0.4) +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    panel.background = element_blank(),
    legend.position = "none",
    axis.title.x = element_blank()
  ) +
  labs(y = "GMST(°C)\nanomalies")

p2 <- clim_bins %>%
  group_by(age) %>%
  summarise(dT_raw = mean(dT_raw, na.rm = TRUE)) %>%
  ggplot(aes(x = age, y = dT_raw, fill = dT_raw > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9")) +
  coord_cartesian(ylim = c(-1.2, 0.7)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = expression("Rate of Change\nẟT(°C) / 100 yr"))


hur_plot_data <- clim_bins %>%
  group_by(age) %>%
  summarise(
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    Hur100_fqcy_sd = sd(Hur100_fqcy, na.rm = TRUE)
  ) %>%
  arrange(age)

p3 <- ggplot(hur_plot_data, aes(x = age)) +
  geom_col(aes(y = Hur100_av, fill = Hur100_av > 10), alpha = 0.6) +
  geom_step(aes(y = Hur100_fqcy), color = "darkblue", linewidth = 0.8) +
  scale_fill_manual(values = c("TRUE" = "#4D4D4D", "FALSE" = "#999999")) +
  coord_cartesian(ylim = c(0, 50)) +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    panel.background = element_blank(),
    legend.position = "none",
    axis.title.x = element_blank()
  ) +
  labs(y = "Hurricane Average\n(events/100 yr)")

# local Florida hurricane series: proportion of Florida-proximal records active in each bin.
# Bars are dark where FL_pact >= 0.5 ("active" regime), as the Schmitt bars are dark above 10.
p3fl <- ggplot(clim_bins, aes(xmin = age - 0.1, xmax = age, ymin = 0, ymax = FL_pact)) +
  geom_rect(aes(fill = FL_pact >= 0.5), alpha = 0.8) +
  annotate("segment", x = 3.5, xend = 6, y = 1.12, yend = 1.12, colour = "grey40", linewidth = 0.3) +
  annotate("text", x = 4.75, y = 1.12, vjust = -0.4, size = 2.6, colour = "grey30",
           label = "one record (Toomey et al. 2013)") +
  scale_fill_manual(values = c("TRUE" = "#08519C", "FALSE" = "#9ECAE1")) +
  scale_x_continuous(breaks = seq(0, 6, 0.5)) +
  coord_cartesian(xlim = c(0, 6), ylim = c(0, 1.25)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Hurricane activity,\nFlorida (prop. records)")

# regime segments drawn over the true span of each bin, (age - 0.1, age]
regime_rows <- c(Thermal = "Thermal", Hydro_YUC = "Hydro (Yucatán)",
                 Hydro_FL = "Hydro (Florida)", Hurricane = "Hurricane (Yucatán)",
                 Hur_FL = "Hurricane (Florida)")
df_segments <- clim_bins %>%
  select(age, all_of(names(regime_rows))) %>%
  pivot_longer(-age, names_to = "variable", values_to = "state") %>%
  arrange(variable, age) %>%
  group_by(variable) %>%
  mutate(segment_id = cumsum(state != lag(state, default = first(state)))) %>%
  group_by(variable, state, segment_id) %>%
  summarise(age_start = min(age) - 0.1, age_end = max(age), .groups = "drop") %>%
  mutate(label_x = (age_start + age_end) / 2,
         row = factor(regime_rows[variable], levels = rev(regime_rows)))

p4 <- ggplot(df_segments) +
  geom_rect(aes(xmin = age_start, xmax = age_end,
                ymin = as.numeric(row) - 0.4, ymax = as.numeric(row) + 0.4,
                fill = state), colour = "grey40", linewidth = 0.2) +
  geom_text(aes(x = label_x, y = as.numeric(row),
                label = ifelse(age_end - age_start >= 0.25, state, "")),
            color = "black", size = 2.4) +
  scale_y_continuous(breaks = seq_along(levels(df_segments$row)),
                     labels = levels(df_segments$row)) +
  scale_fill_manual(values = c(
    "warmer"   = "#DAA520", "cooler"   = "#9ECAE1",
    "LIA"      = "#56B4E9", "CWP&MWA"  = "#E69F00",
    "wet"      = "#999933", "dry"      = "#E8DFA0", "no data" = "#FFFFFF",
    "high"     = "#525252", "enhanced" = "#969696", "calm" = "#D9D9D9",
    "moderate" = "#BDBDBD", "low"      = "#F0F0F0",
    "active"   = "#2171B5", "quiet"    = "#DEEBF7")) +
  scale_x_continuous(breaks = seq(0, 6, 0.5)) +
  coord_cartesian(xlim = c(0, 6)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Age (ka BP)", y = "Nominal Terms")

Figure_3 <- pr / p1 / p2 / p3 / p3fl / p4 +
  plot_layout(ncol = 1, heights = c(1, 1, 1, 1, 0.8, 1.7))
Figure_3

ggsave("Figure_3.pdf", plot = Figure_3, width = 8, height = 11.5, dpi = 300, device = cairo_pdf)
```

Before modelling, we inspect pairwise correlations among numeric variables (Section 2.1) and the distribution of abundance across combinations of climate regimes (Section 3.1.3). No predictor is removed on the basis of these analyses. Redundancy is handled instead by never entering the two hurricane metrics in the same model (Section 3.3) and by assessing overlap among smooth terms after fitting (concurvity, Section 4.2).

## **2.** Exploratory Data Analysis

#### 2.1 Variable correlation plot

We compute Pearson correlations among the four continuous climate predictors over the 60 centennial bins of the climate table (`datos_cli`), so that each bin counts once regardless of how many fossil records it contains, and cross-tabulate the three categorical regimes.

The only strong correlation is between the two hurricane metrics (Hur100_fqcy and Hur100_av, *r* = 0.79), which is why they are never entered in the same model (Section 3.3). The other correlations are weak to moderate (\|*r*\| ≤ 0.36, the largest between GMHT_CPS0_30N and dT_raw). The categorical regimes, by contrast, are strongly nested: the "cooler" thermal regime occurs only under dry or wet hydroclimate, LIA only under variable hydroclimate, and the "medium" hurricane regime occurs almost only within CWP&MWA (6 of 7 bins). This nesting is the origin of the rank-deficient candidates excluded in Section 3.3.2.

**Supplementary Figure S1.** Pearson correlation matrix of the continuous climate predictors across the 60 centennial bins.

```{r corr, fig.width=8, fig.height=7}
library(corrplot)

clim_vars  <- c("GMHT_CPS0_30N", "dT_raw", "Hur100_fqcy", "Hur100_av")
cor_matrix <- cor(datos_cli[, clim_vars])

corrplot(cor_matrix, method = "circle", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black")

pdf("Supplementary_Figure_S1.pdf", width = 8, height = 7)
corrplot(cor_matrix, method = "circle", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black")
invisible(dev.off())

table(Thermal = datos_cli$Thermal, Hydro = datos_cli$Hydro)
table(Thermal = datos_cli$Thermal, Hurricane = datos_cli$Hurricane)
```

## 3. GAM analysis

### 3.1 Modelling data

All models use `merged_df_g`, built in Section 1.3: the source-standardized Florida series, the three single-source site series and the regional composite. The two subsections below assess the Florida source standardization, and Section 3.1.3 describes abundance across climate-regime combinations.

#### 3.1.1 Florida: pooled normalization (sensitivity series)

`datos_fl_norm` applies the same procedure to **Florida** without source standardization, pooling records across sources before normalization. This series is used only in the comparison of Section 3.1.2; no model is fitted to it.

```{r fl-norm-function}

normalize_counts <- function(df, panel_name){
  df %>%
    mutate(
      counts_norm = round(counts / max(counts, na.rm = TRUE), 4),
      # aplicar SV si hay 0 o 1
      counts_norm_sv = ifelse(counts_norm %in% c(0,1),
                              ((counts_norm * (n() - 1)) + 0.5) / n(),
                              counts_norm),
      panel = panel_name
    )
}

datos_fl_norm <- merged_df %>%
  filter(spot == "Florida") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Florida")
```

#### 3.1.2 Florida: effect of source standardization

##### a. Comparison table

We join the pooled series (`norm`) and the source-standardized series (`std`) by bin into `fl_compare`.

```{r florida-comparison}
fl_compare <- full_join(

  datos_fl_norm %>%
    select(age, counts_norm_sv) %>%
    rename(norm = counts_norm_sv),

  datos_fl_std %>%
    select(age, counts_norm_sv) %>%
    rename(std = counts_norm_sv),

  by = "age"

)
```

##### b. Quick assessment

We summarize the differences (std − norm), list the 20 largest, compute the correlation between the two series, and plot both series and their difference.

```{r florida-comparison-visual}

summary(fl_compare$std - fl_compare$norm)

max(abs(fl_compare$std - fl_compare$norm),
    na.rm = TRUE)

fl_compare <- fl_compare %>%
  mutate(
    delta = std - norm
  )

fl_compare %>%
  mutate(
    delta = std - norm,
    abs_delta = abs(delta)
  ) %>%
  arrange(desc(abs_delta)) %>%
  select(age, norm, std, delta) %>%
  slice(1:20)


cor(
  fl_compare$norm,
  fl_compare$std,
  use = "complete.obs"
)


p_fl_a <- ggplot(fl_compare, aes(age)) +
  
  geom_ribbon(
    aes(
      ymin = pmin(norm, std),
      ymax = pmax(norm, std)
    ),
    fill = "grey70",
    alpha = 0.5
  ) +
  
  geom_line(
    aes(y = norm,
        colour = "Original"),
    linewidth = 1
  ) +
  
  geom_line(
    aes(y = std,
        colour = "Standardized"),
    linewidth = 1
  ) +
  
  scale_color_manual(
    values = c(
      "Original" = "black",
      "Standardized" = "red"
    )
  ) +
  
  scale_x_reverse() +
  
  theme_bw() +
  
  labs(
    x = "Age (ka BP)",
    y = "Relative abundance",
    colour = NULL
  )

p_fl_b <- ggplot(
  fl_compare,
  aes(age, delta)
) +
  
  geom_hline(
    yintercept = 0,
    linetype = 2
  ) +
  
  geom_area(
    fill = "steelblue",
    alpha = 0.6
  ) +
  
  scale_x_reverse() +
  
  labs(
    x = "Age (ka BP)",
    y = "Standardized - Original"
  ) +
  
  theme_bw(base_size = 12)

library(patchwork)
figS2 <- p_fl_a / p_fl_b + plot_annotation(tag_levels = "a")
figS2
ggsave("Supplementary_Figure_S2.pdf", plot = figS2, width = 8, height = 8, dpi = 300, device = cairo_pdf)
```

The differences have a median of 0.088, a mean of 0.100 and a range of −0.027 to 0.413. The largest occur at 2.5, 2.4 and 3.2 ka. The two series are correlated at *r* = 0.894. Source standardization therefore preserves the overall temporal pattern but raises relative abundance in most bins (fewer than a quarter of the differences are negative), most strongly at 2.4–2.5 and 3.1–3.2 ka. This comparison concerns the response series only; the GAMs were not refitted with the pooled series.

**Supplementary Figure S2.** (a) Pooled (black) and source-standardized (red) Florida series; the grey band marks their difference. (b) Difference std − norm. Age is in ka.

#### **3.1.3 *Acropora palmata* abundance across climate-regime combinations**

We combine the three regime labels into `ClimateCombo` (Thermal_Hydro_Hurricane) and tabulate their frequency per panel (**Supplementary Table S3**; each panel block ends with its number of bins). The regional composite is one of the panels, so its bins are not additional observations. No missing values occur.

We then retain combinations observed in at least three bins in at least one panel. Because the filter is applied across panels, a retained combination can still have *n* = 1–2 in a given panel (e.g. LIA_variable_low, *n* = 2 in Belize; warmer_variable_low, *n* = 1 in Punta Maroma). Three combinations are excluded: CWP&MWA_variable_medium, warmer_wet_low and LIA_variable_medium.

**Supplementary Table S4** gives *n*, mean, SD and range of `counts_norm_sv` for each panel and retained combination. A panel-level summary (`resumen_panel`) is also computed but not shown.

##### Climate combinations (frequencies)

```{r climate-combinations}

library(forcats)
eda_counts <- merged_df_g %>%
  
  mutate(
    ClimateCombo = interaction(
      Thermal,
      Hydro,
      Hurricane,
      sep = "_"
    )
  )

eda_counts[!complete.cases(eda_counts), ]

eda_counts$ClimateCombo <- reorder(
  eda_counts$ClimateCombo,
  eda_counts$counts_norm_sv,
  median
)

combo_freq <- eda_counts %>%
  count(panel, ClimateCombo)

combo_freq_tbl <- combo_freq %>%
  mutate(ClimateCombo = as.character(ClimateCombo)) %>%
  group_by(panel) %>%
  group_modify(~ bind_rows(.x, tibble::tibble(ClimateCombo = "Total", n = sum(.x$n)))) %>%
  ungroup()

tab_S3 <- combo_freq_tbl %>%
  gt(groupname_col = "panel") %>%
  tab_header(
    title = "Climate combinations (frequencies)",
    subtitle = "Number of centennial bins per panel and climatic combination"
  ) %>%
  fmt_number(columns = n, decimals = 0) %>%
  cols_label(ClimateCombo = "Climate combination", n = "Bins") %>%
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_body(rows = ClimateCombo == "Total"))

tab_S3
gt::gtsave(tab_S3, "Supplementary_Table_S3.docx")
```

##### Climate-combination summary

```{r climate-combination-summary}
combos_validos <- combo_freq %>%
  filter(n >= 3) %>%
  pull(ClimateCombo) %>%
  as.character()

invisible(
  combo_freq %>%
    filter(n >= 3)
)

invisible(
  combo_freq %>%
    filter(n < 3)
)

datos_filtrados <- eda_counts %>%
  filter(
    ClimateCombo %in% combos_validos
  )

datos_filtrados <- datos_filtrados %>%
  group_by(panel) %>%
  mutate(
    ClimateCombo = fct_reorder(
      ClimateCombo,
      counts_norm_sv,
      .fun = median
    )
  ) %>%
  ungroup()

resumen_panel <- datos_filtrados %>%
  group_by(panel) %>%
  summarise(
    n_obs = n(),
    mean_abund = mean(counts_norm_sv, na.rm = TRUE),
    sd_abund = sd(counts_norm_sv, na.rm = TRUE),
    min_abund = min(counts_norm_sv, na.rm = TRUE),
    max_abund = max(counts_norm_sv, na.rm = TRUE),
    .groups = "drop"
  )

resumen_panel_combo <- datos_filtrados %>%
  group_by(
    panel,
    ClimateCombo
  ) %>%
  summarise(
    n_obs = n(),
    mean_abund = mean(counts_norm_sv, na.rm = TRUE),
    sd_abund = sd(counts_norm_sv, na.rm = TRUE),
    min_abund = min(counts_norm_sv, na.rm = TRUE),
    max_abund = max(counts_norm_sv, na.rm = TRUE),
    .groups = "drop"
  )

tab_S4 <- resumen_panel_combo %>%
  gt() %>%
  tab_header(
    title = "Climate-combination summary",
    subtitle = "Standardized Acropora abundance by climatic regime"
  ) %>%
  fmt_number(
    columns = c(
      mean_abund,
      sd_abund,
      min_abund,
      max_abund
    ),
    decimals = 3
  )

tab_S4
gt::gtsave(tab_S4, "Supplementary_Table_S4.docx")
```

##### Coral-Climate Interactions

The figure bellow summarizes the distribution of standardized *A. palmata* abundance across climatic-regime combinations. Climatic states are ordered by median abundance within panels, allowing visual comparison of abundance patterns and variability among environmental conditions.

```{r figS3, fig.width=12, fig.height=10}

figS3 <- ggplot(
  datos_filtrados,
  aes(
    x = fct_reorder(
      ClimateCombo,
      counts_norm_sv,
      .fun = median
    ),
    y = counts_norm_sv,
    fill = panel
  )
) +
  
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.7
  ) +
  
  geom_point(
    position = position_jitter(width = 0.12, seed = 1),  
    size = 2,
    alpha = 0.6,
    shape = 21,
    colour = "black"
  ) +
  
  facet_wrap(
    ~ panel,
    scales = "free_y",
    ncol = 2
  ) +
  
  scale_fill_manual(
    values = custom_colors
  ) +
  
  theme_bw() +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 12
    ),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(
      size = 14,
      face = "bold"
    ),
    axis.title.y = element_text(
      size = 14,
      face = "bold"
    ),
    strip.text = element_text(
      size = 14,
      face = "bold"
    ),
    legend.position = "none"
  ) +
  
  labs(
    x = "Climate combinations (ordered by median abundance)",
    y = "Standardized abundance (counts_norm_sv)"
  )

figS3
ggsave("Supplementary_Figure_S3.pdf", plot = figS3, width = 8, height = 5, dpi = 300, device = cairo_pdf)
```

**Supplementary Figure S3. Standardized *A. palmata* abundance under Holocene climate-regime combinations.** Boxplots of `counts_norm_sv` by combination of thermal, hydroclimatic and hurricane regime, per panel; points are individual centennial bins. Combinations are ordered by their median over all panels pooled, so the order is the same in every facet.

The highest means occur under warmer_variable_high (**Belize** 0.87, *n* = 2; **Punta Maroma** 0.60, *n* = 3; **regional composite** 0.75, *n* = 3; **Saint Croix** 0.97, *n* = 1), except in **Florida**, where this combination has one of the lowest values (0.13, *n* = 2). The lowest means occur under **LIA**\_variable_low (**Florida** 0.10, *n* = 2; **regional composite** 0.18, *n* = 3). Most bins fall under cooler_wet_high (*n* = 3–20 per panel) and cooler_dry_low (*n* = 4–11), with mean abundance of 0.14–0.82 depending on site. The extreme combinations are represented by three or fewer bins, so these patterns are descriptive only.

### 3.2. Marginal GAM effects analysis

Relative abundance is continuous and bounded in (0, 1), so we fit GAMs with a beta distribution and logit link (`mgcv::gam`, family `betar`). Smoothing parameters are estimated by REML. Continuous predictors enter as thin-plate regression splines (mgcv default, basis dimension *k* = 10).

Climate regimes enter as factors with reference levels "warmer" (Thermal), "wet" (Hydro) and "low" (Hurricane). Their coefficients are contrasts on the logit scale relative to the reference, conditional on the other terms. Every model includes a smooth of age, `s(age)`, which absorbs temporal structure not attributed to the climate predictor.

Modelling proceeds in two stages: marginal models, with `s(age)` plus one climate predictor (this section), and combined models, a nested sequence with two branches, one per hurricane metric (Section 3.3).

Candidates are ranked by AIC, ΔAIC and Akaike weights; models within ΔAIC ≤ 2 of the best are considered competitive. Marginal models are fitted without `select = TRUE` and with *k* = 10, whereas combined models use `select = TRUE` and a reduced *k*; AIC values are therefore not comparable between the two stages.

In this document, "linear" models are those whose climate term is a categorical factor (a parametric term), and "smooth" models those whose climate term is a continuous predictor fitted as a spline.

#### 3.2.1 Candidate models

`formulas_mg` defines seven marginal candidates, each consisting of `s(age)` plus one of: `s(GMHT_CPS0_30N)` (M_temp), `s(dT_raw)` (M_dt), `s(Hur100_fqcy)` (M_hf), `s(Hur100_av)` (M_hav), Thermal (M_the), Hydro (M_hyd) or Hurricane (M_hur). `fit_gams_panel()` fits all candidates to one panel, and `compare_gams()` returns AIC, ΔAIC and Akaike weights.

```{r setup-formulas}
formulas_mg <- list(
  M_temp = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N),
  M_dt   = counts_norm_sv ~ s(age) + s(dT_raw),          
  M_hf   = counts_norm_sv ~ s(age) + s(Hur100_fqcy),
  M_hav  = counts_norm_sv ~ s(age) + s(Hur100_av),
  M_the  = counts_norm_sv ~ s(age) + Thermal,
  M_hyd  = counts_norm_sv ~ s(age) + Hydro,
  M_hur  = counts_norm_sv ~ s(age) + Hurricane
)

fit_gams_panel <- function(data, formulas) {
  lapply(formulas, function(f)
    gam(f, data = data, family = betar(link = "logit"), method = "REML"))
}

compare_gams <- function(fits) {
  stats <- lapply(fits, AIC)
  df <- data.frame(Model = names(stats), AIC = unlist(stats))
  df <- df %>%
    mutate(deltaAIC = AIC - min(AIC),
           weight = exp(-0.5 * deltaAIC) / sum(exp(-0.5 * deltaAIC)))
  return(df)
}
```

#### 3.2.2 Model competition (ΔAIC and Akaike weights)

We fit the seven candidates to each panel of `merged_df_g` and plot ΔAIC and Akaike weights by panel.

**Supplementary Figure S4. Information-theoretic comparison of the marginal GAMs by panel.** (a) ΔAIC of each candidate (Supplementary Table S5); the red dashed line marks ΔAIC = 2, and the ΔAIC axis is scaled separately for each panel. (b) Akaike weights, on a common 0–1 scale. Candidates are listed in the same order in every panel.

```{r marginal-results, fig.width=12, fig.height=7}
panels <- unique(merged_df_g$panel)

results <- lapply(panels, function(p) {
  
  df_panel <- merged_df_g %>%
    filter(panel == p)
  
  if(nrow(df_panel) < 3 ||
     all(is.na(df_panel$counts_norm_sv)))
    return(NULL)
  
  fits <- fit_gams_panel(
    df_panel,
    formulas_mg
  )
  
  stats <- compare_gams(fits)
  
  stats$panel <- p
  
  stats
  
})

results <- Filter(
  Negate(is.null),
  results
)

stats_all <- bind_rows(results)

library(patchwork)

site_order <- c("Punta Maroma", "Florida", "Belize", "Saint Croix", "Regional")

plot_model_selection <- function(stats, panel_var, model_levels) {
  d <- stats %>%
    mutate(Panel = factor(.data[[panel_var]], levels = site_order),
           Model = factor(Model, levels = rev(model_levels)))
  base <- function(yvar, ylab) {
    ggplot(d, aes(x = Model, y = .data[[yvar]], colour = Panel)) +
      geom_segment(aes(xend = Model, y = 0, yend = .data[[yvar]]), linewidth = 0.8) +
      geom_point(size = 2.2) +
      coord_flip() +
      scale_x_discrete(drop = FALSE) +
      scale_colour_manual(values = custom_colors[site_order], guide = "none") +
      labs(x = NULL, y = ylab) +
      theme_minimal(base_size = 12) +
      theme(strip.text = element_text(face = "bold"),
            panel.grid.minor = element_blank())
  }
  p_d <- base("deltaAIC", "ΔAIC") +
    geom_hline(yintercept = 2, linetype = "dashed", colour = "red") +
    facet_wrap(~Panel, nrow = 1, drop = FALSE, scales = "free_x")
  p_w <- base("weight", "Akaike weight") +
    facet_wrap(~Panel, nrow = 1, drop = FALSE) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1))
  (p_d / p_w) + plot_annotation(tag_levels = "a")
}

sup_fig4 <- plot_model_selection(stats_all, panel_var = "panel", model_levels = names(formulas_mg))
sup_fig4
ggsave("Supplementary_Figure_S4.pdf", plot = sup_fig4, width = 12, height = 7, dpi = 300, device = cairo_pdf)
```

#### 3.2.3 Marginal GAM model ranking

**Supplementary Table S5. Marginal GAM ranking by panel.** For each panel, the seven candidates are ordered by AIC, with ΔAIC, Akaike weight and rank. (`get_best_models()` refits the same models as Section 3.2.2 on the same data.)

```{r best-gams}

linear_models <- c(
  "M_the",
  "M_hyd",
  "M_hur"
)

smooth_models <- c(
  "M_temp",
  "M_dt",
  "M_hf",
  "M_hav"
)

get_best_models <- function(data, formulas){
  
  fits <- fit_gams_panel(data, formulas)
  
  ranking <- compare_gams(fits) %>%
    arrange(AIC) %>%
    mutate(
      Rank = row_number()
    )
  
  best_linear <- ranking %>%
    filter(Model %in% linear_models) %>%
    slice_min(AIC, n = 1)
  
  best_smooth <- ranking %>%
    filter(Model %in% smooth_models) %>%
    slice_min(AIC, n = 1)
  
  list(
    fits = fits,
    ranking = ranking,
    best_linear = best_linear,
    best_smooth = best_smooth
  )
  
}

best_pm <- get_best_models(datos_pm, formulas_mg)
best_fl <- get_best_models(datos_fl_std, formulas_mg)
best_bz <- get_best_models(datos_bz, formulas_mg)
best_sc <- get_best_models(datos_sc, formulas_mg)
best_reg <- get_best_models(datos_reg, formulas_mg)


tbl_models <- bind_rows(
  
  best_pm$ranking %>%
    mutate(Panel = "Punta Maroma"),
  
  best_fl$ranking %>%
    mutate(Panel = "Florida"),
  
  best_bz$ranking %>%
    mutate(Panel = "Belize"),
  
  best_sc$ranking %>%
    mutate(Panel = "Saint Croix"),
  
  best_reg$ranking %>%
    mutate(Panel = "Regional")
  
)

gt_ranking <- tbl_models %>%
  
  gt(groupname_col = "Panel") %>%
  
  fmt_number(
    columns = c(AIC, deltaAIC, weight),
    decimals = 3
  ) %>%
  
  cols_label(
    Rank = "Rank",
    Model = "Model",
    AIC = "AIC",
    deltaAIC = "ΔAIC",
    weight = "Akaike Weight"
  ) %>%
  
  tab_header(
    title = md("**GAM model selection by panel**")
  )

gt_ranking
gt::gtsave(gt_ranking, "Supplementary_Table_S5.docx")   
```

In **Punta Maroma**, M_hav dominates (weight 0.993; all other models ΔAIC \> 9). In **Belize**, M_hf (weight 0.66) and M_the (ΔAIC 1.93, weight 0.25) are competitive. In the **regional composite**, M_the ranks first (weight 0.68), followed by M_hav at ΔAIC 2.70. In **Florida**, three models lie within ΔAIC ≤ 2 (M_hyd, M_hav, M_temp; weights 0.34, 0.24, 0.16), so no single predictor is clearly supported. In **Saint Croix**, six of seven models lie within ΔAIC ≤ 1.6 (weights 0.10–0.21); with *n* = 18, AIC does not discriminate among predictors.

#### 3.2.4 Best-supported marginal GAMs

**Supplementary Table S6.** Best categorical ("linear") and best continuous ("smooth") marginal model per panel, selected by lowest AIC within each group.

```{r best-gams-winners}
tbl_winners <- bind_rows(
  
  data.frame(
    Panel = "Punta Maroma",
    Best_Linear = best_pm$best_linear$Model,
    Best_Smooth = best_pm$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Florida",
    Best_Linear = best_fl$best_linear$Model,
    Best_Smooth = best_fl$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Belize",
    Best_Linear = best_bz$best_linear$Model,
    Best_Smooth = best_bz$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Saint Croix",
    Best_Linear = best_sc$best_linear$Model,
    Best_Smooth = best_sc$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Regional",
    Best_Linear = best_reg$best_linear$Model,
    Best_Smooth = best_reg$best_smooth$Model
  )
  
)

gt_winners <- tbl_winners %>%
  gt() %>%
  cols_label(
    Panel = "Panel",
    Best_Linear = "Best Linear",
    Best_Smooth = "Best Smooth"
  ) %>%
  tab_header(
    title = md("**Best-supported linear and smooth GAMs**")
  )

gt_winners
gt::gtsave(gt_winners, "Supplementary_Table_S6.docx")   
```

We refit these models and print their full `mgcv` summaries.

```{r best-gams-summary}

print_best_models <- function(best_obj,
                              panel_name,
                              formulas,
                              data){
  
  lin_name <- best_obj$best_linear$Model
  smo_name <- best_obj$best_smooth$Model
  
  cat("\n========================================\n")
  cat(panel_name,"\n")
  cat("========================================\n")
  
  cat("\nBEST LINEAR MODEL:", lin_name, "\n\n")
  
  print(
    summary(
      gam(
        formulas[[lin_name]],
        data = data,
        family = betar(link="logit"),
        method = "REML"
      )
    )
  )
  
  cat("\n----------------------------------------\n")
  
  cat("\nBEST SMOOTH MODEL:", smo_name, "\n\n")
  
  print(
    summary(
      gam(
        formulas[[smo_name]],
        data = data,
        family = betar(link="logit"),
        method = "REML"
      )
    )
  )
  
}

print_best_models(best_pm,
                  "Punta Maroma",
                  formulas_mg,
                  datos_pm)

print_best_models(best_fl,
                  "Florida",
                  formulas_mg,
                  datos_fl_std)

print_best_models(best_bz,
                  "Belize",
                  formulas_mg,
                  datos_bz)

print_best_models(best_sc,
                  "Saint Croix",
                  formulas_mg,
                  datos_sc)

print_best_models(best_reg,
                  "Regional",
                  formulas_mg,
                  datos_reg)
```

We then extract, for each of these models: AIC, deviance explained and adjusted R²; the climate term with the lowest term-level *p*-value (`s(age)` excluded) and that *p*-value; the factor contrasts with *p* \< 0.05; the EDF of each smooth; and a support label ("Supported" if the term-level climate *p* \< 0.05, otherwise "Weak support"). Term-level *p*-values come from a joint Wald test of the whole term (`anova()`): for a categorical regime, all levels are tested together; for a continuous predictor, the spline is tested. Contrasts are relative to the reference levels warmer / wet / low. *p*-values below 10⁻⁴ are reported as "\<1e-04", the precision limit of mgcv's approximation for smooth terms.

**Table 1. Best-supported marginal GAMs by panel.** Highest-ranked categorical-regime (linear) and continuous-climate (smooth) models, with model fit, climate term with the lowest term-level *p*-value, significant contrasts versus the reference level, EDF of the smooth terms and support label.

```{r comparison-table}
library(webshot2)
extract_gam_stats <- function(model, panel, model_type, model_name,
                              alpha = 0.05, p_floor = 1e-4) {
  
  sm <- summary(model)
  av <- anova(model)                       # term-level Wald tests
  
  ## 1. Climate p-values at TERM level (s(age) excluded)
  # Parametric terms: one joint test per factor (all levels together)
  pt_tab <- av$pTerms.table
  p_par  <- if (!is.null(pt_tab) && NROW(pt_tab) > 0)
    setNames(pt_tab[, "p-value"], rownames(pt_tab)) else numeric(0)
  # Smooth terms
  st_tab <- sm$s.table
  st_cl  <- st_tab[rownames(st_tab) != "s(age)", , drop = FALSE]
  p_smo  <- setNames(st_cl[, "p-value"], rownames(st_cl))
  
  p_all  <- c(p_par, p_smo)
  p_all  <- p_all[!is.na(p_all)]
  # mgcv's smooth p-values are approximations; values below ~1e-4 (or slightly
  # negative, from rounding) are not meaningful beyond "very small"
  p_all  <- pmax(p_all, 0)
  
  climate_p   <- if (length(p_all)) min(p_all) else NA_real_
  climate_trm <- if (length(p_all)) names(p_all)[which.min(p_all)] else NA_character_
  sig_terms   <- names(p_all)[p_all < alpha]
  
  ## 2. Significant factor CONTRASTS vs. the reference level (for interpretation)
  ptab  <- sm$p.table
  contr <- ptab[rownames(ptab) != "(Intercept)", , drop = FALSE]
  contr <- contr[!is.na(contr[, 4]) & contr[, 4] < alpha, , drop = FALSE]
  sig_contr <- if (NROW(contr))
    paste0(rownames(contr), " (", sprintf("%.2f", contr[, 1]),
           ", p=", format.pval(contr[, 4], digits = 2, eps = p_floor), ")",
           collapse = "; ") else if (NROW(ptab) <= 1) "n/a (no categorical terms)" else "none"
  
  data.frame(
    Panel             = panel,
    Type              = model_type,
    Model             = model_name,
    AIC               = round(AIC(model), 3),
    DevianceExplained = round(sm$dev.expl, 3),
    R2_Adj            = round(sm$r.sq, 3),
    ClimateTerm       = climate_trm,
    Climate_p         = climate_p,
    Climate_p_label   = if (is.na(climate_p)) "NA" else
      format.pval(climate_p, digits = 2, eps = p_floor),
    SignificantTerms  = if (length(sig_terms)) paste(sig_terms, collapse = "; ")
    else "Not significant",
    SignificantContrasts = sig_contr,
    EDF_Terms         = paste0(rownames(st_tab), " (EDF=",
                               round(st_tab[, "edf"], 2), ")", collapse = "; "),
    Inference         = ifelse(!is.na(climate_p) && climate_p < alpha,
                               "Supported", "Weak support"),
    stringsAsFactors  = FALSE
  )
}

model_labels <- c(
  M_the = "Thermal regime",
  M_hyd = "Hydroclimatic trends",
  M_hur = "Hurricane regime",
  M_temp = "GMHT anomalies",
  M_dt = "thermal rate of change δT",
  M_hf = "Hurricane frequency",
  M_hav = "Hurricane average"
)

build_panel_summary <- function(best_obj,
                                panel_name,
                                formulas,
                                data){
  
  lin_name <- best_obj$best_linear$Model
  smo_name <- best_obj$best_smooth$Model
  
  lin_model <- gam(
    formulas[[lin_name]],
    data = data,
    family = betar(link="logit"),
    method = "REML"
  )
  
  smo_model <- gam(
    formulas[[smo_name]],
    data = data,
    family = betar(link="logit"),
    method = "REML"
  )
  
  bind_rows(
    
    extract_gam_stats(
      lin_model,
      panel_name,
      "Linear",
      lin_name
    ),
    
    extract_gam_stats(
      smo_model,
      panel_name,
      "Smooth",
      smo_name
    )
    
  )
}

tbl_summary <- bind_rows(
  
  build_panel_summary(
    best_pm,
    "Punta Maroma",
    formulas_mg,
    datos_pm
  ),
  
  build_panel_summary(
    best_fl,
    "Florida",
    formulas_mg,
    datos_fl_std
  ),
  
  build_panel_summary(
    best_bz,
    "Belize",
    formulas_mg,
    datos_bz
  ),
  
  build_panel_summary(
    best_sc,
    "Saint Croix",
    formulas_mg,
    datos_sc
  ),
  
  build_panel_summary(
    best_reg,
    "Regional",
    formulas_mg,
    datos_reg
  )
  
)

tbl_summary <- tbl_summary %>%
  mutate(Model_Label = recode(Model, !!!model_labels)) %>%
  select(Panel, Type, Model, Model_Label, AIC, DevianceExplained, R2_Adj,
         ClimateTerm, Climate_p_label, SignificantContrasts, EDF_Terms, Inference)

gt_summary_m <- tbl_summary %>%
  arrange(factor(Panel, levels = c("Punta Maroma", "Florida", "Belize",
                                   "Saint Croix", "Regional")), Type) %>%
  gt(groupname_col = "Panel") %>%
  fmt_number(columns = c(AIC, DevianceExplained, R2_Adj), decimals = 3) %>%
  cols_label(
    Type                 = "Model Type",
    Model                = "Model",
    Model_Label          = "Climate Driver",
    AIC                  = "AIC",
    DevianceExplained    = "Deviance Explained",
    R2_Adj               = "R² Adj",
    ClimateTerm          = "Climate term",
    Climate_p_label      = "Climate p-value (term)",
    SignificantContrasts = "Significant contrasts vs. reference",
    EDF_Terms            = "EDF Terms",
    Inference            = "Inference"
  ) %>%
  cols_width(Model ~ px(90), SignificantContrasts ~ px(240), EDF_Terms ~ px(260)) %>%
  tab_style(style = list(cell_fill(color = "#7298c2"),
                         cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) %>%
  tab_header(title = md("**Best-supported GAMs by panel**"),
             subtitle = md("*Top Linear and Smooth candidates*")) %>%
  tab_footnote(
    footnote = paste("Climate p-value: joint Wald test of the whole term (all factor",
                     "levels together for categorical regimes; the spline for",
                     "continuous predictors). Contrasts are relative to the",
                     "reference levels warmer / wet / low."),
    locations = cells_column_labels(columns = Climate_p_label)) %>%
  opt_row_striping() %>%
  tab_options(table.font.size = 12, data_row.padding = px(5))

gt_summary_m
gt::gtsave(gt_summary_m, "Table_1.docx")
```

Climate terms reach *p* \< 0.05 (term-level test) in the best marginal models of three panels:

- **Punta Maroma:** s(Hur100_av) (EDF 5.21, *p* \< 10⁻⁴; deviance explained 83.4 %) and Thermal (*p* = 0.012), driven by the LIA contrast (−1.81, *p* = 0.003). The thermal-regime effect is not independent of hurricane activity: added to the selected hurricane model it is no longer significant (*p* = 0.22; Section 4.5).

- **Belize:** s(Hur100_fqcy) (EDF 6.16, *p* = 5.7 × 10⁻⁴) and Thermal (*p* = 0.001), with lower abundance in LIA (−2.52, *p* = 2.9 × 10⁻⁴) and CWP&MWA (−2.10, *p* = 0.001) than under the "warmer" reference.

- **Regional composite:** Thermal (*p* = 0.002), driven by the LIA contrast (−2.32, *p* = 1.9 × 10⁻⁴), and s(Hur100_av) (EDF 3.47, *p* = 0.024).

No climate term reaches *p* \< 0.05 in Florida (Hydro, *p* = 0.12; s(Hur100_av), *p* = 0.25) or Saint Croix (s(dT_raw), *p* = 0.080; Hurricane, *p* = 0.21). Wherever the thermal regime is significant, abundance is lower under LIA (and, in Belize, also under CWP&MWA) than under the "warmer" reference.

In the best-supported marginal models, climate terms were significant in Punta Maroma, Belize and the regional composite, but not in Florida or Saint Croix. For Saint Croix, AIC did not discriminate among candidate predictors (*n* = 18).

### 3.3 Combined GAM effects analysis

To capture compounded climatic effects, we fit a nested sequence of additive GAMs, without interaction terms or random effects.

#### 3.3.1 Candidate models

**Supplementary Table S7. Nested sequence of combined GAMs.** Eleven candidates. M1 = `s(age)`; M2 adds `s(GMHT_CPS0_30N)`; M3 adds `s(dT_raw)`. From M3 the sequence splits into two branches, one adding `s(Hur100_fqcy)` and the other `s(Hur100_av)` (M4); each branch then adds Hydro (M5), Thermal (M6) and Hurricane regime (M7) in turn. The two hurricane metrics never appear in the same model.

```{r combined-formulation}
model_table <- data.frame(
  Model = c("M1", "M2", "M3",
            "M4_fqcy", "M5_fqcy", "M6_fqcy", "M7_fqcy",
            "M4_av", "M5_av", "M6_av", "M7_av"),
  Formula = c(
    "counts_norm_sv ~ s(age)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal + Hurricane",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal + Hurricane"
  ),
  Added_Predictors = c(
    "", "GMHT_CPS0_30N", "dT_raw",
    "Hur100_fqcy", "Hydro", "Thermal", "Hurricane",
    "Hur100_av", "Hydro", "Thermal", "Hurricane"
  ),
  Branch = c(
    "Base", "Base", "Base",
    "Hur100_fqcy", "Hur100_fqcy", "Hur100_fqcy", "Hur100_fqcy",
    "Hur100_av", "Hur100_av", "Hur100_av", "Hur100_av"
  ),
  stringsAsFactors = FALSE
)

tab_S7 <- model_table %>%
  gt() %>%
  tab_header(
    title = "Nested sequence of combined GAMs"
  ) %>%
  cols_label(
    Model = "Model",
    Formula = "Formula",
    Added_Predictors = "Added Predictors",
    Branch = "Branch"
  )

tab_S7
gt::gtsave(tab_S7, "Supplementary_Table_S7.docx")   
```

#### 3.3.2 Fitting procedure (ΔAIC and Akaike weights)

`fit_combined_gams_safe()` fits the eleven candidates to each panel in six steps:

1.  It removes rows with missing or infinite numeric values.

2.  It skips a model if *n* \< 8 or *n* ≤ 3 × (number of smooths).

3.  It sets a common basis dimension for all smooths of a model, *k* = max(3, min(10, ⌊(*n* − 1)/number of smooths⌋)). For the four-smooth models this gives *k* = 7 in Punta Maroma (*n* = 31), 6 in Belize (27), 4 in Saint Croix (18), and 10 in Florida (45) and the regional composite (58). Nested models within a panel therefore do not share the same basis size (e.g. in Punta Maroma M1 uses *k* = 10 and M4 uses *k* = 7).

4.  It fits `gam()` (betar, REML) with `select = TRUE`, which penalizes each smooth's null space so that a term can be shrunk to EDF ≈ 0. (`bam()` is used only when *n* \> 100, which does not occur here.)

5.  If the beta fit fails, it refits the model as quasibinomial; such a model has no AIC and is excluded from the ranking.

6.  It excludes rank-deficient fits, i.e. models whose model-matrix rank is lower than the number of coefficients because two parametric terms are aliased. The names of the excluded models are returned in `$excluded`.

The function returns AIC, deviance explained, adjusted R², total EDF, ΔAIC and Akaike weights for each retained model; ΔAIC and weights are computed over the retained candidates only.

**Excluded candidates.** Because the thermal, hydroclimatic and hurricane regimes are defined on the same regional chronology and each site covers only part of the 60 bins, some regime levels are nested within others at site level:

- In Florida, Belize and Saint Croix, the "medium" hurricane regime occurs only within the CWP&MWA thermal regime, so M7_fqcy and M7_av are excluded.

- In Punta Maroma, the "cooler" thermal regime coincides with the wet and dry hydroclimatic states (all LIA and "warmer" bins are "variable"), so M6 and M7 of both branches are excluded.

- No candidate is excluded in the regional composite.

Consequently, the number of candidates compared is 7 in Punta Maroma, 9 in Florida, Belize and Saint Croix, and 11 in the regional composite.

```{r function-fit-combined}

formulas_cb <- list(
  M1 = counts_norm_sv ~ s(age),
  M2 = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N),
  M3 = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw),
  M4_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy),
  M5_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro,
  M6_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal,
  M7_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal + Hurricane,
  M4_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av),
  M5_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro,
  M6_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal,
  M7_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal + Hurricane
)

count_smooth_terms <- function(formula) {
  lengths(regmatches(paste(deparse(formula, width.cutoff = 500L), collapse = " "),
                      gregexpr("s\\(", paste(deparse(formula, width.cutoff = 500L), collapse = " "))))
}

cap_smooth_k <- function(formula, k_max, env = parent.frame()) {
  f_chr <- paste(deparse(formula, width.cutoff = 500L), collapse = " ")
  f_chr <- gsub("s\\(\\s*([A-Za-z0-9_\\.]+)\\s*\\)",
                paste0("s(\\1, k = ", k_max, ")"), f_chr)
  as.formula(f_chr, env = env)
}

safe_k_for <- function(n_obs, n_smooth_terms) {
  max(3, min(10, floor((n_obs - 1) / max(1, n_smooth_terms))))
}

fit_combined_gams_safe <- function(formulas_cb, data, segment_name) {
  
  data_clean <- data %>%
    mutate(across(where(is.numeric), ~replace(., is.infinite(.), NA))) %>%
    drop_na()
  
  n_obs <- nrow(data_clean)
  
  fits <- lapply(names(formulas_cb), function(name) {
    f <- formulas_cb[[name]]
    n_smooth <- count_smooth_terms(f)
    if (n_obs < 8 || n_obs <= n_smooth * 3) {
      message(paste("Skipping model", name, "for", segment_name,
                    "- insufficient observations (n =", n_obs, ") for",
                    n_smooth, "smooth term(s)"))
      return(NULL)
    }
    
    k_max <- safe_k_for(n_obs, n_smooth)
    f_safe <- if (n_smooth > 0) cap_smooth_k(f, k_max) else f

    # rank-deficient categorical design: excluded before fitting
    # (betar fits on aliased factor designs can abort the R session)
    fv <- intersect(c("Thermal", "Hydro", "Hurricane"), all.vars(f))
    if (length(fv)) {
      X <- model.matrix(reformulate(fv), droplevels(data_clean))
      if (qr(X)$rank < ncol(X)) return("rank-deficient")
    }
    
    tryCatch({
      m <- if (n_obs > 100) {
        bam(f_safe, data = data_clean, family = betar(link = "logit"),
            method = "fREML", select = TRUE, discrete = TRUE)
      } else {
        gam(f_safe, data = data_clean,
            family = betar(link = "logit"), method = "REML",
            select = TRUE)
      }
      environment(m$formula) <- environment()
      m
    }, error = function(e) {
      message(paste("Model", name, "fail in", segment_name,
                    "using betar:", e$message, "→ testing quasibinomial"))
      tryCatch({
        m2 <- gam(f_safe, data = data_clean,
                  family = quasibinomial(link = "logit"), method = "REML",
                  select = TRUE)
        environment(m2$formula) <- environment()
        m2
      }, error = function(e2) {
        message(paste("Model", name, "also failed with quasibinomial in", segment_name, ":", e2$message))
        return(NULL)
      })
    })
  })
  names(fits) <- names(formulas_cb)
  
  fits <- fits[!sapply(fits, is.null)]
  pre_excl <- names(fits)[vapply(fits, is.character, logical(1))]
  fits <- fits[!names(fits) %in% pre_excl]
  
 rank_def <- vapply(fits, function(m) m$rank < length(coef(m)), logical(1))
  excluded <- c(pre_excl, names(fits)[rank_def])
  if (length(excluded)) {
    message("Excluded as rank-deficient in ", segment_name, ": ",
            paste(excluded, collapse = ", "))
    fits <- fits[!rank_def]
  }
  
  stats <- lapply(names(fits), function(name) {
    m <- fits[[name]]
    s <- summary(m)
    data.frame(
      Model = name,
      AIC = tryCatch(AIC(m), error = function(e) NA),
      DevianceExplained = s$dev.expl,
      R2_Adj = s$r.sq,
      EDF = sum(m$edf),
      Segment = segment_name,
      stringsAsFactors = FALSE
    )
  })
  
  stats_df <- bind_rows(stats)
  
  if (nrow(stats_df) > 0) {
    
    stats_df <- stats_df %>%
      mutate(
        deltaAIC = AIC - min(AIC, na.rm = TRUE),
        weight = exp(-0.5 * deltaAIC) /
          sum(exp(-0.5 * deltaAIC), na.rm = TRUE)
      ) %>%
      arrange(AIC)
    
  }
  
  if (nrow(stats_df) > 0) {
    p_delta <- ggplot(stats_df, aes(x = reorder(Model, deltaAIC), y = deltaAIC)) +
      geom_col(fill = "steelblue") +
      geom_hline(yintercept = 2, linetype = "dashed", color = "red") +
      coord_flip() +
      labs(title = paste("ΔAIC by model -", segment_name),
           subtitle = "Dashed line = ΔAIC = 2 (competitive threshold)",
           x = "Model", y = "ΔAIC") +
      theme_minimal() +
      theme(legend.position = "bottom")
    
    p_weight <- ggplot(stats_df, aes(x = reorder(Model, -weight), y = weight)) +
      geom_col(fill = "darkgreen") +
      coord_flip() +
      labs(title = paste("AIC weights by model -", segment_name),
           subtitle = "Relative probability each model is best",
           x = "Model", y = "AIC weight") +
      theme_minimal()
    
    plots <- list(delta = p_delta, weight = p_weight)
  } else {
    plots <- NULL
  }
  
  return(list(fits = fits, stats = stats_df, plots = plots,
              excluded = excluded))  
}

datos_pm <- merged_df_g %>% filter(panel == "Punta Maroma")
results_pm <- fit_combined_gams_safe(formulas_cb, datos_pm, "Punta Maroma")

datos_fl <- merged_df_g %>% filter(panel == "Florida")
results_fl <- fit_combined_gams_safe(formulas_cb, datos_fl, "Florida")

datos_bz <- merged_df_g %>% filter(panel == "Belize")
results_bz <- fit_combined_gams_safe(formulas_cb, datos_bz, "Belize")

datos_sc <- merged_df_g %>% filter(panel == "Saint Croix")
results_sc <- fit_combined_gams_safe(formulas_cb, datos_sc, "Saint Croix")

datos_reg <- merged_df_g %>% filter(panel == "Regional")
results_reg <- fit_combined_gams_safe(formulas_cb, datos_reg, "Regional")

```

#### 3.3.3 Model competition and selected models

We stack the model statistics of all panels (`comparison_cb_full`), store the competitive models (ΔAIC ≤ 2; `competitive_models`, not displayed) and plot ΔAIC and Akaike weights by panel.

```{r suppl-fig5, fig.width=12, fig.height=8}
comparison_cb_full <- bind_rows(results_pm$stats,
                           results_fl$stats,
                           results_bz$stats,
                           results_sc$stats,
                           results_reg$stats)

competitive_models <- comparison_cb_full %>%
  filter(deltaAIC <= 2)%>%
  mutate(Segment = factor(Segment,
                          levels = c("Punta Maroma", "Florida", "Belize", "Saint Croix", "Regional")))

sup_fig5 <- plot_model_selection(comparison_cb_full, panel_var = "Segment", model_levels = names(formulas_cb))
sup_fig5
ggsave("Supplementary_Figure_S5.pdf", plot = sup_fig5, width = 12, height = 8, dpi = 300, device = cairo_pdf)
```

**Supplementary Figure S5. Information-theoretic comparison of the combined GAMs by panel.** (a) ΔAIC of each retained candidate (Supplementary Table S7); the red dashed line marks ΔAIC = 2, and the ΔAIC axis is scaled separately for each panel. (b) Akaike weights, on a common 0–1 scale. Candidates are listed in the same order in every panel; rows without a point are candidates excluded as rank-deficient (Section 3.3.2).

```{r delta-aic}
library(dplyr)
comparison_cb_full %>%
  filter(deltaAIC <= 2) %>%
  group_by(Segment) %>%
  mutate(sum_weight = sum(weight)) %>%
  ungroup() %>%
  arrange(Segment, deltaAIC) %>%
  select(Segment, Model, deltaAIC, weight, sum_weight) %>%
  mutate(across(c(deltaAIC, weight, sum_weight), ~ round(.x, 2)))
```

**Table 2. Best-supported combined GAM per panel** (lowest AIC): model, AIC, Akaike weight, deviance explained, adjusted R², climate terms with term-level *p* \< 0.05 (`s(age)` excluded), factor contrasts with *p*-value \< 0.05 versus reference level, and EDF of every smooth.

- **Punta Maroma — M4_av** (Akaike weight 0.81; deviance explained 82.5 %; adjusted R² 0.72). s(Hur100_av) is significant (EDF 4.07, *p* \< 10⁻⁴), as is s(age) (EDF 0.83, *p* = 0.009). s(GMHT_CPS0_30N) (EDF 1.12) and s(dT_raw) (EDF 0.52) are not significant.

- **Florida — M4_av** (weight 0.32; 28.7 %; adjusted R² 0.18). s(Hur100_av) is the only significant term (EDF 1.59, *p* = 0.009). s(GMHT_CPS0_30N) (EDF 1.49, *p* = 0.078) is retained but not significant, and s(age) and s(dT_raw) are shrunk to EDF ≈ 0. The hurricane effect persists when residual autocorrelation is modelled and when the Florida series is not standardized by source (Section 4.5).

- **Belize — M6_fqcy** (weight 0.64; 79.0 %; adjusted R² 0.60). Significant terms: Thermal (joint test, *p* = 0.002), s(GMHT_CPS0_30N) (EDF 1.71, *p* = 0.033) and s(Hur100_fqcy) (EDF 1.55, *p* = 0.032). Within Thermal, abundance is lower than under the "warmer" reference in LIA (−2.39, *p* = 0.003) and CWP&MWA (−1.77, *p* = 0.024). Hydro is not significant as a term, although its "variable" contrast is (+1.16, *p* = 0.039). s(dT_raw) is shrunk to EDF ≈ 0.

- **Saint Croix — M4_fqcy** (weight 0.31; 58.3 %; adjusted R² 0.43). Significant terms: s(dT_raw) (EDF 1.55, *p* = 0.009) and s(Hur100_fqcy) (EDF 0.78, *p* = 0.032). s(GMHT_CPS0_30N) is shrunk to EDF ≈ 0. Because the Saint Croix response is close to binary, these results are treated as exploratory (Sections 4.3 and 4.5).

- **Regional composite — M6_av** (weight 0.30; 43.9 %; adjusted R² 0.27). The only significant climate term is Thermal (joint test, *p* = 0.025), driven by the LIA contrast (−1.96, *p* = 0.003); s(age) is also significant (EDF 2.75, *p* = 2 × 10⁻⁴). s(Hur100_av) and s(dT_raw) are shrunk to EDF ≈ 0, and s(GMHT_CPS0_30N) (EDF 0.60) is not significant.

```{r Table-2}

get_best_combined <- function(results_obj){
  
  ranking <- results_obj$stats %>%
    arrange(AIC)
  
  winner_name <- ranking$Model[1]
  
  list(
    best_model = winner_name,
    ranking = ranking,
    fit = results_obj$fits[[winner_name]]
  )
  
}

best_pm_cb <- get_best_combined(results_pm)
best_fl_cb <- get_best_combined(results_fl)
best_bz_cb <- get_best_combined(results_bz)
best_sc_cb <- get_best_combined(results_sc)
best_reg_cb <- get_best_combined(results_reg)

extract_combined_stats <- function(best_obj, panel_name,
                                   alpha = 0.05, p_floor = 1e-4) {
  
  fmt_p <- function(p) ifelse(p < p_floor, "<1e-04", sprintf("%.2g", p))
  
  model <- best_obj$fit
  sm    <- summary(model)
  av    <- anova(model)
  
  ## Term-level p-values (s(age) excluded)
  pt_tab <- av$pTerms.table
  p_par  <- if (!is.null(pt_tab) && NROW(pt_tab) > 0)
    setNames(pt_tab[, "p-value"], rownames(pt_tab)) else numeric(0)
  st_tab <- sm$s.table
  st_cl  <- st_tab[rownames(st_tab) != "s(age)", , drop = FALSE]
  p_smo  <- setNames(st_cl[, "p-value"], rownames(st_cl))
  
  p_all  <- c(p_par, p_smo)
  p_all  <- pmax(p_all[!is.na(p_all)], 0)
  sig    <- p_all[p_all < alpha]
  
  sig_terms <- if (length(sig))
    paste0(names(sig), " (p=", fmt_p(sig), ")",
           collapse = "; ") else "Not significant"
  climate_p <- if (length(p_all)) min(p_all) else NA_real_
  
ptab  <- sm$p.table
  contr <- ptab[rownames(ptab) != "(Intercept)", , drop = FALSE]
  contr <- contr[!is.na(contr[, 4]) & contr[, 4] < alpha, , drop = FALSE]
  sig_contr <- if (NROW(contr))
    paste0(rownames(contr), " (", sprintf("%.2f", contr[, 1]),
           ", p=", fmt_p(contr[, 4]), ")",
           collapse = "; ") else if (NROW(ptab) <= 1) "n/a (no categorical terms)" else "none"
  
  data.frame(
    Panel                = panel_name,
    Model                = best_obj$best_model,
    AIC                  = round(AIC(model), 3),
    DevianceExplained    = round(sm$dev.expl, 3),
    R2_Adj               = round(sm$r.sq, 3),
    SignificantTerms     = sig_terms,
    Climate_p            = climate_p,
    Climate_p_label      = if (is.na(climate_p)) "NA" else
      fmt_p(climate_p),
    SignificantContrasts = sig_contr,
    EDF_Terms            = paste0(rownames(st_tab), " (EDF=",
                                  round(st_tab[, "edf"], 2), ")", collapse = "; "),
    stringsAsFactors     = FALSE
  )
}

tbl_combined <- bind_rows(
  
  extract_combined_stats(
    best_pm_cb,
    "Punta Maroma"
  ),
  
  extract_combined_stats(
    best_fl_cb,
    "Florida"
  ),
  
  extract_combined_stats(
    best_bz_cb,
    "Belize"
  ),
  
  extract_combined_stats(
    best_sc_cb,
    "Saint Croix"
  ),
  
  extract_combined_stats(
    best_reg_cb,
    "Regional"
  )
  
)

combined_labels <- c(
  
  M1 =
    "Temporal baseline",
  
  M2 =
    "Background temperature model",
  
  M3 =
    "Thermal forcing model",
  
  M4_fqcy =
    "Thermal–disturbance model (frequency)",
  
  M5_fqcy =
    "Thermal–disturbance–hydroclimate model",
  
  M6_fqcy =
    "Integrated climate model (frequency)",
  
  M7_fqcy =
    "Full climate-regime model (frequency)",
  
  M4_av =
    "Thermal–disturbance model (intensity)",
  
  M5_av =
    "Thermal–disturbance–hydroclimate model",
  
  M6_av =
    "Integrated climate model (intensity)",
  
  M7_av =
    "Full climate-regime model (intensity)"
)

tbl_combined <- tbl_combined %>%
  mutate(
    Model_Label =
      recode(Model,
             !!!combined_labels)
  )

tbl_combined <- tbl_combined %>%
  mutate(
    Weight = c(
      best_pm_cb$ranking$weight[1],
      best_fl_cb$ranking$weight[1],
      best_bz_cb$ranking$weight[1],
      best_sc_cb$ranking$weight[1],
      best_reg_cb$ranking$weight[1]
    )
  )


tbl_combined <- tbl_combined %>%
  select(Panel, Model, Model_Label, AIC, Weight, DevianceExplained, R2_Adj,
         SignificantTerms, SignificantContrasts, EDF_Terms)

gt_combined <- tbl_combined %>%
  arrange(factor(Panel, levels = c("Punta Maroma", "Florida", "Belize",
                                   "Saint Croix", "Regional"))) %>%
  gt() %>%
  fmt_number(columns = c(AIC, DevianceExplained, R2_Adj), decimals = 3) %>%
  fmt_number(columns = Weight, decimals = 2) %>%
  cols_label(
    Panel                = "Panel",
    Model                = "Model",
    Model_Label          = "Climate Model",
    AIC                  = "AIC",
    Weight               = "Akaike Weight",
    DevianceExplained    = "Deviance Explained",
    R2_Adj               = "R² Adj",
    SignificantTerms     = "Significant terms (term-level p)",
    SignificantContrasts = "Significant contrasts vs. reference",
    EDF_Terms            = "EDF Terms"
  ) %>%
  cols_width(Model ~ px(90), Model_Label ~ px(220), SignificantTerms ~ px(220),
             SignificantContrasts ~ px(240), EDF_Terms ~ px(300)) %>%
  tab_style(style = list(cell_fill(color = "#7298c2"),
                         cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) %>%
  tab_header(title = md("**Best-supported combined GAMs by panel**"),
             subtitle = md("*Top multivariate model selected by AIC*")) %>%
  tab_footnote(
    footnote = paste("Term-level p: joint Wald test of the whole term (all factor",
                     "levels together; the spline for continuous predictors);",
                     "s(age) excluded. Contrasts are relative to the reference",
                     "levels warmer / wet / low. Rank-deficient candidates were",
                     "excluded before model selection."),
    locations = cells_column_labels(columns = SignificantTerms)) %>%
  opt_row_striping() %>%
  tab_options(table.font.size = 12, data_row.padding = px(5))

gt_combined
gt::gtsave(gt_combined, "Table_2.docx")   

list(`Punta Maroma` = results_pm$excluded, Florida = results_fl$excluded,
     Belize = results_bz$excluded, `Saint Croix` = results_sc$excluded,
     Regional = results_reg$excluded)
```

The selected model is supported in **Punta Maroma** (Akaike weight 0.81) and **Belize** (0.64): in both panels no other candidate lies within ΔAIC ≤ 2. In **Florida** (0.32), **Saint Croix** (0.31) and the **regional composite** (0.30), the competitive set (ΔAIC ≤ 2; Supplementary Fig. S5) includes several models. In **Florida** it comprises three models of the average-hurricane branch (M4_av, M5_av and M6_av; ΔAIC ≤ 1.96; combined weight 0.59), which differ only in whether the hydroclimatic and thermal regimes are included. In **Saint Croix** it comprises two models of the hurricane-frequency branch (M4_fqcy and M5_fqcy; ΔAIC = 0.95; combined weight 0.51), which differ only in the hydroclimatic regime. In the **regional composite** it comprises five models (M6_av, M6_fqcy, M4_av, M7_av and M7_fqcy; ΔAIC ≤ 1.93; combined weight 0.83), spanning both hurricane metrics and models with and without categorical regimes. In these three panels the data do not discriminate between the selected model and simpler or more complex alternatives, so inference should rest on terms that recur across the competitive set rather than on the single top model (Supplementary Table S8). Across panels, the thermal regime is the only categorical predictor that is significant as a whole term in a selected model (Belize and the regional composite), always with lower abundance under LIA than under the "warmer" reference. The hydroclimatic regime is not significant as a term in any selected model.

```{r competitive-set-terms}

term_pvalues <- function(m) {
  sm <- summary(m); av <- anova(m)
  p_par <- if (!is.null(av$pTerms.table) && NROW(av$pTerms.table) > 0)
             setNames(av$pTerms.table[, "p-value"], rownames(av$pTerms.table)) else numeric(0)
  p_smo <- setNames(sm$s.table[, "p-value"], rownames(sm$s.table))
  pmax(c(p_par, p_smo), 0)
}

results_by_panel <- list("Punta Maroma" = results_pm, "Florida" = results_fl, "Belize" = results_bz,
                         "Saint Croix" = results_sc, "Regional" = results_reg)

comp_terms <- purrr::imap_dfr(results_by_panel, function(r, site) {
  cs <- r$stats %>% filter(deltaAIC <= 2) %>% arrange(deltaAIC)
  purrr::map_dfr(seq_len(nrow(cs)), function(i) {
    p <- term_pvalues(r$fits[[cs$Model[i]]])
    tibble::tibble(Panel = site, Model = cs$Model[i], deltaAIC = cs$deltaAIC[i],
                   Weight = cs$weight[i], Term = names(p), p = p)
  })
})

term_order <- c("Thermal", "Hydro", "Hurricane", "s(age)", "s(GMHT_CPS0_30N)",
                "s(dT_raw)", "s(Hur100_av)", "s(Hur100_fqcy)")

comp_wide <- comp_terms %>%
  mutate(p_label = ifelse(p < 1e-4, "<1e-04", sprintf("%.2g", p)),
         p_label = ifelse(p < 0.05, paste0(p_label, " *"), p_label)) %>%
  select(-p) %>%
  tidyr::pivot_wider(names_from = Term, values_from = p_label, values_fill = "–")
comp_wide <- comp_wide[, c("Panel", "Model", "deltaAIC", "Weight",
                           intersect(term_order, names(comp_wide)))]

tab_S8 <- comp_wide %>%
  gt(groupname_col = "Panel") %>%
  fmt_number(columns = c(deltaAIC, Weight), decimals = 2) %>%
  cols_label(deltaAIC = "ΔAIC", Weight = "Akaike weight") %>%
  tab_header(title = "Term-level p-values across the competitive set (ΔAIC ≤ 2)") %>%
  tab_footnote(footnote = "* p < 0.05; – term not in the model. Joint Wald test per term (all levels of a factor together).")

tab_S8
gt::gtsave(tab_S8, "Supplementary_Table_S8.docx")
```

**Supplementary Table S8. Term-level *p*-values across the competitive set.** For each panel, every combined GAM within ΔAIC ≤ 2 of the best model, with its ΔAIC, Akaike weight and the term-level *p*-value of each term (joint Wald test for factors; \* *p* \< 0.05; – term not in the model).

In **Florida**, s(Hur100_av) is significant in all three competitive models (*p* = 0.001–0.02), whereas neither regime factor is significant in any of them. In **Saint Croix**, s(dT_raw) is significant in both competitive models (*p* = 0.009 in each), whereas s(Hur100_fqcy) is significant only in the top model (*p* = 0.032; *p* = 0.072 in M5_fqcy). In the **regional composite**, the thermal regime is significant in all four competitive models that include it (*p* = 0.0005–0.025), always together with a significant s(age); the fifth competitive model, M4_av, contains no regimes and instead retains a significant s(Hur100_av) (*p* = 0.002) with a non-significant s(age). At the regional scale, the thermal regime (together with the temporal trend) and average hurricane activity are therefore alternative, almost equally supported explanations that these data cannot separate. Two further points qualify the regional thermal-regime result: the LIA contrast rests on four centennial bins (0.4–0.8 ka) at the young end of the record, where it may be confounded with other late-Holocene changes; and because the regimes are contiguous periods, the contrast describes an association with the LIA period rather than a demonstrated response to temperature. **Punta Maroma** and **Belize** each have a single competitive model (Table 2).

#### 3.3.4 Full summaries of the selected combined models

We print the complete `mgcv` summaries of the five models in Table 2.

```{r best-combined}
print_best_combined <- function(best_obj,
                                panel_name){
  
  cat("\n====================================\n")
  cat(panel_name, "\n")
  cat("====================================\n")
  
  cat(
    "\nBEST COMBINED MODEL:",
    best_obj$best_model,
    "\n\n"
  )
  
  print(
    summary(
      best_obj$fit
    )
  )
  
}

print_best_combined(best_pm_cb, "Punta Maroma")
print_best_combined(best_fl_cb, "Florida")
print_best_combined(best_bz_cb, "Belize")
print_best_combined(best_sc_cb, "Saint Croix")
print_best_combined(best_reg_cb, "Regional")
```

## 4. Model Quality Control (combined effects)

To ensure methodological transparency and meet standard practices in GAM-based ecological modeling, we examine the five selected combined models through residual diagnostics from gam.check() (Q–Q plot, residuals vs. linear predictor, residual histogram, response vs. fitted values), plots of all fitted terms, concurvity, and temporal autocorrelation of the residuals. The basis-dimension check (k-index) is reported with `k.check()` below each set of diagnostic plots.

### 4.1 Partial effects of the selected models

The five selected models are collected in `best_models` and their summaries are printed again. Smooth estimates with 95 % confidence intervals are extracted with `gratia::smooth_estimates()`, pooled across panels and plotted by predictor (one panel per smooth). The thermal-regime effects of the selected models that include this factor are added as a sixth panel. To put them on the same baseline as the smooths, which mgcv constrains to sum to zero over the fitted observations, the factor effects are centred in the same way: each regime's coefficient minus their mean weighted by the number of bins in each regime. Centring shifts all regimes by the same amount, so differences between regimes (the contrasts reported in Table 2) are unchanged; 95 % confidence intervals are computed from the full coefficient covariance matrix.

```{r Fig4, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

library(gratia)
library(dplyr)
library(ggplot2)

best_models <- list(
  "Punta Maroma" = best_pm_cb$fit,
  "Florida"      = best_fl_cb$fit,
  "Belize"       = best_bz_cb$fit,
  "Saint Croix"  = best_sc_cb$fit,
  "Regional"     = best_reg_cb$fit
)

summary(best_models[["Punta Maroma"]])
summary(best_models[["Florida"]])
summary(best_models[["Belize"]])
summary(best_models[["Saint Croix"]])
summary(best_models[["Regional"]])

df_all <- purrr::map_df(names(best_models), function(site) {
  smooth_estimates(best_models[[site]], ci_level = 0.95) %>%
    add_confint() %>%
    mutate(site = site)
})

df_all <- df_all %>%
  mutate(site = factor(site, levels = names(custom_colors)))

smooth_predictor_cols <- intersect(
  c("age", "GMHT_CPS0_30N", "dT_raw", "Hur100_av", "Hur100_fqcy"),
  names(df_all)
)

df_long <- df_all %>%
  pivot_longer(cols = all_of(smooth_predictor_cols),
               names_to = "predictor", values_to = "xvar") %>%
  filter(!is.na(xvar))

library(patchwork)

site_levels <- c("Punta Maroma", "Florida", "Belize", "Saint Croix", "Regional")
site_cols   <- custom_colors[site_levels]

df_long <- df_long %>%
  mutate(site = factor(as.character(site), levels = site_levels))

pred_order <- intersect(c("age", "GMHT_CPS0_30N", "dT_raw", "Hur100_av", "Hur100_fqcy"),
                        unique(df_long$predictor))

fig4_theme <- theme_minimal(base_size = 14) +
  theme(axis.title = element_text(size = 12),
        axis.text  = element_text(size = 12),
        strip.text = element_text(size = 13))

# One panel per smooth predictor (same look as the former facet_wrap)
smooth_panels <- lapply(seq_along(pred_order), function(i) {
  pr <- pred_order[i]
  ggplot(filter(df_long, predictor == pr),
         aes(x = xvar, y = .estimate, colour = site, fill = site)) +
    geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.15, colour = NA) +
    geom_line(linewidth = 1.3) +
    facet_wrap(~predictor) +
    scale_colour_manual(values = site_cols, limits = site_levels, name = "Site") +
    scale_fill_manual(values = site_cols, limits = site_levels, name = "Site") +
    labs(x = NULL, y = if (i %% 3 == 1) "Partial effect" else NULL) +
    fig4_theme
})

# centred effects of every categorical regime retained in the selected models
factor_order <- list(Thermal   = c("warmer", "cooler", "CWP&MWA", "LIA"),
                     Hydro     = c("wet", "dry", "no data"),
                     Hurricane = c("low", "moderate", "calm", "enhanced", "high", "quiet", "active"))
factor_title <- c(Thermal = "Thermal regime", Hydro = "Hydroclimate", Hurricane = "Hurricane regime")

centred_effects <- function(fac) {
  sites <- names(best_models)[vapply(best_models, function(m)
    any(grepl(paste0("^", fac), names(coef(m)))), logical(1))]
  if (!length(sites)) return(NULL)
  purrr::map_df(sites, function(site) {
    m    <- best_models[[site]]
    idx  <- grep(paste0("^", fac), names(coef(m)))
    b    <- coef(m)[idx]
    V    <- vcov(m)[idx, idx, drop = FALSE]
    lev  <- levels(droplevels(m$model[[fac]]))
    w    <- as.numeric(table(factor(m$model[[fac]], levels = lev))) / nrow(m$model)
    A    <- t(sapply(lev, function(l) as.numeric(names(b) == paste0(fac, l))))  # level -> coef
    Cm   <- A - matrix(colSums(w * A), nrow(A), ncol(A), byrow = TRUE)
    est  <- as.numeric(Cm %*% b)
    se   <- sqrt(pmax(diag(Cm %*% V %*% t(Cm)), 0))
    tibble::tibble(level = lev, est = est, lo = est - 1.96 * se, hi = est + 1.96 * se,
                   site = site)
  }) %>%
    mutate(level     = factor(level, levels = intersect(factor_order[[fac]], unique(level))),
           site      = factor(site, levels = site_levels),
           predictor = factor_title[[fac]])
}

factor_panels <- purrr::compact(lapply(names(factor_order), function(fac) {
  df <- centred_effects(fac)
  if (is.null(df)) return(NULL)
  ggplot(df, aes(x = level, y = est, colour = site)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_pointrange(aes(ymin = lo, ymax = hi),
                    position = position_dodge(width = 0.5), linewidth = 1, size = 0.6) +
    facet_wrap(~predictor) +
    scale_colour_manual(values = site_cols, limits = site_levels, guide = "none") +
    labs(x = NULL, y = NULL) +
    fig4_theme
}))

p_all <- wrap_plots(c(smooth_panels, factor_panels), ncol = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_all
ggsave("Figure_4.pdf", plot = p_all, width = 12, height = 8, dpi = 300, device = cairo_pdf)
```

**Figure 4** (`Figure_4.pdf`). Partial effects of the selected combined GAMs on the logit scale, coloured by panel. The first five panels show the smooth terms with 95 % confidence bands; terms shrunk to EDF ≈ 0 plot as flat lines at zero (s(dT_raw) in Florida, Belize and the regional composite; s(age) in Florida; s(Hur100_av) in the regional composite). The sixth panel shows the thermal-regime effects in the two selected models that include this factor (Belize and the regional composite), with 95 % confidence intervals. All effects are centred to a weighted mean of zero over the fitted bins, so smooth and factor effects share the same baseline and can be compared directly: zero denotes the model's average contribution of that term, and a regime below zero is associated with lower abundance than average. Differences between regimes equal the contrasts in Table 2.

### 4.2 Concurvity analysis

`concurvity(full = TRUE)` measures, for each term, how well it can be approximated by the other terms of the model (0 = not at all, 1 = completely), reported as worst-case, observed and estimated values. Values above about 0.8 are commonly treated as problematic.

```{r concurvity}

library(reshape2)
library(dplyr)
library(gt)

## concurvity() is undefined when a model has at least as many coefficients as observations
## (its matrix decomposition fails); such models are listed as "not computable".
concurvity_results <- lapply(best_models, function(m)
  tryCatch(concurvity(m, full = TRUE), error = function(e) NULL))
not_comp <- names(concurvity_results)[vapply(concurvity_results, is.null, logical(1))]
if (length(not_comp))
  message("Concurvity not computable (coefficients >= observations): ", paste(not_comp, collapse = ", "))

# Supplementary Table S9 = concurvity as a formatted table
conc_tbl <- purrr::imap_dfr(concurvity_results, function(cm, site) {
  if (is.null(cm))
    return(data.frame(Term = "not computable (coefficients >= observations)", worst = NA_real_,
                      observed = NA_real_, estimate = NA_real_, Panel = site))
  as.data.frame(t(cm)) %>%
    tibble::rownames_to_column("Term") %>%
    mutate(Term = ifelse(Term == "para", "parametric terms", Term),
           Panel = site)
}) %>%
  select(Panel, Term, worst, observed, estimate)

tab_S9 <- conc_tbl %>%
  gt(groupname_col = "Panel") %>%
  fmt_number(columns = c(worst, observed, estimate), decimals = 2) %>%
  cols_label(Term = "Term", worst = "Worst", observed = "Observed", estimate = "Estimate") %>%
  tab_header(title = "Concurvity of the selected combined GAMs",
             subtitle = "0 = term not approximated by the other terms; 1 = fully approximated")

tab_S9
gt::gtsave(tab_S9, "Supplementary_Table_S9.docx")
```

**Supplementary Table S9. Concurvity of the selected combined GAMs** (output above). Observed concurvity is high in every panel for s(age) (0.88–0.98) and high in most panels for s(GMHT_CPS0_30N) (0.88–0.95; 0.77 in Florida). The hurricane smooth reaches 0.76–0.87, except in Saint Croix (0.33), and s(dT_raw) ranges from 0.44 (Florida) to 0.78. In Belize and the regional composite, the only selected models with categorical regimes, the parametric block (the regime factors) is almost entirely reproducible by the smooths (≈ 1.00), as expected when each regime occupies contiguous blocks of time.

The temporal trend, background temperature and categorical regimes are therefore largely interchangeable in these data. How variance is partitioned among them, and hence which term `select = TRUE` retains, is not uniquely determined, and the effects in Table 2 and Fig. 4 should be read as conditional on the other terms rather than as independent contributions. The least confounded terms are the hurricane smooth in Saint Croix and s(dT_raw) in Florida and Saint Croix.

### 4.3 GAM diagnostic plots

For each panel we show the four `gam.check()` residual plots and all model terms (`plot(..., all.terms = TRUE)`, one page: smooths with 95 % bands and rug plots, plus factor effects).

**Supplementary Figure S6.** Residual diagnostics and fitted terms of the selected combined GAMs: (a) Punta Maroma, (b) Florida, (c) Belize, (d) regional composite and (e) Saint Croix.

In all panels the response takes few distinct values, because each bin holds only a few records and abundance is normalized to the panel maximum: 2 distinct values in Saint Croix, 4 in Belize, 6 in Punta Maroma, 19 in Florida and 51 in the regional composite. This discreteness produces the diagonal bands in the plots of residuals against the linear predictor and should not be read as a missing covariate. The basis dimension is adequate for every smooth (EDF well below *k′*), except that s(age) has a k-index below 1 with *p* \< 0.05 in Punta Maroma (0.71) and Florida (0.71). Because the EDF of s(age) is far below *k′* in both (0.83 of 6 and ≈ 0 of 9), increasing *k* would not help; the low k-index instead indicates residual pattern along age, consistent with the positive lag-1 autocorrelation detected in these two panels (Section 4.4). Panel by panel:

- **Punta Maroma:** the Q–Q plot follows the reference line (Shapiro–Wilk on deviance residuals, *p* = 0.14); the maximum bin (response 0.98, fitted 0.91) determines the upper end of the response–fitted relationship.
- **Florida:** the residuals have a right tail (Shapiro–Wilk *p* = 0.004), and fitted values span only 0.21–0.54 against observed values of 0.08–0.99, so the model does not reproduce the extremes, consistent with its modest deviance explained (28.7 %).
- **Belize:** the residuals have heavier tails than normal (Shapiro–Wilk *p* = 0.006), with strong banding because most bins hold a single record (response 0.25).
- **Regional composite:** the residuals are close to normal (Shapiro–Wilk *p* = 0.50), show no pattern against the linear predictor, and the response–fitted relationship is the closest to the 1:1 line.
- **Saint Croix:** with 18 bins and only two distinct response values (one or two records per bin), the residuals fall on two bands; the Q–Q plot is acceptable (Shapiro–Wilk *p* = 0.08), but the response is close to binary: it effectively records whether a bin holds a second colony. With this information content, AIC did not discriminate among the marginal predictors (Supplementary Table S5), and the two terms retained in the combined model are no longer significant when the same model is fitted as a binomial GAM of one versus two records (Section 4.5). The Saint Croix results are therefore treated as exploratory, and no regional conclusion is based on them.

##### 4.3.1 Punta Maroma

```{r pm-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}
par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Punta Maroma"]])))
par(mfrow = c(2,3))
plot(best_models[["Punta Maroma"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Punta Maroma"]])   
```

##### **4.3.2 Florida**

```{r fl-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Florida"]])))
par(mfrow = c(2,3))
plot(best_models[["Florida"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Florida"]])   
```

##### **4.3.3 Belize**

```{r bz-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Belize"]])))
par(mfrow = c(2,3))
plot(best_models[["Belize"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Belize"]])   
```

##### 4.3.4 Regional

```{r reg-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Regional"]])))
par(mfrow = c(2,3))
plot(best_models[["Regional"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Regional"]])   
```

##### 4.3.5 Saint Croix

```{r sc-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Saint Croix"]])))
par(mfrow = c(2,3))
plot(best_models[["Saint Croix"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Saint Croix"]])   
```

```{r figS6-save, warning=FALSE, message=FALSE}
# save Supplementary Figure S6 (a-e) as one multi-page PDF, one page per panel
pdf("Supplementary_Figure_S6.pdf", width = 12, height = 10)
for (site in c("Punta Maroma", "Florida", "Belize", "Regional", "Saint Croix")) {
  par(mfrow = c(2, 2), oma = c(0, 0, 2, 0))
  invisible(capture.output(gam.check(best_models[[site]])))
  mtext(paste(site, "- residual diagnostics"), outer = TRUE, font = 2)
  par(oma = c(0, 0, 2, 0))
  plot(best_models[[site]], shade = TRUE, pages = 1, all.terms = TRUE, rug = TRUE)
  mtext(paste(site, "- fitted terms"), outer = TRUE, font = 2)
}
invisible(dev.off())
```

### 4.4 Residual temporal independence

For each selected model we extract Pearson residuals in bin order (young to old) and compute the lag-1 autocorrelation (φ), the Durbin–Watson statistic (`lmtest::dwtest`, a one-sided test against positive autocorrelation) and the Ljung–Box statistic at lag 10. ACF and PACF of the residuals are shown for all panels in one figure (Supplementary Fig. S7).

```{r residual-independency}

data_list <- list(
  "Belize"       = datos_bz,
  "Florida"      = datos_fl,
  "Punta Maroma" = datos_pm,
  "Regional"     = datos_reg,
  "Saint Croix"  = datos_sc
)

library(mgcv)
library(lmtest)

check_temporal_independence <- function(mod) {
  # Extraer residuales tipo pearson
  res_raw <- residuals(mod, type = "pearson")
  if (is.list(res_raw)) res_raw <- unlist(res_raw)
  res_raw <- as.numeric(res_raw)
  
  # AR(1) coeficiente
  phi <- cor(res_raw[-1], res_raw[-length(res_raw)])
  
  # Durbin–Watson
  dw <- lmtest::dwtest(res_raw ~ 1)
  
  # Ljung–Box
  lb <- Box.test(res_raw, lag = 10, type = "Ljung-Box")
  
  list(
    phi_AR1 = phi,
    DW_stat = dw$statistic,
    DW_pval = dw$p.value,
    LB_stat = lb$statistic,
    LB_pval = lb$p.value,
    res_raw = res_raw
  )
}

best_cb_lookup <- list(
  "Punta Maroma" = best_pm_cb,
  "Florida"      = best_fl_cb,
  "Belize"       = best_bz_cb,
  "Saint Croix"  = best_sc_cb,
  "Regional"     = best_reg_cb
)

best_models_table <- tibble(
  Segment = names(best_models),
  Model   = vapply(names(best_models), function(s) best_cb_lookup[[s]]$best_model, character(1))
)

results_resid <- lapply(seq_len(nrow(best_models_table)), function(i) {
  seg <- best_models_table$Segment[i]
  mname <- best_models_table$Model[i]
  cat("Using model:", mname, "for segment:", seg, "\n")

  mod <- best_models[[seg]]

  diag <- check_temporal_independence(mod)
  
  df <- data.frame(Segment = seg,
                   Model = mname,
                   phi_AR1 = diag$phi_AR1,
                   DW_stat = diag$DW_stat,
                   DW_pval = diag$DW_pval,
                   LB_stat = diag$LB_stat,
                   LB_pval = diag$LB_pval)
  
  attr(df, "res_raw") <- diag$res_raw
  df
})

res_summary <- do.call(rbind, results_resid)

tab_S10 <- res_summary %>%
  gt() %>%
  tab_header(
    title = "Residual independence of the selected combined GAMs"
  ) %>%
  fmt_number(columns = c(phi_AR1, DW_stat, DW_pval, LB_stat, LB_pval),
             decimals = 3)

tab_S10
gt::gtsave(tab_S10, "Supplementary_Table_S10.docx")

acf_df <- function(res, type = "ACF", site) {
  acf_obj <- acf(res, plot = FALSE)
  tibble(
    Lag = acf_obj$lag[, , 1],
    Value = acf_obj$acf[, , 1],
    Type = type,
    Site = site
  )
}

pacf_df <- function(res, type = "PACF", site) {
  pacf_obj <- pacf(res, plot = FALSE)
  tibble(
    Lag = pacf_obj$lag,
    Value = pacf_obj$acf,
    Type = type,
    Site = site
  )
}

df_all <- bind_rows(lapply(seq_len(nrow(best_models_table)), function(i) {
  seg <- best_models_table$Segment[i]
  res_raw <- attr(results_resid[[i]], "res_raw")
  bind_rows(
    acf_df(res_raw, "ACF", seg),
    pacf_df(res_raw, "PACF", seg)
  )
}))

acf_ci <- tibble(Site = best_models_table$Segment,
                 ci   = 1.96 / sqrt(vapply(results_resid, function(r) length(attr(r, "res_raw")), numeric(1))))

plot_acf_pacf <- ggplot(df_all, aes(x = Lag, y = Value, fill = Site)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = custom_colors) +
  geom_hline(yintercept = 0, color = "black") +

  geom_hline(data = acf_ci, aes(yintercept =  ci), linetype = "dashed", color = "grey40") +
  geom_hline(data = acf_ci, aes(yintercept = -ci), linetype = "dashed", color = "grey40") +
  facet_grid(Type ~ Site, scales = "free_y") +
  labs(y = "Correlation", x = "Lag") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

plot_acf_pacf
ggsave("Supplementary_Figure_S7.pdf", plot = plot_acf_pacf, width = 12, height = 6, dpi = 300, device = cairo_pdf)   
```

**Supplementary Table S10. Residual independence of the selected combined GAMs** (output above).

- **Ljung–Box:** no panel shows significant autocorrelation up to lag 10 (*p* = 0.08–0.31).

- **Durbin–Watson:** positive lag-1 autocorrelation is detected in Florida (φ = 0.34, DW = 1.23, *p* = 0.004) and Punta Maroma (φ = 0.30, DW = 1.39, *p* = 0.041).

- **Negative autocorrelation:** Belize (φ = −0.35) and Saint Croix (φ = −0.15) show negative lag-1 correlation, which the one-sided Durbin–Watson test does not detect. A value of −0.35 may indicate over-fitting of short-term structure in Belize.

Two caveats apply. Bins without records are absent, so lag 1 does not always correspond to 100 years; this matters most for Saint Croix (18 of 60 bins) and Belize (27 of 60). With *n* = 18–58, both tests have limited power, especially Ljung–Box at lag 10 with *n* = 18.

**Supplementary Figure S7.** ACF and PACF of the residuals of the selected combined GAMs, by panel. Dashed lines mark the approximate 95 % limits for white noise (±1.96/√*n*), from ±0.26 in the regional composite (*n* = 58) to ±0.46 in Saint Croix (*n* = 18). Lags are counted in bins, which are not always consecutive centuries (see text).

### 4.5 Sensitivity analyses

We test the main results of the selected combined models against four alternative analyses:

1.  **Residual autocorrelation (Florida, Punta Maroma).** The two panels with positive lag-1 autocorrelation (Section 4.4) are refitted as Gaussian GAMMs on the logit of the response, with the same terms and basis dimensions, first with independent errors and then with a continuous-time AR(1) structure (`corCAR1` on age, because bins are not always consecutive). mgcv does not provide a beta family with correlated errors, so this is an approximation used as a check, not a replacement for the beta models.
2.  **Florida source standardization.** The combined and marginal candidate sets are refitted to the Florida series pooled across sources (`datos_fl_norm`, Section 3.1.1).
3.  **Saint Croix response.** Because the response takes only two values, the selected model is refitted as a binomial GAM of one versus two records per bin.
4.  **Thermal regime in Punta Maroma.** The M6 and M7 candidates were excluded as rank-deficient in this panel, so the thermal regime is added directly to the selected hurricane model.

```{r sensitivity-analyses, warning=FALSE, message=FALSE}

library(nlme)

fmt_p2 <- function(p) ifelse(p < 1e-4, "<1e-04", sprintf("%.2g", p))
sm_p   <- function(m, term) { st <- summary(m)$s.table
  if (term %in% rownames(st)) st[term, "p-value"] else NA_real_ }
## terms taken from the selected models, so the checks follow the model selection
hur_term  <- function(m) grep("^s\\(Hur100", rownames(summary(m)$s.table), value = TRUE)[1]
clim_smooths <- function(m) setdiff(rownames(summary(m)$s.table), "s(age)")
fl_hur <- hur_term(best_models[["Florida"]]); pm_hur <- hur_term(best_models[["Punta Maroma"]])
sc_terms <- clim_smooths(best_models[["Saint Croix"]])

## (1) Residual autocorrelation: Gaussian GAMM on the logit of the response,
##     without and with a continuous-time AR(1) (corCAR1), because bins are not always consecutive
ar1_check <- function(site, dat, term) {
  d  <- dat %>% arrange(age) %>%
    mutate(ly = qlogis(pmin(pmax(counts_norm_sv, 1e-4), 1 - 1e-4)))
  f  <- update(formula(best_models[[site]]), ly ~ .)
  m0 <- gamm(f, data = d, method = "REML")
  m1 <- gamm(f, data = d, method = "REML", correlation = corCAR1(form = ~ age))
  lrt <- anova(m1$lme, m0$lme)
  phi <- coef(m1$lme$modelStruct$corStruct, unconstrained = FALSE)
  tibble::tibble(
    Check = "Residual AR(1) (Gaussian GAMM on logit, corCAR1)", Panel = site, Term = term,
    Reference = fmt_p2(sm_p(best_models[[site]], term)),
    Sensitivity = paste0(fmt_p2(sm_p(m0$gam, term)), " (independent); ",
                         fmt_p2(sm_p(m1$gam, term)), " (AR(1))"),
    Note = sprintf("phi = %.2f per 0.1 ka; LRT p = %s; dAIC(AR1 - indep) = %.1f",
                   phi^0.1, fmt_p2(lrt[["p-value"]][2]), AIC(m1$lme) - AIC(m0$lme)))
}

res_fl_pool  <- fit_combined_gams_safe(formulas_cb, datos_fl_norm, "Florida (pooled)")
best_fl_pool <- get_best_combined(res_fl_pool)
marg_pool    <- compare_gams(fit_gams_panel(datos_fl_norm, formulas_mg)) %>% arrange(AIC)
marg_pool_m  <- gam(formulas_mg[[marg_pool$Model[1]]], data = datos_fl_norm,
                    family = betar(link = "logit"), method = "REML")
marg_std_m   <- gam(formulas_mg[[marg_pool$Model[1]]], data = datos_fl_std,
                    family = betar(link = "logit"), method = "REML")
marg_term    <- names(term_pvalues(marg_pool_m))[!grepl("s\\(age\\)", names(term_pvalues(marg_pool_m)))][1]

sc_bin  <- datos_sc %>% mutate(two = as.integer(counts >= 2))
m_sc_bn <- tryCatch(gam(update(formula(best_models[["Saint Croix"]]), two ~ .), data = sc_bin,
               family = binomial, method = "REML", select = TRUE), error = function(e) NULL)

m_pm_th <- gam(update(formula(best_models[["Punta Maroma"]]), . ~ . + Thermal), data = datos_pm,
               family = betar(link = "logit"), method = "REML", select = TRUE)
p_pm_marg_th <- anova(gam(formulas_mg$M_the, data = datos_pm, family = betar(link = "logit"),
                          method = "REML"))$pTerms.table["Thermal", "p-value"]

sens_tbl <- bind_rows(
  ar1_check("Florida", datos_fl_std, fl_hur),
  ar1_check("Punta Maroma", datos_pm, pm_hur),
  tibble::tibble(
    Check = "Florida series pooled across sources (combined models)", Panel = "Florida",
    Term = fl_hur, Reference = fmt_p2(sm_p(best_models[["Florida"]], fl_hur)),
    Sensitivity = fmt_p2(sm_p(best_fl_pool$fit, fl_hur)),
    Note = sprintf("selected model %s (weight %.2f) vs %s; deviance explained %.1f %% vs %.1f %%",
                   best_fl_pool$best_model, best_fl_pool$ranking$weight[1], best_fl_cb$best_model,
                   100 * summary(best_fl_pool$fit)$dev.expl, 100 * summary(best_models[["Florida"]])$dev.expl)),
  tibble::tibble(
    Check = "Florida series pooled across sources (marginal models)", Panel = "Florida",
    Term = marg_term, Reference = fmt_p2(term_pvalues(marg_std_m)[marg_term]),
    Sensitivity = fmt_p2(term_pvalues(marg_pool_m)[marg_term]),
    Note = sprintf("best marginal model on the pooled series: %s (weight %.2f); reference = same model on the source-standardized series",
                   marg_pool$Model[1], marg_pool$weight[1])),
  tibble::tibble(
    Check = "Binomial GAM, one vs two records per bin", Panel = "Saint Croix",
    Term = sc_terms,
    Reference = fmt_p2(sapply(sc_terms, function(tm) sm_p(best_models[["Saint Croix"]], tm))),
    Sensitivity = if (is.null(m_sc_bn)) "not estimable" else fmt_p2(sapply(sc_terms, function(tm) sm_p(m_sc_bn, tm))),
    Note = "same terms as the selected model"),
  tibble::tibble(
    Check = "Thermal regime added to the selected hurricane model", Panel = "Punta Maroma",
    Term = "Thermal", Reference = fmt_p2(p_pm_marg_th),
    Sensitivity = fmt_p2(anova(m_pm_th)$pTerms.table["Thermal", "p-value"]),
    Note = sprintf("reference = marginal thermal-regime model; %s in the extended model: p = %s", pm_hur,
                   fmt_p2(sm_p(m_pm_th, pm_hur))))
)

tab_S11 <- sens_tbl %>%
  gt(groupname_col = "Check") %>%
  cols_label(Reference = "p (reference)", Sensitivity = "p (sensitivity)") %>%
  tab_header(title = "Sensitivity analyses of the selected combined GAMs")

tab_S11
gt::gtsave(tab_S11, "Supplementary_Table_S11.docx")
```

**Supplementary Table S11. Sensitivity analyses of the selected combined GAMs.** For each check, the term tested, its *p*-value in the reference analysis and in the sensitivity analysis, and the relevant model statistics.

- **Residual autocorrelation.** The AR(1) parameter is modest in both panels (φ = 0.35 per century in Florida and 0.33 in Punta Maroma) and does not significantly improve the fit (likelihood-ratio *p* = 0.087 and 0.22). Modelling it removes the lag-1 residual correlation and leaves the hurricane effect significant: *p* = 0.046 in Florida and *p* \< 10⁻⁴ in Punta Maroma. In Florida the Gaussian approximation gives a weaker hurricane effect than the beta model even without autocorrelation (*p* = 0.11), so the Florida evidence is real but moderate.
- **Florida source standardization.** With the pooled series, the same combined model is selected (M4_av), and s(Hur100_av) is stronger (*p* = 0.0003; deviance explained 43.7 % vs 28.7 %). The combined-model inference for Florida is therefore robust to how the sources are scaled. The marginal inference is not: on the pooled series the hurricane regime becomes the dominant marginal predictor (*p* = 0.007, lower abundance under "high" than under "low" hurricane activity), whereas on the source-standardized series it is not significant (*p* = 0.79).
- **Saint Croix.** Neither s(dT_raw) (*p* = 0.056) nor s(Hur100_fqcy) (*p* = 0.059) remains significant in the binomial model, which confirms that the Saint Croix results are exploratory.
- **Punta Maroma.** Added to the selected hurricane model, the thermal regime is not significant (*p* = 0.22), whereas s(Hur100_av) remains highly significant (*p* \< 10⁻⁴). The LIA effect found in the marginal thermal model (*p* = 0.012) is therefore not independent of hurricane activity in this panel.

#### 4.5.1 Local hurricane series in Florida

In the Florida panel the hurricane regime (`Hurricane`) is the local Florida series (`Hur_FL`, Section 1.2) in all the analyses above. Here we compare, as single climate terms added to `s(age)`, the local regime, its continuous version (`FL_pact`, proportion of records active) and the two Schmitt et al. (2025) metrics. The comparison is repeated for bins younger than 3.5 ka, where the local series combines two to six records, because older bins rest on a single record.

```{r florida-local-hurricane, warning=FALSE, message=FALSE}
fl_hur <- datos_fl_std %>%
  left_join(datos_cli %>% transmute(age = interval_max, FL_pact, FL_nrec), by = "age")

fl_hur_models <- list(
  "s(age)"                 = counts_norm_sv ~ s(age),
  "s(age) + Hurricane (local regime)"  = counts_norm_sv ~ s(age) + Hurricane,
  "s(age) + FL_pact (local, linear)"   = counts_norm_sv ~ s(age) + FL_pact,
  "s(age) + s(Hur100_av) (Schmitt)"    = counts_norm_sv ~ s(age) + s(Hur100_av, k = 8),
  "s(age) + s(Hur100_fqcy) (Schmitt)"  = counts_norm_sv ~ s(age) + s(Hur100_fqcy, k = 8))

fl_hur_tab <- purrr::map_dfr(c("0-6 ka" = 6, "0-3.5 ka" = 3.5), function(amax) {
  d <- fl_hur %>% filter(age <= amax + 1e-9) %>% droplevels()
  purrr::imap_dfr(fl_hur_models, function(f, nm) {
    m  <- gam(f, data = d, family = betar(link = "logit"), method = "REML")
    pt <- term_pvalues(m); pt <- pt[!grepl("s\\(age\\)", names(pt))]
    tibble::tibble(Model = nm, n = nrow(d), AIC = AIC(m),
                   `p (climate term)` = if (length(pt)) fmt_p2(pt[1]) else "",
                   `Dev. expl. (%)` = round(100 * summary(m)$dev.expl, 1))
  }) %>% mutate(dAIC = AIC - min(AIC))
}, .id = "Window") %>% mutate(AIC = round(AIC, 1), dAIC = round(dAIC, 1))

knitr::kable(fl_hur_tab, caption = "Florida panel: local vs Schmitt et al. (2025) hurricane predictors")
```

### 4.6 Predictive validation and robustness of the coral–climate associations

Every model includes `s(age)`, which represents the temporal structure of each series: the long-term trend and low-frequency variation in abundance, including taphonomic and sampling effects that cannot be separated from ecological change. Climatic effects are therefore estimated conditional on `s(age)`. All analyses in this section compare the selected models with a reference model, `counts_norm_sv ~ s(age)`, and so ask whether climate adds information beyond the temporal structure of the coral record.

For each panel we evaluate three models: the reference model, the best marginal model (lowest AIC, Section 3.2) and the selected combined model (Section 3.3). Four analyses are run:

1.  **Cross-validation.** Each century is left out in turn, the model is refitted to the remaining centuries and the left-out century is predicted (leave-one-out, LOO). In a second version, the centuries within ±200 years of the left-out one are also removed from the fitting data (h-block), so that the prediction cannot borrow information from immediately adjacent centuries. Predictions are scored by their log-likelihood under the fitted beta distribution, and each model is compared with the reference model through the summed difference (ΔELPD; positive = better than the reference), its standard error and the percentage of centuries predicted better than by the reference.
2.  **Circular-shift test.** The complete 60-bin climatic series (all predictors together) is rotated in time by 6–54 bins (600–5,400 years; 49 shifts) relative to the coral record, while the response and `s(age)` stay fixed. Rotation keeps the shape and autocorrelation of both series and the correlations among predictors, and destroys only their alignment. The statistic is the AIC improvement over the reference model; *p* = (1 + number of shifts with an improvement ≥ the observed one) / (1 + number of shifts). The smallest attainable *p* is 0.02. Shifts of ≤ 500 years are excluded because neighbouring centuries are autocorrelated.
3.  **Drop-one influence.** Each century is removed in turn and the model refitted. For every climatic term we report its *p*-value with all data, the largest *p*-value across the drop-one fits, the century whose removal gives that largest value and the percentage of drop-one fits in which the term stays significant (*p* \< 0.05).
4.  **Former cross-validation (withdrawn).** The five-fold cross-validation of the first revision is re-run to document why it was withdrawn (see below).

During refitting the basis dimension of each smooth is capped with `safe_k_for()` from the number of fitting observations, and further reduced if the total number of coefficients would reach that number. A refit is skipped, and the left-out century counted as not predicted, when the fitting data do not contain every level of a categorical predictor or the categorical design is rank-deficient. For panels with at least 21 observations the full-data fits are identical to those of Sections 3.2–3.3. In Saint Croix (n = 18) the marginal smooths use k = 8 instead of the default 10, which changes the AIC of the marginal model by less than 0.01.

```{r validation-robustness, warning=FALSE, message=FALSE}

fac_vars <- c("Thermal", "Hydro", "Hurricane")

# refit a model to a subset with the k cap and the rank / coefficient guards
fit_val <- function(f, d, combined) {
  ns <- count_smooth_terms(f)
  for (v in intersect(fac_vars, all.vars(f))) d[[v]] <- droplevels(d[[v]])
  fv <- intersect(fac_vars, all.vars(f)); n_par <- 1
  if (length(fv)) {
    X <- tryCatch(model.matrix(reformulate(fv), d), error = function(e) NULL)
    if (is.null(X) || qr(X)$rank < ncol(X)) return(NULL)
    n_par <- ncol(X)
  }
  if (ns > 0) {
    k <- safe_k_for(nrow(d), ns)
    while (k > 3 && n_par + ns * (k - 1) >= nrow(d)) k <- k - 1
    if (n_par + ns * (k - 1) >= nrow(d)) return(NULL)
    f <- cap_smooth_k(f, k)
  }
  tryCatch(suppressWarnings(gam(f, data = d, family = betar(link = "logit"),
                                method = "REML", select = combined)),
           error = function(e) NULL)
}

lpd_val <- function(m, test) {
  if (is.null(m)) return(rep(NA_real_, nrow(test)))
  for (v in intersect(fac_vars, all.vars(formula(m))))
    if (any(!as.character(test[[v]]) %in% as.character(unique(m$model[[v]]))))
      return(rep(NA_real_, nrow(test)))
  mu  <- as.numeric(predict(m, newdata = test, type = "response"))
  phi <- m$family$getTheta(TRUE)
  dbeta(test$counts_norm_sv, mu * phi, (1 - mu) * phi, log = TRUE)
}

pointwise_cv <- function(f, d, combined, buffer = 0) {
  sapply(seq_len(nrow(d)), function(i) {
    drop <- if (buffer > 0) which(abs(d$age - d$age[i]) <= buffer + 1e-9) else i
    lpd_val(fit_val(f, d[-drop, ], combined), d[i, , drop = FALSE])
  })
}

val_data <- list("Punta Maroma" = datos_pm, "Florida" = datos_fl_std, "Belize" = datos_bz,
                 "Saint Croix" = datos_sc, "Regional" = datos_reg)
val_marg <- list("Punta Maroma" = best_pm, "Florida" = best_fl, "Belize" = best_bz,
                 "Saint Croix" = best_sc, "Regional" = best_reg)
val_comb <- list("Punta Maroma" = best_pm_cb, "Florida" = best_fl_cb, "Belize" = best_bz_cb,
                 "Saint Croix" = best_sc_cb, "Regional" = best_reg_cb)

val_models <- purrr::imap_dfr(val_data, function(d, site) tibble::tibble(
  Panel = site,
  Type  = c("Reference", "Marginal", "Combined"),
  Model = c("s(age) only", val_marg[[site]]$ranking$Model[1], val_comb[[site]]$best_model)))

val_formula <- function(type, model) switch(type,
  Reference = counts_norm_sv ~ s(age),
  Marginal  = formulas_mg[[model]],
  Combined  = formulas_cb[[model]])

cv_points <- purrr::pmap_dfr(val_models, function(Panel, Type, Model) {
  d <- as.data.frame(val_data[[Panel]]) %>% arrange(age)
  f <- val_formula(Type, Model); cb <- Type == "Combined"
  tibble::tibble(Panel = Panel, Type = Type, Model = Model, age = d$age,
                 LOO = pointwise_cv(f, d, cb, 0), hblock = pointwise_cv(f, d, cb, 0.2))
})

cv_summary <- cv_points %>%
  group_by(Panel, age) %>%
  mutate(dLOO = LOO - LOO[Type == "Reference"],
         dHB  = hblock - hblock[Type == "Reference"]) %>%
  ungroup() %>% filter(Type != "Reference") %>%
  group_by(Panel, Type, Model) %>%
  summarise(
    n            = n(),
    n_pred       = sum(!is.na(dLOO)),
    dELPD_LOO    = sum(dLOO, na.rm = TRUE),
    SE_LOO       = sqrt(sum(!is.na(dLOO))) * sd(dLOO, na.rm = TRUE),
    dELPD_hblock = sum(dHB, na.rm = TRUE),
    SE_hblock    = sqrt(sum(!is.na(dHB))) * sd(dHB, na.rm = TRUE),
    pct_better   = 100 * mean(dLOO > 0, na.rm = TRUE),
    worst_ka     = age[which.min(dLOO)],
    .groups = "drop")

clim_vars <- c("GMHT_CPS0_30N", "dT_raw", "Hur100_fqcy", "Hur100_av", fac_vars)
cli_full  <- datos_cli %>% arrange(interval_max) %>%
  mutate(age = round(interval_max, 1), Hydro = Hydro_YUC) %>% select(age, all_of(clim_vars), Hydro_FL, Hur_FL)
cli_full_fl <- cli_full %>% mutate(Hydro = Hydro_FL, Hurricane = Hur_FL)
for (v in fac_vars) {
  cli_full[[v]]    <- factor(cli_full[[v]],    levels = levels(merged_df_g[[v]]))
  cli_full_fl[[v]] <- factor(cli_full_fl[[v]], levels = levels(merged_df_g[[v]]))
}
n_bins <- nrow(cli_full)
shifts <- 6:(n_bins - 6)

shift_null <- purrr::pmap_dfr(filter(val_models, Type != "Reference"), function(Panel, Type, Model) {
  d   <- as.data.frame(val_data[[Panel]]) %>% mutate(age = round(age, 1))
  f   <- val_formula(Type, Model); cb <- Type == "Combined"
  ref <- AIC(fit_val(counts_norm_sv ~ s(age), d, FALSE))
  obs <- ref - AIC(fit_val(f, d, cb))
  cli_p <- if (Panel == "Florida") cli_full_fl else cli_full
  null <- sapply(shifts, function(s) {
    rot <- cli_p[((seq_len(n_bins) - 1 + s) %% n_bins) + 1, clim_vars]
    rot$age <- cli_p$age
    m <- fit_val(f, merge(d[, c("age", "counts_norm_sv")], rot, by = "age"), cb)
    if (is.null(m) || m$rank < length(coef(m))) NA_real_ else ref - AIC(m)
  })
  tibble::tibble(Panel = Panel, Type = Type, Model = Model, shift = shifts,
                 observed = obs, null = null)
})

shift_summary <- shift_null %>%
  group_by(Panel, Type, Model) %>%
  summarise(dAIC_vs_ref = first(observed),
            n_shifts    = sum(!is.na(null)),
            null_q95    = quantile(null, 0.95, na.rm = TRUE),
            p_shift     = (1 + sum(null >= observed, na.rm = TRUE)) / (1 + sum(!is.na(null))),
            .groups = "drop")
```

**Supplementary Table S12. Cross-validation and circular-shift test of the selected GAMs.** For each panel, the best marginal model and the selected combined model are compared with the reference model `counts_norm_sv ~ s(age)`. ΔAIC vs s(age): AIC improvement over the reference model with all data. ΔELPD: summed difference in out-of-sample log-likelihood relative to the reference model (positive = better), with its standard error, for leave-one-out (LOO) and leave-one-out with a ±200-year buffer (h-block). Centuries better: percentage of left-out centuries predicted better than by the reference model (LOO). Worst century: the century with the largest loss relative to the reference model (LOO). Shift-test *p*: circular-shift test with 49 shifts of 600–5,400 years (minimum attainable *p* = 0.02).

```{r cv circ table}

tab_S12_df <- cv_summary %>%
  left_join(shift_summary, by = c("Panel", "Type", "Model")) %>%
  mutate(Panel = factor(Panel, levels = site_order),
         Type  = factor(Type, levels = c("Marginal", "Combined")),
         `ΔAIC vs s(age)` = sprintf("%.1f", dAIC_vs_ref),
         `LOO ΔELPD (SE)` = sprintf("%.1f (%.1f)", dELPD_LOO, SE_LOO),
         `h-block ΔELPD (SE)` = sprintf("%.1f (%.1f)", dELPD_hblock, SE_hblock),
         `Centuries better (%)` = sprintf("%.0f", pct_better),
         `Worst century (ka)` = sprintf("%.1f", worst_ka),
         `Shift-test p` = sprintf("%.3f", p_shift)) %>%
  arrange(Panel, Type) %>%
  select(Panel, Type, Model, n, `ΔAIC vs s(age)`, `LOO ΔELPD (SE)`, `h-block ΔELPD (SE)`,
         `Centuries better (%)`, `Worst century (ka)`, `Shift-test p`)

tab_S12 <- tab_S12_df %>%
  gt(groupname_col = "Panel") %>%
  tab_header(title = "Cross-validation and circular-shift test of the selected GAMs",
             subtitle = "Reference model: counts_norm_sv ~ s(age)")
tab_S12
gt::gtsave(tab_S12, "Supplementary_Table_S12.docx")
```

```{r figs8}
figS8 <- shift_null %>%
  filter(!is.na(null)) %>%
  mutate(Panel = factor(Panel, levels = site_order),
         Label = paste0(Type, ": ", Model)) %>%
  ggplot(aes(x = null)) +
  geom_histogram(bins = 20, fill = "grey70", colour = "white") +
  geom_vline(aes(xintercept = observed), colour = "red", linewidth = 0.8) +
  facet_wrap(Panel ~ Label, scales = "free", ncol = 4) +
  labs(x = "AIC improvement over s(age) (shifted climate series)", y = "Number of shifts") +
  theme_minimal(base_size = 10)
figS8
ggsave("Supplementary_Figure_S8.pdf", plot = figS8, width = 12, height = 8, device = cairo_pdf)
```

**Supplementary Figure S8. Circular-shift null distributions.** Histograms of the AIC improvement over `s(age)` obtained when the climatic series is shifted by 600–5,400 years relative to the coral record; the red line marks the observed improvement.

**Supplementary Table S13. Drop-one influence of individual centuries.** For each climatic term of the evaluated models: *p*-value with all data, largest *p*-value when one century is removed, the century whose removal gives that value, and the percentage of drop-one fits in which the term remains significant.

```{r drop one infl}
dropone <- purrr::pmap_dfr(filter(val_models, Type != "Reference"), function(Panel, Type, Model) {
  d  <- as.data.frame(val_data[[Panel]]) %>% arrange(age)
  f  <- val_formula(Type, Model); cb <- Type == "Combined"
  p_full <- term_pvalues(fit_val(f, d, cb))
  p_full <- p_full[names(p_full) != "s(age)"]
  p_drop <- sapply(seq_len(nrow(d)), function(i) {
    m <- fit_val(f, d[-i, ], cb)
    if (is.null(m)) return(setNames(rep(NA_real_, length(p_full)), names(p_full)))
    term_pvalues(m)[names(p_full)]
  })
  p_drop <- matrix(p_drop, nrow = length(p_full), dimnames = list(names(p_full), NULL))
  tibble::tibble(Panel = Panel, Type = Type, Model = Model, Term = names(p_full),
                 p_all = p_full,
                 p_max = apply(p_drop, 1, max, na.rm = TRUE),
                 century_max = d$age[apply(p_drop, 1, which.max)],
                 pct_signif = 100 * rowMeans(p_drop < 0.05, na.rm = TRUE))
})

tab_S13 <- dropone %>%
  mutate(Panel = factor(Panel, levels = site_order),
         Type  = factor(Type, levels = c("Marginal", "Combined")),
         `p (all data)` = fmt_p2(p_all),
         `Largest p (drop-one)` = fmt_p2(p_max),
         `Century removed (ka)` = sprintf("%.1f", century_max),
         `Drop-one fits with p < 0.05 (%)` = sprintf("%.0f", pct_signif)) %>%
  arrange(Panel, Type) %>%
  select(Panel, Type, Model, Term, `p (all data)`, `Largest p (drop-one)`,
         `Century removed (ka)`, `Drop-one fits with p < 0.05 (%)`) %>%
  gt(groupname_col = "Panel") %>%
  tab_header(title = "Drop-one influence of individual centuries on the climatic terms")
tab_S13
gt::gtsave(tab_S13, "Supplementary_Table_S13.docx")
```

- **Cross-validation.** In every panel the climate models predict most left-out centuries better than the reference model (60–77 % in LOO), but the summed scores are dominated by one to three centuries at the extremes of the climatic range, where the smooth terms must extrapolate beyond the values seen during fitting (for example, the most recent century at Punta Maroma carries the highest hurricane value of the series, Hur100_av = 45). The summed differences are of the same order as their standard errors, and their sign changes between LOO and h-block in Punta Maroma. With 18–58 centuries per panel, cross-validation therefore does not discriminate among the candidate models; the fitted models describe the sampled record and have limited predictive transferability beyond its climatic range.
- **Circular-shift test.** The marginal associations in Punta Maroma (hurricane average), Belize (hurricane frequency) and the Regional panel (thermal regime) exceed all 49 shifted alignments (*p* = 0.02). For the combined models *p* = 0.06 (Punta Maroma), 0.044 (Belize) and 0.10 (Regional): a model with more smooth terms also fits the shifted series better, so its observed improvement is less exceptional. In Florida (*p* = 0.24–0.30) and Saint Croix (*p* = 0.26–0.30) the observed associations do not differ from the shifted alignments; they cannot be distinguished from a chance alignment of the two series and are treated as exploratory.
- **Drop-one influence.** The thermal regime remains significant in every drop-one fit in Belize (largest *p* = 0.006) and in 93–98 % of the fits in the Regional panel (largest *p* = 0.06 and 0.08, on removing 0.2 and 0.8 ka). In Punta Maroma s(Hur100_av) remains significant in all drop-one fits of the combined model (largest *p* = 0.03) and in 97 % of those of the marginal model (largest *p* = 0.23, on removing 0.2 ka). The hurricane-frequency term is less stable in Belize: significant in 89 % of the marginal fits but in only 52 % of the combined fits. In Florida and Saint Croix several terms remain significant in most drop-one fits; their weakness lies in the circular-shift test, not in the influence of single centuries.
- **Former cross-validation.** Because the Regional panel contains the site records, every observation entered the training data twice, and the score was computed against the mean of 4–6 test values per panel and fold. When those values are nearly identical, as is common with 2–6 distinct response values per site, the denominator approaches zero and the score becomes very negative or undefined (median of fold means −9.05 in Punta Maroma; undefined folds in 11 and 14 of the 20 fold assignments in Belize and Saint Croix). Its results therefore do not measure the predictive performance of the site models, and the analysis is replaced by the cross-validation above.

```{r}
## ---- (4) former five-fold cross-validation (withdrawn) -------------------------
## One model fitted to all panels pooled (each record enters twice through the Regional
## panel), stratified 5-fold split, score = 1 - SSE/SST with SST around the TEST-fold mean.
old_cv <- function(seed) {
  set.seed(seed)
  dd <- as.data.frame(merged_df_g)
  fo <- integer(nrow(dd))
  for (p in unique(dd$panel)) {
    i <- which(dd$panel == p); fo[i] <- sample(rep(1:5, length.out = length(i)))
  }
  purrr::map_dfr(1:5, function(k) {
    m  <- bam(counts_norm_sv ~ s(age) + s(Hur100_av), data = dd[fo != k, ],
              family = betar(link = "logit"), method = "fREML", discrete = TRUE)
    te <- dd[fo == k, ] %>% mutate(pred = predict(m, ., type = "response"))
    te %>% group_by(panel) %>%
      summarise(score = 1 - sum((counts_norm_sv - pred)^2) /
                  sum((counts_norm_sv - mean(counts_norm_sv))^2),
                n_test = n(), .groups = "drop") %>%
      mutate(seed = seed, fold = k)
  })
}
old_cv_res <- purrr::map_dfr(1:20, old_cv)

old_cv_res %>%
  group_by(panel, seed) %>%
  summarise(mean_score = mean(score), n_test = mean(n_test), .groups = "drop") %>%
  group_by(Panel = panel) %>%
  summarise(`Test points per fold` = round(mean(n_test), 1),
            `Median of fold means` = signif(median(mean_score[is.finite(mean_score)]), 3),
            `Range of fold means`  = paste(signif(range(mean_score[is.finite(mean_score)]), 3),
                                           collapse = " to "),
            `Assignments with an undefined fold` = sum(!is.finite(mean_score)),
            .groups = "drop") %>%
  mutate(Panel = factor(Panel, levels = site_order)) %>% arrange(Panel) %>%
  gt() %>%
  tab_header(title = "Former five-fold cross-validation, re-run over 20 random fold assignments")
```

## Session information

R and package versions used to render this document:

```{r session-info}
sessionInfo()
```
