---
title: "6000-yr APAL-Climate Regional Analysis"
author: "Alexis E Medina Valmaseda"
abstract: |
  This document is the executable record (Quarto/R) of all analyses reported in the
  main text. It uses two input files deposited in the repository: `ages_kernel_reg_updt.xlsx`
  (dated *Acropora palmata* fossils) and `df_GAM130326.xlsx` (centennial climate
  predictors). From these it reproduces:

  1. kernel-density reconstructions of *A. palmata* age distributions for each reef
     site and for all records pooled;
  2. binning of fossil ages into 0.1-ka (centennial) intervals and their alignment
     with climate predictors;
  3. exploratory analyses of predictor correlation and of abundance across
     climate-regime combinations;
  4. beta-regression generalized additive models (GAMs) fitted separately to four reef
     sites (Punta Maroma, Florida, Belize, Saint Croix) and to a regional composite,
     first with one climate predictor at a time (marginal models) and then as a nested
     sequence of multi-predictor models, ranked by AIC and Akaike weights;
  5. diagnostics of the selected models (residual plots, concurvity, residual temporal
     autocorrelation).

  Code, outputs and a factual description of each result are shown together, so that
  every value reported in the main text can be traced to the code that produced it.
  Ecological interpretation is given in the main text.
format: html
execute:
  echo: true
---

# REGIONAL ACROPORA PALMATA ANALYSIS

We import the fossil chronology (`ages_kernel_reg_updt.xlsx`) and retain *A. palmata* records (`group == "palmata"`) dated 0–6 ka.

The function `calc_density_summary()` estimates the age distribution of a set of records by repeated subsampling. In each of 1,000 iterations it draws min(40, *n*) records without replacement and computes a Gaussian kernel density estimate over 0–6 ka (bandwidth 0.09 ka, 512 grid points). Across iterations it stores the pointwise median density and the 2.5–97.5 % and 25–75 % quantile envelopes (`signal_df`).

The median curve is then summarized by 0.1-ka bins. `avg_median` is the mean of the median density over the grid points falling in each bin. `sd_median` and `se_median` describe how much the median curve varies within the bin; they are not measures of sampling uncertainty across iterations.

The function is applied to each site and to all records pooled ("Regional (All spots)"), yielding `signal_all` (density curves) and `centennial_density_all` (centennial summaries).

Two properties of this procedure affect interpretation:

1.  **Envelopes are degenerate for small sites.** Because each iteration draws 40 records without replacement, every iteration uses all records for **Saint Croix** (*n* = 24), and all but one for **Belize** (*n* = 41). The envelopes therefore reflect subsampling variability only for **Punta Maroma** (*n* = 64), **Florida** (*n* = 128) and the **pooled** **regional** set (*n* = 257).
2.  **The pooled curve is weighted toward Florida,** which contributes 128 of the 257 records. It differs from the regional composite used in the GAMs (Section 3.1.1), in which each site is weighted equally.

```{r datos-reg, echo=TRUE}
rm(list=ls(all=TRUE))
gc()
#install.packages("here")
library(here)
here()
library(readxl)
library(dplyr)
library(ggplot2)
library(stargazer)

set.seed(20260930)   # REVISED 2026-09-30: fixed seed; the KDE subsampling below is random

datos_reg <- read_excel(here("ages_kernel_reg_updt.xlsx"))

datos_reg_pal <- datos_reg %>%
  filter(group == "palmata", t_ka >= 0 & t_ka <= 6)

calc_density_summary <- function(df, spot_name){
  df$t_ka <- as.numeric(df$t_ka)
  
signalmx <- matrix(ncol = 512, nrow = 1000)
  for(i in 1:nrow(signalmx)){
    signalmx[i,] <- density(df$t_ka[sample(1:nrow(df), min(40, nrow(df)))],
                            bw = 0.09, from = 0, to = 6)$y
  }
  
  x <- seq(0, 5.999, length.out = 512)   
  
signal_df <- data.frame(
    x      = x,
    q025   = apply(signalmx, 2, quantile, 0.025),
    q975   = apply(signalmx, 2, quantile, 0.975),
    q25    = apply(signalmx, 2, quantile, 0.25),
    q75    = apply(signalmx, 2, quantile, 0.75),
    median = apply(signalmx, 2, median),
    spot   = spot_name
  )
  
 breaks <- seq(0, 6, by = 0.1)
  mids   <- breaks[-length(breaks)] + diff(breaks)/2
  
  centennial_summary <- signal_df %>%
    mutate(interval = cut(x, breaks = breaks, include.lowest = TRUE, right = TRUE)) %>%
    group_by(interval) %>%
    summarise(
      avg_median = mean(median, na.rm = TRUE),
      sd_median  = sd(median, na.rm = TRUE),
      n_obs      = n(),
      .groups = "drop"
    ) %>%
    mutate(
      se_median    = sd_median / sqrt(n_obs),
      interval_mid = mids[as.numeric(interval)],
      spot         = spot_name
    )
  
  return(list(signal_df = signal_df, centennial_summary = centennial_summary))
}

spots <- unique(datos_reg_pal$spot)
results_list <- lapply(spots, function(sp){
  df_spot <- datos_reg_pal %>% filter(spot == sp)
  calc_density_summary(df_spot, sp)
})

results_list <- c(results_list, list(calc_density_summary(datos_reg_pal, "Regional (All spots)")))

signal_all <- bind_rows(lapply(results_list, function(x) x$signal_df))
centennial_density_all <- bind_rows(lapply(results_list, function(x) x$centennial_summary))
```

Each record is assigned to a 0.1-ka bin (right-closed intervals; the first interval includes 0), and the upper bound of the bin (`interval_max`) is stored as its centennial age. A second copy of all records is labelled "Regional", and the two sets are stacked into `centennial_counts_all`. This object is still record-level (one row per dated fossil); counts per bin are computed in later chunks.

```{r cent-count, echo=TRUE}
library(dplyr)

breaks <- seq(0, 6, by = 0.1)

datos_reg_pal <- datos_reg %>%
  filter(group == "palmata", t_ka >= 0 & t_ka <= 6)

centennial_counts_sites <- datos_reg_pal %>%
  mutate(interval_bin = cut(
    t_ka,
    breaks = breaks,
    include.lowest = TRUE,
    right = TRUE
  )) %>%
  mutate(interval_max = breaks[match(interval_bin, levels(interval_bin)) + 1])

centennial_counts_regional <- datos_reg %>%
  filter(group == "palmata", t_ka >= 0 & t_ka <= 6) %>%
  mutate(interval_bin = cut(
    t_ka,
    breaks = breaks,
    include.lowest = TRUE,
    right = TRUE
  )) %>%
  mutate(
    spot = "Regional",
    interval_max = breaks[match(interval_bin, levels(interval_bin)) + 1]
  )

centennial_counts_all <- bind_rows(centennial_counts_sites, centennial_counts_regional)
```

## *1. Data Overview*

We count *A. palmata* records per site and 0.1-ka bin and plot them as stacked bars against the upper bin bound. The site colours defined here (`custom_colors`) are reused in all later figures. Bins without records are absent from the count table and appear as gaps.

**Figure 3a** (`Figure_3a.pdf`). Number of *A. palmata* ages per 100-year interval, by reef site.

```{r count-by-site, echo=TRUE}
library(ggplot2)
library(dplyr)

custom_colors <- c(
  "Punta Maroma" = "#F2941A",   #D3410F           
  "Regional (All spots)" = "#4D4D4D",
  "Regional"             = "#4D4D4D",
  "Saint Croix" = "#CC79A7",         
  "Belize" = "#7D9E5F",    
  "Florida" = "#0072B2"
)

breaks <- seq(0, 6, by = 0.1)

centennial_counts_sites <- datos_reg_pal %>%
  mutate(interval_bin = cut(t_ka, breaks = breaks, include.lowest = TRUE, right = TRUE)) %>%
  group_by(spot, interval_bin) %>%
  summarise(counts = n(), .groups = "drop") %>%
  mutate(interval_max = breaks[match(interval_bin, levels(interval_bin)) + 1],
         spot = factor(spot, levels = c("Belize", "Saint Croix", "Florida", "Punta Maroma")))


pr<-ggplot(centennial_counts_sites,
       aes(x = interval_max, y = counts, fill = spot)) +
  geom_bar(stat = "identity", position = "stack", alpha = 0.85) +
  scale_fill_manual(values = custom_colors) +
  labs(
    x = "Time (ka BP)",
    y = "Acropora palmata  \n fossils/100 years",
    fill = "Site"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  ) +
  scale_x_continuous(
    breaks = seq(0, 6, 0.5),
    minor_breaks = seq(0, 6, 0.5)
  ) +
  coord_cartesian(xlim = c(0, 6))
pr
ggsave("Figure_3a.pdf", plot = pr, width = 8, height = 5, dpi = 300, device = cairo_pdf)
```

### *1.1 Acropora palmata* age densities (Site specific KDE)

This section does three things:

1.  It tabulates the centennial density summaries per site (Supplementary Table S1).

2.  It plots the median density curve of every site, with 95 % and interquartile envelopes for the pooled regional curve only.

3.  It defines the intervals highlighted in Fig. 2 (`rects`): **Punta Maroma** 0.71–1.44 ka; **Belize** 1.99–2.72, 3.68–4.20 and 5.54–6.00 ka; **pooled** **regional** curve 2.20–3.00 and 5.10–5.90 ka. The intervals correspond to local *Acropora palmata* presence gaps reported for **Punta Maroma** (Medina-Valmaseda et al., 2025) and **Belize** (Hubbard et al., 2005, Gischler et al., 2023).

    Note: The panel labels assigned to `signal_all` in this chunk are not used by the plot.

```{r dens, echo=TRUE}
library(dplyr)
library(knitr)
library(scales) 
library(colorspace)
library(ggplot2)
library(gt)

rects<-data.frame(
  xmin = c(0.71, 1.99, 2.2,3.68,5.1,5.54),
  xmax = c(1.44, 2.72,3.0, 4.20,5.9,6.00),
  ymin = -Inf,
  ymax = Inf,
  spot = c("Punta Maroma", "Belize","Regional (All spots)","Belize", "Regional (All spots)","Belize"))

signal_all <- signal_all %>%
  mutate(panel = case_when(
    spot == "Punta Maroma" ~ "a) Local (Punta Maroma)",
    spot == "Florida" ~ "b) Local (Florida)",
    spot %in% c("Saint Croix","Belize") ~ "c) Regional (by zones)",
    spot == "Regional (All spots)" ~ "d) Consolidated Regional"
  ))

centennial_density_all <- centennial_density_all %>%
  mutate(panel = case_when(
    spot == "Punta Maroma" ~ "Punta Maroma",
    spot == "Florida" ~ "Florida",
    spot == "Belize" ~ "Belize",
    spot == "Saint Croix" ~ "Saint Croix",
    spot == "Regional (All spots)" ~ "Regional"
  ))

signal_all <- signal_all %>%
  filter(!is.na(panel)) %>%
  mutate(panel = droplevels(factor(panel)))

centennial_density_all <- centennial_density_all %>%
  filter(!is.na(panel)) %>%
  mutate(panel = droplevels(factor(panel)))
centennial_density_all <- centennial_density_all %>%
  filter(!is.na(interval_mid), interval_mid >= 0, interval_mid <= 6)


tabla_resumen <- centennial_density_all %>%
  group_by(panel) %>%
  summarise(
    mean_density = mean(avg_median, na.rm = TRUE),
    sd_density   = mean(sd_median, na.rm = TRUE),
    se_density   = mean(se_median, na.rm = TRUE),
    min_density  = min(avg_median, na.rm = TRUE),
    min_interval = interval_mid[which.min(avg_median)],
    max_density  = max(avg_median, na.rm = TRUE),
    max_interval = interval_mid[which.max(avg_median)],
    n_intervals  = n(),
    .groups = "drop"
  ) %>%
  arrange(panel)

tab_S1 <- tabla_resumen %>%
  gt() %>%
  tab_header(
    title = "Summary of centennial densities by site",
    subtitle = "average, dispersion, and extremes"
  ) %>%
  fmt_number(columns = c(mean_density, sd_density, se_density,
                         min_density, max_density,
                         min_interval, max_interval),
             decimals = 3)

tab_S1
gt::gtsave(tab_S1, "Supplementary_Table_S1.docx")   # REVISED 2026-09-30
```

**Supplementary Table S1. Centennial kernel-density summaries by site.** For each site: mean over the 60 bins of the centennial mean density (`mean_density`); mean within-bin SD and SE of the median curve; minimum and maximum centennial density, with the bin midpoint (ka) at which each occurs.

Each density integrates to one over the 0–6 ka window, so `mean_density` is close to 1/6 (0.153–0.169) in every panel. The informative quantities are the extremes. Maximum density occurs at 0.15 ka in **Punta Maroma** and **Belize**, at 1.05 ka in **Saint Croix**, and at 4.25 ka in **Florida** and in the **pooled** **regional** curve. Minima fall at 5.95 ka (**Punta Maroma**), 2.35 ka (**Belize**), 2.65 ka (**Saint Croix**), 1.55 ka (**Florida**) and 0.65 ka (**pooled regional**).

**Figure 2** (raw version; `Figure_2.pdf`). Median kernel-density curves of *A. palmata* ages per site and for the pooled records. Shaded envelopes (95 % and interquartile) apply to the pooled curve only. Horizontal bars at the top mark the intervals defined above.

```{r fig2m, fig.width=12, fig.height=6,echo=TRUE}
custom_colors_plot <- custom_colors[names(custom_colors) %in% unique(signal_all$spot)]

p<-ggplot(signal_all, aes(x = x)) +
  geom_rect(data = rects,
    aes(xmin = xmin, xmax = xmax, ymin = 0.88, ymax = 0.98, fill = spot),
      inherit.aes = FALSE,
    alpha = 0.5) +
  geom_ribbon(data = subset(signal_all, spot == "Regional (All spots)"),
    aes(ymin = q025, ymax = q975, fill = spot), alpha = 0.15) +
  geom_ribbon(data = subset(signal_all, spot == "Regional (All spots)"),
    aes(ymin = q25, ymax = q75, fill = spot), alpha = 0.20) +
  geom_line(aes(y = median, color = spot), linewidth = 1.5) +
  scale_color_manual(values = custom_colors, guide = "legend") +
  scale_fill_manual(values = custom_colors, guide = "none") +
  guides(color = guide_legend("Region/Source",
                              override.aes = list(fill = custom_colors_plot)),
        shape = guide_legend("Region/Source")) +
  labs(x = "Time (ka BP)", y = "Density") +
  theme(strip.background = element_blank(),
        panel.background = element_blank(),
        legend.position = "bottom") +
  scale_x_continuous(breaks = seq(0, 6, 0.5),
                     minor_breaks = seq(0, 6, 0.5)) +
  coord_cartesian(xlim = c(0, 6))
p
ggsave("Figure_2.pdf", plot = p, width = 8, height = 5, dpi = 300, device = cairo_pdf)
```

### 1.2 Summary of fossil records by site

We summarize the raw record-level ages by site (**Supplementary Table S2**): number of records, mean, SD, minimum and maximum age. `min_interval` and `max_interval` give the upper bound of the bins containing the youngest and oldest record.

```{r count-summ, echo=TRUE}
library(dplyr)
library(purrr)
library(gt)
library(stringr)

tabla_resumen_counts <- centennial_counts_all %>%
  group_by(spot) %>%
  summarise(
    n_records   = n(),                               
    mean_t_ka   = mean(t_ka, na.rm = TRUE),         
    sd_t_ka     = sd(t_ka, na.rm = TRUE),           
    min_t_ka    = min(t_ka, na.rm = TRUE),          
    max_t_ka    = max(t_ka, na.rm = TRUE),          
    min_interval = interval_max[which.min(t_ka)],   
    max_interval = interval_max[which.max(t_ka)],   
    .groups = "drop"  ) %>%
  arrange(spot)


tab_S2 <- tabla_resumen_counts %>%
  gt() %>%
  tab_header(
    title = "Summary of fossil records by site (raw data)",
    subtitle = "Temporal distribution and dispersion per 0.1 ka bin"
  ) %>%
  fmt_number(
    columns = c(mean_t_ka, sd_t_ka, min_t_ka, max_t_ka,
                min_interval, max_interval),
    decimals = 3
  )

tab_S2
gt::gtsave(tab_S2, "Supplementary_Table_S2.docx")   # REVISED 2026-09-30
```

The dataset comprises 257 dated records: **Florida** 128, **Punta Maroma** 64, **Belize** 41 and **Saint Croix** 24. Mean ages range from 2.26 ka (**Punta Maroma**) to 3.64 ka (**Florida**). **Punta Maroma** records span 0.065–5.46 ka and **Saint Croix** records 0.265–5.93 ka; **Belize** and **Florida** records cover the full 0–6 ka window.

### 1.3 Climatic predictor: import & alignment

We import the centennial climate table (`df_GAM130326.xlsx`) and assign each row the upper bound of its 0.1-ka bin. The mapping is positional: `interval_max` is taken from the order in which interval labels first appear, which is correct only if the file has exactly one row per bin, ordered from youngest to oldest.

Climate values are joined to every fossil record by `interval_max`, and the column is renamed `age` (ka). The resulting `merged_df` is record-level and therefore holds climate values only for bins that contain at least one fossil record.

```{r data-integ, echo=TRUE}
datos_cli <- read_excel(here("df_GAM130326.xlsx"))
datos_cli<-as.data.frame(datos_cli)

library(dplyr)

breaks <- seq(0, 6, by = 0.1)

datos_cli <- datos_cli %>%
  mutate(interval = factor(interval, levels = unique(interval))) %>%
  mutate(interval_max = breaks[match(interval, levels(interval)) + 1])

merged_df <- centennial_counts_all %>%
  left_join(datos_cli, by = "interval_max") %>%
  rename( age = interval_max)
```

### 1.4 Holocene trends in climatic variables

We assemble Fig. 3. Panel (a) is the count plot from Section 1. Below it, four panels show, for each of the 60 centennial bins:

- **(b)** the `GMHT_CPS0_30N` temperature anomaly (^o^C);
- **(c)** accumulated rate of change in thermal anomalies (ẟT,^o^C) calculated from the CPS model of GMST (`dT_raw`);
- **(d)** 100 year average reconstruction of hurricane impacts (`Hur100_av` ) as bars (dark grey where \> 10) and the 100 year TC frequency ( `Hur100_fqcy` ) as a step line;
- **(e)** the nominal thermal, hydroclimatic and hurricane regimes, drawn as contiguous time segments.

<!-- REVISED 2026-09-30: Fig. 3 caption now names the file actually saved by the code -->

<!-- REVISED 2026-09-30: climate panels now cover all 60 bins -->

**Figure 3** (`Figure_3.pdf`). Reconstruction of regional coral abundance and paleoclimate trends. The climate panels (b–e) are drawn from the 60-bin climate table, so they cover the full 0–6 ka interval, including bins without fossil records.

```{r fig3m, fig.width=12, fig.height=10,echo=TRUE}

library(dplyr)
library(tidyr)
library(patchwork)

# REVISED 2026-09-30: climate panels drawn from the 60-bin climate table, so bins without fossils are not gaps
clim_bins <- datos_cli %>% rename(age = interval_max)

p1 <- clim_bins %>%
  group_by(age) %>%
  summarise(GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE)) %>%
  ggplot(aes(x = age, y = GMHT_CPS0_30N, fill = GMHT_CPS0_30N > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "#B8860B", "FALSE" = "#0072B2")) +
  coord_cartesian(ylim = c(-1.2, 1.2), xlim = c(0, 6)) +
  scale_x_continuous(position = "top", breaks = seq(0, 6)) +
  geom_hline(yintercept = 1.2, color = "black", linewidth = 0.4) +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    panel.background = element_blank(),
    legend.position = "none",
    axis.title.x = element_blank()
  ) +
  labs(y = "GMST(°C)\nanomalies")

p2 <- clim_bins %>%
  group_by(age) %>%
  summarise(dT_raw = mean(dT_raw, na.rm = TRUE)) %>%
  ggplot(aes(x = age, y = dT_raw, fill = dT_raw > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9")) +
  coord_cartesian(ylim = c(-1.2, 0.7)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = expression("Rate of Change\nẟT(°C) / 100 yr"))


hur_plot_data <- clim_bins %>%
  group_by(age) %>%
  summarise(
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    Hur100_fqcy_sd = sd(Hur100_fqcy, na.rm = TRUE)
  ) %>%
  arrange(age)

p3 <- ggplot(hur_plot_data, aes(x = age)) +
  geom_col(aes(y = Hur100_av, fill = Hur100_av > 10), alpha = 0.6) +
  geom_step(aes(y = Hur100_fqcy), color = "darkblue", linewidth = 0.8) +
  scale_fill_manual(values = c("TRUE" = "#4D4D4D", "FALSE" = "#999999")) +
  coord_cartesian(ylim = c(0, 50)) +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    panel.background = element_blank(),
    legend.position = "none",
    axis.title.x = element_blank()
  ) +
  labs(y = "Hurricane Average\n(events/100 yr)")

df_segments <- clim_bins %>%
  pivot_longer(cols = c(Thermal, Hydro, Hurricane),
               names_to = "variable",
               values_to = "state") %>%
  arrange(variable, age) %>%
  group_by(variable) %>%
  mutate(change = state != lag(state, default = first(state)),
         segment_id = cumsum(change)) %>%
  group_by(variable, state, segment_id) %>%
  summarise(age_start = min(age),
            age_end   = max(age),
            .groups = "drop") %>%
  mutate(label_x = (age_start + age_end)/2)

df_segments <- df_segments %>%
  group_by(variable) %>%
  arrange(age_start) %>%
  mutate(age_end = lead(age_start, default = max(age_end)))

p4<-ggplot(df_segments) +
  geom_rect(aes(xmin = age_start, xmax = age_end,
                ymin = as.numeric(factor(variable)) - 0.4,
                ymax = as.numeric(factor(variable)) + 0.4,
                fill = state)) +
  geom_text(aes(x = label_x,
                y = as.numeric(factor(variable)),
                label = state),
            color = "black", size = 3) +
  scale_y_continuous(breaks = 1:3,
                     labels = levels(factor(df_segments$variable))) +
scale_fill_manual(values = c(
    "warmer"   = "#DAA520",
    "cooler"   = "#56B4E9",
    "LIA"      = "#56B4E9",
    "CWP&MWA"  = "#DAA520",
    "wet"      = "#999933",
    "dry"      = "#CCBB44",
    "variable" = "#DDCC77",
    "low"      = "#FFFFFF",
    "medium"   = "#CCCCCC",
    "high"     = "#999999"))+
  theme_minimal() +
  theme(legend.position = "none")+
  labs(x = "Age (ka)", y = "Nominal Terms")

Figure_3 <- pr/ p1 / p2 / p3 / p4 + plot_layout(ncol = 1, heights = c(1,1, 1, 1, 1.2))
Figure_3

ggsave("Figure_3.pdf", plot = Figure_3, width = 8, height = 10, dpi = 300, device = cairo_pdf)
```

Before modelling, we inspect pairwise correlations among numeric variables (Section 2.1) and the distribution of abundance across combinations of climate regimes (Section 3.1.3). No predictor is removed on the basis of these analyses. Redundancy is handled instead by never entering the two hurricane metrics in the same model (Section 3.3) and by assessing overlap among smooth terms after fitting (concurvity, Section 4.2).

## **2.** Exploratory Data Analysis

#### 2.1 Variable correlation plot

<!-- REVISED 2026-09-30: correlations now on the 60-bin climate table; regime nesting reported -->

We compute Pearson correlations among the four continuous climate predictors over the 60 centennial bins of the climate table (`datos_cli`), so that each bin counts once regardless of how many fossil records it contains, and cross-tabulate the three categorical regimes.

The only strong correlation is between the two hurricane metrics (Hur100_fqcy and Hur100_av, *r* = 0.79), which is why they are never entered in the same model (Section 3.3). The other correlations are weak to moderate (\|*r*\| ≤ 0.36, the largest between GMHT_CPS0_30N and dT_raw). The categorical regimes, by contrast, are strongly nested: the "cooler" thermal regime occurs only under dry or wet hydroclimate, LIA only under variable hydroclimate, and the "medium" hurricane regime occurs almost only within CWP&MWA (6 of 7 bins). This nesting is the origin of the rank-deficient candidates excluded in Section 3.3.2.

**Supplementary Figure S1.** Pearson correlation matrix of the continuous climate predictors across the 60 centennial bins.

```{r corr, fig.width=8, fig.height=7, echo=TRUE}
library(corrplot)

# REVISED 2026-09-30: correlations among the continuous climate predictors over the 60 centennial bins
# (climate table), so that each bin counts once regardless of its number of fossil records
clim_vars  <- c("GMHT_CPS0_30N", "dT_raw", "Hur100_fqcy", "Hur100_av")
cor_matrix <- cor(datos_cli[, clim_vars])

corrplot(cor_matrix, method = "circle", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black")

pdf("Supplementary_Figure_S1.pdf", width = 8, height = 7)
corrplot(cor_matrix, method = "circle", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black")
invisible(dev.off())

# Cross-tabulation of the categorical regimes over the 60 bins
table(Thermal = datos_cli$Thermal, Hydro = datos_cli$Hydro)
table(Thermal = datos_cli$Thermal, Hurricane = datos_cli$Hurricane)
```

## 3. GAM analysis

### 3.1 Modelling data preparation

This chunk makes three lasting changes to `merged_df`:

1.  **Reference levels.** Thermal, Hydro and Hurricane are converted to factors with reference levels "warmer", "wet" and "low". These levels carry through the aggregation in Section 3.1.1 and define every categorical contrast reported in Sections 3.2 and 3.3.

2.  **Sites only.** Rows are restricted to the four reef sites, which drops the duplicated "Regional" records created in the binning step. The regional composite is rebuilt from the site series in Section 3.1.1.

3.  **Record counter.** A column `counts = 1` is added (one per dated record); it is summed per bin in Section 3.1.1.

```         
The per-site objects (`datos_pm`, `datos_fl`, `datos_bz`, `datos_sc`, `datos_reg`) and `merged_df_g` created in this chunk are intermediate: all of them are redefined in Section 3.1.1 before any model is fitted.
```

```{r data-prepa, echo=TRUE}
library(mgcv)
library(dplyr)
library(patchwork)

merged_df <- merged_df %>%
  mutate(
    Thermal   = relevel(as.factor(Thermal), ref = "warmer"),
    Hydro     = relevel(as.factor(Hydro), ref = "wet"),
    Hurricane = relevel(as.factor(Hurricane), ref = "low")
  )%>%
  filter(spot %in% c("Punta Maroma", "Florida", "Belize", "Saint Croix")) %>%
  mutate(panel = spot)

datos_pm <- merged_df %>% filter(spot == "Punta Maroma")
datos_fl <- merged_df %>% filter(spot == "Florida")
datos_bz <- merged_df %>% filter(spot == "Belize")
datos_sc <- merged_df %>% filter(spot == "Saint Croix")
datos_reg <- merged_df %>%
  filter(spot %in% c("Punta Maroma", "Florida", "Belize", "Saint Croix")) %>%
  mutate(spot = "Regional")

merged_df <- merged_df %>%
   mutate(counts = 1)

merged_df_g <- bind_rows(datos_pm, datos_fl, datos_bz, datos_sc, datos_reg)

```

#### 3.1.1 Response variable normalization

For each site, records are summed per bin (`counts`). Climate predictors are averaged within the bin, which leaves them unchanged because they are constant within a bin; regime labels are taken from the first record in the bin.

`normalize_counts()` divides counts by the site maximum (`counts_norm`, rounded to four decimals). Values exactly 0 or 1 are then transformed as y′ = (y(*n* − 1) + 0.5)/*n*, where *n* is the number of bins in the series (`counts_norm_sv`); all other values are unchanged. Only bins containing at least one record exist in the series, so no zeros occur and the correction affects only the maximum bin (e.g. 0.984 in **Punta Maroma**, *n* = 31).

The response is therefore the relative abundance within bins in which a site is represented. Bins without records are not part of any model: 29 of 60 in **Punta Maroma**, 33 in **Belize**, 42 in **Saint Croix**, 15 in **Florida** and 2 in the **regional pool**.

##### a. Florida source-standardization

Florida combines records from several source datasets (column `source`). Therefore, we first sum counts per source and bin, and further divided by each source's maximum and boundary-corrected. Next we sum the standardized values across sources per bin and normalize as above (`datos_fl_std`). Consequently, each source has equal weight at its maximum, independent of its number of records.

The regional composite used in every model is the per-bin sum of the four site-level `counts_norm_sv` series (**Florida** in its source-standardized form), normalized with the same function (`datos_reg`, *n* = 58 bins). Each site contributes equally; this differs from the **pooled "Regional"** series of Figs. 2 and 3a, in which **Florida** records dominate. The five series are stacked into `merged_df_g`.

```{r norm-stand, echo=TRUE}

normalize_counts <- function(df, panel_name){
  
  df %>%
    mutate(
      counts_norm = round(
        counts / max(counts, na.rm = TRUE),
        4
      ),
      
      counts_norm_sv = ifelse(
        counts_norm %in% c(0,1),
        ((counts_norm * (n() - 1)) + 0.5) / n(),
        counts_norm
      ),
      
      panel = panel_name
    )
  
}

#---------------------------------------------------------
# PUNTA MAROMA
#---------------------------------------------------------

datos_pm <- merged_df %>%
  filter(spot == "Punta Maroma") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Punta Maroma")

#---------------------------------------------------------
# BELIZE
#---------------------------------------------------------

datos_bz <- merged_df %>%
  filter(spot == "Belize") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Belize")

#---------------------------------------------------------
# SAINT CROIX
#---------------------------------------------------------

datos_sc <- merged_df %>%
  filter(spot == "Saint Croix") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Saint Croix")

#---------------------------------------------------------
# FLORIDA
# Independently standardized by source dataset 
#---------------------------------------------------------

datos_fl_std <- merged_df %>%
  filter(spot == "Florida") %>%
  
  group_by(source, age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  
  group_by(source) %>%
  mutate(
    
    counts_std = counts / max(counts, na.rm = TRUE),
    
    counts_std = ifelse(
      counts_std %in% c(0,1),
      ((counts_std * (n() - 1)) + 0.5) / n(),
      counts_std
    )
    
  ) %>%
  ungroup() %>%
  
  group_by(age) %>%
  summarise(
    
    counts = sum(counts_std, na.rm = TRUE),
    
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    
    .groups = "drop"
    
  ) %>%
  normalize_counts("Florida")

#---------------------------------------------------------
# REGIONAL
# build upon normalized datasets
#---------------------------------------------------------

datos_reg <- bind_rows(
  datos_pm,
  datos_fl_std,
  datos_bz,
  datos_sc
) %>%
  group_by(age) %>%
  summarise(
    
    counts = sum(counts_norm_sv, na.rm = TRUE),
    
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    
    .groups = "drop"
    
  ) %>%
  normalize_counts("Regional")

#---------------------------------------------------------
# Ready-to use dataset
#---------------------------------------------------------

merged_df_g <- bind_rows(
  datos_pm,
  datos_fl_std,
  datos_bz,
  datos_sc,
  datos_reg
)
```

##### b. Florida: pooled normalization (sensitivity series)

`datos_fl_norm` applies the same procedure to **Florida** without source standardization, pooling records across sources before normalization. This series is used only in the comparison of Section 3.1.2; no model is fitted to it.

```{r fl-norm-function, echo=TRUE}

normalize_counts <- function(df, panel_name){
  df %>%
    mutate(
      counts_norm = round(counts / max(counts, na.rm = TRUE), 4),
      # aplicar SV si hay 0 o 1
      counts_norm_sv = ifelse(counts_norm %in% c(0,1),
                              ((counts_norm * (n() - 1)) + 0.5) / n(),
                              counts_norm),
      panel = panel_name
    )
}

datos_fl_norm <- merged_df %>%
  filter(spot == "Florida") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Florida")
```

#### 3.1.2 Florida: effect of source standardization

##### a. Comparison table

We join the pooled series (`norm`) and the source-standardized series (`std`) by bin into `fl_compare`.

```{r florida-comparison, echo=TRUE}
fl_compare <- full_join(

  datos_fl_norm %>%
    select(age, counts_norm_sv) %>%
    rename(norm = counts_norm_sv),

  datos_fl_std %>%
    select(age, counts_norm_sv) %>%
    rename(std = counts_norm_sv),

  by = "age"

)
```

##### b. Quick assessment

We summarize the differences (std − norm), list the 20 largest, compute the correlation between the two series, and plot both series and their difference.

```{r florida-comparison-visual, echo=TRUE}

summary(fl_compare$std - fl_compare$norm)

max(abs(fl_compare$std - fl_compare$norm),
    na.rm = TRUE)

fl_compare <- fl_compare %>%
  mutate(
    delta = std - norm
  )

fl_compare %>%
  mutate(
    delta = std - norm,
    abs_delta = abs(delta)
  ) %>%
  arrange(desc(abs_delta)) %>%
  select(age, norm, std, delta) %>%
  slice(1:20)


cor(
  fl_compare$norm,
  fl_compare$std,
  use = "complete.obs"
)


p_fl_a <- ggplot(fl_compare, aes(age)) +
  
  geom_ribbon(
    aes(
      ymin = pmin(norm, std),
      ymax = pmax(norm, std)
    ),
    fill = "grey70",
    alpha = 0.5
  ) +
  
  geom_line(
    aes(y = norm,
        colour = "Original"),
    linewidth = 1
  ) +
  
  geom_line(
    aes(y = std,
        colour = "Standardized"),
    linewidth = 1
  ) +
  
  scale_color_manual(
    values = c(
      "Original" = "black",
      "Standardized" = "red"
    )
  ) +
  
  scale_x_reverse() +
  
  theme_bw() +
  
  labs(
    x = "Age (ka BP)",
    y = "Relative abundance",
    colour = NULL
  )

p_fl_b <- ggplot(
  fl_compare,
  aes(age, delta)
) +
  
  geom_hline(
    yintercept = 0,
    linetype = 2
  ) +
  
  geom_area(
    fill = "steelblue",
    alpha = 0.6
  ) +
  
  scale_x_reverse() +
  
  labs(
    x = "Age (ka BP)",
    y = "Standardized - Original"
  ) +
  
  theme_bw(base_size = 12)

# REVISED 2026-09-30: combine as Supplementary Figure S2 (a, b) and save
library(patchwork)
figS2 <- p_fl_a / p_fl_b + plot_annotation(tag_levels = "a")
figS2
ggsave("Supplementary_Figure_S2.pdf", plot = figS2, width = 8, height = 8, dpi = 300, device = cairo_pdf)
```

The differences have a median of 0.088, a mean of 0.100 and a range of −0.027 to 0.413. The largest occur at 2.5, 2.4 and 3.2 ka. The two series are correlated at *r* = 0.894. Source standardization therefore preserves the overall temporal pattern but raises relative abundance in most bins (fewer than a quarter of the differences are negative), most strongly at 2.4–2.5 and 3.1–3.2 ka. This comparison concerns the response series only; the GAMs were not refitted with the pooled series.

**Supplementary Figure S2.** (a) Pooled (black) and source-standardized (red) Florida series; the grey band marks their difference. (b) Difference std − norm. Age is in ka.

#### **3.1.3 *Acropora palmata* abundance across climate-regime combinations**

We combine the three regime labels into `ClimateCombo` (Thermal_Hydro_Hurricane) and tabulate their frequency per panel (**Supplementary Table S3**; each panel block ends with its number of bins). The regional composite is one of the panels, so its bins are not additional observations. No missing values occur.

We then retain combinations observed in at least three bins in at least one panel. Because the filter is applied across panels, a retained combination can still have *n* = 1–2 in a given panel (e.g. LIA_variable_low, *n* = 2 in Belize; warmer_variable_low, *n* = 1 in Punta Maroma). Three combinations are excluded: CWP&MWA_variable_medium, warmer_wet_low and LIA_variable_medium.

**Supplementary Table S4** gives *n*, mean, SD and range of `counts_norm_sv` for each panel and retained combination. A panel-level summary (`resumen_panel`) is also computed but not shown.

##### Climate combinations (frequencies)

```{r climate-combinations, echo=TRUE}

library(forcats)
eda_counts <- merged_df_g %>%
  
  mutate(
    ClimateCombo = interaction(
      Thermal,
      Hydro,
      Hurricane,
      sep = "_"
    )
  )

eda_counts[!complete.cases(eda_counts), ]

eda_counts$ClimateCombo <- reorder(
  eda_counts$ClimateCombo,
  eda_counts$counts_norm_sv,
  median
)

combo_freq <- eda_counts %>%
  count(panel, ClimateCombo)

# REVISED 2026-09-30: Supplementary Table S3 = frequencies with a total row per panel
combo_freq_tbl <- combo_freq %>%
  mutate(ClimateCombo = as.character(ClimateCombo)) %>%
  group_by(panel) %>%
  group_modify(~ bind_rows(.x, tibble::tibble(ClimateCombo = "Total", n = sum(.x$n)))) %>%
  ungroup()

tab_S3 <- combo_freq_tbl %>%
  gt(groupname_col = "panel") %>%
  tab_header(
    title = "Climate combinations (frequencies)",
    subtitle = "Number of centennial bins per panel and climatic combination"
  ) %>%
  fmt_number(columns = n, decimals = 0) %>%
  cols_label(ClimateCombo = "Climate combination", n = "Bins") %>%
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_body(rows = ClimateCombo == "Total"))

tab_S3
gt::gtsave(tab_S3, "Supplementary_Table_S3.docx")
```

##### Climate-combination summary

```{r climate-combination-summary, echo=TRUE}
combos_validos <- combo_freq %>%
  filter(n >= 3) %>%
  pull(ClimateCombo) %>%
  as.character()

invisible(
  combo_freq %>%
    filter(n >= 3)
)

invisible(
  combo_freq %>%
    filter(n < 3)
)

datos_filtrados <- eda_counts %>%
  filter(
    ClimateCombo %in% combos_validos
  )

datos_filtrados <- datos_filtrados %>%
  group_by(panel) %>%
  mutate(
    ClimateCombo = fct_reorder(
      ClimateCombo,
      counts_norm_sv,
      .fun = median
    )
  ) %>%
  ungroup()

resumen_panel <- datos_filtrados %>%
  group_by(panel) %>%
  summarise(
    n_obs = n(),
    mean_abund = mean(counts_norm_sv, na.rm = TRUE),
    sd_abund = sd(counts_norm_sv, na.rm = TRUE),
    min_abund = min(counts_norm_sv, na.rm = TRUE),
    max_abund = max(counts_norm_sv, na.rm = TRUE),
    .groups = "drop"
  )

resumen_panel_combo <- datos_filtrados %>%
  group_by(
    panel,
    ClimateCombo
  ) %>%
  summarise(
    n_obs = n(),
    mean_abund = mean(counts_norm_sv, na.rm = TRUE),
    sd_abund = sd(counts_norm_sv, na.rm = TRUE),
    min_abund = min(counts_norm_sv, na.rm = TRUE),
    max_abund = max(counts_norm_sv, na.rm = TRUE),
    .groups = "drop"
  )

tab_S4 <- resumen_panel_combo %>%
  gt() %>%
  tab_header(
    title = "Climate-combination summary",
    subtitle = "Standardized Acropora abundance by climatic regime"
  ) %>%
  fmt_number(
    columns = c(
      mean_abund,
      sd_abund,
      min_abund,
      max_abund
    ),
    decimals = 3
  )

tab_S4
gt::gtsave(tab_S4, "Supplementary_Table_S4.docx")   # REVISED 2026-09-30
```

##### Coral-Climate Interactions

The figure bellow summarizes the distribution of standardized *A. palmata* abundance across climatic-regime combinations. Climatic states are ordered by median abundance within panels, allowing visual comparison of abundance patterns and variability among environmental conditions.

```{r figS3, fig.width=12, fig.height=10,echo=TRUE}

figS3 <- ggplot(
  datos_filtrados,
  aes(
    x = fct_reorder(
      ClimateCombo,
      counts_norm_sv,
      .fun = median
    ),
    y = counts_norm_sv,
    fill = panel
  )
) +
  
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.7
  ) +
  
  geom_point(
    position = position_jitter(width = 0.12, seed = 1),   # REVISED 2026-09-30: reproducible jitter
    size = 2,
    alpha = 0.6,
    shape = 21,
    colour = "black"
  ) +
  
  facet_wrap(
    ~ panel,
    scales = "free_y",
    ncol = 2
  ) +
  
  scale_fill_manual(
    values = custom_colors
  ) +
  
  theme_bw() +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 12
    ),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(
      size = 14,
      face = "bold"
    ),
    axis.title.y = element_text(
      size = 14,
      face = "bold"
    ),
    strip.text = element_text(
      size = 14,
      face = "bold"
    ),
    legend.position = "none"
  ) +
  
  labs(
    x = "Climate combinations (ordered by median abundance)",
    y = "Standardized abundance (counts_norm_sv)"
  )

figS3
ggsave("Supplementary_Figure_S3.pdf", plot = figS3, width = 8, height = 5, dpi = 300, device = cairo_pdf)
```

**Supplementary Figure S3. Standardized *A. palmata* abundance under Holocene climate-regime combinations.** Boxplots of `counts_norm_sv` by combination of thermal, hydroclimatic and hurricane regime, per panel; points are individual centennial bins. Combinations are ordered by their median over all panels pooled, so the order is the same in every facet.

The highest means occur under warmer_variable_high (**Belize** 0.87, *n* = 2; **Punta Maroma** 0.60, *n* = 3; **regional composite** 0.75, *n* = 3; **Saint Croix** 0.97, *n* = 1), except in **Florida**, where this combination has one of the lowest values (0.13, *n* = 2). The lowest means occur under **LIA**\_variable_low (**Florida** 0.10, *n* = 2; **regional composite** 0.18, *n* = 3). Most bins fall under cooler_wet_high (*n* = 3–20 per panel) and cooler_dry_low (*n* = 4–11), with mean abundance of 0.14–0.82 depending on site. The extreme combinations are represented by three or fewer bins, so these patterns are descriptive only.

### 3.2. Marginal GAM effects analysis

Relative abundance is continuous and bounded in (0, 1), so we fit GAMs with a beta distribution and logit link (`mgcv::gam`, family `betar`). Smoothing parameters are estimated by REML. Continuous predictors enter as thin-plate regression splines (mgcv default, basis dimension *k* = 10).

Climate regimes enter as factors with reference levels "warmer" (Thermal), "wet" (Hydro) and "low" (Hurricane). Their coefficients are contrasts on the logit scale relative to the reference, conditional on the other terms. Every model includes a smooth of age, `s(age)`, which absorbs temporal structure not attributed to the climate predictor.

Modelling proceeds in two stages: marginal models, with `s(age)` plus one climate predictor (this section), and combined models, a nested sequence with two branches, one per hurricane metric (Section 3.3).

Candidates are ranked by AIC, ΔAIC and Akaike weights; models within ΔAIC ≤ 2 of the best are considered competitive. Marginal models are fitted without `select = TRUE` and with *k* = 10, whereas combined models use `select = TRUE` and a reduced *k*; AIC values are therefore not comparable between the two stages.

In this document, "linear" models are those whose climate term is a categorical factor (a parametric term), and "smooth" models those whose climate term is a continuous predictor fitted as a spline.

#### 3.2.1 Candidate models

`formulas_mg` defines seven marginal candidates, each consisting of `s(age)` plus one of: `s(GMHT_CPS0_30N)` (M_temp), `s(dT_raw)` (M_dt), `s(Hur100_fqcy)` (M_hf), `s(Hur100_av)` (M_hav), Thermal (M_the), Hydro (M_hyd) or Hurricane (M_hur). `fit_gams_panel()` fits all candidates to one panel, and `compare_gams()` returns AIC, ΔAIC and Akaike weights.

```{r setup-formulas, echo=TRUE}
formulas_mg <- list(
  M_temp = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N),
  M_dt   = counts_norm_sv ~ s(age) + s(dT_raw),          
  M_hf   = counts_norm_sv ~ s(age) + s(Hur100_fqcy),
  M_hav  = counts_norm_sv ~ s(age) + s(Hur100_av),
  M_the  = counts_norm_sv ~ s(age) + Thermal,
  M_hyd  = counts_norm_sv ~ s(age) + Hydro,
  M_hur  = counts_norm_sv ~ s(age) + Hurricane
)

fit_gams_panel <- function(data, formulas) {
  lapply(formulas, function(f)
    gam(f, data = data, family = betar(link = "logit"), method = "REML"))
}

compare_gams <- function(fits) {
  stats <- lapply(fits, AIC)
  df <- data.frame(Model = names(stats), AIC = unlist(stats))
  df <- df %>%
    mutate(deltaAIC = AIC - min(AIC),
           weight = exp(-0.5 * deltaAIC) / sum(exp(-0.5 * deltaAIC)))
  return(df)
}
```

#### 3.2.2 Model competition (ΔAIC and Akaike weights)

We fit the seven candidates to each panel of `merged_df_g` and plot ΔAIC and Akaike weights by panel.

<!-- REVISED 2026-09-30: new layout shared with Supplementary Figure S5 -->

**Supplementary Figure S4. Information-theoretic comparison of the marginal GAMs by panel.** (a) ΔAIC of each candidate (Supplementary Table S5); the red dashed line marks ΔAIC = 2, and the ΔAIC axis is scaled separately for each panel. (b) Akaike weights, on a common 0–1 scale. Candidates are listed in the same order in every panel.

```{r marginal-results, echo=TRUE, fig.width=12, fig.height=7}
panels <- unique(merged_df_g$panel)

results <- lapply(panels, function(p) {
  
  df_panel <- merged_df_g %>%
    filter(panel == p)
  
  if(nrow(df_panel) < 3 ||
     all(is.na(df_panel$counts_norm_sv)))
    return(NULL)
  
  fits <- fit_gams_panel(
    df_panel,
    formulas_mg
  )
  
  stats <- compare_gams(fits)
  
  stats$panel <- p
  
  stats
  
})

results <- Filter(
  Negate(is.null),
  results
)

stats_all <- bind_rows(results)

library(patchwork)

# REVISED 2026-09-30: one plotting function for Supplementary Figures S4 (marginal) and S5 (combined),
# so both share layout, panel order and a fixed candidate order. Lollipops instead of bars, so that
# the best model (ΔAIC = 0) is visible as a point and excluded candidates are simply absent.
site_order <- c("Punta Maroma", "Florida", "Belize", "Saint Croix", "Regional")

plot_model_selection <- function(stats, panel_var, model_levels) {
  d <- stats %>%
    mutate(Panel = factor(.data[[panel_var]], levels = site_order),
           Model = factor(Model, levels = rev(model_levels)))
  base <- function(yvar, ylab) {
    ggplot(d, aes(x = Model, y = .data[[yvar]], colour = Panel)) +
      geom_segment(aes(xend = Model, y = 0, yend = .data[[yvar]]), linewidth = 0.8) +
      geom_point(size = 2.2) +
      coord_flip() +
      scale_x_discrete(drop = FALSE) +
      scale_colour_manual(values = custom_colors[site_order], guide = "none") +
      labs(x = NULL, y = ylab) +
      theme_minimal(base_size = 12) +
      theme(strip.text = element_text(face = "bold"),
            panel.grid.minor = element_blank())
  }
  p_d <- base("deltaAIC", "ΔAIC") +
    geom_hline(yintercept = 2, linetype = "dashed", colour = "red") +
    facet_wrap(~Panel, nrow = 1, drop = FALSE, scales = "free_x")
  p_w <- base("weight", "Akaike weight") +
    facet_wrap(~Panel, nrow = 1, drop = FALSE) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1))
  (p_d / p_w) + plot_annotation(tag_levels = "a")
}

sup_fig4 <- plot_model_selection(stats_all, panel_var = "panel", model_levels = names(formulas_mg))
sup_fig4
ggsave("Supplementary_Figure_S4.pdf", plot = sup_fig4, width = 12, height = 7, dpi = 300, device = cairo_pdf)
```

#### 3.2.3 Marginal GAM model ranking

**Supplementary Table S5. Marginal GAM ranking by panel.** For each panel, the seven candidates are ordered by AIC, with ΔAIC, Akaike weight and rank. (`get_best_models()` refits the same models as Section 3.2.2 on the same data.)

```{r best-gams, echo=TRUE}

linear_models <- c(
  "M_the",
  "M_hyd",
  "M_hur"
)

smooth_models <- c(
  "M_temp",
  "M_dt",
  "M_hf",
  "M_hav"
)

get_best_models <- function(data, formulas){
  
  fits <- fit_gams_panel(data, formulas)
  
  ranking <- compare_gams(fits) %>%
    arrange(AIC) %>%
    mutate(
      Rank = row_number()
    )
  
  best_linear <- ranking %>%
    filter(Model %in% linear_models) %>%
    slice_min(AIC, n = 1)
  
  best_smooth <- ranking %>%
    filter(Model %in% smooth_models) %>%
    slice_min(AIC, n = 1)
  
  list(
    fits = fits,
    ranking = ranking,
    best_linear = best_linear,
    best_smooth = best_smooth
  )
  
}

best_pm <- get_best_models(datos_pm, formulas_mg)
best_fl <- get_best_models(datos_fl_std, formulas_mg)
best_bz <- get_best_models(datos_bz, formulas_mg)
best_sc <- get_best_models(datos_sc, formulas_mg)
best_reg <- get_best_models(datos_reg, formulas_mg)


tbl_models <- bind_rows(
  
  best_pm$ranking %>%
    mutate(Panel = "Punta Maroma"),
  
  best_fl$ranking %>%
    mutate(Panel = "Florida"),
  
  best_bz$ranking %>%
    mutate(Panel = "Belize"),
  
  best_sc$ranking %>%
    mutate(Panel = "Saint Croix"),
  
  best_reg$ranking %>%
    mutate(Panel = "Regional")
  
)

gt_ranking <- tbl_models %>%
  
  gt(groupname_col = "Panel") %>%
  
  fmt_number(
    columns = c(AIC, deltaAIC, weight),
    decimals = 3
  ) %>%
  
  cols_label(
    Rank = "Rank",
    Model = "Model",
    AIC = "AIC",
    deltaAIC = "ΔAIC",
    weight = "Akaike Weight"
  ) %>%
  
  tab_header(
    title = md("**GAM model selection by panel**")
  )

gt_ranking
gt::gtsave(gt_ranking, "Supplementary_Table_S5.docx")   # REVISED 2026-09-30
```

In **Punta Maroma**, M_hav dominates (weight 0.993; all other models ΔAIC \> 9). In **Belize**, M_hf (weight 0.66) and M_the (ΔAIC 1.93, weight 0.25) are competitive. In the **regional composite**, M_the ranks first (weight 0.68), followed by M_hav at ΔAIC 2.70. In **Florida**, three models lie within ΔAIC ≤ 2 (M_hyd, M_hav, M_temp; weights 0.34, 0.24, 0.16), so no single predictor is clearly supported. In **Saint Croix**, six of seven models lie within ΔAIC ≤ 1.6 (weights 0.10–0.21); with *n* = 18, AIC does not discriminate among predictors.

#### 3.2.4 Best-supported marginal GAMs

**Supplementary Table S6.** Best categorical ("linear") and best continuous ("smooth") marginal model per panel, selected by lowest AIC within each group.

```{r best-gams-winners, echo=TRUE}
tbl_winners <- bind_rows(
  
  data.frame(
    Panel = "Punta Maroma",
    Best_Linear = best_pm$best_linear$Model,
    Best_Smooth = best_pm$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Florida",
    Best_Linear = best_fl$best_linear$Model,
    Best_Smooth = best_fl$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Belize",
    Best_Linear = best_bz$best_linear$Model,
    Best_Smooth = best_bz$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Saint Croix",
    Best_Linear = best_sc$best_linear$Model,
    Best_Smooth = best_sc$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Regional",
    Best_Linear = best_reg$best_linear$Model,
    Best_Smooth = best_reg$best_smooth$Model
  )
  
)

gt_winners <- tbl_winners %>%
  gt() %>%
  cols_label(
    Panel = "Panel",
    Best_Linear = "Best Linear",
    Best_Smooth = "Best Smooth"
  ) %>%
  tab_header(
    title = md("**Best-supported linear and smooth GAMs**")
  )

gt_winners
gt::gtsave(gt_winners, "Supplementary_Table_S6.docx")   # REVISED 2026-09-30
```

We refit these models and print their full `mgcv` summaries.

```{r best-gams-summary, echo=TRUE}

print_best_models <- function(best_obj,
                              panel_name,
                              formulas,
                              data){
  
  lin_name <- best_obj$best_linear$Model
  smo_name <- best_obj$best_smooth$Model
  
  cat("\n========================================\n")
  cat(panel_name,"\n")
  cat("========================================\n")
  
  cat("\nBEST LINEAR MODEL:", lin_name, "\n\n")
  
  print(
    summary(
      gam(
        formulas[[lin_name]],
        data = data,
        family = betar(link="logit"),
        method = "REML"
      )
    )
  )
  
  cat("\n----------------------------------------\n")
  
  cat("\nBEST SMOOTH MODEL:", smo_name, "\n\n")
  
  print(
    summary(
      gam(
        formulas[[smo_name]],
        data = data,
        family = betar(link="logit"),
        method = "REML"
      )
    )
  )
  
}

print_best_models(best_pm,
                  "Punta Maroma",
                  formulas_mg,
                  datos_pm)

print_best_models(best_fl,
                  "Florida",
                  formulas_mg,
                  datos_fl_std)

print_best_models(best_bz,
                  "Belize",
                  formulas_mg,
                  datos_bz)

print_best_models(best_sc,
                  "Saint Croix",
                  formulas_mg,
                  datos_sc)

print_best_models(best_reg,
                  "Regional",
                  formulas_mg,
                  datos_reg)

```

<!-- REVISED 2026-09-30: Table 1 now uses the term-level (joint Wald) test -->

We then extract, for each of these models: AIC, deviance explained and adjusted R²; the climate term with the lowest term-level *p*-value (`s(age)` excluded) and that *p*-value; the factor contrasts with *p* \< 0.05; the EDF of each smooth; and a support label ("Supported" if the term-level climate *p* \< 0.05, otherwise "Weak support"). Term-level *p*-values come from a joint Wald test of the whole term (`anova()`): for a categorical regime, all levels are tested together; for a continuous predictor, the spline is tested. Contrasts are relative to the reference levels warmer / wet / low. *p*-values below 10⁻⁴ are reported as "\<1e-04", the precision limit of mgcv's approximation for smooth terms.

<!-- REVISED 2026-09-30: removed the old underlined Table 1 caption, superseded by the caption below -->

**Table 1. Best-supported marginal GAMs by panel.** Highest-ranked categorical-regime (linear) and continuous-climate (smooth) models, with model fit, climate term with the lowest term-level *p*-value, significant contrasts versus the reference level, EDF of the smooth terms and support label.

```{r comparison-table, echo=TRUE}
library(webshot2)
extract_gam_stats <- function(model, panel, model_type, model_name,
                              alpha = 0.05, p_floor = 1e-4) {
  
  sm <- summary(model)
  av <- anova(model)                       # term-level Wald tests
  
  ## 1. Climate p-values at TERM level (s(age) excluded)
  # Parametric terms: one joint test per factor (all levels together)
  pt_tab <- av$pTerms.table
  p_par  <- if (!is.null(pt_tab) && NROW(pt_tab) > 0)
    setNames(pt_tab[, "p-value"], rownames(pt_tab)) else numeric(0)
  # Smooth terms
  st_tab <- sm$s.table
  st_cl  <- st_tab[rownames(st_tab) != "s(age)", , drop = FALSE]
  p_smo  <- setNames(st_cl[, "p-value"], rownames(st_cl))
  
  p_all  <- c(p_par, p_smo)
  p_all  <- p_all[!is.na(p_all)]
  # mgcv's smooth p-values are approximations; values below ~1e-4 (or slightly
  # negative, from rounding) are not meaningful beyond "very small"
  p_all  <- pmax(p_all, 0)
  
  climate_p   <- if (length(p_all)) min(p_all) else NA_real_
  climate_trm <- if (length(p_all)) names(p_all)[which.min(p_all)] else NA_character_
  sig_terms   <- names(p_all)[p_all < alpha]
  
  ## 2. Significant factor CONTRASTS vs. the reference level (for interpretation)
  ptab  <- sm$p.table
  contr <- ptab[rownames(ptab) != "(Intercept)", , drop = FALSE]
  contr <- contr[!is.na(contr[, 4]) & contr[, 4] < alpha, , drop = FALSE]
  sig_contr <- if (NROW(contr))
    paste0(rownames(contr), " (", sprintf("%.2f", contr[, 1]),
           ", p=", format.pval(contr[, 4], digits = 2, eps = p_floor), ")",
           collapse = "; ") else "none"
  
  data.frame(
    Panel             = panel,
    Type              = model_type,
    Model             = model_name,
    AIC               = round(AIC(model), 3),
    DevianceExplained = round(sm$dev.expl, 3),
    R2_Adj            = round(sm$r.sq, 3),
    ClimateTerm       = climate_trm,
    Climate_p         = climate_p,
    Climate_p_label   = if (is.na(climate_p)) "NA" else
      format.pval(climate_p, digits = 2, eps = p_floor),
    SignificantTerms  = if (length(sig_terms)) paste(sig_terms, collapse = "; ")
    else "Not significant",
    SignificantContrasts = sig_contr,
    EDF_Terms         = paste0(rownames(st_tab), " (EDF=",
                               round(st_tab[, "edf"], 2), ")", collapse = "; "),
    Inference         = ifelse(!is.na(climate_p) && climate_p < alpha,
                               "Supported", "Weak support"),
    stringsAsFactors  = FALSE
  )
}

# ---- 2. Labels, per-panel builder and table assembly (unchanged from original) ----
model_labels <- c(
  M_the = "Thermal regime",
  M_hyd = "Hydroclimatic trends",
  M_hur = "Hurricane regime",
  M_temp = "GMHT anomalies",
  M_dt = "thermal rate of change δT",
  M_hf = "Hurricane frequency",
  M_hav = "Hurricane average"
)

build_panel_summary <- function(best_obj,
                                panel_name,
                                formulas,
                                data){
  
  lin_name <- best_obj$best_linear$Model
  smo_name <- best_obj$best_smooth$Model
  
  lin_model <- gam(
    formulas[[lin_name]],
    data = data,
    family = betar(link="logit"),
    method = "REML"
  )
  
  smo_model <- gam(
    formulas[[smo_name]],
    data = data,
    family = betar(link="logit"),
    method = "REML"
  )
  
  bind_rows(
    
    extract_gam_stats(
      lin_model,
      panel_name,
      "Linear",
      lin_name
    ),
    
    extract_gam_stats(
      smo_model,
      panel_name,
      "Smooth",
      smo_name
    )
    
  )
}

tbl_summary <- bind_rows(
  
  build_panel_summary(
    best_pm,
    "Punta Maroma",
    formulas_mg,
    datos_pm
  ),
  
  build_panel_summary(
    best_fl,
    "Florida",
    formulas_mg,
    datos_fl_std
  ),
  
  build_panel_summary(
    best_bz,
    "Belize",
    formulas_mg,
    datos_bz
  ),
  
  build_panel_summary(
    best_sc,
    "Saint Croix",
    formulas_mg,
    datos_sc
  ),
  
  build_panel_summary(
    best_reg,
    "Regional",
    formulas_mg,
    datos_reg
  )
  
)

tbl_summary <- tbl_summary %>%
  mutate(Model_Label = recode(Model, !!!model_labels)) %>%
  select(Panel, Type, Model, Model_Label, AIC, DevianceExplained, R2_Adj,
         ClimateTerm, Climate_p_label, SignificantContrasts, EDF_Terms, Inference)

gt_summary_m <- tbl_summary %>%
  arrange(factor(Panel, levels = c("Punta Maroma", "Florida", "Belize",
                                   "Saint Croix", "Regional")), Type) %>%
  gt(groupname_col = "Panel") %>%
  fmt_number(columns = c(AIC, DevianceExplained, R2_Adj), decimals = 3) %>%
  cols_label(
    Type                 = "Model Type",
    Model                = "Model",
    Model_Label          = "Climate Driver",
    AIC                  = "AIC",
    DevianceExplained    = "Deviance Explained",
    R2_Adj               = "R² Adj",
    ClimateTerm          = "Climate term",
    Climate_p_label      = "Climate p-value (term)",
    SignificantContrasts = "Significant contrasts vs. reference",
    EDF_Terms            = "EDF Terms",
    Inference            = "Inference"
  ) %>%
  cols_width(Model ~ px(90), SignificantContrasts ~ px(240), EDF_Terms ~ px(260)) %>%
  tab_style(style = list(cell_fill(color = "#7298c2"),
                         cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) %>%
  tab_header(title = md("**Best-supported GAMs by panel**"),
             subtitle = md("*Top Linear and Smooth candidates*")) %>%
  tab_footnote(
    footnote = paste("Climate p-value: joint Wald test of the whole term (all factor",
                     "levels together for categorical regimes; the spline for",
                     "continuous predictors). Contrasts are relative to the",
                     "reference levels warmer / wet / low."),
    locations = cells_column_labels(columns = Climate_p_label)) %>%
  opt_row_striping() %>%
  tab_options(table.font.size = 12, data_row.padding = px(5))

gt_summary_m
gt::gtsave(gt_summary_m, "Table_1.docx")
```

<!-- REVISED 2026-09-30: Table 1 results rewritten with term-level p-values from the new render -->

Climate terms reach *p* \< 0.05 (term-level test) in the best marginal models of three panels:

- **Punta Maroma:** s(Hur100_av) (EDF 5.21, *p* \< 10⁻⁴; deviance explained 83.4 %) and Thermal (*p* = 0.012), driven by the LIA contrast (−1.81, *p* = 0.003).

- **Belize:** s(Hur100_fqcy) (EDF 6.16, *p* = 5.7 × 10⁻⁴) and Thermal (*p* = 0.001), with lower abundance in LIA (−2.52, *p* = 2.9 × 10⁻⁴) and CWP&MWA (−2.10, *p* = 0.001) than under the "warmer" reference.

- **Regional composite:** Thermal (*p* = 0.002), driven by the LIA contrast (−2.32, *p* = 1.9 × 10⁻⁴), and s(Hur100_av) (EDF 3.47, *p* = 0.024).

No climate term reaches *p* \< 0.05 in Florida (Hydro, *p* = 0.12; s(Hur100_av), *p* = 0.25) or Saint Croix (s(dT_raw), *p* = 0.080; Hurricane, *p* = 0.21). Wherever the thermal regime is significant, abundance is lower under LIA (and, in Belize, also under CWP&MWA) than under the "warmer" reference.

<!-- REVISED 2026-09-30: old underlined conclusion replaced -->

In the best-supported marginal models, climate terms were significant in Punta Maroma, Belize and the regional composite, but not in Florida or Saint Croix. For Saint Croix, AIC did not discriminate among candidate predictors (*n* = 18).

### 3.3 Combined GAM effects analysis

<!-- REVISED 2026-09-30: typos fixed -->

To capture compounded climatic effects, we fit a nested sequence of additive GAMs, without interaction terms or random effects.

#### 3.3.1 Candidate models

**Supplementary Table S7. Nested sequence of combined GAMs.** Eleven candidates. M1 = `s(age)`; M2 adds `s(GMHT_CPS0_30N)`; M3 adds `s(dT_raw)`. From M3 the sequence splits into two branches, one adding `s(Hur100_fqcy)` and the other `s(Hur100_av)` (M4); each branch then adds Hydro (M5), Thermal (M6) and Hurricane regime (M7) in turn. The two hurricane metrics never appear in the same model.

```{r combined-formulation, echo=TRUE}
model_table <- data.frame(
  Model = c("M1", "M2", "M3",
            "M4_fqcy", "M5_fqcy", "M6_fqcy", "M7_fqcy",
            "M4_av", "M5_av", "M6_av", "M7_av"),
  Formula = c(
    "counts_norm_sv ~ s(age)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal + Hurricane",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal + Hurricane"
  ),
  Added_Predictors = c(
    "", "GMHT_CPS0_30N", "dT_raw",
    "Hur100_fqcy", "Hydro", "Thermal", "Hurricane",
    "Hur100_av", "Hydro", "Thermal", "Hurricane"
  ),
  Branch = c(
    "Base", "Base", "Base",
    "Hur100_fqcy", "Hur100_fqcy", "Hur100_fqcy", "Hur100_fqcy",
    "Hur100_av", "Hur100_av", "Hur100_av", "Hur100_av"
  ),
  stringsAsFactors = FALSE
)

tab_S7 <- model_table %>%
  gt() %>%
  tab_header(
    title = "Nested sequence of combined GAMs"
  ) %>%
  cols_label(
    Model = "Model",
    Formula = "Formula",
    Added_Predictors = "Added Predictors",
    Branch = "Branch"
  )

tab_S7
gt::gtsave(tab_S7, "Supplementary_Table_S7.docx")   # REVISED 2026-09-30
```

#### 3.3.2 Fitting procedure (ΔAIC and Akaike weights)

<!-- REVISED 2026-09-30: step 6 (exclusion of rank-deficient fits) added; list restructured -->

`fit_combined_gams_safe()` fits the eleven candidates to each panel in six steps:

1.  It removes rows with missing or infinite numeric values.

2.  It skips a model if *n* \< 8 or *n* ≤ 3 × (number of smooths).

3.  It sets a common basis dimension for all smooths of a model, *k* = max(3, min(10, ⌊(*n* − 1)/number of smooths⌋)). For the four-smooth models this gives *k* = 7 in Punta Maroma (*n* = 31), 6 in Belize (27), 4 in Saint Croix (18), and 10 in Florida (45) and the regional composite (58). Nested models within a panel therefore do not share the same basis size (e.g. in Punta Maroma M1 uses *k* = 10 and M4 uses *k* = 7).

4.  It fits `gam()` (betar, REML) with `select = TRUE`, which penalizes each smooth's null space so that a term can be shrunk to EDF ≈ 0. (`bam()` is used only when *n* \> 100, which does not occur here.)

5.  If the beta fit fails, it refits the model as quasibinomial; such a model has no AIC and is excluded from the ranking.

6.  It excludes rank-deficient fits, i.e. models whose model-matrix rank is lower than the number of coefficients because two parametric terms are aliased. The names of the excluded models are returned in `$excluded`.

The function returns AIC, deviance explained, adjusted R², total EDF, ΔAIC and Akaike weights for each retained model; ΔAIC and weights are computed over the retained candidates only.

**Excluded candidates.** Because the thermal, hydroclimatic and hurricane regimes are defined on the same regional chronology and each site covers only part of the 60 bins, some regime levels are nested within others at site level:

- In Florida, Belize and Saint Croix, the "medium" hurricane regime occurs only within the CWP&MWA thermal regime, so M7_fqcy and M7_av are excluded.

- In Punta Maroma, the "cooler" thermal regime coincides with the wet and dry hydroclimatic states (all LIA and "warmer" bins are "variable"), so M6 and M7 of both branches are excluded.

- No candidate is excluded in the regional composite.

Consequently, the number of candidates compared is 7 in Punta Maroma, 9 in Florida, Belize and Saint Croix, and 11 in the regional composite.

```{r function-fit-combined, echo=TRUE}

formulas_cb <- list(
  M1 = counts_norm_sv ~ s(age),
  M2 = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N),
  M3 = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw),
  M4_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy),
  M5_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro,
  M6_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal,
  M7_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal + Hurricane,
  M4_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av),
  M5_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro,
  M6_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal,
  M7_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal + Hurricane
)

count_smooth_terms <- function(formula) {
  lengths(regmatches(paste(deparse(formula, width.cutoff = 500L), collapse = " "),
                      gregexpr("s\\(", paste(deparse(formula, width.cutoff = 500L), collapse = " "))))
}

cap_smooth_k <- function(formula, k_max, env = parent.frame()) {
  f_chr <- paste(deparse(formula, width.cutoff = 500L), collapse = " ")
  f_chr <- gsub("s\\(\\s*([A-Za-z0-9_\\.]+)\\s*\\)",
                paste0("s(\\1, k = ", k_max, ")"), f_chr)
  as.formula(f_chr, env = env)
}

safe_k_for <- function(n_obs, n_smooth_terms) {
  max(3, min(10, floor((n_obs - 1) / max(1, n_smooth_terms))))
}

fit_combined_gams_safe <- function(formulas_cb, data, segment_name) {
  
  data_clean <- data %>%
    mutate(across(where(is.numeric), ~replace(., is.infinite(.), NA))) %>%
    drop_na()
  
  n_obs <- nrow(data_clean)
  
  fits <- lapply(names(formulas_cb), function(name) {
    f <- formulas_cb[[name]]
    n_smooth <- count_smooth_terms(f)
    if (n_obs < 8 || n_obs <= n_smooth * 3) {
      message(paste("Skipping model", name, "for", segment_name,
                    "- insufficient observations (n =", n_obs, ") for",
                    n_smooth, "smooth term(s)"))
      return(NULL)
    }
    
    k_max <- safe_k_for(n_obs, n_smooth)
    f_safe <- if (n_smooth > 0) cap_smooth_k(f, k_max) else f
    
    tryCatch({
      m <- if (n_obs > 100) {
        bam(f_safe, data = data_clean, family = betar(link = "logit"),
            method = "fREML", select = TRUE, discrete = TRUE)
      } else {
        gam(f_safe, data = data_clean,
            family = betar(link = "logit"), method = "REML",
            select = TRUE)
      }
      environment(m$formula) <- environment()
      m
    }, error = function(e) {
      message(paste("Model", name, "fail in", segment_name,
                    "using betar:", e$message, "→ testing quasibinomial"))
      tryCatch({
        m2 <- gam(f_safe, data = data_clean,
                  family = quasibinomial(link = "logit"), method = "REML",
                  select = TRUE)
        environment(m2$formula) <- environment()
        m2
      }, error = function(e2) {
        message(paste("Model", name, "also failed with quasibinomial in", segment_name, ":", e2$message))
        return(NULL)
      })
    })
  })
  names(fits) <- names(formulas_cb)
  
  fits <- fits[!sapply(fits, is.null)]
  
 rank_def <- vapply(fits, function(m) m$rank < length(coef(m)), logical(1))
  excluded <- names(fits)[rank_def]
  if (length(excluded)) {
    message("Excluded as rank-deficient in ", segment_name, ": ",
            paste(excluded, collapse = ", "))
    fits <- fits[!rank_def]
  }
  
  stats <- lapply(names(fits), function(name) {
    m <- fits[[name]]
    s <- summary(m)
    data.frame(
      Model = name,
      AIC = tryCatch(AIC(m), error = function(e) NA),
      DevianceExplained = s$dev.expl,
      R2_Adj = s$r.sq,
      EDF = sum(m$edf),
      Segment = segment_name,
      stringsAsFactors = FALSE
    )
  })
  
  stats_df <- bind_rows(stats)
  
  if (nrow(stats_df) > 0) {
    
    stats_df <- stats_df %>%
      mutate(
        deltaAIC = AIC - min(AIC, na.rm = TRUE),
        weight = exp(-0.5 * deltaAIC) /
          sum(exp(-0.5 * deltaAIC), na.rm = TRUE)
      ) %>%
      arrange(AIC)
    
  }
  
  if (nrow(stats_df) > 0) {
    p_delta <- ggplot(stats_df, aes(x = reorder(Model, deltaAIC), y = deltaAIC)) +
      geom_col(fill = "steelblue") +
      geom_hline(yintercept = 2, linetype = "dashed", color = "red") +
      coord_flip() +
      labs(title = paste("ΔAIC by model -", segment_name),
           subtitle = "Dashed line = ΔAIC = 2 (competitive threshold)",
           x = "Model", y = "ΔAIC") +
      theme_minimal() +
      theme(legend.position = "bottom")
    
    p_weight <- ggplot(stats_df, aes(x = reorder(Model, -weight), y = weight)) +
      geom_col(fill = "darkgreen") +
      coord_flip() +
      labs(title = paste("AIC weights by model -", segment_name),
           subtitle = "Relative probability each model is best",
           x = "Model", y = "AIC weight") +
      theme_minimal()
    
    plots <- list(delta = p_delta, weight = p_weight)
  } else {
    plots <- NULL
  }
  
  return(list(fits = fits, stats = stats_df, plots = plots,
              excluded = excluded))  
}

datos_pm <- merged_df_g %>% filter(panel == "Punta Maroma")
results_pm <- fit_combined_gams_safe(formulas_cb, datos_pm, "Punta Maroma")

datos_fl <- merged_df_g %>% filter(panel == "Florida")
results_fl <- fit_combined_gams_safe(formulas_cb, datos_fl, "Florida")

datos_bz <- merged_df_g %>% filter(panel == "Belize")
results_bz <- fit_combined_gams_safe(formulas_cb, datos_bz, "Belize")

datos_sc <- merged_df_g %>% filter(panel == "Saint Croix")
results_sc <- fit_combined_gams_safe(formulas_cb, datos_sc, "Saint Croix")

datos_reg <- merged_df_g %>% filter(panel == "Regional")
results_reg <- fit_combined_gams_safe(formulas_cb, datos_reg, "Regional")

```

#### 3.3.3 Model competition and selected models

We stack the model statistics of all panels (`comparison_cb_full`), store the competitive models (ΔAIC ≤ 2; `competitive_models`, not displayed) and plot ΔAIC and Akaike weights by panel.

```{r suppl-fig5, fig.width=12, fig.height=8, echo=TRUE}
comparison_cb_full <- bind_rows(results_pm$stats,
                           results_fl$stats,
                           results_bz$stats,
                           results_sc$stats,
                           results_reg$stats)

competitive_models <- comparison_cb_full %>%
  filter(deltaAIC <= 2)%>%
  mutate(Segment = factor(Segment,
                          levels = c("Punta Maroma", "Florida", "Belize", "Saint Croix", "Regional")))

# REVISED 2026-09-30: same plotting function and layout as Supplementary Figure S4
sup_fig5 <- plot_model_selection(comparison_cb_full, panel_var = "Segment", model_levels = names(formulas_cb))
sup_fig5
ggsave("Supplementary_Figure_S5.pdf", plot = sup_fig5, width = 12, height = 8, dpi = 300, device = cairo_pdf)
```

<!-- REVISED 2026-09-30: excluded candidates noted -->

**Supplementary Figure S5. Information-theoretic comparison of the combined GAMs by panel.** (a) ΔAIC of each retained candidate (Supplementary Table S7); the red dashed line marks ΔAIC = 2, and the ΔAIC axis is scaled separately for each panel. (b) Akaike weights, on a common 0–1 scale. Candidates are listed in the same order in every panel; rows without a point are candidates excluded as rank-deficient (Section 3.3.2).

```{r delta-aic, echo=TRUE}
library(dplyr)
comparison_cb_full %>%
  filter(deltaAIC <= 2) %>%
  group_by(Segment) %>%
  mutate(sum_weight = sum(weight)) %>%
  ungroup() %>%
  arrange(Segment, deltaAIC) %>%
  select(Segment, Model, deltaAIC, weight, sum_weight) %>%
  mutate(across(c(deltaAIC, weight, sum_weight), ~ round(.x, 2)))
```

<!-- REVISED 2026-09-30: label and typo fixes; Florida GMHT p added -->

**Table 2. Best-supported combined GAM per panel** (lowest AIC): model, AIC, Akaike weight, deviance explained, adjusted R², climate terms with term-level *p* \< 0.05 (`s(age)` excluded), factor contrasts with *p*-value \< 0.05 versus reference level, and EDF of every smooth.

- **Punta Maroma — M4_av** (Akaike weight 0.81; deviance explained 82.5 %; adjusted R² 0.72). s(Hur100_av) is significant (EDF 4.07, *p* \< 10⁻⁴), as is s(age) (EDF 0.83, *p* = 0.009). s(GMHT_CPS0_30N) (EDF 1.12) and s(dT_raw) (EDF 0.52) are not significant.

- **Florida — M4_av** (weight 0.32; 28.7 %; adjusted R² 0.18). s(Hur100_av) is the only significant term (EDF 1.59, *p* = 0.009). s(GMHT_CPS0_30N) (EDF 1.49, *p* = 0.078) is retained but not significant, and s(age) and s(dT_raw) are shrunk to EDF ≈ 0.

- **Belize — M6_fqcy** (weight 0.64; 79.0 %; adjusted R² 0.60). Significant terms: Thermal (joint test, *p* = 0.002), s(GMHT_CPS0_30N) (EDF 1.71, *p* = 0.033) and s(Hur100_fqcy) (EDF 1.55, *p* = 0.032). Within Thermal, abundance is lower than under the "warmer" reference in LIA (−2.39, *p* = 0.003) and CWP&MWA (−1.77, *p* = 0.024). Hydro is not significant as a term, although its "variable" contrast is (+1.16, *p* = 0.039). s(dT_raw) is shrunk to EDF ≈ 0.

- **Saint Croix — M4_fqcy** (weight 0.31; 58.3 %; adjusted R² 0.43). Significant terms: s(dT_raw) (EDF 1.55, *p* = 0.009) and s(Hur100_fqcy) (EDF 0.78, *p* = 0.032). s(GMHT_CPS0_30N) is shrunk to EDF ≈ 0.

- **Regional composite — M6_av** (weight 0.30; 43.9 %; adjusted R² 0.27). The only significant climate term is Thermal (joint test, *p* = 0.025), driven by the LIA contrast (−1.96, *p* = 0.003); s(age) is also significant (EDF 2.75, *p* = 2 × 10⁻⁴). s(Hur100_av) and s(dT_raw) are shrunk to EDF ≈ 0, and s(GMHT_CPS0_30N) (EDF 0.60) is not significant.

```{r Table-2, echo=TRUE}

get_best_combined <- function(results_obj){
  
  ranking <- results_obj$stats %>%
    arrange(AIC)
  
  winner_name <- ranking$Model[1]
  
  list(
    best_model = winner_name,
    ranking = ranking,
    fit = results_obj$fits[[winner_name]]
  )
  
}

best_pm_cb <- get_best_combined(results_pm)
best_fl_cb <- get_best_combined(results_fl)
best_bz_cb <- get_best_combined(results_bz)
best_sc_cb <- get_best_combined(results_sc)
best_reg_cb <- get_best_combined(results_reg)

extract_combined_stats <- function(best_obj, panel_name,
                                   alpha = 0.05, p_floor = 1e-4) {
  
  fmt_p <- function(p) ifelse(p < p_floor, "<1e-04", sprintf("%.2g", p))
  
  model <- best_obj$fit
  sm    <- summary(model)
  av    <- anova(model)
  
  ## Term-level p-values (s(age) excluded)
  pt_tab <- av$pTerms.table
  p_par  <- if (!is.null(pt_tab) && NROW(pt_tab) > 0)
    setNames(pt_tab[, "p-value"], rownames(pt_tab)) else numeric(0)
  st_tab <- sm$s.table
  st_cl  <- st_tab[rownames(st_tab) != "s(age)", , drop = FALSE]
  p_smo  <- setNames(st_cl[, "p-value"], rownames(st_cl))
  
  p_all  <- c(p_par, p_smo)
  p_all  <- pmax(p_all[!is.na(p_all)], 0)
  sig    <- p_all[p_all < alpha]
  
  sig_terms <- if (length(sig))
    paste0(names(sig), " (p=", fmt_p(sig), ")",
           collapse = "; ") else "Not significant"
  climate_p <- if (length(p_all)) min(p_all) else NA_real_
  
ptab  <- sm$p.table
  contr <- ptab[rownames(ptab) != "(Intercept)", , drop = FALSE]
  contr <- contr[!is.na(contr[, 4]) & contr[, 4] < alpha, , drop = FALSE]
  sig_contr <- if (NROW(contr))
    paste0(rownames(contr), " (", sprintf("%.2f", contr[, 1]),
           ", p=", fmt_p(contr[, 4]), ")",
           collapse = "; ") else "none"
  
  data.frame(
    Panel                = panel_name,
    Model                = best_obj$best_model,
    AIC                  = round(AIC(model), 3),
    DevianceExplained    = round(sm$dev.expl, 3),
    R2_Adj               = round(sm$r.sq, 3),
    SignificantTerms     = sig_terms,
    Climate_p            = climate_p,
    Climate_p_label      = if (is.na(climate_p)) "NA" else
      fmt_p(climate_p),
    SignificantContrasts = sig_contr,
    EDF_Terms            = paste0(rownames(st_tab), " (EDF=",
                                  round(st_tab[, "edf"], 2), ")", collapse = "; "),
    stringsAsFactors     = FALSE
  )
}

tbl_combined <- bind_rows(
  
  extract_combined_stats(
    best_pm_cb,
    "Punta Maroma"
  ),
  
  extract_combined_stats(
    best_fl_cb,
    "Florida"
  ),
  
  extract_combined_stats(
    best_bz_cb,
    "Belize"
  ),
  
  extract_combined_stats(
    best_sc_cb,
    "Saint Croix"
  ),
  
  extract_combined_stats(
    best_reg_cb,
    "Regional"
  )
  
)

combined_labels <- c(
  
  M1 =
    "Temporal baseline",
  
  M2 =
    "Background temperature model",
  
  M3 =
    "Thermal forcing model",
  
  M4_fqcy =
    "Thermal–disturbance model (frequency)",
  
  M5_fqcy =
    "Thermal–disturbance–hydroclimate model",
  
  M6_fqcy =
    "Integrated climate model (frequency)",
  
  M7_fqcy =
    "Full climate-regime model (frequency)",
  
  M4_av =
    "Thermal–disturbance model (intensity)",
  
  M5_av =
    "Thermal–disturbance–hydroclimate model",
  
  M6_av =
    "Integrated climate model (intensity)",
  
  M7_av =
    "Full climate-regime model (intensity)"
)

tbl_combined <- tbl_combined %>%
  mutate(
    Model_Label =
      recode(Model,
             !!!combined_labels)
  )

tbl_combined <- tbl_combined %>%
  mutate(
    Weight = c(
      best_pm_cb$ranking$weight[1],
      best_fl_cb$ranking$weight[1],
      best_bz_cb$ranking$weight[1],
      best_sc_cb$ranking$weight[1],
      best_reg_cb$ranking$weight[1]
    )
  )


tbl_combined <- tbl_combined %>%
  select(Panel, Model, Model_Label, AIC, Weight, DevianceExplained, R2_Adj,
         SignificantTerms, SignificantContrasts, EDF_Terms)

gt_combined <- tbl_combined %>%
  arrange(factor(Panel, levels = c("Punta Maroma", "Florida", "Belize",
                                   "Saint Croix", "Regional"))) %>%
  gt() %>%
  fmt_number(columns = c(AIC, DevianceExplained, R2_Adj), decimals = 3) %>%
  fmt_number(columns = Weight, decimals = 2) %>%
  cols_label(
    Panel                = "Panel",
    Model                = "Model",
    Model_Label          = "Climate Model",
    AIC                  = "AIC",
    Weight               = "Akaike Weight",
    DevianceExplained    = "Deviance Explained",
    R2_Adj               = "R² Adj",
    SignificantTerms     = "Significant terms (term-level p)",
    SignificantContrasts = "Significant contrasts vs. reference",
    EDF_Terms            = "EDF Terms"
  ) %>%
  cols_width(Model ~ px(90), Model_Label ~ px(220), SignificantTerms ~ px(220),
             SignificantContrasts ~ px(240), EDF_Terms ~ px(300)) %>%
  tab_style(style = list(cell_fill(color = "#7298c2"),
                         cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) %>%
  tab_header(title = md("**Best-supported combined GAMs by panel**"),
             subtitle = md("*Top multivariate model selected by AIC*")) %>%
  tab_footnote(
    footnote = paste("Term-level p: joint Wald test of the whole term (all factor",
                     "levels together; the spline for continuous predictors);",
                     "s(age) excluded. Contrasts are relative to the reference",
                     "levels warmer / wet / low. Rank-deficient candidates were",
                     "excluded before model selection."),
    locations = cells_column_labels(columns = SignificantTerms)) %>%
  opt_row_striping() %>%
  tab_options(table.font.size = 12, data_row.padding = px(5))

gt_combined
gt::gtsave(gt_combined, "Table_2.docx")   

list(`Punta Maroma` = results_pm$excluded, Florida = results_fl$excluded,
     Belize = results_bz$excluded, `Saint Croix` = results_sc$excluded,
     Regional = results_reg$excluded)
```

The selected model is supported in **Punta Maroma** (Akaike weight 0.81) and **Belize** (0.64): in both panels no other candidate lies within ΔAIC ≤ 2. In **Florida** (0.32), **Saint Croix** (0.31) and the **regional composite** (0.30), the competitive set (ΔAIC ≤ 2; Supplementary Fig. S5) includes several models. In **Florida** it comprises three models of the average-hurricane branch (M4_av, M5_av and M6_av; ΔAIC ≤ 1.96; combined weight 0.59), which differ only in whether the hydroclimatic and thermal regimes are included. In **Saint Croix** it comprises two models of the hurricane-frequency branch (M4_fqcy and M5_fqcy; ΔAIC = 0.95; combined weight 0.51), which differ only in the hydroclimatic regime. In the **regional composite** it comprises five models (M6_av, M6_fqcy, M4_av, M7_av and M7_fqcy; ΔAIC ≤ 1.93; combined weight 0.83), spanning both hurricane metrics and models with and without categorical regimes. In these three panels the data do not discriminate between the selected model and simpler or more complex alternatives, so inference should rest on terms that recur across the competitive set rather than on the single top model (Supplementary Table S8). Across panels, the thermal regime is the only categorical predictor that is significant as a whole term in a selected model (Belize and the regional composite), always with lower abundance under LIA than under the "warmer" reference. The hydroclimatic regime is not significant as a term in any selected model.

```{r competitive-set-terms, echo=TRUE}
# REVISED 2026-09-30: Supplementary Table S8 = term-level p-values in every competitive model (ΔAIC ≤ 2)
term_pvalues <- function(m) {
  sm <- summary(m); av <- anova(m)
  p_par <- if (!is.null(av$pTerms.table) && NROW(av$pTerms.table) > 0)
             setNames(av$pTerms.table[, "p-value"], rownames(av$pTerms.table)) else numeric(0)
  p_smo <- setNames(sm$s.table[, "p-value"], rownames(sm$s.table))
  pmax(c(p_par, p_smo), 0)
}

results_by_panel <- list("Punta Maroma" = results_pm, "Florida" = results_fl, "Belize" = results_bz,
                         "Saint Croix" = results_sc, "Regional" = results_reg)

comp_terms <- purrr::imap_dfr(results_by_panel, function(r, site) {
  cs <- r$stats %>% filter(deltaAIC <= 2) %>% arrange(deltaAIC)
  purrr::map_dfr(seq_len(nrow(cs)), function(i) {
    p <- term_pvalues(r$fits[[cs$Model[i]]])
    tibble::tibble(Panel = site, Model = cs$Model[i], deltaAIC = cs$deltaAIC[i],
                   Weight = cs$weight[i], Term = names(p), p = p)
  })
})

term_order <- c("Thermal", "Hydro", "Hurricane", "s(age)", "s(GMHT_CPS0_30N)",
                "s(dT_raw)", "s(Hur100_av)", "s(Hur100_fqcy)")

comp_wide <- comp_terms %>%
  mutate(p_label = ifelse(p < 1e-4, "<1e-04", sprintf("%.2g", p)),
         p_label = ifelse(p < 0.05, paste0(p_label, " *"), p_label)) %>%
  select(-p) %>%
  tidyr::pivot_wider(names_from = Term, values_from = p_label, values_fill = "–")
comp_wide <- comp_wide[, c("Panel", "Model", "deltaAIC", "Weight",
                           intersect(term_order, names(comp_wide)))]

tab_S8 <- comp_wide %>%
  gt(groupname_col = "Panel") %>%
  fmt_number(columns = c(deltaAIC, Weight), decimals = 2) %>%
  cols_label(deltaAIC = "ΔAIC", Weight = "Akaike weight") %>%
  tab_header(title = "Term-level p-values across the competitive set (ΔAIC ≤ 2)") %>%
  tab_footnote(footnote = "* p < 0.05; – term not in the model. Joint Wald test per term (all levels of a factor together).")

tab_S8
gt::gtsave(tab_S8, "Supplementary_Table_S8.docx")
```

<!-- REVISED 2026-09-30: new table: terms across the competitive set -->

**Supplementary Table S8. Term-level *p*-values across the competitive set.** For each panel, every combined GAM within ΔAIC ≤ 2 of the best model, with its ΔAIC, Akaike weight and the term-level *p*-value of each term (joint Wald test for factors; \* *p* \< 0.05; – term not in the model).

In **Florida**, s(Hur100_av) is significant in all three competitive models (*p* = 0.001–0.02), whereas neither regime factor is significant in any of them. In **Saint Croix**, s(dT_raw) is significant in both competitive models (*p* = 0.009 in each), whereas s(Hur100_fqcy) is significant only in the top model (*p* = 0.032; *p* = 0.072 in M5_fqcy). In the **regional composite**, the thermal regime is significant in all four competitive models that include it (*p* = 0.0005–0.025), always together with a significant s(age); the fifth competitive model, M4_av, contains no regimes and instead retains a significant s(Hur100_av) (*p* = 0.002) with a non-significant s(age). At the regional scale, the thermal regime (together with the temporal trend) and average hurricane activity are therefore alternative, almost equally supported explanations that these data cannot separate. **Punta Maroma** and **Belize** each have a single competitive model (Table 2).

#### 3.3.4 Full summaries of the selected combined models

We print the complete `mgcv` summaries of the five models in Table 2.

```{r best-combined, echo=TRUE}
print_best_combined <- function(best_obj,
                                panel_name){
  
  cat("\n====================================\n")
  cat(panel_name, "\n")
  cat("====================================\n")
  
  cat(
    "\nBEST COMBINED MODEL:",
    best_obj$best_model,
    "\n\n"
  )
  
  print(
    summary(
      best_obj$fit
    )
  )
  
}

print_best_combined(best_pm_cb, "Punta Maroma")
print_best_combined(best_fl_cb, "Florida")
print_best_combined(best_bz_cb, "Belize")
print_best_combined(best_sc_cb, "Saint Croix")
print_best_combined(best_reg_cb, "Regional")
```

## 4. Model Quality Control (combined effects)

To ensure methodological transparency and meet standard practices in GAM-based ecological modeling, we examine the five selected combined models through residual diagnostics from gam.check() (Q–Q plot, residuals vs. linear predictor, residual histogram, response vs. fitted values), plots of all fitted terms, concurvity, and temporal autocorrelation of the residuals. The basis-dimension check (k-index) is reported with `k.check()` below each set of diagnostic plots.

### 4.1 Partial effects of the selected models

The five selected models are collected in `best_models` and their summaries are printed again. Smooth estimates with 95 % confidence intervals are extracted with `gratia::smooth_estimates()`, pooled across panels and plotted by predictor (one panel per smooth). The thermal-regime effects of the selected models that include this factor are added as a sixth panel. To put them on the same baseline as the smooths, which mgcv constrains to sum to zero over the fitted observations, the factor effects are centred in the same way: each regime's coefficient minus their mean weighted by the number of bins in each regime. Centring shifts all regimes by the same amount, so differences between regimes (the contrasts reported in Table 2) are unchanged; 95 % confidence intervals are computed from the full coefficient covariance matrix.

```{r Fig4, echo=TRUE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

library(gratia)
library(dplyr)
library(ggplot2)

best_models <- list(
  "Punta Maroma" = best_pm_cb$fit,
  "Florida"      = best_fl_cb$fit,
  "Belize"       = best_bz_cb$fit,
  "Saint Croix"  = best_sc_cb$fit,
  "Regional"     = best_reg_cb$fit
)

summary(best_models[["Punta Maroma"]])
summary(best_models[["Florida"]])
summary(best_models[["Belize"]])
summary(best_models[["Saint Croix"]])
summary(best_models[["Regional"]])

df_all <- purrr::map_df(names(best_models), function(site) {
  smooth_estimates(best_models[[site]], ci_level = 0.95) %>%
    add_confint() %>%
    mutate(site = site)
})

df_all <- df_all %>%
  mutate(site = factor(site, levels = names(custom_colors)))

smooth_predictor_cols <- intersect(
  c("age", "GMHT_CPS0_30N", "dT_raw", "Hur100_av", "Hur100_fqcy"),
  names(df_all)
)

df_long <- df_all %>%
  pivot_longer(cols = all_of(smooth_predictor_cols),
               names_to = "predictor", values_to = "xvar") %>%
  filter(!is.na(xvar))

# REVISED 2026-09-30 ---- Figure 4: smooth partial effects (5 panels) + thermal-regime effects (6th panel) ----
library(patchwork)

site_levels <- c("Punta Maroma", "Florida", "Belize", "Saint Croix", "Regional")
site_cols   <- custom_colors[site_levels]

df_long <- df_long %>%
  mutate(site = factor(as.character(site), levels = site_levels))

pred_order <- intersect(c("age", "GMHT_CPS0_30N", "dT_raw", "Hur100_av", "Hur100_fqcy"),
                        unique(df_long$predictor))

fig4_theme <- theme_minimal(base_size = 14) +
  theme(axis.title = element_text(size = 12),
        axis.text  = element_text(size = 12),
        strip.text = element_text(size = 13))

# One panel per smooth predictor (same look as the former facet_wrap)
smooth_panels <- lapply(seq_along(pred_order), function(i) {
  pr <- pred_order[i]
  ggplot(filter(df_long, predictor == pr),
         aes(x = xvar, y = .estimate, colour = site, fill = site)) +
    geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.15, colour = NA) +
    geom_line(linewidth = 1.3) +
    facet_wrap(~predictor) +
    scale_colour_manual(values = site_cols, limits = site_levels, name = "Site") +
    scale_fill_manual(values = site_cols, limits = site_levels, name = "Site") +
    labs(x = NULL, y = if (i %% 3 == 1) "Partial effect" else NULL) +
    fig4_theme
})

# Thermal-regime effects of the selected models that contain Thermal (Belize and Regional),
# CENTRED like the smooths: mgcv constrains each smooth to sum to zero over the fitted
# observations, so the factor effect is centred the same way (weighted by the number of
# bins in each regime). Differences between regimes are unchanged by the centring;
# 95 % CIs use the full coefficient covariance matrix.
thermal_sites <- names(best_models)[vapply(best_models, function(m)
  any(grepl("^Thermal", names(coef(m)))), logical(1))]

thermal_df <- purrr::map_df(thermal_sites, function(site) {
  m    <- best_models[[site]]
  idx  <- grep("^Thermal", names(coef(m)))
  b    <- coef(m)[idx]
  V    <- vcov(m)[idx, idx, drop = FALSE]
  lev  <- levels(droplevels(m$model$Thermal))                 # reference first
  w    <- as.numeric(table(factor(m$model$Thermal, levels = lev))) / nrow(m$model)
  A    <- t(sapply(lev, function(l) as.numeric(names(b) == paste0("Thermal", l))))  # level -> coef
  Cm   <- A - matrix(colSums(w * A), nrow(A), ncol(A), byrow = TRUE)                 # centring
  est  <- as.numeric(Cm %*% b)
  se   <- sqrt(pmax(diag(Cm %*% V %*% t(Cm)), 0))
  tibble::tibble(level = lev, est = est, lo = est - 1.96 * se, hi = est + 1.96 * se,
                 site = site)
}) %>%
  mutate(level     = factor(level, levels = intersect(c("warmer", "cooler", "CWP&MWA", "LIA"),
                                                      unique(level))),
         site      = factor(site, levels = site_levels),
         predictor = "Thermal regime")

thermal_panel <- ggplot(thermal_df, aes(x = level, y = est, colour = site)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(ymin = lo, ymax = hi),
                  position = position_dodge(width = 0.5), linewidth = 1, size = 0.6) +
  facet_wrap(~predictor) +
  scale_colour_manual(values = site_cols, limits = site_levels, guide = "none") +
  labs(x = NULL, y = NULL) +
  fig4_theme

p_all <- wrap_plots(c(smooth_panels, list(thermal_panel)), ncol = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_all
ggsave("Figure_4.pdf", plot = p_all, width = 12, height = 8, dpi = 300, device = cairo_pdf)
```

<!-- REVISED 2026-09-30: Fig. 4 caption updated for the new Florida model and file name -->

**Figure 4** (`Figure_4.pdf`). Partial effects of the selected combined GAMs on the logit scale, coloured by panel. The first five panels show the smooth terms with 95 % confidence bands; terms shrunk to EDF ≈ 0 plot as flat lines at zero (s(dT_raw) in Florida, Belize and the regional composite; s(age) in Florida; s(Hur100_av) in the regional composite). The sixth panel shows the thermal-regime effects in the two selected models that include this factor (Belize and the regional composite), with 95 % confidence intervals. All effects are centred to a weighted mean of zero over the fitted bins, so smooth and factor effects share the same baseline and can be compared directly: zero denotes the model's average contribution of that term, and a regime below zero is associated with lower abundance than average. Differences between regimes equal the contrasts in Table 2.

### 4.2 Concurvity analysis

`concurvity(full = TRUE)` measures, for each term, how well it can be approximated by the other terms of the model (0 = not at all, 1 = completely), reported as worst-case, observed and estimated values. Values above about 0.8 are commonly treated as problematic.

```{r concurvity, echo=TRUE}

library(reshape2)
library(dplyr)
library(gt)

concurvity_results <- lapply(best_models, function(m) concurvity(m, full = TRUE))

# REVISED 2026-09-30: Supplementary Table S9 = concurvity as a formatted table
conc_tbl <- purrr::imap_dfr(concurvity_results, function(cm, site) {
  as.data.frame(t(cm)) %>%
    tibble::rownames_to_column("Term") %>%
    mutate(Term = ifelse(Term == "para", "parametric terms", Term),
           Panel = site)
}) %>%
  select(Panel, Term, worst, observed, estimate)

tab_S9 <- conc_tbl %>%
  gt(groupname_col = "Panel") %>%
  fmt_number(columns = c(worst, observed, estimate), decimals = 2) %>%
  cols_label(Term = "Term", worst = "Worst", observed = "Observed", estimate = "Estimate") %>%
  tab_header(title = "Concurvity of the selected combined GAMs",
             subtitle = "0 = term not approximated by the other terms; 1 = fully approximated")

tab_S9
gt::gtsave(tab_S9, "Supplementary_Table_S9.docx")
```

<!-- REVISED 2026-09-30: renumbered; values updated with the new Florida model (M4_av) -->

**Supplementary Table S9. Concurvity of the selected combined GAMs** (output above). Observed concurvity is high in every panel for s(age) (0.88–0.98) and high in most panels for s(GMHT_CPS0_30N) (0.88–0.95; 0.77 in Florida). The hurricane smooth reaches 0.76–0.87, except in Saint Croix (0.33), and s(dT_raw) ranges from 0.44 (Florida) to 0.78. In Belize and the regional composite, the only selected models with categorical regimes, the parametric block (the regime factors) is almost entirely reproducible by the smooths (≈ 1.00), as expected when each regime occupies contiguous blocks of time.

The temporal trend, background temperature and categorical regimes are therefore largely interchangeable in these data. How variance is partitioned among them, and hence which term `select = TRUE` retains, is not uniquely determined, and the effects in Table 2 and Fig. 4 should be read as conditional on the other terms rather than as independent contributions. The least confounded terms are the hurricane smooth in Saint Croix and s(dT_raw) in Florida and Saint Croix.

### 4.3 GAM diagnostic plots

For each panel we show the four `gam.check()` residual plots and all model terms (`plot(..., all.terms = TRUE)`, one page: smooths with 95 % bands and rug plots, plus factor effects).

<!-- REVISED 2026-09-30: figure and table labels standardized to journal style throughout (Figure N / Table N; Supplementary Figure SN / Supplementary Table SN; Fig. and Supplementary Fig. in running text) -->

**Supplementary Figure S6.** Residual diagnostics and fitted terms of the selected combined GAMs: (a) Punta Maroma, (b) Florida, (c) Belize, (d) regional composite and (e) Saint Croix.

<!-- REVISED 2026-09-30: diagnostic assessment written from the plots, k.check() and residual summaries -->

In all panels the response takes few distinct values, because each bin holds only a few records and abundance is normalized to the panel maximum: 2 distinct values in Saint Croix, 4 in Belize, 6 in Punta Maroma, 19 in Florida and 51 in the regional composite. This discreteness produces the diagonal bands in the plots of residuals against the linear predictor and should not be read as a missing covariate. The basis dimension is adequate for every smooth (EDF well below *k′*), except that s(age) has a k-index below 1 with *p* \< 0.05 in Punta Maroma (0.71) and Florida (0.71). Because the EDF of s(age) is far below *k′* in both (0.83 of 6 and ≈ 0 of 9), increasing *k* would not help; the low k-index instead indicates residual pattern along age, consistent with the positive lag-1 autocorrelation detected in these two panels (Section 4.4). Panel by panel:

- **Punta Maroma:** the Q–Q plot follows the reference line (Shapiro–Wilk on deviance residuals, *p* = 0.14); the maximum bin (response 0.98, fitted 0.91) determines the upper end of the response–fitted relationship.
- **Florida:** the residuals have a right tail (Shapiro–Wilk *p* = 0.004), and fitted values span only 0.21–0.54 against observed values of 0.08–0.99, so the model does not reproduce the extremes, consistent with its modest deviance explained (28.7 %).
- **Belize:** the residuals have heavier tails than normal (Shapiro–Wilk *p* = 0.006), with strong banding because most bins hold a single record (response 0.25).
- **Regional composite:** the residuals are close to normal (Shapiro–Wilk *p* = 0.50), show no pattern against the linear predictor, and the response–fitted relationship is the closest to the 1:1 line.
- **Saint Croix:** with 18 bins and only two distinct response values (one or two records per bin), the residuals fall on two bands; the Q–Q plot is acceptable (Shapiro–Wilk *p* = 0.08), but the response is close to binary, which limits what a beta-regression smooth can resolve in this panel.

##### 4.3.1 Punta Maroma

```{r pm-diagn, echo=TRUE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}
par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Punta Maroma"]])))
par(mfrow = c(2,3))
plot(best_models[["Punta Maroma"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Punta Maroma"]])   # REVISED 2026-09-30: basis-dimension check reported
```

##### **4.3.2 Florida**

```{r fl-diagn, echo=TRUE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Florida"]])))
par(mfrow = c(2,3))
plot(best_models[["Florida"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Florida"]])   # REVISED 2026-09-30: basis-dimension check reported
```

##### **4.3.3 Belize**

```{r bz-diagn, echo=TRUE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Belize"]])))
par(mfrow = c(2,3))
plot(best_models[["Belize"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Belize"]])   # REVISED 2026-09-30: basis-dimension check reported
```

##### 4.3.4 Regional

```{r reg-diagn, echo=TRUE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Regional"]])))
par(mfrow = c(2,3))
plot(best_models[["Regional"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Regional"]])   # REVISED 2026-09-30: basis-dimension check reported
```

##### 4.3.5 Saint Croix

<!-- REVISED 2026-09-30: development note removed -->

```{r sc-diagn, echo=TRUE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Saint Croix"]])))
par(mfrow = c(2,3))
plot(best_models[["Saint Croix"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Saint Croix"]])   # REVISED 2026-09-30: basis-dimension check reported
```

```{r figS6-save, echo=TRUE, warning=FALSE, message=FALSE}
# REVISED 2026-09-30: save Supplementary Figure S6 (a-e) as one multi-page PDF, one page per panel
pdf("Supplementary_Figure_S6.pdf", width = 12, height = 10)
for (site in c("Punta Maroma", "Florida", "Belize", "Regional", "Saint Croix")) {
  par(mfrow = c(2, 2), oma = c(0, 0, 2, 0))
  invisible(capture.output(gam.check(best_models[[site]])))
  mtext(paste(site, "- residual diagnostics"), outer = TRUE, font = 2)
  par(oma = c(0, 0, 2, 0))
  plot(best_models[[site]], shade = TRUE, pages = 1, all.terms = TRUE, rug = TRUE)
  mtext(paste(site, "- fitted terms"), outer = TRUE, font = 2)
}
invisible(dev.off())
```

### 4.4 Residual temporal independence

For each selected model we extract Pearson residuals in bin order (young to old) and compute the lag-1 autocorrelation (φ), the Durbin–Watson statistic (`lmtest::dwtest`, a one-sided test against positive autocorrelation) and the Ljung–Box statistic at lag 10. ACF and PACF of the residuals are shown for all panels in one figure (Supplementary Fig. S7).

```{r residual-independency, echo=TRUE}

data_list <- list(
  "Belize"       = datos_bz,
  "Florida"      = datos_fl,
  "Punta Maroma" = datos_pm,
  "Regional"     = datos_reg,
  "Saint Croix"  = datos_sc
)

library(mgcv)
library(lmtest)

check_temporal_independence <- function(mod) {
  # Extraer residuales tipo pearson
  res_raw <- residuals(mod, type = "pearson")
  if (is.list(res_raw)) res_raw <- unlist(res_raw)
  res_raw <- as.numeric(res_raw)
  
  # AR(1) coeficiente
  phi <- cor(res_raw[-1], res_raw[-length(res_raw)])
  
  # Durbin–Watson
  dw <- lmtest::dwtest(res_raw ~ 1)
  
  # Ljung–Box
  lb <- Box.test(res_raw, lag = 10, type = "Ljung-Box")
  
  list(
    phi_AR1 = phi,
    DW_stat = dw$statistic,
    DW_pval = dw$p.value,
    LB_stat = lb$statistic,
    LB_pval = lb$p.value,
    res_raw = res_raw
  )
}

best_cb_lookup <- list(
  "Punta Maroma" = best_pm_cb,
  "Florida"      = best_fl_cb,
  "Belize"       = best_bz_cb,
  "Saint Croix"  = best_sc_cb,
  "Regional"     = best_reg_cb
)

best_models_table <- tibble(
  Segment = names(best_models),
  Model   = vapply(names(best_models), function(s) best_cb_lookup[[s]]$best_model, character(1))
)

results_resid <- lapply(seq_len(nrow(best_models_table)), function(i) {
  seg <- best_models_table$Segment[i]
  mname <- best_models_table$Model[i]
  cat("Using model:", mname, "for segment:", seg, "\n")

  mod <- best_models[[seg]]

  diag <- check_temporal_independence(mod)
  
  df <- data.frame(Segment = seg,
                   Model = mname,
                   phi_AR1 = diag$phi_AR1,
                   DW_stat = diag$DW_stat,
                   DW_pval = diag$DW_pval,
                   LB_stat = diag$LB_stat,
                   LB_pval = diag$LB_pval)
  
  attr(df, "res_raw") <- diag$res_raw
  df
})

res_summary <- do.call(rbind, results_resid)

tab_S10 <- res_summary %>%
  gt() %>%
  tab_header(
    title = "Residual independence of the selected combined GAMs"
  ) %>%
  fmt_number(columns = c(phi_AR1, DW_stat, DW_pval, LB_stat, LB_pval),
             decimals = 3)

tab_S10
gt::gtsave(tab_S10, "Supplementary_Table_S10.docx")   # REVISED 2026-09-30

# REVISED 2026-09-30: per-panel base ACF/PACF PDF removed; Supplementary Figure S7 below is the saved figure

acf_df <- function(res, type = "ACF", site) {
  acf_obj <- acf(res, plot = FALSE)
  tibble(
    Lag = acf_obj$lag[, , 1],
    Value = acf_obj$acf[, , 1],
    Type = type,
    Site = site
  )
}

pacf_df <- function(res, type = "PACF", site) {
  pacf_obj <- pacf(res, plot = FALSE)
  tibble(
    Lag = pacf_obj$lag,
    Value = pacf_obj$acf,
    Type = type,
    Site = site
  )
}

df_all <- bind_rows(lapply(seq_len(nrow(best_models_table)), function(i) {
  seg <- best_models_table$Segment[i]
  res_raw <- attr(results_resid[[i]], "res_raw")
  bind_rows(
    acf_df(res_raw, "ACF", seg),
    pacf_df(res_raw, "PACF", seg)
  )
}))

acf_ci <- tibble(Site = best_models_table$Segment,
                 ci   = 1.96 / sqrt(vapply(results_resid, function(r) length(attr(r, "res_raw")), numeric(1))))

plot_acf_pacf <- ggplot(df_all, aes(x = Lag, y = Value, fill = Site)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = custom_colors) +
  geom_hline(yintercept = 0, color = "black") +
  # REVISED 2026-09-30: panel-specific approximate 95 % limits (±1.96/sqrt(n)) instead of fixed ±0.2
  geom_hline(data = acf_ci, aes(yintercept =  ci), linetype = "dashed", color = "grey40") +
  geom_hline(data = acf_ci, aes(yintercept = -ci), linetype = "dashed", color = "grey40") +
  facet_grid(Type ~ Site, scales = "free_y") +
  labs(y = "Correlation", x = "Lag") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

plot_acf_pacf
ggsave("Supplementary_Figure_S7.pdf", plot = plot_acf_pacf, width = 12, height = 6, dpi = 300, device = cairo_pdf)   # REVISED 2026-09-30
```

<!-- REVISED 2026-09-30: renumbered; Florida (M4_av) values updated -->

**Supplementary Table S10. Residual independence of the selected combined GAMs** (output above).

- **Ljung–Box:** no panel shows significant autocorrelation up to lag 10 (*p* = 0.08–0.31).

- **Durbin–Watson:** positive lag-1 autocorrelation is detected in Florida (φ = 0.34, DW = 1.23, *p* = 0.004) and Punta Maroma (φ = 0.30, DW = 1.39, *p* = 0.041).

- **Negative autocorrelation:** Belize (φ = −0.35) and Saint Croix (φ = −0.15) show negative lag-1 correlation, which the one-sided Durbin–Watson test does not detect. A value of −0.35 may indicate over-fitting of short-term structure in Belize.

Two caveats apply. Bins without records are absent, so lag 1 does not always correspond to 100 years; this matters most for Saint Croix (18 of 60 bins) and Belize (27 of 60). With *n* = 18–58, both tests have limited power, especially Ljung–Box at lag 10 with *n* = 18.

<!-- REVISED 2026-09-30: dashed lines are now the panel-specific 95 % limits -->

**Supplementary Figure S7.** ACF and PACF of the residuals of the selected combined GAMs, by panel. Dashed lines mark the approximate 95 % limits for white noise (±1.96/√*n*), from ±0.26 in the regional composite (*n* = 58) to ±0.46 in Saint Croix (*n* = 18). Lags are counted in bins, which are not always consecutive centuries (see text).

## Session information

<!-- REVISED 2026-09-30: R and package versions recorded for reproducibility -->

R and package versions used to render this document:

```{r session-info, echo=TRUE}
sessionInfo()
```
