---
title: "6000-yr APAL-Climate Regional Analysis"
author: "Alexis E Medina Valmaseda"
abstract: |
  This document is the executable record (Quarto/R) of all analyses reported in the
  main text. It uses three input files deposited in the repository: `ages_kernel_reg_updt2.xlsx`
  (dated *Acropora palmata* fossils, audited against the original age publications),
  `df_GAM130326_regimes_v3.xlsx` (centennial climate predictors) and `Marine20_0-15ka.csv`
  (the Marine20 radiocarbon calibration curve, used to propagate age uncertainty). From these it reproduces:

  1. binning of fossil ages into 0.1-ka (centennial) intervals, their alignment with
     climate predictors and the normalization of the response variable;
  2. kernel-density reconstructions of *A. palmata* age distributions for each reef
     site and for the regional composite, weighted so that they have the shape of the
     modelled series;
  3. exploratory analyses of predictor correlation and of abundance across
     climate-regime combinations;
  4. beta-regression generalized additive models (GAMs) fitted separately to four reef
     sites (Punta Maroma, Florida, Belize, Saint Croix) and to a regional composite,
     first with one climate predictor at a time (marginal models) and then as a nested
     sequence of multi-predictor models, ranked by AIC and Akaike weights;
  5. diagnostics of the selected models (residual plots, concurvity, residual temporal
     autocorrelation) and sensitivity analyses;
  6. validation and robustness tests of the selected models (cross-validation against an
     s(age)-only reference model, circular-shift test, drop-one influence).

  Code, outputs and a factual description of each result are shown together, so that
  every value reported in the main text can be traced to the code that produced it.
  Ecological interpretation is given in the main text.
format: html
execute:
  echo: true
---

# REGIONAL ACROPORA PALMATA ANALYSIS

We import the fossil chronology (`ages_kernel_reg_updt2.xlsx`, sheet `Registros`) and retain *A. palmata* records (`group == "palmata"`) dated 0–6 ka. The site colours defined here (`custom_colors`) are used in all figures.

**Age reference, calibration and data corrections.** The chronology was audited record by record against the original publications of each age (workbook `coral_ages_audit_M3.xlsx`; sheet `Trazabilidad` of `ages_kernel_reg_updt2.xlsx` lists, for every record, its source, method, published age, analytical uncertainty, calibration and the change with respect to the previous version, `ages_kernel_reg_updt.xlsx`). All ages are expressed in ka BP (before 1950 CE), the reference of the climate series, and all radiocarbon ages are calibrated with Marine20 using a local reservoir correction (ΔR), propagated in quadrature.

- **Punta Maroma.** U-Th ages (Medina-Valmaseda et al. 2025) are published relative to the year of chemical separation (2021) and are converted to BP (0.071 ka younger). Ages from Blanchon et al. (2017) were already in years BP. One sample (SURF7(a), \~1956 CE) post-dates 1950; it keeps its age (−0.006 ka) and is assigned to the youngest bin. Two samples dated on separate subsamples (M1(a), M5(b)) retain both ages.

- **Saint Croix.** Radiocarbon ages (Hubbard et al. 2005, 2013), originally calibrated with Marine98 and no local correction, are recalibrated with Marine20 and ΔR = 64 ± 32 ¹⁴C yr (DiNapoli et al. 2020). For samples without a published conventional age (Hubbard et al. 2013), the conventional age and its uncertainty are recovered as the values that reproduce the published calibrated age and 2σ range under the original calibration. One sample dated at two laboratories (BI2-05) enters once, with the error-weighted mean of its conventional ages; one published calibrated age inconsistent with its conventional age (BI1-13) is replaced by the recalibrated value; samples listed in both source publications are counted once. Four U-Th ages are not modified.

- **Belize.** U-Th ages (Gischler et al. 2023) are in years BP and are not modified. Radiocarbon ages are recalibrated with Marine20 and ΔR = −248 ± 35 ¹⁴C yr (Greer et al. 2020), from the conventional ages in Gischler & Hudson (2004) or, for the atoll cores, from conventional ages recovered by inverting the original Marine98 calibration. Because this ΔR is a single, strongly negative estimate, we repeated the analysis using a regional ΔR of −114 ± 70 ¹⁴C yr (inverse-variance weighted mean of the ten nearest entries of the Marine Reservoir Database, with their standard deviation as uncertainty). Seven *A. cervicornis* samples previously coded as *A. palmata* were reclassified and excluded from the *A. palmata* series, and one typographical error in the compiled table was corrected from the original publication.

- **Florida.** We used in the analysis radiocarbon ages recalibrated with Marine20 by L. T. Toth (pers. comm.); three samples absent from that compilation were recalibrated with the same local correction. U-Th ages were already in years BP. Seventeen rows that repeated a sample already present under another source database were removed. Records were standardized by collection (main reef tract, nearshore ridge, Dry Tortugas) as before.

- **Effect.** *A. palmata* records in 0–6 ka: 257 → 236 (Punta Maroma 64, Florida 111, Belize 39, Saint Croix 22). Radiocarbon ages carry 95 % calibrated ranges of 0.3–0.5 kyr (3–5 centennial bins), so for the radiocarbon-dated sites the 0.1-kyr bin is an aggregation unit, not the resolving power of the records.

- **Calibration curve and age uncertainty.** The age uncertainty shown in Fig. 2 is computed within this document, not taken from the source publications: the Marine20 curve (Heaton et al. 2020; https://doi.org/10.1017/RDC.2020.68), deposited in the repository as `Marine20_0-15ka.csv` (0–15,000 cal BP, columns cal BP, ¹⁴C age and its 1σ error), is combined with the conventional age, its error and the local ΔR with its error of each radiocarbon age (sheet `Trazabilidad`) to obtain its calibrated probability distribution, from which the ages are drawn (Section 1.4). No external calibration software is needed to reproduce it.

<!-- Verificar antes de citar: DiNapoli et al. 2020, Greer et al. 2020. Heaton et al. 2020 = Marine20, Radiocarbon 62: 779–820. -->

```{r datos-reg}
rm(list=ls(all=TRUE))
gc()
#install.packages("here")
library(here)
here()
library(readxl)
library(dplyr)
library(ggplot2)
library(stargazer)

set.seed(20260930)
datos_reg <- read_excel(here("ages_kernel_reg_updt2.xlsx"), sheet = "Registros")

datos_reg_pal <- datos_reg %>%
  # post-1950 ages (t_ka < 0) enter the youngest bin
  filter(group == "palmata", t_ka > -0.1 & t_ka <= 6)

custom_colors <- c(
  "Punta Maroma" = "#F2941A",
  "Regional (All spots)" = "#4D4D4D",
  "Regional"             = "#4D4D4D",
  "Saint Croix" = "#CC79A7",
  "Belize" = "#7D9E5F",
  "Florida" = "#0072B2"
)
```

Each record is assigned to a 0.1-ka bin (right-closed intervals; the first interval includes 0), and the upper bound of the bin (`interval_max`) is stored as its centennial age. A second copy of all records is labelled "Regional", and the two sets are stacked into `centennial_counts_all`. This object is still record-level (one row per dated fossil); counts per bin are computed in later chunks.

```{r cent-count}
library(dplyr)

breaks <- seq(0, 6, by = 0.1)

datos_reg_pal <- datos_reg %>%
  # post-1950 ages (t_ka < 0) enter the youngest bin
  filter(group == "palmata", t_ka > -0.1 & t_ka <= 6)

centennial_counts_sites <- datos_reg_pal %>%
  mutate(interval_bin = cut(
    pmax(t_ka, 0),
    breaks = breaks,
    include.lowest = TRUE,
    right = TRUE
  )) %>%
  mutate(interval_max = breaks[match(interval_bin, levels(interval_bin)) + 1])

centennial_counts_regional <- datos_reg %>%
  # post-1950 ages (t_ka < 0) enter the youngest bin
  filter(group == "palmata", t_ka > -0.1 & t_ka <= 6) %>%
  mutate(interval_bin = cut(
    pmax(t_ka, 0),
    breaks = breaks,
    include.lowest = TRUE,
    right = TRUE
  )) %>%
  mutate(
    spot = "Regional",
    interval_max = breaks[match(interval_bin, levels(interval_bin)) + 1]
  )

centennial_counts_all <- bind_rows(centennial_counts_sites, centennial_counts_regional)
```

## *1. Data Overview*

### 1.1 Summary of fossil records by site

We summarize the raw record-level ages by site (**Supplementary Table S1**): number of records, mean, SD, minimum and maximum age. `min_interval` and `max_interval` give the upper bound of the bins containing the youngest and oldest record.

```{r count-summ}
library(dplyr)
library(purrr)
library(gt)
library(stringr)

tabla_resumen_counts <- centennial_counts_all %>%
  group_by(spot) %>%
  summarise(
    n_records   = n(),                               
    mean_t_ka   = mean(t_ka, na.rm = TRUE),         
    sd_t_ka     = sd(t_ka, na.rm = TRUE),           
    min_t_ka    = min(t_ka, na.rm = TRUE),          
    max_t_ka    = max(t_ka, na.rm = TRUE),          
    min_interval = interval_max[which.min(t_ka)],   
    max_interval = interval_max[which.max(t_ka)],   
    .groups = "drop"  ) %>%
  arrange(spot)


tab_S1 <- tabla_resumen_counts %>%
  gt() %>%
  tab_header(
    title = "Summary of fossil records by site (raw data)",
    subtitle = "Temporal distribution and dispersion per 0.1 ka bin"
  ) %>%
  fmt_number(
    columns = c(mean_t_ka, sd_t_ka, min_t_ka, max_t_ka,
                min_interval, max_interval),
    decimals = 3
  )

tab_S1
gt::gtsave(tab_S1, "Supplementary_Table_S1.docx")   
```

The dataset comprises 236 dated records: **Florida** 111, **Punta Maroma** 64, **Belize** 39 and **Saint Croix** 22. Mean ages range from 2.20 ka (**Punta Maroma**) to 3.56 ka (**Florida**). **Punta Maroma** records span −0.006 ka (one post-1950 colony, assigned to the youngest bin) to 5.46 ka, **Belize** 0–5.54 ka, **Saint Croix** 0.073–5.81 ka and **Florida** 0.069–6.00 ka.

### 1.2 Climatic predictor: import & alignment

We import the centennial climate table (`df_GAM130326_regimes_v3.xlsx`) and assign each row the upper bound of its 0.1-ka bin. Ages are in ka BP (before 1950 CE), the convention of the two continuous climate sources (Kaufman et al. 2020; Schmitt et al. 2025).

The three categorical regimes were coded from the published chronologies compiled in Supplementary Material 2, converted to ka BP, assigning each bin to the period that contains its midpoint (a midpoint on a boundary goes to the younger period):

- **Thermal** (published chronology, not the GMST series): warmer 0–0.1 ka (post-1850 CE); LIA 0.1–0.7 ka (\~1300–1850 CE); CWP&MWA 0.7–1.3 ka (700–1250 CE); cooler 1.3–4.8 ka; warmer 4.8–6.0 ka (mid-Holocene warmth).
- **Hurricane** (Schmitt et al. 2025 only, the source of `Hur100_av` and `Hur100_fqcy`): high 0–0.2 ka (rise after 1800 CE); calm 0.2–0.6 ka (1400–1800 CE); enhanced 0.6–1.1 and 1.2–1.4 ka (900–1400 and 600–800 CE); moderate 1.1–1.2 and 1.4–4.0 ka (mean 9–14 events per century); low 4.0–6.0 ka (\~7 events per century; the record begins at 5.7 ka).
- **Hydro**, by region. Yucatán (`Hydro`): wet 0–0.8 ka, dry 0.8–1.7 ka and wet 1.7–3.5 ka (Curtis et al. 1996, boundaries calibrated with IntCal20), wet 5.2–5.7 ka (Serrato Marks et al. 2021), no data elsewhere. Florida (`Hydro_FL`): dry 4.6–6.0 ka (Pollock et al. 2017; Donders 2014), wet 2.8–4.6 ka and dry 0–2.8 ka (Glaser et al. 2013), with multi-century droughts near 1.0 and 0.4 ka (Willard & Bernhardt 2011).
- **Hurricane, Florida** (`Hur_FL`, `FL_pact`, `FL_nrec`). Schmitt et al. (2025) record storms reaching the Yucatán coast, and Florida-proximal records do not follow it (agreement in 34 % of the bins younger than 3.5 ka). The Florida panel therefore uses a local series built from six records on the Atlantic side of Florida: the western Great Bahama Bank / Florida Straits (Toomey et al. 2013; 0–6 ka), the Everglades (Yao et al. 2020; 0–3.5 ka), southwest Florida (Ercolani et al., via Oliva et al. 2017), Andros and Abaco (van Hengstum et al., via Oliva et al. 2017) and Abaco (Winkler et al. 2020). All ages are cal BP (1950). For each bin, `FL_pact` is the proportion of the records covering the bin that report an active interval at its midpoint, `FL_nrec` the number of those records, and `Hur_FL` is "active" when `FL_pact` ≥ 0.5 and "quiet" otherwise. Older than 3.5 ka the series rests on a single record (Toomey et al. 2013). Records from the Gulf coast of Florida were not used, because they sample Gulf landfalls. For Hydro and the hurricane regime, the Florida panel uses the Florida series; Punta Maroma, Belize, Saint Croix and the regional composite use the Yucatán series (Hurricane = Schmitt et al. 2025). The continuous hurricane metrics (`Hur100_av`, `Hur100_fqcy`) are those of Schmitt et al. (2025) in every panel. In the Yucatán Hydro series, undocumented bins are coded as a separate level, "no data", so that no bins are lost. The mapping is positional: `interval_max` is taken from the order in which interval labels first appear, which is correct only if the file has exactly one row per bin, ordered from youngest to oldest.

Climate values are joined to every fossil record by `interval_max`, and the column is renamed `age` (ka). The resulting `merged_df` is record-level and therefore holds climate values only for bins that contain at least one fossil record.

```{r data-integ}
datos_cli <- read_excel(here("df_GAM130326_regimes_v4.xlsx"))
datos_cli<-as.data.frame(datos_cli)

library(dplyr)

breaks <- seq(0, 6, by = 0.1)

datos_cli <- datos_cli %>%
  mutate(interval = factor(interval, levels = unique(interval))) %>%
  mutate(interval_max = breaks[match(interval, levels(interval)) + 1])

datos_cli <- datos_cli %>%
  mutate(Hydro_YUC = ifelse(is.na(Hydro), "no data", Hydro))

# Hydro and hurricane regime by region: Florida records take the Florida series,
# all other records the Yucatán series (Hurricane = Schmitt et al. 2025)
merged_df <- centennial_counts_all %>%
  left_join(datos_cli, by = "interval_max") %>%
  mutate(Hydro     = ifelse(spot == "Florida", Hydro_FL, Hydro_YUC),
         Hurricane = ifelse(spot == "Florida", Hur_FL, Hurricane)) %>%
  select(-Hydro_FL, -Hydro_YUC, -Hur_FL, -FL_nrec) %>%
  rename( age = interval_max)
```

### 1.3 Response variable: centennial counts and normalization

The response variable of all models is built here, before any figure, so that Figs. 2 and 3 show the same series that enter the GAMs. This chunk makes three lasting changes to `merged_df`:

1.  **Reference levels.** Thermal, Hydro and Hurricane are converted to factors with reference levels "warmer", "wet" and "low" ("quiet" for the local Florida hurricane series). These levels carry through the aggregation below and define every categorical contrast reported in Sections 3.2 and 3.3.

2.  **Sites only.** Rows are restricted to the four reef sites, which drops the duplicated "Regional" records created in the binning step. The regional composite is rebuilt from the site series below.

3.  **Record counter.** A column `counts = 1` is added (one per dated record); it is summed per bin below.

```         
The per-site objects (`datos_pm`, `datos_fl`, `datos_bz`, `datos_sc`, `datos_reg`) and `merged_df_g` created in this chunk are intermediate: all of them are redefined below, before any model is fitted.
```

```{r data-prepa}
library(mgcv)
library(dplyr)
library(patchwork)

merged_df <- merged_df %>%
  mutate(
    Thermal   = relevel(as.factor(Thermal), ref = "warmer"),
    Hydro     = relevel(as.factor(Hydro), ref = "wet"),
    # "low" is the reference of the Schmitt regimes; in Florida (local series) the
    # first level present, "quiet", is the reference
    Hurricane = factor(Hurricane, levels = c("low", "moderate", "calm", "enhanced", "high",
                                             "quiet", "active"))
  )%>%
  filter(spot %in% c("Punta Maroma", "Florida", "Belize", "Saint Croix")) %>%
  mutate(panel = spot)

datos_pm <- merged_df %>% filter(spot == "Punta Maroma")
datos_fl <- merged_df %>% filter(spot == "Florida")
datos_bz <- merged_df %>% filter(spot == "Belize")
datos_sc <- merged_df %>% filter(spot == "Saint Croix")
datos_reg <- merged_df %>%
  filter(spot %in% c("Punta Maroma", "Florida", "Belize", "Saint Croix")) %>%
  mutate(spot = "Regional")

merged_df <- merged_df %>%
   mutate(counts = 1)

merged_df_g <- bind_rows(datos_pm, datos_fl, datos_bz, datos_sc, datos_reg)

```

#### Normalization

For each site, records are summed per bin (`counts`). Climate predictors are averaged within the bin, which leaves them unchanged because they are constant within a bin; regime labels are taken from the first record in the bin.

`normalize_counts()` divides counts by the site maximum (`counts_norm`, rounded to four decimals). Values exactly 0 or 1 are then transformed as y′ = (y(*n* − 1) + 0.5)/*n*, where *n* is the number of bins in the series (`counts_norm_sv`); all other values are unchanged. Only bins containing at least one record exist in the series, so no zeros occur and the correction affects only the maximum bin (e.g. 0.984 in **Punta Maroma**, *n* = 31).

The response is therefore the relative abundance within bins in which a site is represented. Bins without records are not part of any model: 29 of 60 in **Punta Maroma**, 36 in **Belize**, 41 in **Saint Croix**, 16 in **Florida** and 3 in the **regional composite**.

#### Florida source-standardization

Florida combines records from several source datasets (column `source`). Therefore, we first sum counts per source and bin, and further divided by each source's maximum and boundary-corrected. Next we sum the standardized values across sources per bin and normalize as above (`datos_fl_std`). Consequently, each source has equal weight at its maximum, independent of its number of records.

The regional composite used in every model is the per-bin sum of the four site-level `counts_norm_sv` series (**Florida** in its source-standardized form), normalized with the same function (`datos_reg`, *n* = 57 bins). Each site therefore enters the composite with its maximum bin equal to one, so its contribution depends on how its records are distributed among bins rather than on its number of records (Section 1.4). The five series are stacked into `merged_df_g`, which is used by every model and by Figs. 2 and 3a.

```{r norm-stand}

normalize_counts <- function(df, panel_name){
  
  df %>%
    mutate(
      counts_norm = round(
        counts / max(counts, na.rm = TRUE),
        4
      ),
      
      counts_norm_sv = ifelse(
        counts_norm %in% c(0,1),
        ((counts_norm * (n() - 1)) + 0.5) / n(),
        counts_norm
      ),
      
      panel = panel_name
    )
  
}

#---------------------------------------------------------
# PUNTA MAROMA
#---------------------------------------------------------

datos_pm <- merged_df %>%
  filter(spot == "Punta Maroma") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Punta Maroma")

#---------------------------------------------------------
# BELIZE
#---------------------------------------------------------

datos_bz <- merged_df %>%
  filter(spot == "Belize") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Belize")

#---------------------------------------------------------
# SAINT CROIX
#---------------------------------------------------------

datos_sc <- merged_df %>%
  filter(spot == "Saint Croix") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Saint Croix")

#---------------------------------------------------------
# FLORIDA
# Independently standardized by source dataset 
#---------------------------------------------------------

datos_fl_std <- merged_df %>%
  filter(spot == "Florida") %>%
  
  group_by(source, age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  
  group_by(source) %>%
  mutate(
    
    counts_std = counts / max(counts, na.rm = TRUE),
    
    counts_std = ifelse(
      counts_std %in% c(0,1),
      ((counts_std * (n() - 1)) + 0.5) / n(),
      counts_std
    )
    
  ) %>%
  ungroup() %>%
  
  group_by(age) %>%
  summarise(
    
    counts = sum(counts_std, na.rm = TRUE),
    
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    
    .groups = "drop"
    
  ) %>%
  normalize_counts("Florida")

#---------------------------------------------------------
# REGIONAL
# build upon normalized datasets
#---------------------------------------------------------

datos_reg <- bind_rows(
  datos_pm,
  datos_fl_std,
  datos_bz,
  datos_sc
) %>%
  group_by(age) %>%
  summarise(
    
    counts = sum(counts_norm_sv, na.rm = TRUE),
    
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    
    .groups = "drop"
    
  ) %>%
  select(-Hydro) %>%
  left_join(datos_cli %>% transmute(age = interval_max, Hydro = Hydro_YUC, Hur_YUC = Hurricane),
            by = "age") %>%
  mutate(Hydro = factor(Hydro, levels = levels(merged_df$Hydro)),
         Hurricane = factor(Hur_YUC, levels = levels(merged_df$Hurricane))) %>%
  select(-Hur_YUC) %>%
  normalize_counts("Regional")

#---------------------------------------------------------
# Ready-to use dataset
#---------------------------------------------------------

merged_df_g <- bind_rows(
  datos_pm,
  datos_fl_std,
  datos_bz,
  datos_sc,
  datos_reg
)
```

### 1.4 *Acropora palmata* age densities (weighted KDE)

Kernel density estimates (KDE) are computed from the individual ages, weighted so that the density has the shape of the modelled series of Section 1.3. Each record receives the weight it carries in `counts_norm_sv`: in Punta Maroma, Belize and Saint Croix, `counts_norm_sv / counts` of its bin; in Florida, the product of its source-standardization factor and the Florida normalization factor; in the regional composite, the site weight multiplied by the regional normalization factor. The weights of the records in each bin therefore add up to `counts_norm_sv` of that bin (checked in the code). For a site with a single source the weights are constant except in the maximum bin, so the weighted KDE has the same shape as the KDE of the raw ages; the weighting changes the shape only for Florida (source standardization) and for the regional composite.

Each density is a Gaussian KDE over 0–6 ka (bandwidth 0.09 ka, 512 points) with a reflection boundary correction (Silverman 1986; Jones 1993): ages are mirrored at 0 and 6 ka, so that the part of each kernel that would fall outside the window is folded back into it. Without the correction the density is underestimated within about two bandwidths of each edge, by up to one half at the edge itself, which would understate the modern peak and the abundance of the oldest centuries; with it each curve integrates to one over 0–6 ka. Post-1950 ages count at 0 ka, as in the binning. The bandwidth is a smoothing choice, not the resolving power of the record: the 95 % range of a radiocarbon age spans 0.3–0.5 ka (3–5 centennial bins), that of a U-Th age 0.02–0.05 ka. Two sources of uncertainty are estimated separately:

- **Sampling** (which fossils were recovered and dated): 1,000 bootstrap resamples of the records with replacement, keeping their weights and stratified by source in Florida and by site in the regional composite. For each resample we keep the density curve and its mean in each 0.1-ka bin. `signal_all` holds the estimate and the 2.5–97.5 % and 25–75 % bootstrap envelopes of each curve; `centennial_density_all` holds the bin means of the estimate and their bootstrap SD and 95 % interval.
- **Age** (where each dated fossil falls in time): 1,000 Monte Carlo draws of every age, with the record weights held at their central values; draws older than 6 ka leave the window. U-Th ages are drawn from a normal distribution with SD equal to half the reported 2σ error. Radiocarbon ages are drawn from their calibrated probability distribution, computed here from the conventional age, its error and the local ΔR with its error (combined in quadrature) under Marine20 (Heaton et al. 2020; https://doi.org/10.1017/RDC.2020.68; curve deposited as `Marine20_0-15ka.csv`), and shifted so that its median equals the audited age (the shift is ≤ 1 yr except for a few Florida ages calibrated with other software by their authors). The inputs are read from the sheet `Trazabilidad` of `ages_kernel_reg_updt2.xlsx`. Three Florida radiocarbon ages without a ΔR error take the median ΔR error of Florida. The 2.5–97.5 % range of the draws is the age envelope (`age_q025`, `age_q975` in `signal_all`) and the 95 % range of each age is shown in Fig. 2.

Because a KDE integrates to one, the regional density distributes a fixed total among the four sites in proportion to the sum of their weights, which depends on how their records are spread among bins, not on how many records they have. Saint Croix, whose 22 records occupy 19 bins with at most two records each, carries 26 % of the regional weight with 9 % of the records, and Punta Maroma 12 % with 27 % of the records (`site_share`). This is a property of the per-site normalization used in the models, not of the KDE, and it applies equally to the regional GAMs.

```{r dens}
library(dplyr)
library(gt)

## ---- record-level weights reproducing the modelled response -------------------
## The weights of the records in a bin add up to counts_norm_sv of that bin, so the
## weighted KDE has the shape of the series that enter the GAMs.
bin_upper <- function(t) breaks[as.integer(cut(pmax(t, 0), breaks, include.lowest = TRUE, right = TRUE)) + 1]

recs <- datos_reg_pal %>%
  mutate(t_ka = as.numeric(t_ka), age = round(bin_upper(t_ka), 1))

# Punta Maroma, Belize, Saint Croix: weight = counts_norm_sv / counts of the bin
w_single <- bind_rows(datos_pm, datos_bz, datos_sc) %>%
  transmute(spot = panel, age = round(age, 1), w_site = counts_norm_sv / counts)

# Florida: per-source standardization (as in datos_fl_std), then the Florida normalization
w_fl_src <- recs %>%
  filter(spot == "Florida") %>%
  count(source, age, name = "n_src") %>%
  group_by(source) %>%
  mutate(std = n_src / max(n_src),
         std = ifelse(std %in% c(0, 1), (std * (n() - 1) + 0.5) / n(), std),
         w_src = std / n_src) %>%
  ungroup() %>%
  select(source, age, w_src)
w_fl_bin <- datos_fl_std %>% transmute(age = round(age, 1), w_bin = counts_norm_sv / counts)

# Regional composite: site weights rescaled by the regional normalization
w_reg_bin <- datos_reg %>% transmute(age = round(age, 1), w_reg = counts_norm_sv / counts)

recs <- recs %>%
  left_join(w_single, by = c("spot", "age")) %>%
  left_join(w_fl_src, by = c("source", "age")) %>%
  left_join(w_fl_bin, by = "age") %>%
  left_join(w_reg_bin, by = "age") %>%
  mutate(w_site = ifelse(spot == "Florida", w_src * w_bin, w_site),
         w_regional = w_site * w_reg) %>%
  select(-w_src, -w_bin, -w_reg)

# check: the weights of each bin reproduce the modelled series
chk_site <- recs %>% group_by(spot, age) %>% summarise(w = sum(w_site), .groups = "drop") %>%
  inner_join(bind_rows(datos_pm, datos_fl_std, datos_bz, datos_sc) %>%
               transmute(spot = panel, age = round(age, 1), y = counts_norm_sv), by = c("spot", "age"))
chk_reg <- recs %>% group_by(age) %>% summarise(w = sum(w_regional), .groups = "drop") %>%
  inner_join(datos_reg %>% transmute(age = round(age, 1), y = counts_norm_sv), by = "age")
stopifnot(nrow(chk_site) == sum(merged_df_g$panel != "Regional"),
          nrow(chk_reg) == sum(merged_df_g$panel == "Regional"),
          max(abs(chk_site$w - chk_site$y)) < 1e-3, max(abs(chk_reg$w - chk_reg$y)) < 1e-3)

# share of each site in the regional density (weights sum to one)
site_share <- recs %>% group_by(spot) %>%
  summarise(records = n(), pct_records = 100 * n() / nrow(recs),
            pct_weight = 100 * sum(w_regional) / sum(recs$w_regional), .groups = "drop")
site_share

## ---- weighted KDE with stratified bootstrap -------------------------------------
## Records are resampled with replacement within strata (sources in Florida, sites in the
## regional composite), keeping their weights. bw = 0.09 ka, 512 points over 0-6 ka.
## Boundary correction by reflection (Silverman 1986; Jones 1993): ages are mirrored at
## 0 and 6 ka, so the kernel mass that would fall outside the window is folded back
## inside it. Without it the density is underestimated within ~2 bandwidths of each edge
## (by up to half at the edge itself). Output on 512 points over 0-6 ka.
## As in the binning, post-1950 ages (t < 0) count at 0 ka, and ages > 6 ka (Monte Carlo
## draws that leave the window) are dropped.
kde_reflect <- function(t, w, bw, n = 512) {
  keep <- t <= 6
  t <- pmax(t[keep], 0); w <- w[keep] / sum(w[keep])
  d <- density(c(t, -t, 12 - t), weights = rep(w, 3) / 3, bw = bw, from = 0, to = 6, n = n)
  d$y <- 3 * d$y
  d
}

calc_density_summary <- function(t, w, spot_name, strata = rep(1, length(t)),
                                 B = 1000, bw = 0.09) {
  kde <- function(i) kde_reflect(t[i], w[i], bw = bw)
  full <- kde(seq_along(t))
  idx  <- split(seq_along(t), strata)
  boot <- t(vapply(seq_len(B), function(b) {
    i <- unlist(lapply(idx, function(ix) ix[sample.int(length(ix), length(ix), replace = TRUE)]))
    kde(i)$y
  }, numeric(512)))

  signal_df <- data.frame(
    x = full$x, estimate = full$y,
    q025 = apply(boot, 2, quantile, 0.025), q975 = apply(boot, 2, quantile, 0.975),
    q25  = apply(boot, 2, quantile, 0.25),  q75  = apply(boot, 2, quantile, 0.75),
    spot = spot_name)

  # centennial summary: estimate and bootstrap SD of the bin-mean density
  bin <- cut(full$x, breaks = breaks, include.lowest = TRUE, right = TRUE)
  bin_mean <- function(y) tapply(y, bin, mean)
  boot_bins <- apply(boot, 1, bin_mean)                 # 60 bins x B
  centennial_summary <- data.frame(
    interval_mid = breaks[-length(breaks)] + 0.05,
    density      = as.numeric(bin_mean(full$y)),
    boot_sd      = apply(boot_bins, 1, sd),
    boot_q025    = apply(boot_bins, 1, quantile, 0.025),
    boot_q975    = apply(boot_bins, 1, quantile, 0.975),
    spot         = spot_name)

  list(signal_df = signal_df, centennial_summary = centennial_summary)
}

site_levels <- c("Punta Maroma", "Belize", "Saint Croix", "Florida")
results_list <- lapply(site_levels, function(sp) {
  r <- recs %>% filter(spot == sp)
  calc_density_summary(r$t_ka, r$w_site, sp,
                       strata = if (sp == "Florida") r$source else rep(1, nrow(r)))
})
results_list <- c(results_list, list(
  calc_density_summary(recs$t_ka, recs$w_regional, "Regional composite", strata = recs$spot)))

signal_all <- bind_rows(lapply(results_list, `[[`, "signal_df"))
centennial_density_all <- bind_rows(lapply(results_list, `[[`, "centennial_summary"))

## ---- Supplementary Table S2 ---------------------------------------------------
tabla_resumen <- centennial_density_all %>%
  mutate(spot = factor(spot, levels = c(site_levels, "Regional composite"))) %>%
  group_by(Site = spot) %>%
  summarise(
    mean_boot_sd  = mean(boot_sd),
    max_density   = max(density),
    max_interval  = interval_mid[which.max(density)],
    min_density   = min(density),
    min_interval  = interval_mid[which.min(density)],
    .groups = "drop")

tab_S2 <- tabla_resumen %>%
  gt() %>%
  tab_header(title = "Centennial weighted kernel-density summaries by site",
             subtitle = "Density of the modelled series and its bootstrap uncertainty") %>%
  fmt_number(columns = c(mean_boot_sd, max_density, min_density), decimals = 3) %>%
  fmt_number(columns = c(max_interval, min_interval), decimals = 2)
tab_S2
gt::gtsave(tab_S2, "Supplementary_Table_S2.docx")
```

**Supplementary Table S2. Centennial weighted kernel-density summaries.** For each site and the regional composite: mean bootstrap SD of the bin-mean density over the 60 bins (`mean_boot_sd`), and the maximum and minimum bin-mean density with the bin midpoint (ka) at which each occurs.

Maximum density occurs at 0.05 ka (the youngest bin) in **Punta Maroma**, **Belize** and the **regional composite**, at 0.85 ka in **Saint Croix** and at 5.95 ka in **Florida**, where many records fall at 5.8–6.0 ka. Minima fall at 5.95 ka (**Punta Maroma**), 2.45 ka (**Belize** and **Saint Croix**), 1.45 ka (**Florida**) and 0.45 ka (**regional composite**). With the audited ages the youngest bin holds the maximum of the regional curve with or without the weights.

**Intervals without records and low-density intervals.** Two kinds of interval are marked in Fig. 2, both computed from the audited ages:

- **Published presence gaps** (coloured bars) for **Punta Maroma** (Medina-Valmaseda et al., 2025) and **Belize** (Hubbard et al., 2005; Gischler et al., 2023). Each is drawn as the record-free interval of the audited ages that overlaps the published gap (`site_gaps`), so its limits follow the corrected chronology.
- **Low-density intervals** (grey shading), in every panel, including the regional composite, for which no gaps have been published: intervals of at least 0.2 ka in which the upper 95 % limits of both the sampling and the age envelopes stay below the density of a uniform distribution over 0–6 ka (1/6 per ka; dashed line). These are centuries with fewer records than expected under a uniform distribution of the dated fossils, irrespective of which fossils were dated and of their age uncertainty (`low_density`). The rule recovers independently the published gaps of Punta Maroma and Belize. In the regional composite it identifies a single interval, during the Little Ice Age; the regional intervals marked in the previous version (2.20–3.00 and 5.10–5.90 ka) do not meet it with the audited ages.

```{r fig2-uncert}
library(readxl)
## ---- record metadata (method, analytical uncertainty, calibration inputs) -------
## from the traceability sheet of the chronology, linked by record ID
trace <- read_excel(here("ages_kernel_reg_updt2.xlsx"), sheet = "Trazabilidad") %>%
  select(ID, method, err_2s_ka, conv14C, conv14C_err, dR, dR_err) %>%
  mutate(across(c(err_2s_ka, conv14C, conv14C_err, dR, dR_err), as.numeric))
recs <- recs %>% left_join(trace, by = "ID")
stopifnot(!any(is.na(recs$method)), !any(is.na(recs$err_2s_ka)))

## ---- age uncertainty: Monte Carlo draws of every age ------------------------------
## U-Th: normal, sd = 2-sigma error / 2. Radiocarbon: draws from the calibrated
## probability distribution (Marine20; conventional age, its error and the local dR,
## errors combined in quadrature), shifted so that its median equals the audited age
## (the shift is <= 1 yr except where the published calibration used other software).
## Three Florida 14C ages without a dR error take the median dR error of Florida.
m20  <- read.csv(here("Marine20_0-15ka.csv"), comment.char = "#")
g    <- 0:15000
m20_mu <- approx(m20$cal_BP, m20$c14_age, g)$y
m20_sd <- approx(m20$cal_BP, m20$sigma, g)$y
dre_fl <- median(recs$dR_err[recs$spot == "Florida" & recs$method == "14C"], na.rm = TRUE)

N_age <- 1000
draw_age <- function(i) {
  r <- recs[i, ]
  if (r$method != "14C") return(rnorm(N_age, r$t_ka, r$err_2s_ka / 2))
  dre <- ifelse(is.na(r$dR_err), dre_fl, r$dR_err)
  s2  <- r$conv14C_err^2 + dre^2 + m20_sd^2
  p   <- exp(-(r$conv14C - r$dR - m20_mu)^2 / (2 * s2)) / sqrt(s2)
  p   <- p / sum(p)
  med <- g[which(cumsum(p) >= 0.5)[1]]
  x   <- sample(g, N_age, replace = TRUE, prob = p) + runif(N_age, -0.5, 0.5)
  (x - med) / 1000 + r$t_ka
}
set.seed(20261003)
age_draws <- t(vapply(seq_len(nrow(recs)), draw_age, numeric(N_age)))   # records x draws

## 95 % range of each age (for the record panels)
recs <- recs %>%
  mutate(lo95 = apply(age_draws, 1, quantile, 0.025),
         hi95 = apply(age_draws, 1, quantile, 0.975))

## ---- weighted KDE under age uncertainty (weights held at their central values) ------
kde_age_env <- function(idx, w, bw = 0.09) {
  dens <- vapply(seq_len(N_age), function(j)
    kde_reflect(age_draws[idx, j], w, bw = bw)$y,
    numeric(512))
  data.frame(age_q025 = apply(dens, 1, quantile, 0.025),
             age_q975 = apply(dens, 1, quantile, 0.975))
}
age_env <- bind_rows(
  lapply(site_levels, function(sp) {
    i <- which(recs$spot == sp)
    cbind(spot = sp, x = signal_all$x[signal_all$spot == sp], kde_age_env(i, recs$w_site[i]))
  }),
  cbind(spot = "Regional composite", x = signal_all$x[signal_all$spot == "Regional composite"],
        kde_age_env(seq_len(nrow(recs)), recs$w_regional)))
signal_all <- signal_all %>% left_join(age_env, by = c("spot", "x"))

## ---- intervals without records -----------------------------------------------------
## Sites: the record-free interval of the audited ages that overlaps each published gap
## (Punta Maroma: Medina-Valmaseda et al. 2025; Belize: Hubbard et al. 2005, Gischler et al. 2023).
pub_gaps <- data.frame(spot = c("Punta Maroma", "Belize", "Belize", "Belize"),
                       pub_min = c(0.71, 1.99, 3.68, 5.54), pub_max = c(1.44, 2.72, 4.20, 6.00))
gap_from_records <- function(t, a, b) {
  e <- sort(unique(c(t, 6)))
  k <- which(head(e, -1) < b & tail(e, -1) > a)
  k <- k[which.max(diff(e)[k])]
  c(e[k], e[k + 1])
}
site_gaps <- pub_gaps %>% rowwise() %>%
  mutate(g = list(gap_from_records(recs$t_ka[recs$spot == spot], pub_min, pub_max)),
         xmin = g[1], xmax = g[2]) %>% ungroup() %>% select(-g)

## Regional composite: no published gaps. Low-density intervals are recalculated as the
## intervals (>= 0.2 ka) in which the upper 95 % limit of BOTH envelopes (sampling and age)
## stays below the density of a uniform distribution over 0-6 ka (1/6 per ka).
low_runs <- function(x, flag, min_len = 0.2) {
  r <- rle(flag); end <- cumsum(r$lengths); start <- end - r$lengths + 1
  out <- data.frame(xmin = x[start], xmax = x[end])[r$values, , drop = FALSE]
  out[out$xmax - out$xmin >= min_len, , drop = FALSE]
}
low_density <- bind_rows(lapply(c(site_levels, "Regional composite"), function(sp) {
  d <- signal_all %>% filter(spot == sp)
  cbind(spot = sp, low_runs(d$x, pmax(d$q975, d$age_q975) < 1 / 6))
}))
reg_gaps <- low_density %>% filter(spot == "Regional composite")

site_gaps
low_density
```

**Figure 2** (`Figure_2.pdf`). *Acropora palmata* age distributions and their uncertainty. (a–d) Each site: weighted kernel density (bandwidth 0.09 ka, reflection-corrected at 0 and 6 ka; line) with its 95 % sampling envelope (bootstrap of the records; light) and 95 % age envelope (Monte Carlo of the ages; dark); below, every dated fossil with its 95 % age range, by dating method (circles U-Th, triangles ¹⁴C). (e) Regional composite. Dashed horizontal line (every panel): density expected if the dated fossils were spread evenly over 0–6 ka (uniform distribution, 1/6 ≈ 0.17 ka⁻¹), the reference used to define the low-density intervals. Grey shading: low-density intervals (both upper 95 % limits below the uniform density). Coloured bars: published presence gaps, bounded by the audited ages. In Punta Maroma (U-Th only) the age envelope is narrower than the line at this scale: its chronology is effectively fixed and its uncertainty is almost entirely sampling. Where the age envelope is as wide as the sampling envelope (Saint Croix, dated only by radiocarbon), the timing of density peaks and troughs is not resolved at the centennial scale.

```{r fig2m, fig.width=8.5, fig.height=10}
library(patchwork)
## ---- Figure 2 ---------------------------------------------------------------------
## Per site: weighted KDE (bw 0.09 ka) with its 95 % sampling (bootstrap) and age (Monte Carlo)
## envelopes, and below it every age with its 95 % range, by dating method.
## Grey shading: low-density intervals (both upper 95 % limits below the uniform density);
## coloured bars: published presence gaps, bounded by the audited ages.
meth_shape <- c("U-Th" = 16, "14C" = 17)
env_alpha  <- c("Sampling (bootstrap)" = 0.22, "Age (Monte Carlo)" = 0.5)
panel_lab  <- c(site_levels, "Regional composite")
panel_col  <- c(custom_colors[site_levels], "Regional composite" = unname(custom_colors["Regional"]))
n_lab <- recs %>% count(spot, method) %>% group_by(spot) %>%
  summarise(lab = paste0("n = ", sum(n), " (", paste(paste(n, sub("14C", "\u00b9\u2074C", method)), collapse = ", "), ")"), .groups = "drop")
n_lab <- bind_rows(n_lab, data.frame(spot = "Regional composite", lab = paste0("n = ", nrow(recs), " records, 4 sites")))

x_common <- list(scale_x_continuous(breaks = 0:6, minor_breaks = seq(0, 6, 0.5), expand = c(0.005, 0)),
                 coord_cartesian(xlim = c(0, 6)),
                 theme_classic(base_size = 9),
                 theme(plot.tag = element_text(face = "bold"), legend.position = "bottom",
                       panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3)))

kde_panel <- function(sp, show_x = FALSE) {
  d   <- signal_all %>% filter(spot == sp)
  col <- unname(panel_col[sp])
  top <- max(d$q975, d$age_q975) * 1.05
  env <- bind_rows(
    data.frame(x = d$x, lo = d$q025, hi = d$q975, env = "Sampling (bootstrap)"),
    data.frame(x = d$x, lo = d$age_q025, hi = d$age_q975, env = "Age (Monte Carlo)"))
  env$env <- factor(env$env, names(env_alpha))
  ld  <- low_density %>% filter(spot == sp)
  sg  <- site_gaps %>% filter(spot == sp)
  p <- ggplot() +
    geom_rect(data = ld, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf), fill = "grey88") +
    geom_ribbon(data = env, aes(x, ymin = lo, ymax = hi, alpha = env), fill = col) +
    geom_hline(yintercept = 1 / 6, linetype = "dashed", linewidth = 0.3, colour = "grey40") +
    geom_line(data = d, aes(x, estimate), colour = col, linewidth = 0.7) +
    scale_alpha_manual(values = env_alpha, name = "95 % envelope", drop = FALSE) +
    labs(x = NULL, y = expression("Density (ka"^-1*")"), title = sp, subtitle = n_lab$lab[n_lab$spot == sp]) +
    x_common + theme(axis.text.x = if (show_x) element_text() else element_blank(),
                     plot.title = element_text(colour = col, face = "bold", size = 10, margin = margin(b = 1)),
                     plot.subtitle = element_text(colour = "grey30", size = 7.5, margin = margin(b = 2)))
  if (nrow(sg)) p <- p + geom_rect(data = sg, aes(xmin = xmin, xmax = xmax),
                                    ymin = top * 1.13, ymax = top * 1.2, fill = col, alpha = 0.8)
  p + scale_y_continuous(limits = c(0, top * 1.22), expand = c(0, 0))
}

rec_panel <- function(sp, show_x = TRUE) {
  d <- recs %>% filter(spot == sp) %>%
    mutate(method = factor(method, c("U-Th", "14C")),
           y = as.numeric(method) + (rank(t_ka, ties.method = "first") %% 4 - 1.5) * 0.12)
  ggplot(d, aes(y = y)) +
    geom_errorbarh(aes(xmin = pmax(lo95, 0), xmax = pmin(hi95, 6)), height = 0, colour = panel_col[sp],
                   alpha = 0.55, linewidth = 0.35) +
    geom_point(aes(x = pmax(t_ka, 0), shape = method), colour = panel_col[sp], size = 0.9) +
    scale_shape_manual(values = meth_shape, labels = c("U-Th", "\u00b9\u2074C"), name = "Dating method", drop = FALSE) +
    scale_y_continuous(breaks = 1:2, labels = c("U-Th", "\u00b9\u2074C"), limits = c(0.55, 2.45)) +
    labs(x = NULL, y = NULL) +
    x_common + theme(axis.text.x = if (show_x) element_text() else element_blank(),
                     axis.line.y = element_blank(), axis.ticks.y = element_blank())
}

tags <- setNames(letters[1:5], panel_lab)
## sites in the order used by every figure of this document (site_levels)
pl <- list(kde_panel(site_levels[1]) + labs(tag = "a"), kde_panel(site_levels[2]) + labs(tag = "b"),
           rec_panel(site_levels[1]), rec_panel(site_levels[2]),
           kde_panel(site_levels[3]) + labs(tag = "c"), kde_panel(site_levels[4]) + labs(tag = "d"),
           rec_panel(site_levels[3]), rec_panel(site_levels[4]),
           kde_panel("Regional composite", show_x = TRUE) + labs(tag = "e", x = "Age (ka BP)"))
design <- "
AB
CD
EF
GH
II"
fig2 <- wrap_plots(pl, design = design, heights = c(3, 1.1, 3, 1.1, 3), guides = "collect") &
  theme(legend.position = "bottom") &
  guides(alpha = guide_legend(override.aes = list(fill = "grey30")),
         shape = guide_legend(override.aes = list(colour = "grey30", size = 2)))

fig2
ggsave("Figure_2.pdf", plot = fig2, width = 8.5, height = 10, dpi = 300, device = cairo_pdf)
```

**Supplementary Figure S1** (`Supplementary_Figure_S1.pdf`). Sensitivity of the weighted kernel densities to the bandwidth (0.05, 0.09 and 0.15 ka), with the 95 % sampling (light) and age (dark) envelopes of each. Narrower bandwidths resolve more peaks, but the age envelope widens relative to the sampling envelope, most in the radiocarbon-dated sites. Peaks and troughs that persist at 0.15 ka are not artefacts of the smoothing choice.

```{r figS1-bandwidth, fig.width=10, fig.height=10}
bw_set <- c(0.05, 0.09, 0.15)
bw_env <- bind_rows(lapply(bw_set, function(bw) {
  set.seed(20261003)
  bind_rows(lapply(c(site_levels, "Regional composite"), function(sp) {
    reg <- sp == "Regional composite"
    i   <- if (reg) seq_len(nrow(recs)) else which(recs$spot == sp)
    w   <- if (reg) recs$w_regional else recs$w_site[i]
    st  <- if (reg) recs$spot else if (sp == "Florida") recs$source[i] else rep(1, length(i))
    smp <- calc_density_summary(recs$t_ka[i], w, sp, strata = st, B = 1000, bw = bw)$signal_df
    cbind(smp[, c("x", "estimate", "q025", "q975", "spot")], kde_age_env(i, w, bw = bw), bw = bw)
  }))
}))
bw_env$spot <- factor(bw_env$spot, panel_lab)  # site_levels, as in every figure
figS1 <- ggplot(bw_env, aes(x)) +
  geom_ribbon(aes(ymin = q025, ymax = q975, fill = spot), alpha = 0.22) +
  geom_ribbon(aes(ymin = age_q025, ymax = age_q975, fill = spot), alpha = 0.5) +
  geom_line(aes(y = estimate, colour = spot), linewidth = 0.5) +
  scale_fill_manual(values = panel_col, guide = "none") +
  scale_colour_manual(values = panel_col, guide = "none") +
  facet_grid(spot ~ paste("bandwidth", bw, "ka"), scales = "free_y") +
  scale_x_continuous(breaks = 0:6, expand = c(0.005, 0)) +
  labs(x = "Age (ka BP)", y = expression("Density (ka"^-1*")")) +
  theme_classic(base_size = 9) + theme(strip.background = element_blank(), strip.text = element_text(face = "bold"))
figS1
ggsave("Supplementary_Figure_S1.pdf", plot = figS1, width = 10, height = 10, dpi = 300, device = cairo_pdf)

## relative width of the envelopes (sum of 95 % ranges / sum of the estimate), by bandwidth
bw_env %>% group_by(Panel = spot, bw) %>%
  summarise(sampling = sum(q975 - q025) / sum(estimate), age = sum(age_q975 - age_q025) / sum(estimate),
            .groups = "drop") %>%
  mutate(age_over_sampling = age / sampling) %>% mutate(across(where(is.numeric), ~ round(.x, 2)))
```

### 1.5 Abundance and Holocene trends in climatic variables

Panel (a) shows the modelled series of Section 1.3: the relative abundance `counts_norm_sv` of each site per 0.1-ka bin (Florida source-standardized), stacked by site. The height of each stack is the per-bin sum from which the regional composite is normalized. Bins without records have no bar. The legend gives the number of records per site.

**Figure 3a** (`Figure_3a.pdf`). Relative abundance of *A. palmata* per 100-year interval, stacked by reef site.

```{r count-by-site}
library(ggplot2)
library(dplyr)

n_records <- table(datos_reg_pal$spot)
site_labels <- setNames(paste0(site_levels, " (n = ", n_records[site_levels], ")"), site_levels)

norm_stack <- bind_rows(datos_pm, datos_fl_std, datos_bz, datos_sc) %>%
  transmute(age, spot = factor(panel, levels = rev(site_levels)), counts_norm_sv)

pr <- ggplot(norm_stack, aes(x = age, y = counts_norm_sv, fill = spot)) +
  geom_col(position = "stack", width = 0.09, alpha = 0.85) +
  scale_fill_manual(values = custom_colors[site_levels], labels = site_labels,
                    breaks = site_levels) +
  labs(x = "Time (ka BP)",
       y = "Relative abundance\n(stacked by site)",
       fill = "Site") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom") +
  scale_x_continuous(breaks = seq(0, 6, 0.5), minor_breaks = seq(0, 6, 0.5)) +
  coord_cartesian(xlim = c(0, 6))
pr
ggsave("Figure_3a.pdf", plot = pr, width = 8, height = 5, dpi = 300, device = cairo_pdf)
```

We assemble Fig. 3. Panel (a) is Fig. 3a. Below it, four panels show, for each of the 60 centennial bins:

- **(b)** the `GMHT_CPS0_30N` temperature anomaly (^o^C);
- **(c)** accumulated rate of change in thermal anomalies (ẟT,^o^C) calculated from the CPS model of GMST (`dT_raw`);
- **(d)** 100 year average reconstruction of hurricane impacts (`Hur100_av` ) as bars (dark grey where \> 10) and the 100 year TC frequency ( `Hur100_fqcy` ) as a step line;
- **(e)** local hurricane activity in Florida (`FL_pact`): proportion of the Florida-proximal paleotempestological records that report an active interval in each bin, as bars (dark blue where ≥ 0.5, the threshold that defines the "active" regime); bins older than 3.5 ka rest on a single record (Toomey et al. 2013);
- **(f)** the nominal thermal, hydroclimatic (Yucatán and Florida series) and hurricane (Yucatán and Florida series) regimes, drawn as contiguous time segments over the full span of each bin.

**Figure 3** (`Figure_3.pdf`). Reconstruction of regional coral abundance and paleoclimate trends. The climate panels (b–f) are drawn from the 60-bin climate table, so they cover the full 0–6 ka interval, including bins without fossil records.

```{r fig3m, fig.width=12, fig.height=11.5}

library(dplyr)
library(tidyr)
library(patchwork)

clim_bins <- datos_cli %>% rename(age = interval_max)

p1 <- clim_bins %>%
  group_by(age) %>%
  summarise(GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE)) %>%
  ggplot(aes(x = age, y = GMHT_CPS0_30N, fill = GMHT_CPS0_30N > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "#B8860B", "FALSE" = "#0072B2")) +
  coord_cartesian(ylim = c(-1.2, 1.2), xlim = c(0, 6)) +
  scale_x_continuous(position = "top", breaks = seq(0, 6)) +
  geom_hline(yintercept = 1.2, color = "black", linewidth = 0.4) +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    panel.background = element_blank(),
    legend.position = "none",
    axis.title.x = element_blank()
  ) +
  labs(y = "GMST(°C)\nanomalies")

p2 <- clim_bins %>%
  group_by(age) %>%
  summarise(dT_raw = mean(dT_raw, na.rm = TRUE)) %>%
  ggplot(aes(x = age, y = dT_raw, fill = dT_raw > 0)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "#E69F00", "FALSE" = "#56B4E9")) +
  coord_cartesian(ylim = c(-1.2, 0.7)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  ) +
  labs(y = expression("Rate of Change\nẟT(°C) / 100 yr"))


hur_plot_data <- clim_bins %>%
  group_by(age) %>%
  summarise(
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    Hur100_fqcy_sd = sd(Hur100_fqcy, na.rm = TRUE)
  ) %>%
  arrange(age)

p3 <- ggplot(hur_plot_data, aes(x = age)) +
  geom_col(aes(y = Hur100_av, fill = Hur100_av > 10), alpha = 0.6) +
  geom_step(aes(y = Hur100_fqcy), color = "darkblue", linewidth = 0.8) +
  scale_fill_manual(values = c("TRUE" = "#4D4D4D", "FALSE" = "#999999")) +
  coord_cartesian(ylim = c(0, 50)) +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    panel.background = element_blank(),
    legend.position = "none",
    axis.title.x = element_blank()
  ) +
  labs(y = "Hurricane Average\n(events/100 yr)")

# local Florida hurricane series: proportion of Florida-proximal records active in each bin.
# Bars are dark where FL_pact >= 0.5 ("active" regime), as the Schmitt bars are dark above 10.
p3fl <- ggplot(clim_bins, aes(xmin = age - 0.1, xmax = age, ymin = 0, ymax = FL_pact)) +
  geom_rect(aes(fill = FL_pact >= 0.5), alpha = 0.8) +
  annotate("segment", x = 3.5, xend = 6, y = 1.12, yend = 1.12, colour = "grey40", linewidth = 0.3) +
  annotate("text", x = 4.75, y = 1.12, vjust = -0.4, size = 2.6, colour = "grey30",
           label = "one record (Toomey et al. 2013)") +
  scale_fill_manual(values = c("TRUE" = "#08519C", "FALSE" = "#9ECAE1")) +
  scale_x_continuous(breaks = seq(0, 6, 0.5)) +
  coord_cartesian(xlim = c(0, 6), ylim = c(0, 1.25)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.text.x = element_blank()) +
  labs(y = "Hurricane activity,\nFlorida (prop. records)")

# regime segments drawn over the true span of each bin, (age - 0.1, age]
regime_rows <- c(Thermal = "Thermal", Hydro_YUC = "Hydro (Yucatán)",
                 Hydro_FL = "Hydro (Florida)", Hurricane = "Hurricane (Yucatán)",
                 Hur_FL = "Hurricane (Florida)")
df_segments <- clim_bins %>%
  select(age, all_of(names(regime_rows))) %>%
  pivot_longer(-age, names_to = "variable", values_to = "state") %>%
  arrange(variable, age) %>%
  group_by(variable) %>%
  mutate(segment_id = cumsum(state != lag(state, default = first(state)))) %>%
  group_by(variable, state, segment_id) %>%
  summarise(age_start = min(age) - 0.1, age_end = max(age), .groups = "drop") %>%
  mutate(label_x = (age_start + age_end) / 2,
         row = factor(regime_rows[variable], levels = rev(regime_rows)))

p4 <- ggplot(df_segments) +
  geom_rect(aes(xmin = age_start, xmax = age_end,
                ymin = as.numeric(row) - 0.4, ymax = as.numeric(row) + 0.4,
                fill = state), colour = "grey40", linewidth = 0.2) +
  geom_text(aes(x = label_x, y = as.numeric(row),
                label = ifelse(age_end - age_start >= 0.25, state, "")),
            color = "black", size = 2.4) +
  scale_y_continuous(breaks = seq_along(levels(df_segments$row)),
                     labels = levels(df_segments$row)) +
  scale_fill_manual(values = c(
    "warmer"   = "#DAA520", "cooler"   = "#9ECAE1",
    "LIA"      = "#56B4E9", "CWP&MWA"  = "#E69F00",
    "wet"      = "#999933", "dry"      = "#E8DFA0", "no data" = "#FFFFFF",
    "high"     = "#525252", "enhanced" = "#969696", "calm" = "#D9D9D9",
    "moderate" = "#BDBDBD", "low"      = "#F0F0F0",
    "active"   = "#2171B5", "quiet"    = "#DEEBF7")) +
  scale_x_continuous(breaks = seq(0, 6, 0.5)) +
  coord_cartesian(xlim = c(0, 6)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Age (ka BP)", y = "Nominal Terms")

Figure_3 <- pr / p1 / p2 / p3 / p3fl / p4 +
  plot_layout(ncol = 1, heights = c(1, 1, 1, 1, 0.8, 1.7))
Figure_3

ggsave("Figure_3.pdf", plot = Figure_3, width = 8, height = 11.5, dpi = 300, device = cairo_pdf)
```

Before modelling, we inspect pairwise correlations among numeric variables (Section 2.1) and the distribution of abundance across combinations of climate regimes (Section 3.1.3). No predictor is removed on the basis of these analyses. Redundancy is handled instead by never entering the two hurricane metrics in the same model (Section 3.3) and by assessing overlap among smooth terms after fitting (concurvity, Section 4.2).

## **2.** Exploratory Data Analysis

#### 2.1 Variable correlation plot

We compute Pearson correlations among the four continuous climate predictors over the 60 centennial bins of the climate table (`datos_cli`), so that each bin counts once regardless of how many fossil records it contains, and cross-tabulate the three categorical regimes.

The only strong correlation is between the two hurricane metrics (Hur100_fqcy and Hur100_av, *r* = 0.79), which is why they are never entered in the same model (Section 3.3). The other correlations are weak to moderate (\|*r*\| ≤ 0.36, the largest between GMHT_CPS0_30N and dT_raw).

The categorical regimes, by contrast, are partly nested, because they are defined on the same chronology (Yucatán hydroclimate and Schmitt et al. 2025 hurricane series for the Yucatán, Belize and Saint Croix panels and the regional composite; Florida series for the Florida panel):

- **Yucatán hydroclimate.** LIA bins are all wet (6 of 6) and "warmer" bins are never dry (6 wet, 7 without data); CWP&MWA bins are mostly dry (5 of 6). The 20 bins without hydroclimate data all fall in the "cooler" or "warmer" regimes and under low or moderate hurricane activity.
- **Hurricane regime (Schmitt et al. 2025).** "calm" occurs only within LIA (4 of 4 bins), "enhanced" mostly within CWP&MWA (5 of 7), "low" only within the "cooler" and "warmer" regimes, and "high" in two bins (LIA and warmer).
- **Florida series.** The Florida hydroclimate is wet only within the "cooler" regime (18 of 35 bins); all CWP&MWA, LIA and "warmer" bins are dry. All "warmer" bins are quiet in the Florida hurricane series.

With the current data no combined candidate is rank-deficient (Section 3.3), but the nesting means that regime contrasts are partly confounded with one another (for example, LIA with the wet Yucatán hydroclimate and with the calm hurricane regime) and must be read jointly.

**Supplementary Figure S2.** Pearson correlation matrix of the continuous climate predictors across the 60 centennial bins.

```{r corr, fig.width=8, fig.height=7}
library(corrplot)

clim_vars  <- c("GMHT_CPS0_30N", "dT_raw", "Hur100_fqcy", "Hur100_av")
cor_matrix <- cor(datos_cli[, clim_vars])

corrplot(cor_matrix, method = "circle", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black")

pdf("Supplementary_Figure_S2.pdf", width = 8, height = 7)
corrplot(cor_matrix, method = "circle", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black")
invisible(dev.off())

## regime nesting (NA = bins without hydroclimate data)
table(Thermal = datos_cli$Thermal, Hydro = datos_cli$Hydro, useNA = "ifany")
table(Thermal = datos_cli$Thermal, Hurricane = datos_cli$Hurricane)
table(Hydro = datos_cli$Hydro, Hurricane = datos_cli$Hurricane, useNA = "ifany")
## Florida series
table(Thermal = datos_cli$Thermal, Hydro_FL = datos_cli$Hydro_FL)
table(Thermal = datos_cli$Thermal, Hur_FL = datos_cli$Hur_FL)
```

## 3. GAM analysis

### 3.1 Modelling data

All models use `merged_df_g`, built in Section 1.3: the source-standardized Florida series, the three single-source site series and the regional composite. The two subsections below assess the Florida source standardization, and Section 3.1.3 describes abundance across climate-regime combinations.

#### 3.1.1 Florida: pooled normalization (sensitivity series)

`datos_fl_norm` applies the same procedure to **Florida** without source standardization, pooling records across sources before normalization. This series is used in the comparison of Section 3.1.2 and in the sensitivity analysis of Section 4.5, where the selected Florida models are refitted to it; it is not used in model selection.

```{r fl-norm-function}

normalize_counts <- function(df, panel_name){
  df %>%
    mutate(
      counts_norm = round(counts / max(counts, na.rm = TRUE), 4),
      # aplicar SV si hay 0 o 1
      counts_norm_sv = ifelse(counts_norm %in% c(0,1),
                              ((counts_norm * (n() - 1)) + 0.5) / n(),
                              counts_norm),
      panel = panel_name
    )
}

datos_fl_norm <- merged_df %>%
  filter(spot == "Florida") %>%
  group_by(age) %>%
  summarise(
    counts = sum(counts, na.rm = TRUE),
    Hur100_av = mean(Hur100_av, na.rm = TRUE),
    Hur100_fqcy = mean(Hur100_fqcy, na.rm = TRUE),
    GMHT_CPS0_30N = mean(GMHT_CPS0_30N, na.rm = TRUE),
    dT_raw = mean(dT_raw, na.rm = TRUE),
    Hydro = first(Hydro),
    Thermal = first(Thermal),
    Hurricane = first(Hurricane),
    .groups = "drop"
  ) %>%
  normalize_counts("Florida")
```

#### 3.1.2 Florida: effect of source standardization

##### a. Comparison table

We join the pooled series (`norm`) and the source-standardized series (`std`) by bin into `fl_compare`.

```{r florida-comparison}
fl_compare <- full_join(

  datos_fl_norm %>%
    select(age, counts_norm_sv) %>%
    rename(norm = counts_norm_sv),

  datos_fl_std %>%
    select(age, counts_norm_sv) %>%
    rename(std = counts_norm_sv),

  by = "age"

)
```

##### b. Quick assessment

We summarize the differences (std − norm), list the 20 largest, compute the correlation between the two series, and plot both series and their difference.

```{r florida-comparison-visual}

summary(fl_compare$std - fl_compare$norm)

max(abs(fl_compare$std - fl_compare$norm),
    na.rm = TRUE)

fl_compare <- fl_compare %>%
  mutate(
    delta = std - norm
  )

fl_compare %>%
  mutate(
    delta = std - norm,
    abs_delta = abs(delta)
  ) %>%
  arrange(desc(abs_delta)) %>%
  select(age, norm, std, delta) %>%
  slice(1:20)


cor(
  fl_compare$norm,
  fl_compare$std,
  use = "complete.obs"
)


p_fl_a <- ggplot(fl_compare, aes(age)) +
  
  geom_ribbon(
    aes(
      ymin = pmin(norm, std),
      ymax = pmax(norm, std)
    ),
    fill = "grey70",
    alpha = 0.5
  ) +
  
  geom_line(
    aes(y = norm,
        colour = "Original"),
    linewidth = 1
  ) +
  
  geom_line(
    aes(y = std,
        colour = "Standardized"),
    linewidth = 1
  ) +
  
  scale_color_manual(
    values = c(
      "Original" = "black",
      "Standardized" = "red"
    )
  ) +
  
  scale_x_reverse() +
  
  theme_bw() +
  
  labs(
    x = "Age (ka BP)",
    y = "Relative abundance",
    colour = NULL
  )

p_fl_b <- ggplot(
  fl_compare,
  aes(age, delta)
) +
  
  geom_hline(
    yintercept = 0,
    linetype = 2
  ) +
  
  geom_area(
    fill = "steelblue",
    alpha = 0.6
  ) +
  
  scale_x_reverse() +
  
  labs(
    x = "Age (ka BP)",
    y = "Standardized - Original"
  ) +
  
  theme_bw(base_size = 12)

library(patchwork)
figS3 <- p_fl_a / p_fl_b + plot_annotation(tag_levels = "a")
figS3
ggsave("Supplementary_Figure_S3.pdf", plot = figS3, width = 8, height = 8, dpi = 300, device = cairo_pdf)
```

The differences have a median of 0.100, a mean of 0.121 and a range of −0.211 to 0.489. The largest occur at 2.5, 3.3 and 5.9–6.0 ka. The two series are correlated at *r* = 0.794. Source standardization therefore preserves the overall temporal pattern but raises relative abundance in most bins (13 of the 44 differences, 30 %, are negative), most strongly at 2.4–2.5, 3.2–3.3 and 5.9–6.0 ka. This comparison concerns the response series; the GAMs refitted to the pooled series are compared with the source-standardized ones in Section 4.5.

**Supplementary Figure S3.** (a) Pooled (black) and source-standardized (red) Florida series; the grey band marks their difference. (b) Difference std − norm. Age is in ka.

#### **3.1.3 *Acropora palmata* abundance across climate-regime combinations**

We combine the three regime labels into `ClimateCombo` (Thermal_Hydro_Hurricane) and tabulate their frequency per panel (**Supplementary Table S3**; each panel block ends with its number of bins). The regional composite is one of the panels, so its bins are not additional observations. No missing values occur.

We then retain combinations observed in at least three bins in at least one panel. Because the filter is applied across panels, a retained combination can still have *n* = 1–2 in a given panel (e.g. LIA_wet_calm, *n* = 1 in Belize; warmer_wet_low, *n* = 1 in Punta Maroma). Nine combinations are excluded, each with one or two bins per panel: cooler_dry_enhanced, CWP&MWA_dry_quiet, LIA_dry_active, LIA_dry_quiet, CWP&MWA_wet_enhanced, LIA_wet_enhanced, CWP&MWA_dry_moderate, LIA_wet_high and warmer_wet_high. The filter retains 27 of 31 bins in Punta Maroma, 40 of 44 in Florida, 20 of 24 in Belize, 15 of 19 in Saint Croix and 51 of 57 in the regional composite.

**The filter excludes the youngest bin (0–0.1 ka).** Its combination, warmer_wet_high, occurs in a single bin, and this bin holds the maximum relative abundance in Punta Maroma (0.98), Saint Croix (0.97) and the regional composite (0.99), and one of the highest values in Belize (0.75; maximum 0.98). Supplementary Table S4 and Supplementary Figure S4 therefore describe abundance across regimes *excluding* the modern peak; that peak is shown in Figs. 2 and 3, and its influence on the models is assessed by refitting them without it (Section 4.5).

**Supplementary Table S4** gives *n*, mean, SD and range of `counts_norm_sv` for each panel and retained combination. A panel-level summary (`resumen_panel`) is also computed but not shown.

##### Climate combinations (frequencies)

```{r climate-combinations}

library(forcats)
eda_counts <- merged_df_g %>%
  
  mutate(
    ClimateCombo = interaction(
      Thermal,
      Hydro,
      Hurricane,
      sep = "_"
    )
  )

eda_counts[!complete.cases(eda_counts), ]

eda_counts$ClimateCombo <- reorder(
  eda_counts$ClimateCombo,
  eda_counts$counts_norm_sv,
  median
)

combo_freq <- eda_counts %>%
  count(panel, ClimateCombo)

combo_freq_tbl <- combo_freq %>%
  mutate(ClimateCombo = as.character(ClimateCombo)) %>%
  group_by(panel) %>%
  group_modify(~ bind_rows(.x, tibble::tibble(ClimateCombo = "Total", n = sum(.x$n)))) %>%
  ungroup()

tab_S3 <- combo_freq_tbl %>%
  gt(groupname_col = "panel") %>%
  tab_header(
    title = "Climate combinations (frequencies)",
    subtitle = "Number of centennial bins per panel and climatic combination"
  ) %>%
  fmt_number(columns = n, decimals = 0) %>%
  cols_label(ClimateCombo = "Climate combination", n = "Bins") %>%
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_body(rows = ClimateCombo == "Total"))

tab_S3
gt::gtsave(tab_S3, "Supplementary_Table_S3.docx")
```

##### Climate-combination summary

```{r climate-combination-summary}
combos_validos <- combo_freq %>%
  filter(n >= 3) %>%
  pull(ClimateCombo) %>%
  as.character() %>%
  unique()

## excluded combinations and the bins they leave out of Table S4 / Fig. S4
combo_freq %>% filter(!as.character(ClimateCombo) %in% combos_validos) %>%
  group_by(ClimateCombo) %>% summarise(bins = paste(panel, n, collapse = "; "), .groups = "drop")
eda_counts %>% group_by(panel) %>%
  summarise(bins = n(), retained = sum(as.character(ClimateCombo) %in% combos_validos), .groups = "drop")

invisible(
  combo_freq %>%
    filter(n >= 3)
)

invisible(
  combo_freq %>%
    filter(n < 3)
)

datos_filtrados <- eda_counts %>%
  filter(
    ClimateCombo %in% combos_validos
  )

datos_filtrados <- datos_filtrados %>%
  group_by(panel) %>%
  mutate(
    ClimateCombo = fct_reorder(
      ClimateCombo,
      counts_norm_sv,
      .fun = median
    )
  ) %>%
  ungroup()

resumen_panel <- datos_filtrados %>%
  group_by(panel) %>%
  summarise(
    n_obs = n(),
    mean_abund = mean(counts_norm_sv, na.rm = TRUE),
    sd_abund = sd(counts_norm_sv, na.rm = TRUE),
    min_abund = min(counts_norm_sv, na.rm = TRUE),
    max_abund = max(counts_norm_sv, na.rm = TRUE),
    .groups = "drop"
  )

resumen_panel_combo <- datos_filtrados %>%
  group_by(
    panel,
    ClimateCombo
  ) %>%
  summarise(
    n_obs = n(),
    mean_abund = mean(counts_norm_sv, na.rm = TRUE),
    sd_abund = sd(counts_norm_sv, na.rm = TRUE),
    min_abund = min(counts_norm_sv, na.rm = TRUE),
    max_abund = max(counts_norm_sv, na.rm = TRUE),
    .groups = "drop"
  )

tab_S4 <- resumen_panel_combo %>%
  gt() %>%
  tab_header(
    title = "Climate-combination summary",
    subtitle = "Standardized Acropora abundance by climatic regime"
  ) %>%
  fmt_number(
    columns = c(
      mean_abund,
      sd_abund,
      min_abund,
      max_abund
    ),
    decimals = 3
  )

tab_S4
gt::gtsave(tab_S4, "Supplementary_Table_S4.docx")
```

##### Coral-Climate Interactions

The figure below summarizes the distribution of standardized *A. palmata* abundance across the retained climate-regime combinations. Combinations are ordered by their median abundance over all panels pooled, so the order is the same in every facet.

```{r figS4, fig.width=12, fig.height=10}

figS4 <- ggplot(
  datos_filtrados,
  aes(
    x = fct_reorder(
      ClimateCombo,
      counts_norm_sv,
      .fun = median
    ),
    y = counts_norm_sv,
    fill = panel
  )
) +
  
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.7
  ) +
  
  geom_point(
    position = position_jitter(width = 0.12, seed = 1),  
    size = 2,
    alpha = 0.6,
    shape = 21,
    colour = "black"
  ) +
  
  facet_wrap(
    ~ panel,
    scales = "free_y",
    ncol = 2
  ) +
  
  scale_fill_manual(
    values = custom_colors
  ) +
  
  theme_bw() +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 12
    ),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(
      size = 14,
      face = "bold"
    ),
    axis.title.y = element_text(
      size = 14,
      face = "bold"
    ),
    strip.text = element_text(
      size = 14,
      face = "bold"
    ),
    legend.position = "none"
  ) +
  
  labs(
    x = "Climate combinations (ordered by median abundance)",
    y = "Standardized abundance (counts_norm_sv)"
  )

figS4
ggsave("Supplementary_Figure_S4.pdf", plot = figS4, width = 12, height = 10, dpi = 300, device = cairo_pdf)
```

**Supplementary Figure S4. Standardized *A. palmata* abundance under Holocene climate-regime combinations.** Boxplots of `counts_norm_sv` by combination of thermal, hydroclimatic and hurricane regime, per panel; points are individual centennial bins. Combinations are ordered by their median over all panels pooled, so the order is the same in every facet.

Among the retained combinations (Supplementary Table S4), mean abundance is highest under warmer_no data_low in the **regional composite** (0.32, *n* = 6), under warmer_wet_low in **Belize** (0.63, *n* = 2), under cooler_no data_moderate in **Punta Maroma** (0.19, *n* = 4) and under cooler_wet_active in **Florida** (0.51, *n* = 10). It is lowest under LIA_wet_calm in the **regional composite** (0.08, *n* = 3), under LIA_wet_calm and cooler_no data_low in **Belize** (0.25, *n* = 1 and 4), under warmer_wet_low in **Punta Maroma** (0.08, *n* = 1) and under CWP&MWA_dry_active and cooler_wet_quiet in **Florida** (0.29–0.30, *n* = 3–5). Most bins fall under cooler_wet_moderate (*n* = 2–18 per panel) and cooler_no data_low (*n* = 3–8). In **Saint Croix** the response takes essentially two values: 13 of the 15 retained bins hold a single record (0.50), so differences among combinations rest on two bins with two records each (CWP&MWA_dry_enhanced and warmer_wet_low). Most combinations are represented by fewer than five bins per panel, so these patterns are descriptive only.

### 3.2. Marginal GAM effects analysis

Relative abundance is continuous and bounded in (0, 1), so we fit GAMs with a beta distribution and logit link (`mgcv::gam`, family `betar`). Smoothing parameters are estimated by REML. Continuous predictors enter as thin-plate regression splines (mgcv default, basis dimension *k* = 10).

Climate regimes enter as factors with reference levels "warmer" (Thermal), "wet" (Hydro) and "low" (Hurricane). Their coefficients are contrasts on the logit scale relative to the reference, conditional on the other terms. Every model includes a smooth of age, `s(age)`, which absorbs temporal structure not attributed to the climate predictor.

Modelling proceeds in two stages: marginal models, with `s(age)` plus one climate predictor (this section), and combined models, a nested sequence with two branches, one per hurricane metric (Section 3.3).

Candidates are ranked by AIC, ΔAIC and Akaike weights; models within ΔAIC ≤ 2 of the best are considered competitive. Marginal models are fitted without `select = TRUE` and with *k* = 10, whereas combined models use `select = TRUE` and a reduced *k*; AIC values are therefore not comparable between the two stages.

In this document, "linear" models are those whose climate term is a categorical factor (a parametric term), and "smooth" models those whose climate term is a continuous predictor fitted as a spline.

#### 3.2.1 Candidate models

`formulas_mg` defines seven marginal candidates, each consisting of `s(age)` plus one of: `s(GMHT_CPS0_30N)` (M_temp), `s(dT_raw)` (M_dt), `s(Hur100_fqcy)` (M_hf), `s(Hur100_av)` (M_hav), Thermal (M_the), Hydro (M_hyd) or Hurricane (M_hur). `fit_gams_panel()` fits all candidates to one panel, and `compare_gams()` returns AIC, ΔAIC and Akaike weights.

**Hydroclimate in panels with gaps in its series.** The Yucatán hydroclimate series, used for Punta Maroma, Belize, Saint Croix and the regional composite, has no data for 20 of the 60 bins (level "no data"). That level is not a climatic state: as a factor level it contrasts periods with and without a hydroclimate record, and in preliminary fits it produced effects that had no climatic meaning (e.g. Saint Croix, where the hydroclimate model ranked first only through the "no data" contrast). In every panel whose series has such bins, the Hydro term is therefore removed from all candidates (`drop_hydro()`): M_hyd is not fitted, and in the combined sequence (Section 3.3) M5 is dropped and M6 and M7 keep the thermal and hurricane regimes without Hydro. Florida, whose hydroclimate series has no gaps, keeps all candidates. The marginal comparison therefore involves six candidates in Punta Maroma, Belize, Saint Croix and the regional composite, and seven in Florida.

```{r setup-formulas}
formulas_mg <- list(
  M_temp = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N),
  M_dt   = counts_norm_sv ~ s(age) + s(dT_raw),          
  M_hf   = counts_norm_sv ~ s(age) + s(Hur100_fqcy),
  M_hav  = counts_norm_sv ~ s(age) + s(Hur100_av),
  M_the  = counts_norm_sv ~ s(age) + Thermal,
  M_hyd  = counts_norm_sv ~ s(age) + Hydro,
  M_hur  = counts_norm_sv ~ s(age) + Hurricane
)

## ---- Hydroclimate in panels with gaps in its series ---------------------------------
## The Yucatán hydroclimate series (used for Punta Maroma, Belize, Saint Croix and the
## regional composite) has no data for part of the record (level "no data"). That level is
## not a climatic state: as a factor level it contrasts periods with and without a
## hydroclimate record and can produce spurious effects. In any panel whose series has
## such bins, the Hydro term is therefore removed from every candidate; candidates that
## become duplicates of another (M5 = M4) or lose their only climate term (M_hyd) are
## dropped, and M6 / M7 keep the thermal and hurricane regimes without Hydro. Florida,
## whose hydroclimate series has no gaps, keeps all candidates.
hydro_has_gaps <- function(data) any(as.character(data$Hydro) == "no data", na.rm = TRUE)
drop_hydro <- function(formulas, data) {
  if (!hydro_has_gaps(data)) return(formulas)
  has_h <- vapply(formulas, function(f) "Hydro" %in% all.vars(f), logical(1))
  red   <- lapply(formulas, function(f) if ("Hydro" %in% all.vars(f)) update(f, . ~ . - Hydro) else f)
  key   <- vapply(red, function(f) paste(sort(attr(terms(f), "term.labels")), collapse = " + "), "")
  red[!(has_h & (duplicated(key) | key == "s(age)"))]
}

fit_gams_panel <- function(data, formulas) {
  formulas <- drop_hydro(formulas, data)
  lapply(formulas, function(f)
    gam(f, data = data, family = betar(link = "logit"), method = "REML"))
}

compare_gams <- function(fits) {
  stats <- lapply(fits, AIC)
  df <- data.frame(Model = names(stats), AIC = unlist(stats))
  df <- df %>%
    mutate(deltaAIC = AIC - min(AIC),
           weight = exp(-0.5 * deltaAIC) / sum(exp(-0.5 * deltaAIC)))
  return(df)
}
```

#### 3.2.2 Model competition (ΔAIC and Akaike weights)

We fit the seven candidates to each panel of `merged_df_g` and plot ΔAIC and Akaike weights by panel.

**Supplementary Figure S5. Information-theoretic comparison of the marginal GAMs by panel.** (a) ΔAIC of each candidate (Supplementary Table S5); the red dashed line marks ΔAIC = 2, and the ΔAIC axis is scaled separately for each panel. (b) Akaike weights, on a common 0–1 scale. Candidates are listed in the same order in every panel.

```{r marginal-results, fig.width=12, fig.height=7}
panels <- unique(merged_df_g$panel)

results <- lapply(panels, function(p) {
  
  df_panel <- merged_df_g %>%
    filter(panel == p)
  
  if(nrow(df_panel) < 3 ||
     all(is.na(df_panel$counts_norm_sv)))
    return(NULL)
  
  fits <- fit_gams_panel(
    df_panel,
    formulas_mg
  )
  
  stats <- compare_gams(fits)
  
  stats$panel <- p
  
  stats
  
})

results <- Filter(
  Negate(is.null),
  results
)

stats_all <- bind_rows(results)

library(patchwork)

site_order <- c("Punta Maroma", "Florida", "Belize", "Saint Croix", "Regional")

plot_model_selection <- function(stats, panel_var, model_levels) {
  d <- stats %>%
    mutate(Panel = factor(.data[[panel_var]], levels = site_order),
           Model = factor(Model, levels = rev(model_levels)))
  base <- function(yvar, ylab) {
    ggplot(d, aes(x = Model, y = .data[[yvar]], colour = Panel)) +
      geom_segment(aes(xend = Model, y = 0, yend = .data[[yvar]]), linewidth = 0.8) +
      geom_point(size = 2.2) +
      coord_flip() +
      scale_x_discrete(drop = FALSE) +
      scale_colour_manual(values = custom_colors[site_order], guide = "none") +
      labs(x = NULL, y = ylab) +
      theme_minimal(base_size = 12) +
      theme(strip.text = element_text(face = "bold"),
            panel.grid.minor = element_blank())
  }
  p_d <- base("deltaAIC", "ΔAIC") +
    geom_hline(yintercept = 2, linetype = "dashed", colour = "red") +
    facet_wrap(~Panel, nrow = 1, drop = FALSE, scales = "free_x")
  p_w <- base("weight", "Akaike weight") +
    facet_wrap(~Panel, nrow = 1, drop = FALSE) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1))
  (p_d / p_w) + plot_annotation(tag_levels = "a")
}

sup_fig4 <- plot_model_selection(stats_all, panel_var = "panel", model_levels = names(formulas_mg))
sup_fig4
ggsave("Supplementary_Figure_S5.pdf", plot = sup_fig4, width = 12, height = 7, dpi = 300, device = cairo_pdf)
```

#### 3.2.3 Marginal GAM model ranking

**Supplementary Table S5. Marginal GAM ranking by panel.** For each panel, the seven candidates are ordered by AIC, with ΔAIC, Akaike weight and rank. (`get_best_models()` refits the same models as Section 3.2.2 on the same data.)

```{r best-gams}

linear_models <- c(
  "M_the",
  "M_hyd",
  "M_hur"
)

smooth_models <- c(
  "M_temp",
  "M_dt",
  "M_hf",
  "M_hav"
)

get_best_models <- function(data, formulas){
  
  fits <- fit_gams_panel(data, formulas)
  
  ranking <- compare_gams(fits) %>%
    arrange(AIC) %>%
    mutate(
      Rank = row_number()
    )
  
  best_linear <- ranking %>%
    filter(Model %in% linear_models) %>%
    slice_min(AIC, n = 1)
  
  best_smooth <- ranking %>%
    filter(Model %in% smooth_models) %>%
    slice_min(AIC, n = 1)
  
  list(
    fits = fits,
    ranking = ranking,
    best_linear = best_linear,
    best_smooth = best_smooth
  )
  
}

best_pm <- get_best_models(datos_pm, formulas_mg)
best_fl <- get_best_models(datos_fl_std, formulas_mg)
best_bz <- get_best_models(datos_bz, formulas_mg)
best_sc <- get_best_models(datos_sc, formulas_mg)
best_reg <- get_best_models(datos_reg, formulas_mg)


tbl_models <- bind_rows(
  
  best_pm$ranking %>%
    mutate(Panel = "Punta Maroma"),
  
  best_fl$ranking %>%
    mutate(Panel = "Florida"),
  
  best_bz$ranking %>%
    mutate(Panel = "Belize"),
  
  best_sc$ranking %>%
    mutate(Panel = "Saint Croix"),
  
  best_reg$ranking %>%
    mutate(Panel = "Regional")
  
)

gt_ranking <- tbl_models %>%
  
  gt(groupname_col = "Panel") %>%
  
  fmt_number(
    columns = c(AIC, deltaAIC, weight),
    decimals = 3
  ) %>%
  
  cols_label(
    Rank = "Rank",
    Model = "Model",
    AIC = "AIC",
    deltaAIC = "ΔAIC",
    weight = "Akaike Weight"
  ) %>%
  
  tab_header(
    title = md("**GAM model selection by panel**")
  )

gt_ranking
gt::gtsave(gt_ranking, "Supplementary_Table_S5.docx")   
```

In **Punta Maroma**, M_hav dominates (weight 0.95); M_the follows at ΔAIC 6.5 and all other models at ΔAIC \> 9. In the **regional composite**, M_hav also ranks first (weight 0.94), followed by M_hf at ΔAIC 6.2. In **Belize**, M_temp (weight 0.42) and M_hur (ΔAIC 0.59, weight 0.31) are competitive, with M_hav at ΔAIC 2.1. In **Florida**, six of the seven models lie within ΔAIC ≤ 1.5 (weights 0.10–0.21), so no single predictor is supported. In **Saint Croix**, five of the six candidates lie within ΔAIC ≤ 1.6 (weights 0.13–0.28); with *n* = 19 bins whose response takes essentially two values (Section 3.1.3), AIC does not discriminate among predictors.

#### 3.2.4 Best-supported marginal GAMs

**Supplementary Table S6.** Best categorical ("linear") and best continuous ("smooth") marginal model per panel, selected by lowest AIC within each group.

```{r best-gams-winners}
tbl_winners <- bind_rows(
  
  data.frame(
    Panel = "Punta Maroma",
    Best_Linear = best_pm$best_linear$Model,
    Best_Smooth = best_pm$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Florida",
    Best_Linear = best_fl$best_linear$Model,
    Best_Smooth = best_fl$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Belize",
    Best_Linear = best_bz$best_linear$Model,
    Best_Smooth = best_bz$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Saint Croix",
    Best_Linear = best_sc$best_linear$Model,
    Best_Smooth = best_sc$best_smooth$Model
  ),
  
  data.frame(
    Panel = "Regional",
    Best_Linear = best_reg$best_linear$Model,
    Best_Smooth = best_reg$best_smooth$Model
  )
  
)

gt_winners <- tbl_winners %>%
  gt() %>%
  cols_label(
    Panel = "Panel",
    Best_Linear = "Best Linear",
    Best_Smooth = "Best Smooth"
  ) %>%
  tab_header(
    title = md("**Best-supported linear and smooth GAMs**")
  )

gt_winners
gt::gtsave(gt_winners, "Supplementary_Table_S6.docx")   
```

We refit these models and print their full `mgcv` summaries.

```{r best-gams-summary}

print_best_models <- function(best_obj,
                              panel_name,
                              formulas,
                              data){
  
  lin_name <- best_obj$best_linear$Model
  smo_name <- best_obj$best_smooth$Model
  
  cat("\n========================================\n")
  cat(panel_name,"\n")
  cat("========================================\n")
  
  cat("\nBEST LINEAR MODEL:", lin_name, "\n\n")
  
  print(
    summary(
      gam(
        formulas[[lin_name]],
        data = data,
        family = betar(link="logit"),
        method = "REML"
      )
    )
  )
  
  cat("\n----------------------------------------\n")
  
  cat("\nBEST SMOOTH MODEL:", smo_name, "\n\n")
  
  print(
    summary(
      gam(
        formulas[[smo_name]],
        data = data,
        family = betar(link="logit"),
        method = "REML"
      )
    )
  )
  
}

print_best_models(best_pm,
                  "Punta Maroma",
                  formulas_mg,
                  datos_pm)

print_best_models(best_fl,
                  "Florida",
                  formulas_mg,
                  datos_fl_std)

print_best_models(best_bz,
                  "Belize",
                  formulas_mg,
                  datos_bz)

print_best_models(best_sc,
                  "Saint Croix",
                  formulas_mg,
                  datos_sc)

print_best_models(best_reg,
                  "Regional",
                  formulas_mg,
                  datos_reg)
```

We then extract, for each of these models: AIC, deviance explained and adjusted R²; the climate term with the lowest term-level *p*-value (`s(age)` excluded) and that *p*-value; the factor contrasts with *p* \< 0.05; the EDF of each smooth; and a support label ("Supported" if the term-level climate *p* \< 0.05, otherwise "Weak support"). Term-level *p*-values come from a joint Wald test of the whole term (`anova()`): for a categorical regime, all levels are tested together; for a continuous predictor, the spline is tested. Contrasts are relative to the reference levels warmer / wet / low. *p*-values below 10⁻⁴ are reported as "\<1e-04", the precision limit of mgcv's approximation for smooth terms.

**Table 1. Best-supported marginal GAMs by panel.** Highest-ranked categorical-regime (linear) and continuous-climate (smooth) models, with model fit, climate term with the lowest term-level *p*-value, significant contrasts versus the reference level, EDF of the smooth terms and support label.

```{r comparison-table}
library(webshot2)
extract_gam_stats <- function(model, panel, model_type, model_name,
                              alpha = 0.05, p_floor = 1e-4) {
  
  sm <- summary(model)
  av <- anova(model)                       # term-level Wald tests
  
  ## 1. Climate p-values at TERM level (s(age) excluded)
  # Parametric terms: one joint test per factor (all levels together)
  pt_tab <- av$pTerms.table
  p_par  <- if (!is.null(pt_tab) && NROW(pt_tab) > 0)
    setNames(pt_tab[, "p-value"], rownames(pt_tab)) else numeric(0)
  # Smooth terms
  st_tab <- sm$s.table
  st_cl  <- st_tab[rownames(st_tab) != "s(age)", , drop = FALSE]
  p_smo  <- setNames(st_cl[, "p-value"], rownames(st_cl))
  
  p_all  <- c(p_par, p_smo)
  p_all  <- p_all[!is.na(p_all)]
  # mgcv's smooth p-values are approximations; values below ~1e-4 (or slightly
  # negative, from rounding) are not meaningful beyond "very small"
  p_all  <- pmax(p_all, 0)
  
  climate_p   <- if (length(p_all)) min(p_all) else NA_real_
  climate_trm <- if (length(p_all)) names(p_all)[which.min(p_all)] else NA_character_
  sig_terms   <- names(p_all)[p_all < alpha]
  
  ## 2. Significant factor CONTRASTS vs. the reference level (for interpretation)
  ptab  <- sm$p.table
  contr <- ptab[rownames(ptab) != "(Intercept)", , drop = FALSE]
  contr <- contr[!is.na(contr[, 4]) & contr[, 4] < alpha, , drop = FALSE]
  sig_contr <- if (NROW(contr))
    paste0(rownames(contr), " (", sprintf("%.2f", contr[, 1]),
           ", p=", format.pval(contr[, 4], digits = 2, eps = p_floor), ")",
           collapse = "; ") else if (NROW(ptab) <= 1) "n/a (no categorical terms)" else "none"
  
  data.frame(
    Panel             = panel,
    Type              = model_type,
    Model             = model_name,
    AIC               = round(AIC(model), 3),
    DevianceExplained = round(sm$dev.expl, 3),
    R2_Adj            = round(sm$r.sq, 3),
    ClimateTerm       = climate_trm,
    Climate_p         = climate_p,
    Climate_p_label   = if (is.na(climate_p)) "NA" else
      format.pval(climate_p, digits = 2, eps = p_floor),
    SignificantTerms  = if (length(sig_terms)) paste(sig_terms, collapse = "; ")
    else "Not significant",
    SignificantContrasts = sig_contr,
    EDF_Terms         = paste0(rownames(st_tab), " (EDF=",
                               round(st_tab[, "edf"], 2), ")", collapse = "; "),
    Inference         = ifelse(!is.na(climate_p) && climate_p < alpha,
                               "Supported", "Weak support"),
    stringsAsFactors  = FALSE
  )
}

model_labels <- c(
  M_the = "Thermal regime",
  M_hyd = "Hydroclimatic trends",
  M_hur = "Hurricane regime",
  M_temp = "GMHT anomalies",
  M_dt = "thermal rate of change δT",
  M_hf = "TC frequency",
  M_hav = "Smoothed TC frequency"
)

build_panel_summary <- function(best_obj,
                                panel_name,
                                formulas,
                                data){
  
  lin_name <- best_obj$best_linear$Model
  smo_name <- best_obj$best_smooth$Model
  
  lin_model <- gam(
    formulas[[lin_name]],
    data = data,
    family = betar(link="logit"),
    method = "REML"
  )
  
  smo_model <- gam(
    formulas[[smo_name]],
    data = data,
    family = betar(link="logit"),
    method = "REML"
  )
  
  bind_rows(
    
    extract_gam_stats(
      lin_model,
      panel_name,
      "Linear",
      lin_name
    ),
    
    extract_gam_stats(
      smo_model,
      panel_name,
      "Smooth",
      smo_name
    )
    
  )
}

tbl_summary <- bind_rows(
  
  build_panel_summary(
    best_pm,
    "Punta Maroma",
    formulas_mg,
    datos_pm
  ),
  
  build_panel_summary(
    best_fl,
    "Florida",
    formulas_mg,
    datos_fl_std
  ),
  
  build_panel_summary(
    best_bz,
    "Belize",
    formulas_mg,
    datos_bz
  ),
  
  build_panel_summary(
    best_sc,
    "Saint Croix",
    formulas_mg,
    datos_sc
  ),
  
  build_panel_summary(
    best_reg,
    "Regional",
    formulas_mg,
    datos_reg
  )
  
)

tbl_summary <- tbl_summary %>%
  mutate(Model_Label = recode(Model, !!!model_labels)) %>%
  select(Panel, Type, Model, Model_Label, AIC, DevianceExplained, R2_Adj,
         ClimateTerm, Climate_p_label, SignificantContrasts, EDF_Terms, Inference)

gt_summary_m <- tbl_summary %>%
  arrange(factor(Panel, levels = c("Punta Maroma", "Florida", "Belize",
                                   "Saint Croix", "Regional")), Type) %>%
  gt(groupname_col = "Panel") %>%
  fmt_number(columns = c(AIC, DevianceExplained, R2_Adj), decimals = 3) %>%
  cols_label(
    Type                 = "Model Type",
    Model                = "Model",
    Model_Label          = "Climate Driver",
    AIC                  = "AIC",
    DevianceExplained    = "Deviance Explained",
    R2_Adj               = "R² Adj",
    ClimateTerm          = "Climate term",
    Climate_p_label      = "Climate p-value (term)",
    SignificantContrasts = "Significant contrasts vs. reference",
    EDF_Terms            = "EDF Terms",
    Inference            = "Inference"
  ) %>%
  cols_width(Model ~ px(90), SignificantContrasts ~ px(240), EDF_Terms ~ px(260)) %>%
  tab_style(style = list(cell_fill(color = "#7298c2"),
                         cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) %>%
  tab_header(title = md("**Best-supported GAMs by panel**"),
             subtitle = md("*Top Linear and Smooth candidates*")) %>%
  tab_footnote(
    footnote = paste("Climate p-value: joint Wald test of the whole term (all factor",
                     "levels together for categorical regimes; the spline for",
                     "continuous predictors). Contrasts are relative to the",
                     "reference levels warmer / wet / low."),
    locations = cells_column_labels(columns = Climate_p_label)) %>%
  opt_row_striping() %>%
  tab_options(table.font.size = 12, data_row.padding = px(5))

gt_summary_m
gt::gtsave(gt_summary_m, "Table_1.docx")
```

Climate terms reach *p* \< 0.05 (term-level test) in the best marginal models of three panels:

- **Punta Maroma:** s(Hur100_av) (EDF 3.46, *p* \< 10⁻⁴; deviance explained 90.0 %) and Thermal (*p* \< 10⁻⁴), driven by lower abundance under LIA (−3.79, *p* \< 10⁻⁴) and, less so, under the "cooler" regime (−1.48, *p* = 0.031) than under the "warmer" reference. The thermal-regime effect is not independent of hurricane activity: added to the selected hurricane model it is not significant (*p* = 0.99; Section 4.5).
- **Regional composite:** s(Hur100_av) (EDF 3.01, *p* \< 10⁻⁴; deviance explained 52.5 %) and Hurricane (*p* = 2.3 × 10⁻⁴), with lower abundance under the "calm" regime (−2.41, *p* = 0.043) than under the "low" reference.
- **Belize:** s(GMHT_CPS0_30N) (EDF 1.0, i.e. linear; *p* = 0.0097; deviance explained 48.9 %). The hurricane regime is not significant at term level (*p* = 0.057), although the "high" contrast is (+4.22, *p* = 0.019; the two youngest bins, 0–0.2 ka).

No climate term reaches *p* \< 0.05 in Florida (Hydro, *p* = 0.66; s(GMHT_CPS0_30N), *p* = 0.39) or Saint Croix (Hurricane, *p* = 0.29; s(dT_raw), *p* = 0.21).

In the marginal models, therefore, hurricane activity (s(Hur100_av)) is the best-supported single predictor in Punta Maroma and the regional composite; the thermal regime is significant only in Punta Maroma and does not add to hurricane activity there; Belize is associated with the temperature anomaly; and neither Florida nor Saint Croix supports any climate predictor.

### 3.3 Combined GAM effects analysis

To capture compounded climatic effects, we fit a nested sequence of additive GAMs, without interaction terms or random effects.

#### 3.3.1 Candidate models

**Supplementary Table S7. Nested sequence of combined GAMs.** Eleven candidates. M1 = `s(age)`; M2 adds `s(GMHT_CPS0_30N)`; M3 adds `s(dT_raw)`. From M3 the sequence splits into two branches, one adding `s(Hur100_fqcy)` and the other `s(Hur100_av)` (M4); each branch then adds Hydro (M5), Thermal (M6) and Hurricane regime (M7) in turn. The two hurricane metrics never appear in the same model. In the panels whose hydroclimate series has gaps (all except Florida), Hydro is removed from every candidate (Section 3.2.1): M5 is dropped, M6 becomes M4 + Thermal and M7 becomes M4 + Thermal + Hurricane, so nine candidates are compared.

```{r combined-formulation}
model_table <- data.frame(
  Model = c("M1", "M2", "M3",
            "M4_fqcy", "M5_fqcy", "M6_fqcy", "M7_fqcy",
            "M4_av", "M5_av", "M6_av", "M7_av"),
  Formula = c(
    "counts_norm_sv ~ s(age)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal + Hurricane",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av)",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal",
    "counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal + Hurricane"
  ),
  Added_Predictors = c(
    "", "GMHT_CPS0_30N", "dT_raw",
    "Hur100_fqcy", "Hydro", "Thermal", "Hurricane",
    "Hur100_av", "Hydro", "Thermal", "Hurricane"
  ),
  Branch = c(
    "Base", "Base", "Base",
    "Hur100_fqcy", "Hur100_fqcy", "Hur100_fqcy", "Hur100_fqcy",
    "Hur100_av", "Hur100_av", "Hur100_av", "Hur100_av"
  ),
  stringsAsFactors = FALSE
)

tab_S7 <- model_table %>%
  gt() %>%
  tab_header(
    title = "Nested sequence of combined GAMs"
  ) %>%
  cols_label(
    Model = "Model",
    Formula = "Formula",
    Added_Predictors = "Added Predictors",
    Branch = "Branch"
  )

tab_S7
gt::gtsave(tab_S7, "Supplementary_Table_S7.docx")   
```

#### 3.3.2 Fitting procedure (ΔAIC and Akaike weights)

`fit_combined_gams_safe()` first removes Hydro where its series has gaps (`drop_hydro()`, Section 3.2.1) and then fits the remaining candidates to each panel in six steps:

1.  It removes rows with missing or infinite numeric values.

2.  It skips a model if *n* \< 8 or *n* ≤ 3 × (number of smooths).

3.  It sets a common basis dimension for all smooths of a model, *k* = max(3, min(10, ⌊(*n* − 1)/number of smooths⌋)). For the four-smooth models this gives *k* = 7 in Punta Maroma (*n* = 31), 5 in Belize (24), 4 in Saint Croix (19), and 10 in Florida (44) and the regional composite (57). Nested models within a panel therefore do not share the same basis size (e.g. in Punta Maroma M1 uses *k* = 10 and M4 uses *k* = 7).

4.  It fits `gam()` (betar, REML) with `select = TRUE`, which penalizes each smooth's null space so that a term can be shrunk to EDF ≈ 0. (`bam()` is used only when *n* \> 100, which does not occur here.)

5.  If the beta fit fails, it refits the model as quasibinomial; such a model has no AIC and is excluded from the ranking.

6.  It excludes rank-deficient fits, i.e. models whose model-matrix rank is lower than the number of coefficients because two parametric terms are aliased. The names of the excluded models are returned in `$excluded`.

The function returns AIC, deviance explained, adjusted R², total EDF, ΔAIC and Akaike weights for each retained model; ΔAIC and weights are computed over the retained candidates only.

**Candidates compared.** With the current data no candidate is rank-deficient in any panel, so the comparison involves 11 candidates in Florida and 9 in the other panels (after removing Hydro). The regimes are nevertheless partly nested (Section 2.1), so contrasts between regime levels in M6 and M7 are partly confounded and are interpreted jointly.

```{r function-fit-combined}

formulas_cb <- list(
  M1 = counts_norm_sv ~ s(age),
  M2 = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N),
  M3 = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw),
  M4_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy),
  M5_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro,
  M6_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal,
  M7_fqcy = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_fqcy) + Hydro + Thermal + Hurricane,
  M4_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av),
  M5_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro,
  M6_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal,
  M7_av = counts_norm_sv ~ s(age) + s(GMHT_CPS0_30N) + s(dT_raw) + s(Hur100_av) + Hydro + Thermal + Hurricane
)

count_smooth_terms <- function(formula) {
  lengths(regmatches(paste(deparse(formula, width.cutoff = 500L), collapse = " "),
                      gregexpr("s\\(", paste(deparse(formula, width.cutoff = 500L), collapse = " "))))
}

cap_smooth_k <- function(formula, k_max, env = parent.frame()) {
  f_chr <- paste(deparse(formula, width.cutoff = 500L), collapse = " ")
  f_chr <- gsub("s\\(\\s*([A-Za-z0-9_\\.]+)\\s*\\)",
                paste0("s(\\1, k = ", k_max, ")"), f_chr)
  as.formula(f_chr, env = env)
}

safe_k_for <- function(n_obs, n_smooth_terms) {
  max(3, min(10, floor((n_obs - 1) / max(1, n_smooth_terms))))
}

fit_combined_gams_safe <- function(formulas_cb, data, segment_name) {

  formulas_cb <- drop_hydro(formulas_cb, data)   # Hydro removed where its series has gaps
  
  data_clean <- data %>%
    mutate(across(where(is.numeric), ~replace(., is.infinite(.), NA))) %>%
    drop_na()
  
  n_obs <- nrow(data_clean)
  
  fits <- lapply(names(formulas_cb), function(name) {
    f <- formulas_cb[[name]]
    n_smooth <- count_smooth_terms(f)
    if (n_obs < 8 || n_obs <= n_smooth * 3) {
      message(paste("Skipping model", name, "for", segment_name,
                    "- insufficient observations (n =", n_obs, ") for",
                    n_smooth, "smooth term(s)"))
      return(NULL)
    }
    
    k_max <- safe_k_for(n_obs, n_smooth)
    f_safe <- if (n_smooth > 0) cap_smooth_k(f, k_max) else f

    # rank-deficient categorical design: excluded before fitting
    # (betar fits on aliased factor designs can abort the R session)
    fv <- intersect(c("Thermal", "Hydro", "Hurricane"), all.vars(f))
    if (length(fv)) {
      X <- model.matrix(reformulate(fv), droplevels(data_clean))
      if (qr(X)$rank < ncol(X)) return("rank-deficient")
    }
    
    tryCatch({
      m <- if (n_obs > 100) {
        bam(f_safe, data = data_clean, family = betar(link = "logit"),
            method = "fREML", select = TRUE, discrete = TRUE)
      } else {
        gam(f_safe, data = data_clean,
            family = betar(link = "logit"), method = "REML",
            select = TRUE)
      }
      environment(m$formula) <- environment()
      m
    }, error = function(e) {
      message(paste("Model", name, "fail in", segment_name,
                    "using betar:", e$message, "→ testing quasibinomial"))
      tryCatch({
        m2 <- gam(f_safe, data = data_clean,
                  family = quasibinomial(link = "logit"), method = "REML",
                  select = TRUE)
        environment(m2$formula) <- environment()
        m2
      }, error = function(e2) {
        message(paste("Model", name, "also failed with quasibinomial in", segment_name, ":", e2$message))
        return(NULL)
      })
    })
  })
  names(fits) <- names(formulas_cb)
  
  fits <- fits[!sapply(fits, is.null)]
  pre_excl <- names(fits)[vapply(fits, is.character, logical(1))]
  fits <- fits[!names(fits) %in% pre_excl]
  
 rank_def <- vapply(fits, function(m) m$rank < length(coef(m)), logical(1))
  excluded <- c(pre_excl, names(fits)[rank_def])
  if (length(excluded)) {
    message("Excluded as rank-deficient in ", segment_name, ": ",
            paste(excluded, collapse = ", "))
    fits <- fits[!rank_def]
  }
  
  stats <- lapply(names(fits), function(name) {
    m <- fits[[name]]
    s <- summary(m)
    data.frame(
      Model = name,
      AIC = tryCatch(AIC(m), error = function(e) NA),
      DevianceExplained = s$dev.expl,
      R2_Adj = s$r.sq,
      EDF = sum(m$edf),
      Segment = segment_name,
      stringsAsFactors = FALSE
    )
  })
  
  stats_df <- bind_rows(stats)
  
  if (nrow(stats_df) > 0) {
    
    stats_df <- stats_df %>%
      mutate(
        deltaAIC = AIC - min(AIC, na.rm = TRUE),
        weight = exp(-0.5 * deltaAIC) /
          sum(exp(-0.5 * deltaAIC), na.rm = TRUE)
      ) %>%
      arrange(AIC)
    
  }
  
  if (nrow(stats_df) > 0) {
    p_delta <- ggplot(stats_df, aes(x = reorder(Model, deltaAIC), y = deltaAIC)) +
      geom_col(fill = "steelblue") +
      geom_hline(yintercept = 2, linetype = "dashed", color = "red") +
      coord_flip() +
      labs(title = paste("ΔAIC by model -", segment_name),
           subtitle = "Dashed line = ΔAIC = 2 (competitive threshold)",
           x = "Model", y = "ΔAIC") +
      theme_minimal() +
      theme(legend.position = "bottom")
    
    p_weight <- ggplot(stats_df, aes(x = reorder(Model, -weight), y = weight)) +
      geom_col(fill = "darkgreen") +
      coord_flip() +
      labs(title = paste("AIC weights by model -", segment_name),
           subtitle = "Relative probability each model is best",
           x = "Model", y = "AIC weight") +
      theme_minimal()
    
    plots <- list(delta = p_delta, weight = p_weight)
  } else {
    plots <- NULL
  }
  
  return(list(fits = fits, stats = stats_df, plots = plots,
              excluded = excluded))  
}

datos_pm <- merged_df_g %>% filter(panel == "Punta Maroma")
results_pm <- fit_combined_gams_safe(formulas_cb, datos_pm, "Punta Maroma")

datos_fl <- merged_df_g %>% filter(panel == "Florida")
results_fl <- fit_combined_gams_safe(formulas_cb, datos_fl, "Florida")

datos_bz <- merged_df_g %>% filter(panel == "Belize")
results_bz <- fit_combined_gams_safe(formulas_cb, datos_bz, "Belize")

datos_sc <- merged_df_g %>% filter(panel == "Saint Croix")
results_sc <- fit_combined_gams_safe(formulas_cb, datos_sc, "Saint Croix")

datos_reg <- merged_df_g %>% filter(panel == "Regional")
results_reg <- fit_combined_gams_safe(formulas_cb, datos_reg, "Regional")

```

#### 3.3.3 Model competition and selected models

We stack the model statistics of all panels (`comparison_cb_full`), store the competitive models (ΔAIC ≤ 2; `competitive_models`, not displayed) and plot ΔAIC and Akaike weights by panel.

```{r suppl-fig5, fig.width=12, fig.height=8}
comparison_cb_full <- bind_rows(results_pm$stats,
                           results_fl$stats,
                           results_bz$stats,
                           results_sc$stats,
                           results_reg$stats)

competitive_models <- comparison_cb_full %>%
  filter(deltaAIC <= 2)%>%
  mutate(Segment = factor(Segment,
                          levels = c("Punta Maroma", "Florida", "Belize", "Saint Croix", "Regional")))

sup_fig5 <- plot_model_selection(comparison_cb_full, panel_var = "Segment", model_levels = names(formulas_cb))
sup_fig5
ggsave("Supplementary_Figure_S6.pdf", plot = sup_fig5, width = 12, height = 8, dpi = 300, device = cairo_pdf)
```

**Supplementary Figure S6. Information-theoretic comparison of the combined GAMs by panel.** (a) ΔAIC of each retained candidate (Supplementary Table S7); the red dashed line marks ΔAIC = 2, and the ΔAIC axis is scaled separately for each panel. (b) Akaike weights, on a common 0–1 scale. Candidates are listed in the same order in every panel; rows without a point are candidates not fitted in that panel (Hydro removed where its series has gaps, Section 3.2.1; rank-deficient candidates, Section 3.3.2).

```{r delta-aic}
library(dplyr)
comparison_cb_full %>%
  filter(deltaAIC <= 2) %>%
  group_by(Segment) %>%
  mutate(sum_weight = sum(weight)) %>%
  ungroup() %>%
  arrange(Segment, deltaAIC) %>%
  select(Segment, Model, deltaAIC, weight, sum_weight) %>%
  mutate(across(c(deltaAIC, weight, sum_weight), ~ round(.x, 2)))
```

**Table 2. Best-supported combined GAM per panel** (lowest AIC): model, AIC, Akaike weight, deviance explained, adjusted R², climate terms with term-level *p* \< 0.05 (`s(age)` excluded), factor contrasts with *p*-value \< 0.05 versus reference level, and EDF of every smooth.

- **Punta Maroma — M4_av** (Akaike weight 0.88; deviance explained 89.8 %; adjusted R² 0.87). s(Hur100_av) is the only significant term (EDF 3.27, *p* \< 10⁻⁴); s(age), s(GMHT_CPS0_30N) and s(dT_raw) are shrunk to EDF ≈ 0.
- **Florida — M4_fqcy** (weight 0.23; 16.5 %; adjusted R² 0.11). No term is significant: s(Hur100_fqcy) (EDF 0.70, *p* = 0.081) and s(GMHT_CPS0_30N) (EDF 1.30, *p* = 0.12) are retained but not significant, and s(age) and s(dT_raw) are shrunk to EDF ≈ 0.
- **Belize — M4_av** (weight 0.62; 68.6 %; adjusted R² 0.59). Significant terms: s(GMHT_CPS0_30N) (EDF 0.94, *p* \< 10⁻⁴), s(Hur100_av) (EDF 1.80, *p* = 6.3 × 10⁻⁴) and s(dT_raw) (EDF 0.87, *p* = 0.006); s(age) is shrunk to EDF ≈ 0.
- **Saint Croix — M3** (weight 0.27; 57.0 %; adjusted R² 0.38). Only s(age) is significant (EDF 1.71, *p* = 0.005); s(GMHT_CPS0_30N) (*p* = 0.076) and s(dT_raw) (*p* = 0.051) are not. This panel is not interpreted (Section 3.3.5).
- **Regional composite — M4_av** (weight 0.71; 53.2 %; adjusted R² 0.44). s(Hur100_av) is the only significant term (EDF 2.70, *p* = 2.3 × 10⁻⁶); s(GMHT_CPS0_30N) (EDF 0.56) is not significant, and s(age) and s(dT_raw) are shrunk to EDF ≈ 0.

No model with a categorical regime (M6, M7) is selected in any panel.

```{r Table-2}

get_best_combined <- function(results_obj){
  
  ranking <- results_obj$stats %>%
    arrange(AIC)
  
  winner_name <- ranking$Model[1]
  
  list(
    best_model = winner_name,
    ranking = ranking,
    fit = results_obj$fits[[winner_name]]
  )
  
}

best_pm_cb <- get_best_combined(results_pm)
best_fl_cb <- get_best_combined(results_fl)
best_bz_cb <- get_best_combined(results_bz)
best_sc_cb <- get_best_combined(results_sc)
best_reg_cb <- get_best_combined(results_reg)

extract_combined_stats <- function(best_obj, panel_name,
                                   alpha = 0.05, p_floor = 1e-4) {
  
  fmt_p <- function(p) ifelse(p < p_floor, "<1e-04", sprintf("%.2g", p))
  
  model <- best_obj$fit
  sm    <- summary(model)
  av    <- anova(model)
  
  ## Term-level p-values (s(age) excluded)
  pt_tab <- av$pTerms.table
  p_par  <- if (!is.null(pt_tab) && NROW(pt_tab) > 0)
    setNames(pt_tab[, "p-value"], rownames(pt_tab)) else numeric(0)
  st_tab <- sm$s.table
  st_cl  <- st_tab[rownames(st_tab) != "s(age)", , drop = FALSE]
  p_smo  <- setNames(st_cl[, "p-value"], rownames(st_cl))
  
  p_all  <- c(p_par, p_smo)
  p_all  <- pmax(p_all[!is.na(p_all)], 0)
  sig    <- p_all[p_all < alpha]
  
  sig_terms <- if (length(sig))
    paste0(names(sig), " (p=", fmt_p(sig), ")",
           collapse = "; ") else "Not significant"
  climate_p <- if (length(p_all)) min(p_all) else NA_real_
  
ptab  <- sm$p.table
  contr <- ptab[rownames(ptab) != "(Intercept)", , drop = FALSE]
  contr <- contr[!is.na(contr[, 4]) & contr[, 4] < alpha, , drop = FALSE]
  sig_contr <- if (NROW(contr))
    paste0(rownames(contr), " (", sprintf("%.2f", contr[, 1]),
           ", p=", fmt_p(contr[, 4]), ")",
           collapse = "; ") else if (NROW(ptab) <= 1) "n/a (no categorical terms)" else "none"
  
  data.frame(
    Panel                = panel_name,
    Model                = best_obj$best_model,
    AIC                  = round(AIC(model), 3),
    DevianceExplained    = round(sm$dev.expl, 3),
    R2_Adj               = round(sm$r.sq, 3),
    SignificantTerms     = sig_terms,
    Climate_p            = climate_p,
    Climate_p_label      = if (is.na(climate_p)) "NA" else
      fmt_p(climate_p),
    SignificantContrasts = sig_contr,
    EDF_Terms            = paste0(rownames(st_tab), " (EDF=",
                                  round(st_tab[, "edf"], 2), ")", collapse = "; "),
    stringsAsFactors     = FALSE
  )
}

tbl_combined <- bind_rows(
  
  extract_combined_stats(
    best_pm_cb,
    "Punta Maroma"
  ),
  
  extract_combined_stats(
    best_fl_cb,
    "Florida"
  ),
  
  extract_combined_stats(
    best_bz_cb,
    "Belize"
  ),
  
  extract_combined_stats(
    best_sc_cb,
    "Saint Croix"
  ),
  
  extract_combined_stats(
    best_reg_cb,
    "Regional"
  )
  
)

combined_labels <- c(
  
  M1 =
    "Temporal baseline",
  
  M2 =
    "Background temperature model",
  
  M3 =
    "Thermal forcing model",
  
  M4_fqcy =
    "Thermal–disturbance model (frequency)",
  
  M5_fqcy =
    "Thermal–disturbance–hydroclimate model",
  
  M6_fqcy =
    "Integrated climate model (frequency)",
  
  M7_fqcy =
    "Full climate-regime model (frequency)",
  
  M4_av =
    "Thermal–disturbance model (smoothed frequency)",
  
  M5_av =
    "Thermal–disturbance–hydroclimate model",
  
  M6_av =
    "Integrated climate model (smoothed frequency)",
  
  M7_av =
    "Full climate-regime model (smoothed frequency)"
)

tbl_combined <- tbl_combined %>%
  mutate(
    Model_Label =
      recode(Model,
             !!!combined_labels)
  )

tbl_combined <- tbl_combined %>%
  mutate(
    Weight = c(
      best_pm_cb$ranking$weight[1],
      best_fl_cb$ranking$weight[1],
      best_bz_cb$ranking$weight[1],
      best_sc_cb$ranking$weight[1],
      best_reg_cb$ranking$weight[1]
    )
  )


tbl_combined <- tbl_combined %>%
  select(Panel, Model, Model_Label, AIC, Weight, DevianceExplained, R2_Adj,
         SignificantTerms, SignificantContrasts, EDF_Terms)

gt_combined <- tbl_combined %>%
  arrange(factor(Panel, levels = c("Punta Maroma", "Florida", "Belize",
                                   "Saint Croix", "Regional"))) %>%
  gt() %>%
  fmt_number(columns = c(AIC, DevianceExplained, R2_Adj), decimals = 3) %>%
  fmt_number(columns = Weight, decimals = 2) %>%
  cols_label(
    Panel                = "Panel",
    Model                = "Model",
    Model_Label          = "Climate Model",
    AIC                  = "AIC",
    Weight               = "Akaike Weight",
    DevianceExplained    = "Deviance Explained",
    R2_Adj               = "R² Adj",
    SignificantTerms     = "Significant terms (term-level p)",
    SignificantContrasts = "Significant contrasts vs. reference",
    EDF_Terms            = "EDF Terms"
  ) %>%
  cols_width(Model ~ px(90), Model_Label ~ px(220), SignificantTerms ~ px(220),
             SignificantContrasts ~ px(240), EDF_Terms ~ px(300)) %>%
  tab_style(style = list(cell_fill(color = "#7298c2"),
                         cell_text(color = "white", weight = "bold")),
            locations = cells_column_labels()) %>%
  tab_header(title = md("**Best-supported combined GAMs by panel**"),
             subtitle = md("*Top multivariate model selected by AIC*")) %>%
  tab_footnote(
    footnote = paste("Term-level p: joint Wald test of the whole term (all factor",
                     "levels together; the spline for continuous predictors);",
                     "s(age) excluded. Contrasts are relative to the reference",
                     "levels warmer / wet / low. Rank-deficient candidates were",
                     "excluded before model selection."),
    locations = cells_column_labels(columns = SignificantTerms)) %>%
  opt_row_striping() %>%
  tab_options(table.font.size = 12, data_row.padding = px(5))

gt_combined
gt::gtsave(gt_combined, "Table_2.docx")   

list(`Punta Maroma` = results_pm$excluded, Florida = results_fl$excluded,
     Belize = results_bz$excluded, `Saint Croix` = results_sc$excluded,
     Regional = results_reg$excluded)
```

The selected model is supported in **Punta Maroma** (Akaike weight 0.88), **Belize** (0.62) and the **regional composite** (0.71): in these panels no other candidate lies within ΔAIC ≤ 2. In **Florida** (0.23) the competitive set (ΔAIC ≤ 2; Supplementary Fig. S6) comprises five models (M4_fqcy, M1, M2, M3 and M4_av; combined weight 0.81) and includes the model with `s(age)` alone (ΔAIC 0.56). In **Saint Croix** (0.27) it comprises four models (M3, M2, M1 and M4_av; ΔAIC ≤ 1.04; combined weight 0.78), again including `s(age)` alone (ΔAIC 0.94). In these two panels the data do not distinguish models with climate terms from the temporal trend alone, so they provide no support for a climate effect (Supplementary Table S8).

```{r competitive-set-terms}

term_pvalues <- function(m) {
  sm <- summary(m); av <- anova(m)
  p_par <- if (!is.null(av$pTerms.table) && NROW(av$pTerms.table) > 0)
             setNames(av$pTerms.table[, "p-value"], rownames(av$pTerms.table)) else numeric(0)
  p_smo <- setNames(sm$s.table[, "p-value"], rownames(sm$s.table))
  pmax(c(p_par, p_smo), 0)
}

results_by_panel <- list("Punta Maroma" = results_pm, "Florida" = results_fl, "Belize" = results_bz,
                         "Saint Croix" = results_sc, "Regional" = results_reg)

comp_terms <- purrr::imap_dfr(results_by_panel, function(r, site) {
  cs <- r$stats %>% filter(deltaAIC <= 2) %>% arrange(deltaAIC)
  purrr::map_dfr(seq_len(nrow(cs)), function(i) {
    p <- term_pvalues(r$fits[[cs$Model[i]]])
    tibble::tibble(Panel = site, Model = cs$Model[i], deltaAIC = cs$deltaAIC[i],
                   Weight = cs$weight[i], Term = names(p), p = p)
  })
})

term_order <- c("Thermal", "Hydro", "Hurricane", "s(age)", "s(GMHT_CPS0_30N)",
                "s(dT_raw)", "s(Hur100_av)", "s(Hur100_fqcy)")

comp_wide <- comp_terms %>%
  mutate(p_label = ifelse(p < 1e-4, "<1e-04", sprintf("%.2g", p)),
         p_label = ifelse(p < 0.05, paste0(p_label, " *"), p_label)) %>%
  select(-p) %>%
  tidyr::pivot_wider(names_from = Term, values_from = p_label, values_fill = "–")
comp_wide <- comp_wide[, c("Panel", "Model", "deltaAIC", "Weight",
                           intersect(term_order, names(comp_wide)))]

tab_S8 <- comp_wide %>%
  gt(groupname_col = "Panel") %>%
  fmt_number(columns = c(deltaAIC, Weight), decimals = 2) %>%
  cols_label(deltaAIC = "ΔAIC", Weight = "Akaike weight") %>%
  tab_header(title = "Term-level p-values across the competitive set (ΔAIC ≤ 2)") %>%
  tab_footnote(footnote = "* p < 0.05; – term not in the model. Joint Wald test per term (all levels of a factor together).")

tab_S8
gt::gtsave(tab_S8, "Supplementary_Table_S8.docx")
```

**Supplementary Table S8. Term-level *p*-values across the competitive set.** For each panel, every combined GAM within ΔAIC ≤ 2 of the best model, with its ΔAIC, Akaike weight and the term-level *p*-value of each term (joint Wald test for factors; \* *p* \< 0.05; – term not in the model).

In **Florida**, no climate term is significant in any competitive model (s(Hur100_fqcy), *p* = 0.081 in M4_fqcy; s(Hur100_av), *p* = 0.39 in M4_av). In **Saint Croix**, s(age) is significant in M1–M3 (*p* = 0.005–0.025) and s(Hur100_av) only in M4_av (*p* = 0.034; ΔAIC 1.04). Across panels, average hurricane activity, s(Hur100_av), is the only climate term that is both selected and significant in Punta Maroma, Belize and the regional composite; in Belize the two temperature terms are also significant.

#### 3.3.4 Full summaries of the selected combined models

We print the complete `mgcv` summaries of the five models in Table 2.

```{r best-combined}
print_best_combined <- function(best_obj,
                                panel_name){
  
  cat("\n====================================\n")
  cat(panel_name, "\n")
  cat("====================================\n")
  
  cat(
    "\nBEST COMBINED MODEL:",
    best_obj$best_model,
    "\n\n"
  )
  
  print(
    summary(
      best_obj$fit
    )
  )
  
}

print_best_combined(best_pm_cb, "Punta Maroma")
print_best_combined(best_fl_cb, "Florida")
print_best_combined(best_bz_cb, "Belize")
print_best_combined(best_sc_cb, "Saint Croix")
print_best_combined(best_reg_cb, "Regional")
```

#### 3.3.5 Saint Croix: limitations and decision

We do not draw site-level conclusions for **Saint Croix**. Its models are reported for completeness and labelled as not interpretable, for six reasons that concern the data rather than the results:

1.  **Dating.** All 22 ages are radiocarbon. The 95 % calibrated range of each spans 3–5 centennial bins (Section 1.4), so the bin of an individual record is uncertain: in 1,000 Monte Carlo draws of the ages, a Saint Croix age stays in its central bin in 37 % of draws on average (Punta Maroma, dated by U-Th: 87 %).
2.  **Response.** 16 of the 19 occupied bins hold a single record, so `counts_norm_sv` takes essentially two values (0.50 and 0.97; Section 3.1.3). The beta model then contrasts a handful of bins with two records against the rest.
3.  **Marginal models.** Five of the six candidates lie within ΔAIC ≤ 1.6 (Section 3.2.3), and no climate term is significant in the best models. Before Hydro was removed, the top-ranked model owed its rank only to the "no data" level of the hydroclimate series.
4.  **Combined models.** The selected model (M3) contains no significant climate term, and the competitive set includes `s(age)` alone (Section 3.3.3).
5.  **Age and sampling uncertainty.** Across replicates of the ages and of the records, the selected model changes from replicate to replicate (the most frequent is chosen in 30–36 %), the climate models improve on `s(age)` in 34–57 % of the replicates and no climate term is significant in more than 32 % of them (Section 4.5.2).
6.  **Density.** At the narrowest bandwidth (0.05 ka) the age envelope of the Saint Croix density is as wide as its sampling envelope (Supplementary Fig. S1).

Saint Croix is kept in the regional composite, because removing a site after examining its results would be a post hoc selection and because it is the only eastern Caribbean record. Its weight in the composite (26 % of the weight with 9 % of the records; Section 1.4) is examined directly: the regional models are refitted without Saint Croix in Section 4.5.

## 4. Model Quality Control (combined effects)

To ensure methodological transparency and meet standard practices in GAM-based ecological modeling, we examine the five selected combined models through residual diagnostics from gam.check() (Q–Q plot, residuals vs. linear predictor, residual histogram, response vs. fitted values), plots of all fitted terms, concurvity, and temporal autocorrelation of the residuals. The basis-dimension check (k-index) is reported with `k.check()` below each set of diagnostic plots.

### 4.1 Partial effects of the selected models

The five selected models are collected in `best_models` and their summaries are printed again. Smooth estimates with 95 % confidence intervals are extracted with `gratia::smooth_estimates()`, pooled across panels and plotted by predictor (one panel per smooth). Saint Croix is not shown, because it is not interpreted (Section 3.3.5); its terms are plotted with its diagnostics (Supplementary Fig. S7). If a selected model includes a categorical regime, its effects are added as an extra panel; with the audited ages no selected model includes one. To put such effects on the same baseline as the smooths, which mgcv constrains to sum to zero over the fitted observations, the factor effects are centred in the same way: each regime's coefficient minus their mean weighted by the number of bins in each regime. Centring shifts all regimes by the same amount, so differences between regimes (the contrasts reported in Table 2) are unchanged; 95 % confidence intervals are computed from the full coefficient covariance matrix.

```{r Fig4, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

library(gratia)
library(dplyr)
library(ggplot2)

best_models <- list(
  "Punta Maroma" = best_pm_cb$fit,
  "Florida"      = best_fl_cb$fit,
  "Belize"       = best_bz_cb$fit,
  "Saint Croix"  = best_sc_cb$fit,
  "Regional"     = best_reg_cb$fit
)

summary(best_models[["Punta Maroma"]])
summary(best_models[["Florida"]])
summary(best_models[["Belize"]])
summary(best_models[["Saint Croix"]])
summary(best_models[["Regional"]])

df_all <- purrr::map_df(names(best_models), function(site) {
  smooth_estimates(best_models[[site]], ci_level = 0.95) %>%
    add_confint() %>%
    mutate(site = site)
})

df_all <- df_all %>%
  mutate(site = factor(site, levels = names(custom_colors)))

smooth_predictor_cols <- intersect(
  c("age", "GMHT_CPS0_30N", "dT_raw", "Hur100_av", "Hur100_fqcy"),
  names(df_all)
)

df_long <- df_all %>%
  pivot_longer(cols = all_of(smooth_predictor_cols),
               names_to = "predictor", values_to = "xvar") %>%
  filter(!is.na(xvar))

library(patchwork)

site_levels <- c("Punta Maroma", "Florida", "Belize", "Saint Croix", "Regional")
site_cols   <- custom_colors[site_levels]

df_long <- df_long %>%
  mutate(site = factor(as.character(site), levels = site_levels)) %>%
  # Saint Croix is not interpreted (Section 3.3.5); its terms are shown in Fig. S7 only
  filter(site != "Saint Croix")
fig4_sites <- setdiff(site_levels, "Saint Croix")

pred_order <- intersect(c("age", "GMHT_CPS0_30N", "dT_raw", "Hur100_av", "Hur100_fqcy"),
                        unique(df_long$predictor))

fig4_theme <- theme_minimal(base_size = 14) +
  theme(axis.title = element_text(size = 12),
        axis.text  = element_text(size = 12),
        strip.text = element_text(size = 13))

# One panel per smooth predictor (same look as the former facet_wrap)
smooth_panels <- lapply(seq_along(pred_order), function(i) {
  pr <- pred_order[i]
  ggplot(filter(df_long, predictor == pr),
         aes(x = xvar, y = .estimate, colour = site, fill = site)) +
    geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci), alpha = 0.15, colour = NA) +
    geom_line(linewidth = 1.3) +
    facet_wrap(~predictor) +
    scale_colour_manual(values = site_cols, limits = fig4_sites, name = "Site") +
    scale_fill_manual(values = site_cols, limits = fig4_sites, name = "Site") +
    labs(x = NULL, y = if (i %% 3 == 1) "Partial effect" else NULL) +
    fig4_theme
})

# centred effects of every categorical regime retained in the selected models
factor_order <- list(Thermal   = c("warmer", "cooler", "CWP&MWA", "LIA"),
                     Hydro     = c("wet", "dry", "no data"),
                     Hurricane = c("low", "moderate", "calm", "enhanced", "high", "quiet", "active"))
factor_title <- c(Thermal = "Thermal regime", Hydro = "Hydroclimate", Hurricane = "Hurricane regime")

centred_effects <- function(fac) {
  sites <- names(best_models)[vapply(best_models, function(m)
    any(grepl(paste0("^", fac), names(coef(m)))), logical(1))]
  if (!length(sites)) return(NULL)
  purrr::map_df(sites, function(site) {
    m    <- best_models[[site]]
    idx  <- grep(paste0("^", fac), names(coef(m)))
    b    <- coef(m)[idx]
    V    <- vcov(m)[idx, idx, drop = FALSE]
    lev  <- levels(droplevels(m$model[[fac]]))
    w    <- as.numeric(table(factor(m$model[[fac]], levels = lev))) / nrow(m$model)
    A    <- do.call(rbind, lapply(lev, function(l) as.numeric(names(b) == paste0(fac, l))))  # level -> coef (matrix also with a single coefficient)
    Cm   <- A - matrix(colSums(w * A), nrow(A), ncol(A), byrow = TRUE)
    est  <- as.numeric(Cm %*% b)
    se   <- sqrt(pmax(diag(Cm %*% V %*% t(Cm)), 0))
    tibble::tibble(level = lev, est = est, lo = est - 1.96 * se, hi = est + 1.96 * se,
                   site = site)
  }) %>%
    mutate(level     = factor(level, levels = intersect(factor_order[[fac]], unique(level))),
           site      = factor(site, levels = site_levels),
           predictor = factor_title[[fac]])
}

factor_panels <- purrr::compact(lapply(names(factor_order), function(fac) {
  df <- centred_effects(fac)
  if (is.null(df)) return(NULL)
  ggplot(df, aes(x = level, y = est, colour = site)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_pointrange(aes(ymin = lo, ymax = hi),
                    position = position_dodge(width = 0.5), linewidth = 1, size = 0.6) +
    facet_wrap(~predictor) +
    scale_colour_manual(values = site_cols, limits = site_levels, guide = "none") +
    labs(x = NULL, y = NULL) +
    fig4_theme
}))

p_all <- wrap_plots(c(smooth_panels, factor_panels), ncol = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p_all
ggsave("Figure_4.pdf", plot = p_all, width = 12, height = 8, dpi = 300, device = cairo_pdf)
```

**Figure 4** (`Figure_4.pdf`). Partial effects of the selected combined GAMs on the logit scale, coloured by panel (Punta Maroma, Florida, Belize and the regional composite; Saint Croix is not interpreted, Section 3.3.5). Each panel shows one smooth term with its 95 % confidence band. Terms shrunk to EDF ≈ 0 plot as flat lines at zero: s(age) in all four panels, s(GMHT_CPS0_30N) in Punta Maroma, and s(dT_raw) in Punta Maroma, Florida and the regional composite. Smooths are centred to a mean of zero over the fitted bins, so zero denotes the model's average contribution of that term and a curve below zero is associated with lower abundance than average.

### 4.2 Concurvity analysis

`concurvity(full = TRUE)` measures, for each term, how well it can be approximated by the other terms of the model (0 = not at all, 1 = completely), reported as worst-case, observed and estimated values. Values above about 0.8 are commonly treated as problematic.

```{r concurvity}

library(reshape2)
library(dplyr)
library(gt)

## concurvity() is undefined when a model has at least as many coefficients as observations
## (its matrix decomposition fails); such models are listed as "not computable".
concurvity_results <- lapply(best_models, function(m)
  tryCatch(concurvity(m, full = TRUE), error = function(e) NULL))
not_comp <- names(concurvity_results)[vapply(concurvity_results, is.null, logical(1))]
if (length(not_comp))
  message("Concurvity not computable (coefficients >= observations): ", paste(not_comp, collapse = ", "))

# Supplementary Table S9 = concurvity as a formatted table
conc_tbl <- purrr::imap_dfr(concurvity_results, function(cm, site) {
  if (is.null(cm))
    return(data.frame(Term = "not computable (coefficients >= observations)", worst = NA_real_,
                      observed = NA_real_, estimate = NA_real_, Panel = site))
  as.data.frame(t(cm)) %>%
    tibble::rownames_to_column("Term") %>%
    mutate(Term = ifelse(Term == "para", "parametric terms", Term),
           Panel = site)
}) %>%
  select(Panel, Term, worst, observed, estimate)

tab_S9 <- conc_tbl %>%
  gt(groupname_col = "Panel") %>%
  fmt_number(columns = c(worst, observed, estimate), decimals = 2) %>%
  cols_label(Term = "Term", worst = "Worst", observed = "Observed", estimate = "Estimate") %>%
  tab_header(title = "Concurvity of the selected combined GAMs",
             subtitle = "0 = term not approximated by the other terms; 1 = fully approximated")

tab_S9
gt::gtsave(tab_S9, "Supplementary_Table_S9.docx")
```

**Supplementary Table S9. Concurvity of the selected combined GAMs** (output above). No selected model contains a parametric (factor) term. Observed concurvity is high for s(GMHT_CPS0_30N) in every panel (0.85–0.90), for the hurricane smooth (0.80–0.84) and for s(age) (0.70–0.90, except 0.57 in Saint Croix), and moderate for s(dT_raw) (0.56–0.72).

The temporal trend, background temperature and hurricane activity are therefore partly interchangeable in these data: the hurricane series itself has a temporal structure (lowest in the mid-Holocene, highest in the last centuries). With `select = TRUE`, s(age) is shrunk to EDF ≈ 0 in Punta Maroma, Florida, Belize and the regional composite, so the hurricane smooth carries the temporal pattern in those models. Which term `select = TRUE` retains is not uniquely determined, and the effects in Table 2 and Fig. 4 should be read as conditional on the other terms rather than as independent contributions. The least confounded term is s(dT_raw). The dependence of the hurricane effect on the youngest bin, where both abundance and hurricane activity peak, is examined in Section 4.5.2.

### 4.3 GAM diagnostic plots

For each panel we show the four `gam.check()` residual plots and all model terms (`plot(..., all.terms = TRUE)`, one page: smooths with 95 % bands and rug plots, plus factor effects).

**Supplementary Figure S7.** Residual diagnostics and fitted terms of the selected combined GAMs: (a) Punta Maroma, (b) Florida, (c) Belize, (d) regional composite and (e) Saint Croix.

In all panels the response takes few distinct values, because each bin holds only a few records and abundance is normalized to the panel maximum: 2 distinct values in Saint Croix, 4 in Belize, 5 in Punta Maroma, 14 in Florida and 46 in the regional composite. This discreteness produces the diagonal bands in the plots of residuals against the linear predictor and should not be read as a missing covariate; for the same reason, departures of the deviance residuals from normality (Shapiro–Wilk *p* = 0.001–0.04 in every panel) are expected and do not by themselves indicate misspecification. The basis dimension is adequate for every smooth (EDF well below *k′*); the only k-index below 1 with *p* \< 0.05 is s(age) in Florida (0.76, *p* = 0.048), whose EDF is ≈ 0 of 9, so increasing *k* would not help; it reflects residual pattern along age, consistent with the positive lag-1 autocorrelation in this panel (Section 4.4). Panel by panel:

- **Punta Maroma:** the fitted values span 0.11–0.96 against observed values of 0.08–0.98; the maximum bin (0–0.1 ka; response 0.98, fitted 0.96) determines the upper end of the response–fitted relationship.
- **Florida:** fitted values span only 0.24–0.48 against observed values of 0.08–0.99, so the model does not reproduce the extremes, consistent with its low deviance explained (16.5 %) and the absence of significant climate terms.
- **Belize:** strong banding, because most bins hold a single record (response 0.25); fitted values span 0.21–0.86 against 0.25–0.98.
- **Regional composite:** fitted values span 0.18–0.93 against 0.03–0.99; the residuals show no pattern against the linear predictor.
- **Saint Croix:** with 19 bins and two distinct response values (one or two records per bin), the residuals fall on two bands. The response effectively records whether a bin holds a second colony; this panel is not interpreted (Section 3.3.5).

##### 4.3.1 Punta Maroma

```{r pm-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}
par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Punta Maroma"]])))
par(mfrow = c(2,3))
plot(best_models[["Punta Maroma"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Punta Maroma"]])   
```

##### **4.3.2 Florida**

```{r fl-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Florida"]])))
par(mfrow = c(2,3))
plot(best_models[["Florida"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Florida"]])   
```

##### **4.3.3 Belize**

```{r bz-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Belize"]])))
par(mfrow = c(2,3))
plot(best_models[["Belize"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Belize"]])   
```

##### 4.3.4 Regional

```{r reg-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Regional"]])))
par(mfrow = c(2,3))
plot(best_models[["Regional"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Regional"]])   
```

##### 4.3.5 Saint Croix

```{r sc-diagn, warning=FALSE, message=FALSE, fig.width=12, fig.height=10}

par(mfrow = c(2,2))
invisible(capture.output(gam.check(best_models[["Saint Croix"]])))
par(mfrow = c(2,3))
plot(best_models[["Saint Croix"]], shade=TRUE, pages=1, all.terms=TRUE, rug=TRUE)
k.check(best_models[["Saint Croix"]])   
```

```{r figS7-save, warning=FALSE, message=FALSE}
# save Supplementary Figure S7 (a-e) as one multi-page PDF, one page per panel
pdf("Supplementary_Figure_S7.pdf", width = 12, height = 10)
for (site in c("Punta Maroma", "Florida", "Belize", "Regional", "Saint Croix")) {
  par(mfrow = c(2, 2), oma = c(0, 0, 2, 0))
  invisible(capture.output(gam.check(best_models[[site]])))
  mtext(paste(site, "- residual diagnostics"), outer = TRUE, font = 2)
  par(oma = c(0, 0, 2, 0))
  plot(best_models[[site]], shade = TRUE, pages = 1, all.terms = TRUE, rug = TRUE)
  mtext(paste(site, "- fitted terms"), outer = TRUE, font = 2)
}
invisible(dev.off())
```

### 4.4 Residual temporal independence

For each selected model we extract Pearson residuals in bin order (young to old) and compute the lag-1 autocorrelation (φ), the Durbin–Watson statistic (`lmtest::dwtest`, a one-sided test against positive autocorrelation) and the Ljung–Box statistic at lag 10. ACF and PACF of the residuals are shown for all panels in one figure (Supplementary Fig. S8).

```{r residual-independency}

data_list <- list(
  "Belize"       = datos_bz,
  "Florida"      = datos_fl,
  "Punta Maroma" = datos_pm,
  "Regional"     = datos_reg,
  "Saint Croix"  = datos_sc
)

library(mgcv)
library(lmtest)

check_temporal_independence <- function(mod) {
  # Extraer residuales tipo pearson
  res_raw <- residuals(mod, type = "pearson")
  if (is.list(res_raw)) res_raw <- unlist(res_raw)
  res_raw <- as.numeric(res_raw)
  
  # AR(1) coeficiente
  phi <- cor(res_raw[-1], res_raw[-length(res_raw)])
  
  # Durbin–Watson
  dw <- lmtest::dwtest(res_raw ~ 1)
  
  # Ljung–Box
  lb <- Box.test(res_raw, lag = 10, type = "Ljung-Box")
  
  list(
    phi_AR1 = phi,
    DW_stat = dw$statistic,
    DW_pval = dw$p.value,
    LB_stat = lb$statistic,
    LB_pval = lb$p.value,
    res_raw = res_raw
  )
}

best_cb_lookup <- list(
  "Punta Maroma" = best_pm_cb,
  "Florida"      = best_fl_cb,
  "Belize"       = best_bz_cb,
  "Saint Croix"  = best_sc_cb,
  "Regional"     = best_reg_cb
)

best_models_table <- tibble(
  Segment = names(best_models),
  Model   = vapply(names(best_models), function(s) best_cb_lookup[[s]]$best_model, character(1))
)

results_resid <- lapply(seq_len(nrow(best_models_table)), function(i) {
  seg <- best_models_table$Segment[i]
  mname <- best_models_table$Model[i]
  cat("Using model:", mname, "for segment:", seg, "\n")

  mod <- best_models[[seg]]

  diag <- check_temporal_independence(mod)
  
  df <- data.frame(Segment = seg,
                   Model = mname,
                   phi_AR1 = diag$phi_AR1,
                   DW_stat = diag$DW_stat,
                   DW_pval = diag$DW_pval,
                   LB_stat = diag$LB_stat,
                   LB_pval = diag$LB_pval)
  
  attr(df, "res_raw") <- diag$res_raw
  df
})

res_summary <- do.call(rbind, results_resid)

tab_S10 <- res_summary %>%
  gt() %>%
  tab_header(
    title = "Residual independence of the selected combined GAMs"
  ) %>%
  fmt_number(columns = c(phi_AR1, DW_stat, DW_pval, LB_stat, LB_pval),
             decimals = 3)

tab_S10
gt::gtsave(tab_S10, "Supplementary_Table_S10.docx")

acf_df <- function(res, type = "ACF", site) {
  acf_obj <- acf(res, plot = FALSE)
  tibble(
    Lag = acf_obj$lag[, , 1],
    Value = acf_obj$acf[, , 1],
    Type = type,
    Site = site
  )
}

pacf_df <- function(res, type = "PACF", site) {
  pacf_obj <- pacf(res, plot = FALSE)
  tibble(
    Lag = pacf_obj$lag,
    Value = pacf_obj$acf,
    Type = type,
    Site = site
  )
}

df_all <- bind_rows(lapply(seq_len(nrow(best_models_table)), function(i) {
  seg <- best_models_table$Segment[i]
  res_raw <- attr(results_resid[[i]], "res_raw")
  bind_rows(
    acf_df(res_raw, "ACF", seg),
    pacf_df(res_raw, "PACF", seg)
  )
}))

acf_ci <- tibble(Site = best_models_table$Segment,
                 ci   = 1.96 / sqrt(vapply(results_resid, function(r) length(attr(r, "res_raw")), numeric(1))))

plot_acf_pacf <- ggplot(df_all, aes(x = Lag, y = Value, fill = Site)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = custom_colors) +
  geom_hline(yintercept = 0, color = "black") +

  geom_hline(data = acf_ci, aes(yintercept =  ci), linetype = "dashed", color = "grey40") +
  geom_hline(data = acf_ci, aes(yintercept = -ci), linetype = "dashed", color = "grey40") +
  facet_grid(Type ~ Site, scales = "free_y") +
  labs(y = "Correlation", x = "Lag") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

plot_acf_pacf
ggsave("Supplementary_Figure_S8.pdf", plot = plot_acf_pacf, width = 12, height = 6, dpi = 300, device = cairo_pdf)   
```

**Supplementary Table S10. Residual independence of the selected combined GAMs** (output above).

- **Ljung–Box:** no panel shows significant autocorrelation up to lag 10 (*p* = 0.09–0.66).
- **Durbin–Watson:** positive lag-1 autocorrelation is detected only in Florida (φ = 0.35, DW = 1.26, *p* = 0.005); Punta Maroma (φ = 0.05), Belize and the regional composite show none.
- **Negative autocorrelation:** Saint Croix (φ = −0.51) and Belize (φ = −0.13) show negative lag-1 correlation, which the one-sided Durbin–Watson test does not detect. In Saint Croix it reflects the alternation of one- and two-record bins of a near-binary response.

Two caveats apply. Bins without records are absent, so lag 1 does not always correspond to 100 years; this matters most for Saint Croix (19 of 60 bins) and Belize (24 of 60). With *n* = 19–57, both tests have limited power, especially Ljung–Box at lag 10 with *n* = 19.

**Supplementary Figure S8.** ACF and PACF of the residuals of the selected combined GAMs, by panel. Dashed lines mark the approximate 95 % limits for white noise (±1.96/√*n*), from ±0.26 in the regional composite (*n* = 57) to ±0.45 in Saint Croix (*n* = 19). Lags are counted in bins, which are not always consecutive centuries (see text).

### 4.5 Sensitivity analyses

We test the main results of the selected combined models against four alternative analyses:

1.  **Residual autocorrelation (Florida, Punta Maroma).** Florida, the panel with positive lag-1 autocorrelation (Section 4.4), and Punta Maroma, the panel with the strongest climate effect, are refitted as Gaussian GAMMs on the logit of the response, with the same terms and basis dimensions, first with independent errors and then with a continuous-time AR(1) structure (`corCAR1` on age, because bins are not always consecutive). mgcv does not provide a beta family with correlated errors, so this is an approximation used as a check, not a replacement for the beta models.
2.  **Florida source standardization.** The combined and marginal candidate sets are refitted to the Florida series pooled across sources (`datos_fl_norm`, Section 3.1.1).
3.  **Saint Croix response.** Because the response takes only two values, the selected model is refitted as a binomial GAM of one versus two records per bin.
4.  **Thermal regime in Punta Maroma.** The thermal regime is significant in the marginal model of this panel (Section 3.2.4) but is not part of the selected combined model, so it is added directly to the selected hurricane model.

```{r sensitivity-analyses, warning=FALSE, message=FALSE}

library(nlme)

fmt_p2 <- function(p) ifelse(p < 1e-4, "<1e-04", sprintf("%.2g", p))
sm_p   <- function(m, term) { st <- summary(m)$s.table
  if (term %in% rownames(st)) st[term, "p-value"] else NA_real_ }
## terms taken from the selected models, so the checks follow the model selection
hur_term  <- function(m) grep("^s\\(Hur100", rownames(summary(m)$s.table), value = TRUE)[1]
clim_smooths <- function(m) setdiff(rownames(summary(m)$s.table), "s(age)")
sig_terms <- function(m) {          # climate terms with term-level p < 0.05 ("none" otherwise)
  p <- term_pvalues(m); p <- p[!grepl("s\\(age\\)", names(p)) & p < 0.05]
  if (!length(p)) "none" else paste0(names(p), " (", fmt_p2(p), ")", collapse = "; ")
}
fl_hur <- hur_term(best_models[["Florida"]]); pm_hur <- hur_term(best_models[["Punta Maroma"]])
sc_terms <- clim_smooths(best_models[["Saint Croix"]])

## (1) Residual autocorrelation: Gaussian GAMM on the logit of the response,
##     without and with a continuous-time AR(1) (corCAR1), because bins are not always consecutive
ar1_check <- function(site, dat, term) {
  d  <- dat %>% arrange(age) %>%
    mutate(ly = qlogis(pmin(pmax(counts_norm_sv, 1e-4), 1 - 1e-4)))
  f  <- update(formula(best_models[[site]]), ly ~ .)
  m0 <- gamm(f, data = d, method = "REML")
  m1 <- gamm(f, data = d, method = "REML", correlation = corCAR1(form = ~ age))
  lrt <- anova(m1$lme, m0$lme)
  phi <- coef(m1$lme$modelStruct$corStruct, unconstrained = FALSE)
  tibble::tibble(
    Check = "Residual AR(1) (Gaussian GAMM on logit, corCAR1)", Panel = site, Term = term,
    Reference = fmt_p2(sm_p(best_models[[site]], term)),
    Sensitivity = paste0(fmt_p2(sm_p(m0$gam, term)), " (independent); ",
                         fmt_p2(sm_p(m1$gam, term)), " (AR(1))"),
    Note = sprintf("phi = %.2f per 0.1 ka; LRT p = %s; dAIC(AR1 - indep) = %.1f",
                   phi^0.1, fmt_p2(lrt[["p-value"]][2]), AIC(m1$lme) - AIC(m0$lme)))
}

res_fl_pool  <- fit_combined_gams_safe(formulas_cb, datos_fl_norm, "Florida (pooled)")
best_fl_pool <- get_best_combined(res_fl_pool)
marg_pool    <- compare_gams(fit_gams_panel(datos_fl_norm, formulas_mg)) %>% arrange(AIC)
marg_pool_m  <- gam(formulas_mg[[marg_pool$Model[1]]], data = datos_fl_norm,
                    family = betar(link = "logit"), method = "REML")
marg_std_m   <- gam(formulas_mg[[marg_pool$Model[1]]], data = datos_fl_std,
                    family = betar(link = "logit"), method = "REML")
marg_term    <- names(term_pvalues(marg_pool_m))[!grepl("s\\(age\\)", names(term_pvalues(marg_pool_m)))][1]

sc_bin  <- datos_sc %>% mutate(two = as.integer(counts >= 2))
m_sc_bn <- tryCatch(gam(update(formula(best_models[["Saint Croix"]]), two ~ .), data = sc_bin,
               family = binomial, method = "REML", select = TRUE), error = function(e) NULL)

m_pm_th <- gam(update(formula(best_models[["Punta Maroma"]]), . ~ . + Thermal), data = datos_pm,
               family = betar(link = "logit"), method = "REML", select = TRUE)
p_pm_marg_th <- anova(gam(formulas_mg$M_the, data = datos_pm, family = betar(link = "logit"),
                          method = "REML"))$pTerms.table["Thermal", "p-value"]

sens_tbl <- bind_rows(
  ar1_check("Florida", datos_fl_std, fl_hur),
  ar1_check("Punta Maroma", datos_pm, pm_hur),
  tibble::tibble(
    Check = "Florida series pooled across sources (combined models)", Panel = "Florida",
    Term = "climate terms with p < 0.05", Reference = sig_terms(best_models[["Florida"]]),
    Sensitivity = sig_terms(best_fl_pool$fit),
    Note = sprintf("selected model %s (weight %.2f) vs %s; deviance explained %.1f %% vs %.1f %%",
                   best_fl_pool$best_model, best_fl_pool$ranking$weight[1], best_fl_cb$best_model,
                   100 * summary(best_fl_pool$fit)$dev.expl, 100 * summary(best_models[["Florida"]])$dev.expl)),
  tibble::tibble(
    Check = "Florida series pooled across sources (marginal models)", Panel = "Florida",
    Term = marg_term, Reference = fmt_p2(term_pvalues(marg_std_m)[marg_term]),
    Sensitivity = fmt_p2(term_pvalues(marg_pool_m)[marg_term]),
    Note = sprintf("best marginal model on the pooled series: %s (weight %.2f); reference = same model on the source-standardized series",
                   marg_pool$Model[1], marg_pool$weight[1])),
  tibble::tibble(
    Check = "Binomial GAM, one vs two records per bin", Panel = "Saint Croix",
    Term = sc_terms,
    Reference = fmt_p2(sapply(sc_terms, function(tm) sm_p(best_models[["Saint Croix"]], tm))),
    Sensitivity = if (is.null(m_sc_bn)) "not estimable" else fmt_p2(sapply(sc_terms, function(tm) sm_p(m_sc_bn, tm))),
    Note = "same terms as the selected model"),
  tibble::tibble(
    Check = "Thermal regime added to the selected hurricane model", Panel = "Punta Maroma",
    Term = "Thermal", Reference = fmt_p2(p_pm_marg_th),
    Sensitivity = fmt_p2(anova(m_pm_th)$pTerms.table["Thermal", "p-value"]),
    Note = sprintf("reference = marginal thermal-regime model; %s in the extended model: p = %s", pm_hur,
                   fmt_p2(sm_p(m_pm_th, pm_hur))))
)

tab_S11 <- sens_tbl %>%
  gt(groupname_col = "Check") %>%
  cols_label(Reference = "p (reference)", Sensitivity = "p (sensitivity)") %>%
  tab_header(title = "Sensitivity analyses of the selected combined GAMs")

tab_S11
gt::gtsave(tab_S11, "Supplementary_Table_S11.docx")
```

**Supplementary Table S11. Sensitivity analyses of the selected combined GAMs.** For each check, the term tested, its *p*-value in the reference analysis and in the sensitivity analysis, and the relevant model statistics.

- **Residual autocorrelation.** The AR(1) parameter is small (φ = 0.22 per century in Florida and 0.07 in Punta Maroma) and does not improve the fit (likelihood-ratio *p* = 0.32 and 0.72). In Punta Maroma s(Hur100_av) remains significant with independent and with AR(1) errors (*p* \< 10⁻⁴ in both). In Florida s(Hur100_fqcy), already not significant in the beta model (*p* = 0.081), is not significant in either Gaussian model (*p* = 0.24 and 0.46).
- **Florida source standardization.** The Florida result depends on how the sources are scaled. With the pooled series the selected model is M5_av (weight 0.34; deviance explained 34.0 % vs 16.5 %), in which the only significant climate term is the Florida hydroclimate (*p* = 0.0027) and s(Hur100_av) is not significant (*p* = 0.11); the best marginal model is also the hydroclimate model (weight 0.80; *p* = 0.0025, against 0.66 on the source-standardized series). Pooling lets the most abundant source dominate the series, so we retain the source-standardized series and report no climate effect for Florida.
- **Saint Croix.** In the binomial model of one versus two records per bin, neither s(GMHT_CPS0_30N) (*p* = 0.27) nor s(dT_raw) (*p* = 0.23) is significant, consistent with the absence of climate effects in this panel.
- **Punta Maroma.** Added to the selected hurricane model, the thermal regime is not significant (*p* = 0.99), whereas s(Hur100_av) remains highly significant (*p* \< 10⁻⁴). The LIA effect of the marginal thermal model (*p* \< 10⁻⁴) is therefore not independent of hurricane activity in this panel.

#### 4.5.1 Local hurricane series in Florida

In the Florida panel the hurricane regime (`Hurricane`) is the local Florida series (`Hur_FL`, Section 1.2) in all the analyses above. Here we compare, as single climate terms added to `s(age)`, the local regime, its continuous version (`FL_pact`, proportion of records active) and the two Schmitt et al. (2025) metrics. The comparison is repeated for bins younger than 3.5 ka, where the local series combines two to six records, because older bins rest on a single record.

```{r florida-local-hurricane, warning=FALSE, message=FALSE}
fl_hur <- datos_fl_std %>%
  left_join(datos_cli %>% transmute(age = interval_max, FL_pact, FL_nrec), by = "age")

fl_hur_models <- list(
  "s(age)"                 = counts_norm_sv ~ s(age),
  "s(age) + Hurricane (local regime)"  = counts_norm_sv ~ s(age) + Hurricane,
  "s(age) + FL_pact (local, linear)"   = counts_norm_sv ~ s(age) + FL_pact,
  "s(age) + s(Hur100_av) (Schmitt)"    = counts_norm_sv ~ s(age) + s(Hur100_av, k = 8),
  "s(age) + s(Hur100_fqcy) (Schmitt)"  = counts_norm_sv ~ s(age) + s(Hur100_fqcy, k = 8))

fl_hur_tab <- purrr::map_dfr(c("0-6 ka" = 6, "0-3.5 ka" = 3.5), function(amax) {
  d <- fl_hur %>% filter(age <= amax + 1e-9) %>% droplevels()
  purrr::imap_dfr(fl_hur_models, function(f, nm) {
    m  <- gam(f, data = d, family = betar(link = "logit"), method = "REML")
    pt <- term_pvalues(m); pt <- pt[!grepl("s\\(age\\)", names(pt))]
    tibble::tibble(Model = nm, n = nrow(d), AIC = AIC(m),
                   `p (climate term)` = if (length(pt)) fmt_p2(pt[1]) else "",
                   `Dev. expl. (%)` = round(100 * summary(m)$dev.expl, 1))
  }) %>% mutate(dAIC = AIC - min(AIC))
}, .id = "Window") %>% mutate(AIC = round(AIC, 1), dAIC = round(dAIC, 1))

knitr::kable(fl_hur_tab, caption = "Florida panel: local vs Schmitt et al. (2025) hurricane predictors")
```

#### 4.5.2 Age uncertainty, sampling, binning, the modern peak and Saint Croix

These analyses answer how much the combined-model results depend on (i) the uncertainty of the individual ages, (ii) which fossils were sampled, (iii) the width and phase of the bins, (iv) the youngest bin, which holds the modern abundance peak, (v) the information criterion and (vi) the inclusion of Saint Croix in the regional composite. All of them rebuild the five series with `build_series()`, which applies the rules of Section 1.3 to any set of records; applied to the audited ages it reproduces `merged_df_g` exactly (checked in the code).

```{r sensitivity-m3, warning=FALSE, message=FALSE}
library(dplyr)
library(mgcv)

## ---- series builder: the rules of Section 1.3 applied to any set of records ----------
## rec: one row per dated record with spot, source and t_ka. Bins of width `binw` starting
## at `offset` (0.1 ka and 0 reproduce the main analysis; other widths keep only the
## numeric climate predictors, averaged over the 0.1-ka bins they contain).
cli_tab <- datos_cli %>%
  transmute(age = interval_max, GMHT_CPS0_30N, dT_raw, Hur100_fqcy, Hur100_av,
            Thermal, Hydro_YUC, Hydro_FL, Hur_YUC = Hurricane, Hur_FL)
site_names <- c("Punta Maroma", "Florida", "Belize", "Saint Croix")

build_series <- function(rec, binw = 0.1, offset = 0, sites = site_names) {
  ## same bin edges as the main analysis (breaks), so ages join exactly to the climate table
  ub  <- if (binw == 0.1 && offset == 0) breaks[-1] else seq(binw + offset, 6 + 1e-9, by = binw)
  rec <- rec %>% filter(t_ka > -0.1, t_ka <= 6) %>%
    mutate(age = ub[pmin(length(ub), pmax(1, ceiling(round((pmax(t_ka, 1e-6) - offset) / binw, 6))))])
  cli <- if (binw == 0.1 && offset == 0) cli_tab else
    bind_rows(lapply(ub, function(u) cli_tab %>% filter(age > u - binw + 1e-6, age <= u + 1e-6) %>%
      summarise(age = u, across(c(GMHT_CPS0_30N, dT_raw, Hur100_fqcy, Hur100_av), mean))))
  one_site <- function(sp) {
    r <- rec %>% filter(spot == sp)
    cnt <- if (sp == "Florida") {
      r %>% count(source, age) %>% group_by(source) %>%
        mutate(std = n / max(n), std = ifelse(std %in% c(0, 1), (std * (n() - 1) + 0.5) / n(), std)) %>%
        ungroup() %>% group_by(age) %>% summarise(counts = sum(std), .groups = "drop")
    } else r %>% count(age, name = "counts")
    out <- cnt %>% left_join(cli, by = "age")
    if (binw == 0.1 && offset == 0)
      out <- out %>% mutate(Hydro = if (sp == "Florida") Hydro_FL else Hydro_YUC,
                            Hurricane = if (sp == "Florida") Hur_FL else Hur_YUC)
    normalize_counts(out, sp)
  }
  s   <- lapply(sites, one_site)
  reg <- bind_rows(s) %>% group_by(age) %>% summarise(counts = sum(counts_norm_sv), .groups = "drop") %>%
    left_join(cli, by = "age")
  if (binw == 0.1 && offset == 0) reg <- reg %>% mutate(Hydro = Hydro_YUC, Hurricane = Hur_YUC)
  out <- bind_rows(s, normalize_counts(reg, "Regional"))
  if (binw == 0.1 && offset == 0)
    out <- out %>% mutate(Thermal = relevel(factor(Thermal), ref = "warmer"),
                          Hydro = relevel(factor(Hydro), ref = "wet"),
                          Hurricane = factor(Hurricane, levels = levels(merged_df_g$Hurricane)))
  out
}

## check: applied to the audited ages, the builder reproduces the modelled series exactly
chk <- build_series(recs %>% select(spot, source, t_ka)) %>% transmute(panel, age = round(age, 3), y = counts_norm_sv) %>%
  full_join(merged_df_g %>% transmute(panel, age = round(age, 3), y0 = counts_norm_sv), by = c("panel", "age"))
stopifnot(!anyNA(chk$y), !anyNA(chk$y0), max(abs(chk$y - chk$y0)) < 1e-9)

## ---- lean refit of the candidates (same k rule and options as fit_combined_gams_safe) ---
fit_set <- function(d, fset) {
  d <- d %>% filter(!is.na(counts_norm_sv), !is.na(GMHT_CPS0_30N), !is.na(Hur100_av))
  n <- nrow(d)
  fits <- lapply(fset, function(f) {
    ns <- count_smooth_terms(f)
    if (n < 8 || n <= ns * 3) return(NULL)
    tryCatch(gam(cap_smooth_k(f, safe_k_for(n, ns)), data = d, family = betar(link = "logit"),
                 method = "REML", select = TRUE), error = function(e) NULL)
  })
  fits[!vapply(fits, is.null, logical(1))]
}
smooth_p <- function(m, term) {
  st <- summary(m)$s.table
  if (term %in% rownames(st)) st[term, "p-value"] else NA_real_
}
summ_fits <- function(fits) {
  aic <- vapply(fits, AIC, numeric(1)); best <- names(which.min(aic))
  data.frame(best = best, dAIC_M1 = unname(aic["M1"] - min(aic)),
             p_hav  = if (!is.null(fits$M4_av))   smooth_p(fits$M4_av, "s(Hur100_av)") else NA,
             p_temp = if (!is.null(fits$M4_av))   smooth_p(fits$M4_av, "s(GMHT_CPS0_30N)") else NA,
             p_dt   = if (!is.null(fits$M4_av))   smooth_p(fits$M4_av, "s(dT_raw)") else NA,
             p_hf   = if (!is.null(fits$M4_fqcy)) smooth_p(fits$M4_fqcy, "s(Hur100_fqcy)") else NA)
}
smooth_set <- formulas_cb[c("M1", "M2", "M3", "M4_fqcy", "M4_av")]
panel_names <- c(site_names, "Regional")

## ---- (a) the modern peak: selected models refitted without the youngest bin -----------
modern_tbl <- bind_rows(lapply(panel_names, function(p) {
  d  <- merged_df_g %>% filter(panel == p)
  fs <- list(M1 = formulas_cb$M1, selected = formula(best_models[[p]]))
  r_all <- fit_set(d, fs); r_young <- fit_set(d %>% filter(age > 0.15), fs)
  clim  <- setdiff(rownames(summary(r_all$selected)$s.table), "s(age)")
  data.frame(Panel = p, Selected = NA_character_,
             `dAIC vs s(age), all bins` = AIC(r_all$M1) - AIC(r_all$selected),
             `dAIC vs s(age), without 0-0.1 ka` = AIC(r_young$M1) - AIC(r_young$selected),
             `Climate terms p < 0.05, all bins` = paste(clim[sapply(clim, function(t) smooth_p(r_all$selected, t)) < 0.05], collapse = ", "),
             `Climate terms p < 0.05, without 0-0.1 ka` = paste(clim[sapply(clim, function(t) smooth_p(r_young$selected, t)) < 0.05], collapse = ", "),
             check.names = FALSE)
}))
modern_tbl$Selected <- c(best_pm_cb$best_model, best_fl_cb$best_model, best_bz_cb$best_model,
                         best_sc_cb$best_model, best_reg_cb$best_model)
modern_tbl %>% mutate(across(where(is.numeric), ~ round(.x, 1)))

## ---- (b) the regional composite without Saint Croix -------------------------------------
reg_noSC <- build_series(recs %>% select(spot, source, t_ka), sites = setdiff(site_names, "Saint Croix")) %>%
  filter(panel == "Regional")
res_noSC  <- fit_combined_gams_safe(formulas_cb, reg_noSC, "Regional without Saint Croix")
best_noSC <- get_best_combined(res_noSC)
marg_noSC <- compare_gams(fit_gams_panel(reg_noSC, formulas_mg)) %>% arrange(AIC)
noSC_tbl <- tibble::tibble(
  Composite = c("Four sites", "Without Saint Croix"),
  Bins = c(nrow(datos_reg), nrow(reg_noSC)),
  `Best marginal` = c(get_best_models(datos_reg, formulas_mg)$ranking$Model[1], marg_noSC$Model[1]),
  `Best combined` = c(best_reg_cb$best_model, best_noSC$best_model),
  `Deviance explained (%)` = round(100 * c(summary(best_reg_cb$fit)$dev.expl, summary(best_noSC$fit)$dev.expl), 1),
  `p s(Hur100_av)` = c(smooth_p(best_reg_cb$fit, "s(Hur100_av)"), smooth_p(best_noSC$fit, "s(Hur100_av)")))
noSC_tbl

## ---- (c) small-sample correction: AICc ---------------------------------------------------
aicc_tbl <- bind_rows(lapply(list(`Punta Maroma` = results_pm, Florida = results_fl, Belize = results_bz,
                                  `Saint Croix` = results_sc, Regional = results_reg), function(r) {
  n <- nrow(r$fits[[1]]$model)
  k <- vapply(r$fits, function(m) attr(logLik(m), "df"), numeric(1))
  a <- vapply(r$fits, AIC, numeric(1)); ac <- a + 2 * k * (k + 1) / (n - k - 1)
  data.frame(n = n, `Best by AIC` = names(which.min(a)), `Best by AICc` = names(which.min(ac)),
             `dAICc of the AIC-selected model` = round(ac[which.min(a)] - min(ac), 2),
             check.names = FALSE, row.names = NULL)
}), .id = "Panel")
aicc_tbl

## ---- (d) bin width and bin phase (audited ages; numeric-predictor candidates) -------------
bin_tbl <- bind_rows(lapply(list(c(0.1, 0), c(0.2, 0), c(0.2, 0.1)), function(bp) {
  ser <- build_series(recs %>% select(spot, source, t_ka), binw = bp[1], offset = bp[2])
  bind_rows(lapply(panel_names, function(p) cbind(Panel = p, `Bin (ka)` = bp[1], Offset = bp[2],
                                                  bins = sum(ser$panel == p), summ_fits(fit_set(ser %>% filter(panel == p), smooth_set)))))
}))
bin_tbl %>% mutate(across(c(dAIC_M1, starts_with("p_")), ~ signif(.x, 2)))

## ---- (e) age and sampling uncertainty propagated to the GAMs (cached) ----------------------
## Age: each replicate takes one Monte Carlo draw of every age (age_draws, Section 1.4).
## Sampling: records resampled with replacement within sites (and Florida sources), ages fixed.
## Both: the two combined. For each replicate the five series are rebuilt (build_series) and
## the candidates without categorical regimes (M1-M4) are refitted; no M6/M7 model is selected
## with the audited ages, and factor models on near-binary responses can abort mgcv.
## The result is stored in M3_sensitivity/gam_uncertainty_mc.rds; set rerun_mc <- TRUE to recompute.
rerun_mc <- FALSE
B_mc     <- 300
mc_file  <- here("M3_sensitivity", "gam_uncertainty_mc.rds")
if (!file.exists(mc_file) || rerun_mc) {
  rec0  <- recs %>% select(spot, source, t_ka)
  strat <- paste(rec0$spot, ifelse(rec0$spot == "Florida", rec0$source, ""))
  idx   <- split(seq_len(nrow(rec0)), strat)
  one_rep <- function(b, scen) {
    r <- rec0
    if (scen %in% c("Age", "Both")) r$t_ka <- age_draws[, b]
    if (scen %in% c("Sampling", "Both"))
      r <- r[unlist(lapply(idx, function(i) i[sample.int(length(i), length(i), replace = TRUE)])), ]
    ser <- build_series(r)
    bind_rows(lapply(panel_names, function(p)
      cbind(Panel = p, Scenario = scen, rep = b, summ_fits(fit_set(ser %>% filter(panel == p), smooth_set)))))
  }
  set.seed(20261004)
  t0 <- Sys.time()
  mc_res <- bind_rows(lapply(c("Age", "Sampling", "Both"), function(sc)
    bind_rows(lapply(seq_len(B_mc), function(b) tryCatch(one_rep(b, sc), error = function(e) NULL)))))
  attr(mc_res, "info") <- list(B = B_mc, seed = 20261004, date = Sys.time(), minutes = as.numeric(difftime(Sys.time(), t0, units = "mins")))
  dir.create(dirname(mc_file), showWarnings = FALSE)
  saveRDS(mc_res, mc_file)
}
mc_res <- readRDS(mc_file)
attr(mc_res, "info")[c("B", "seed", "minutes")]

mc_tbl <- mc_res %>%
  group_by(Panel = factor(Panel, panel_names), Scenario = factor(Scenario, c("Age", "Sampling", "Both"))) %>%
  summarise(Replicates = n(),
            `Top model (share)` = { tb <- sort(table(best), decreasing = TRUE); sprintf("%s (%.0f %%)", names(tb)[1], 100 * tb[1] / n()) },
            `dAIC vs s(age) > 2 (%)` = round(100 * mean(dAIC_M1 > 2)),
            `dAIC vs s(age), median [95 %]` = sprintf("%.1f [%.1f, %.1f]", median(dAIC_M1), quantile(dAIC_M1, 0.025), quantile(dAIC_M1, 0.975)),
            `s(Hur100_av) p < 0.05 (%)` = round(100 * mean(p_hav < 0.05, na.rm = TRUE)),
            `s(GMHT) p < 0.05 (%)` = round(100 * mean(p_temp < 0.05, na.rm = TRUE)),
            `s(dT_raw) p < 0.05 (%)` = round(100 * mean(p_dt < 0.05, na.rm = TRUE)),
            `s(Hur100_fqcy) p < 0.05 (%)` = round(100 * mean(p_hf < 0.05, na.rm = TRUE)),
            .groups = "drop")
mc_tbl

tab_S12 <- mc_tbl %>%
  gt(groupname_col = "Panel") %>%
  tab_header(title = "Propagation of age and sampling uncertainty to the combined GAMs",
             subtitle = paste0(attr(mc_res, "info")$B, " replicates per scenario; candidates M1-M4"))
tab_S12
gtsave(tab_S12, "Supplementary_Table_S12.docx")
```

**Supplementary Table S12. Propagation of age and sampling uncertainty to the combined GAMs** (output above). For each panel and scenario: number of replicates, model selected most often among M1–M4 and its share, percentage of replicates in which the selected model improves on `s(age)` by more than 2 AIC units, median and 95 % range of that improvement, and percentage of replicates in which each climate smooth is significant (*p* \< 0.05) in its M4 model.

- **The modern peak (a).** Removing the youngest bin (0–0.1 ka) reduces the AIC advantage of the selected model over `s(age)` from 56.5 to 4.4 in Punta Maroma and from 30.0 to 3.4 in the regional composite, while s(Hur100_av) remains significant in both. In these two panels most of the support for the hurricane effect therefore comes from the coincidence, in the last century, of the abundance maximum with the highest hurricane activity of the series (Hur100_av = 45); the association persists without it, but is much weaker. Belize does not depend on the youngest bin (12.5 → 15.1; the three climate terms remain significant). Florida and Saint Croix show no climate effect with or without it.
- **Information criterion (c).** AICc, which penalizes complexity more strongly in small samples, selects the same model as AIC in Punta Maroma, Belize and the regional composite. In Florida it prefers `s(age)` alone (the AIC-selected model is 0.32 AICc units behind), and in Saint Croix a simpler model (M2; the AIC-selected M3 is 4.3 units behind), consistent with the absence of climate effects in these panels.
- **Bin width and phase (d).** With 0.2-ka bins, starting at 0 or at 0.1 ka, the hurricane smooth remains significant in Punta Maroma (*p* \< 10⁻⁴ in both phases; improvement over `s(age)` 35 and 28 AIC units) and in the regional composite (*p* ≤ 3 × 10⁻⁴; 21 and 31 units), and in Belize (*p* ≤ 9 × 10⁻⁴; 23 and 24 units), although the M4 branch that is selected (average or frequency) changes with the phase. The temperature smooths of Belize are less stable (s(GMHT_CPS0_30N) significant in one phase only). Florida and Saint Croix show no consistent climate effect at either bin width.
- **Regional composite without Saint Croix (b).** Without Saint Croix the composite has 55 bins instead of 57, and the results are unchanged: the best marginal model is again the hurricane-average model, and the selected combined model is again M4_av (weight 0.93; deviance explained 57.2 % vs 53.2 %; improvement over `s(age)` 31.5 vs 30.0 AIC units), with s(Hur100_av) significant (*p* = 1.9 × 10⁻⁶ vs 2.3 × 10⁻⁶). s(GMHT_CPS0_30N) becomes marginally significant (*p* = 0.046). The regional conclusions therefore do not depend on Saint Croix or on its large weight in the composite.
- **Age and sampling uncertainty (e).** Of 300 replicates per scenario, two age replicates failed to fit and were dropped (298 retained). In **Punta Maroma** and the **regional composite** the hurricane effect is robust to both sources of uncertainty: the selected model improves on `s(age)` by more than 2 AIC units in 96–100 % of the replicates and s(Hur100_av) is significant in 91–100 % of them, in every scenario. Age uncertainty alone barely moves these results (Punta Maroma: improvement 54.7 \[44.8, 64.2\] AIC units, against 29.3 \[7.4, 54.7\] under resampling of the records), because their chronology is dominated by U-Th ages or, in the composite, averaged over four sites; sampling is the larger source of uncertainty. In **Belize** the association holds under age uncertainty (improvement \> 2 in 100 % of the replicates; s(Hur100_av) significant in 89 %), but weakens when the records are resampled (67–68 %; s(Hur100_av) significant in 44–45 %, s(GMHT_CPS0_30N) in 31–41 %): with 39 records, the Belize result depends on which fossils were dated. In **Florida** and **Saint Croix** the climate models improve on `s(age)` in a minority of replicates (23–24 % and 34–57 %) and no climate term is significant in more than 32 % of them, confirming the absence of a climate effect in these panels. Across scenarios, the hurricane metric that is selected (average or frequency) varies in some replicates, but one of the two is significant in at least 91 % of the replicates in Punta Maroma and the regional composite.

### 4.6 Predictive validation and robustness of the coral–climate associations

Every model includes `s(age)`, which represents the temporal structure of each series: the long-term trend and low-frequency variation in abundance, including taphonomic and sampling effects that cannot be separated from ecological change. Climatic effects are therefore estimated conditional on `s(age)`. All analyses in this section compare the selected models with a reference model, `counts_norm_sv ~ s(age)`, and so ask whether climate adds information beyond the temporal structure of the coral record.

For each panel we evaluate three models: the reference model, the best marginal model (lowest AIC, Section 3.2) and the selected combined model (Section 3.3). Four analyses are run:

1.  **Cross-validation.** Each century is left out in turn, the model is refitted to the remaining centuries and the left-out century is predicted (leave-one-out, LOO). In a second version, the centuries within ±200 years of the left-out one are also removed from the fitting data (h-block), so that the prediction cannot borrow information from immediately adjacent centuries. Predictions are scored by their log-likelihood under the fitted beta distribution, and each model is compared with the reference model through the summed difference (ΔELPD; positive = better than the reference), its standard error and the percentage of centuries predicted better than by the reference.
2.  **Circular-shift test.** The complete 60-bin climatic series (all predictors together) is rotated in time by 6–54 bins (600–5,400 years; 49 shifts) relative to the coral record, while the response and `s(age)` stay fixed. Rotation keeps the shape and autocorrelation of both series and the correlations among predictors, and destroys only their alignment. The statistic is the AIC improvement over the reference model; *p* = (1 + number of shifts with an improvement ≥ the observed one) / (1 + number of shifts). The smallest attainable *p* is 0.02. Shifts of ≤ 500 years are excluded because neighbouring centuries are autocorrelated.
3.  **Drop-one influence.** Each century is removed in turn and the model refitted. For every climatic term we report its *p*-value with all data, the largest *p*-value across the drop-one fits, the century whose removal gives that largest value and the percentage of drop-one fits in which the term stays significant (*p* \< 0.05).
4.  **Former cross-validation (withdrawn).** The five-fold cross-validation of the first revision is re-run to document why it was withdrawn (see below).

During refitting the basis dimension of each smooth is capped with `safe_k_for()` from the number of fitting observations, and further reduced if the total number of coefficients would reach that number. A refit is skipped, and the left-out century counted as not predicted, when the fitting data do not contain every level of a categorical predictor or the categorical design is rank-deficient. For panels with at least 21 observations the full-data fits are identical to those of Sections 3.2–3.3. In Saint Croix (n = 19) the marginal smooths use k = 9 instead of the default 10, which changes the AIC of the marginal model by 0.01.

```{r validation-robustness, warning=FALSE, message=FALSE}

fac_vars <- c("Thermal", "Hydro", "Hurricane")

# refit a model to a subset with the k cap and the rank / coefficient guards
fit_val <- function(f, d, combined) {
  ns <- count_smooth_terms(f)
  for (v in intersect(fac_vars, all.vars(f))) d[[v]] <- droplevels(d[[v]])
  fv <- intersect(fac_vars, all.vars(f)); n_par <- 1
  if (length(fv)) {
    X <- tryCatch(model.matrix(reformulate(fv), d), error = function(e) NULL)
    if (is.null(X) || qr(X)$rank < ncol(X)) return(NULL)
    n_par <- ncol(X)
  }
  if (ns > 0) {
    k <- safe_k_for(nrow(d), ns)
    while (k > 3 && n_par + ns * (k - 1) >= nrow(d)) k <- k - 1
    if (n_par + ns * (k - 1) >= nrow(d)) return(NULL)
    f <- cap_smooth_k(f, k)
  }
  tryCatch(suppressWarnings(gam(f, data = d, family = betar(link = "logit"),
                                method = "REML", select = combined)),
           error = function(e) NULL)
}

lpd_val <- function(m, test) {
  if (is.null(m)) return(rep(NA_real_, nrow(test)))
  for (v in intersect(fac_vars, all.vars(formula(m))))
    if (any(!as.character(test[[v]]) %in% as.character(unique(m$model[[v]]))))
      return(rep(NA_real_, nrow(test)))
  mu  <- as.numeric(predict(m, newdata = test, type = "response"))
  phi <- m$family$getTheta(TRUE)
  dbeta(test$counts_norm_sv, mu * phi, (1 - mu) * phi, log = TRUE)
}

pointwise_cv <- function(f, d, combined, buffer = 0) {
  sapply(seq_len(nrow(d)), function(i) {
    drop <- if (buffer > 0) which(abs(d$age - d$age[i]) <= buffer + 1e-9) else i
    lpd_val(fit_val(f, d[-drop, ], combined), d[i, , drop = FALSE])
  })
}

val_data <- list("Punta Maroma" = datos_pm, "Florida" = datos_fl_std, "Belize" = datos_bz,
                 "Saint Croix" = datos_sc, "Regional" = datos_reg)
val_marg <- list("Punta Maroma" = best_pm, "Florida" = best_fl, "Belize" = best_bz,
                 "Saint Croix" = best_sc, "Regional" = best_reg)
val_comb <- list("Punta Maroma" = best_pm_cb, "Florida" = best_fl_cb, "Belize" = best_bz_cb,
                 "Saint Croix" = best_sc_cb, "Regional" = best_reg_cb)

val_models <- purrr::imap_dfr(val_data, function(d, site) tibble::tibble(
  Panel = site,
  Type  = c("Reference", "Marginal", "Combined"),
  Model = c("s(age) only", val_marg[[site]]$ranking$Model[1], val_comb[[site]]$best_model)))

val_formula <- function(type, model, data) {
  f <- switch(type,
    Reference = counts_norm_sv ~ s(age),
    Marginal  = formulas_mg[[model]],
    Combined  = formulas_cb[[model]])
  # as in the fitted models (drop_hydro), Hydro is removed where its series has gaps
  if (hydro_has_gaps(data) && "Hydro" %in% all.vars(f)) f <- update(f, . ~ . - Hydro)
  f
}

cv_points <- purrr::pmap_dfr(val_models, function(Panel, Type, Model) {
  d <- as.data.frame(val_data[[Panel]]) %>% arrange(age)
  f <- val_formula(Type, Model, d); cb <- Type == "Combined"
  tibble::tibble(Panel = Panel, Type = Type, Model = Model, age = d$age,
                 LOO = pointwise_cv(f, d, cb, 0), hblock = pointwise_cv(f, d, cb, 0.2))
})

cv_summary <- cv_points %>%
  group_by(Panel, age) %>%
  mutate(dLOO = LOO - LOO[Type == "Reference"],
         dHB  = hblock - hblock[Type == "Reference"]) %>%
  ungroup() %>% filter(Type != "Reference") %>%
  group_by(Panel, Type, Model) %>%
  summarise(
    n            = n(),
    n_pred       = sum(!is.na(dLOO)),
    dELPD_LOO    = sum(dLOO, na.rm = TRUE),
    SE_LOO       = sqrt(sum(!is.na(dLOO))) * sd(dLOO, na.rm = TRUE),
    dELPD_hblock = sum(dHB, na.rm = TRUE),
    SE_hblock    = sqrt(sum(!is.na(dHB))) * sd(dHB, na.rm = TRUE),
    pct_better   = 100 * mean(dLOO > 0, na.rm = TRUE),
    worst_ka     = age[which.min(dLOO)],
    .groups = "drop")

clim_vars <- c("GMHT_CPS0_30N", "dT_raw", "Hur100_fqcy", "Hur100_av", fac_vars)
cli_full  <- datos_cli %>% arrange(interval_max) %>%
  mutate(age = round(interval_max, 1), Hydro = Hydro_YUC) %>% select(age, all_of(clim_vars), Hydro_FL, Hur_FL)
cli_full_fl <- cli_full %>% mutate(Hydro = Hydro_FL, Hurricane = Hur_FL)
for (v in fac_vars) {
  cli_full[[v]]    <- factor(cli_full[[v]],    levels = levels(merged_df_g[[v]]))
  cli_full_fl[[v]] <- factor(cli_full_fl[[v]], levels = levels(merged_df_g[[v]]))
}
n_bins <- nrow(cli_full)
shifts <- 6:(n_bins - 6)

shift_null <- purrr::pmap_dfr(filter(val_models, Type != "Reference"), function(Panel, Type, Model) {
  d   <- as.data.frame(val_data[[Panel]]) %>% mutate(age = round(age, 1))
  f   <- val_formula(Type, Model, d); cb <- Type == "Combined"
  ref <- AIC(fit_val(counts_norm_sv ~ s(age), d, FALSE))
  obs <- ref - AIC(fit_val(f, d, cb))
  cli_p <- if (Panel == "Florida") cli_full_fl else cli_full
  null <- sapply(shifts, function(s) {
    rot <- cli_p[((seq_len(n_bins) - 1 + s) %% n_bins) + 1, clim_vars]
    rot$age <- cli_p$age
    m <- fit_val(f, merge(d[, c("age", "counts_norm_sv")], rot, by = "age"), cb)
    if (is.null(m) || m$rank < length(coef(m))) NA_real_ else ref - AIC(m)
  })
  tibble::tibble(Panel = Panel, Type = Type, Model = Model, shift = shifts,
                 observed = obs, null = null)
})

shift_summary <- shift_null %>%
  group_by(Panel, Type, Model) %>%
  summarise(dAIC_vs_ref = first(observed),
            n_shifts    = sum(!is.na(null)),
            null_q95    = quantile(null, 0.95, na.rm = TRUE),
            p_shift     = (1 + sum(null >= observed, na.rm = TRUE)) / (1 + sum(!is.na(null))),
            .groups = "drop")
```

**Supplementary Table S13. Cross-validation and circular-shift test of the selected GAMs.** For each panel, the best marginal model and the selected combined model are compared with the reference model `counts_norm_sv ~ s(age)`. ΔAIC vs s(age): AIC improvement over the reference model with all data. ΔELPD: summed difference in out-of-sample log-likelihood relative to the reference model (positive = better), with its standard error, for leave-one-out (LOO) and leave-one-out with a ±200-year buffer (h-block). Centuries better: percentage of left-out centuries predicted better than by the reference model (LOO). Worst century: the century with the largest loss relative to the reference model (LOO). Shift-test *p*: circular-shift test with 49 shifts of 600–5,400 years (minimum attainable *p* = 0.02).

```{r cv circ table}

tab_S13_df <- cv_summary %>%
  left_join(shift_summary, by = c("Panel", "Type", "Model")) %>%
  mutate(Panel = factor(Panel, levels = site_order),
         Type  = factor(Type, levels = c("Marginal", "Combined")),
         `ΔAIC vs s(age)` = sprintf("%.1f", dAIC_vs_ref),
         `LOO ΔELPD (SE)` = sprintf("%.1f (%.1f)", dELPD_LOO, SE_LOO),
         `h-block ΔELPD (SE)` = sprintf("%.1f (%.1f)", dELPD_hblock, SE_hblock),
         `Centuries better (%)` = sprintf("%.0f", pct_better),
         `Worst century (ka)` = sprintf("%.1f", worst_ka),
         `Shift-test p` = sprintf("%.3f", p_shift)) %>%
  arrange(Panel, Type) %>%
  select(Panel, Type, Model, n, `ΔAIC vs s(age)`, `LOO ΔELPD (SE)`, `h-block ΔELPD (SE)`,
         `Centuries better (%)`, `Worst century (ka)`, `Shift-test p`)

tab_S13 <- tab_S13_df %>%
  gt(groupname_col = "Panel") %>%
  tab_header(title = "Cross-validation and circular-shift test of the selected GAMs",
             subtitle = "Reference model: counts_norm_sv ~ s(age)")
tab_S13
gt::gtsave(tab_S13, "Supplementary_Table_S13.docx")
```

```{r figs8}
figS9 <- shift_null %>%
  filter(!is.na(null)) %>%
  mutate(Panel = factor(Panel, levels = site_order),
         Label = paste0(Type, ": ", Model)) %>%
  ggplot(aes(x = null)) +
  geom_histogram(bins = 20, fill = "grey70", colour = "white") +
  geom_vline(aes(xintercept = observed), colour = "red", linewidth = 0.8) +
  facet_wrap(Panel ~ Label, scales = "free", ncol = 4) +
  labs(x = "AIC improvement over s(age) (shifted climate series)", y = "Number of shifts") +
  theme_minimal(base_size = 10)
figS9
ggsave("Supplementary_Figure_S9.pdf", plot = figS9, width = 12, height = 8, device = cairo_pdf)
```

**Supplementary Figure S9. Circular-shift null distributions.** Histograms of the AIC improvement over `s(age)` obtained when the climatic series is shifted by 600–5,400 years relative to the coral record; the red line marks the observed improvement.

**Supplementary Table S14. Drop-one influence of individual centuries.** For each climatic term of the evaluated models: *p*-value with all data, largest *p*-value when one century is removed, the century whose removal gives that value, and the percentage of drop-one fits in which the term remains significant.

```{r drop one infl}
dropone <- purrr::pmap_dfr(filter(val_models, Type != "Reference"), function(Panel, Type, Model) {
  d  <- as.data.frame(val_data[[Panel]]) %>% arrange(age)
  f  <- val_formula(Type, Model, d); cb <- Type == "Combined"
  p_full <- term_pvalues(fit_val(f, d, cb))
  p_full <- p_full[names(p_full) != "s(age)"]
  p_drop <- sapply(seq_len(nrow(d)), function(i) {
    m <- fit_val(f, d[-i, ], cb)
    if (is.null(m)) return(setNames(rep(NA_real_, length(p_full)), names(p_full)))
    term_pvalues(m)[names(p_full)]
  })
  p_drop <- matrix(p_drop, nrow = length(p_full), dimnames = list(names(p_full), NULL))
  tibble::tibble(Panel = Panel, Type = Type, Model = Model, Term = names(p_full),
                 p_all = p_full,
                 p_max = apply(p_drop, 1, max, na.rm = TRUE),
                 century_max = d$age[apply(p_drop, 1, which.max)],
                 pct_signif = 100 * rowMeans(p_drop < 0.05, na.rm = TRUE))
})

tab_S14 <- dropone %>%
  mutate(Panel = factor(Panel, levels = site_order),
         Type  = factor(Type, levels = c("Marginal", "Combined")),
         `p (all data)` = fmt_p2(p_all),
         `Largest p (drop-one)` = fmt_p2(p_max),
         `Century removed (ka)` = sprintf("%.1f", century_max),
         `Drop-one fits with p < 0.05 (%)` = sprintf("%.0f", pct_signif)) %>%
  arrange(Panel, Type) %>%
  select(Panel, Type, Model, Term, `p (all data)`, `Largest p (drop-one)`,
         `Century removed (ka)`, `Drop-one fits with p < 0.05 (%)`) %>%
  gt(groupname_col = "Panel") %>%
  tab_header(title = "Drop-one influence of individual centuries on the climatic terms")
tab_S14
gt::gtsave(tab_S14, "Supplementary_Table_S14.docx")
```

- **Cross-validation.** In Punta Maroma and the regional composite, the panels with a significant hurricane effect, the climate models predict most left-out centuries better than the reference model (68–87 % in LOO) and the summed differences are positive in LOO and h-block, but of the same order as their standard errors (e.g. Punta Maroma combined: ΔELPD = 50.2 ± 36.9 in LOO, 36.4 ± 22.0 in h-block). In Belize the combined model predicts 67 % of the left-out centuries better, but the summed differences are negative and dominated by a few centuries at the extremes of the climatic range (worst: 5.2 ka), where the smooths must extrapolate. In Florida and Saint Croix the climate models are no better than the reference model. With 19–57 centuries per panel, cross-validation does not discriminate firmly among models; the fitted models describe the sampled record and have limited predictive transferability beyond its climatic range.
- **Circular-shift test.** The associations with hurricane activity in Punta Maroma (marginal and combined) and the regional composite (marginal and combined) exceed all 49 shifted alignments (*p* = 0.02, the smallest attainable). In Belize *p* = 0.12 (marginal) and 0.08 (combined): the alignment is not distinguishable from a chance alignment at α = 0.05. In Florida (*p* = 0.60–0.78) and Saint Croix (*p* = 0.38) the observed associations do not differ from the shifted alignments.
- **Drop-one influence.** In Punta Maroma s(Hur100_av) remains significant in every drop-one fit of both models (largest *p* = 0.015 marginal and 0.003 combined, on removing 0.1 ka). In the regional composite it remains significant in 98 % of the marginal fits and all combined fits (largest *p* = 0.073 and 0.034, again on removing 0.1 ka). In Belize the three terms of the combined model remain significant in 96–100 % of the fits (largest *p* = 0.38 for s(dT_raw), on removing 5.2 ka; 0.064 for s(Hur100_av), on removing 0.2 ka). In every panel the century with most influence on the hurricane term is the youngest (0.1–0.2 ka), which is examined further in Section 4.5.2.
- **Former cross-validation.** Because the Regional panel contains the site records, every observation entered the training data twice, and the score was computed against the mean of 4–6 test values per panel and fold. When those values are nearly identical, as is common with 2–6 distinct response values per site, the denominator approaches zero and the score becomes very negative or undefined (median of fold means −9.05 in Punta Maroma; undefined folds in 11 and 14 of the 20 fold assignments in Belize and Saint Croix). Its results therefore do not measure the predictive performance of the site models, and the analysis is replaced by the cross-validation above.

```{r}
## ---- (4) former five-fold cross-validation (withdrawn) -------------------------
## One model fitted to all panels pooled (each record enters twice through the Regional
## panel), stratified 5-fold split, score = 1 - SSE/SST with SST around the TEST-fold mean.
old_cv <- function(seed) {
  set.seed(seed)
  dd <- as.data.frame(merged_df_g)
  fo <- integer(nrow(dd))
  for (p in unique(dd$panel)) {
    i <- which(dd$panel == p); fo[i] <- sample(rep(1:5, length.out = length(i)))
  }
  purrr::map_dfr(1:5, function(k) {
    m  <- bam(counts_norm_sv ~ s(age) + s(Hur100_av), data = dd[fo != k, ],
              family = betar(link = "logit"), method = "fREML", discrete = TRUE)
    te <- dd[fo == k, ] %>% mutate(pred = predict(m, ., type = "response"))
    te %>% group_by(panel) %>%
      summarise(score = 1 - sum((counts_norm_sv - pred)^2) /
                  sum((counts_norm_sv - mean(counts_norm_sv))^2),
                n_test = n(), .groups = "drop") %>%
      mutate(seed = seed, fold = k)
  })
}
old_cv_res <- purrr::map_dfr(1:20, old_cv)

old_cv_res %>%
  group_by(panel, seed) %>%
  summarise(mean_score = mean(score), n_test = mean(n_test), .groups = "drop") %>%
  group_by(Panel = panel) %>%
  summarise(`Test points per fold` = round(mean(n_test), 1),
            `Median of fold means` = signif(median(mean_score[is.finite(mean_score)]), 3),
            `Range of fold means`  = paste(signif(range(mean_score[is.finite(mean_score)]), 3),
                                           collapse = " to "),
            `Assignments with an undefined fold` = sum(!is.finite(mean_score)),
            .groups = "drop") %>%
  mutate(Panel = factor(Panel, levels = site_order)) %>% arrange(Panel) %>%
  gt() %>%
  tab_header(title = "Former five-fold cross-validation, re-run over 20 random fold assignments")
```

## Session information

R and package versions used to render this document:

```{r session-info}
sessionInfo()
```
